### 2023-11-17 project_description.qmd
project description

---
title: "Norwegian ecosystem condition indicators from unsystematic field data"
subtitle: "Project description"
format: pdf
bibliography: mybibfile_projectDescription.bib
author:
  - name: Anders Kolstad
date: "`r Sys.Date()`"
---

\begin{center}
\begin{LARGE}
Background 
\end{LARGE}
\end{center}

:::{.column-margin}
::: {.callout-note appearance="minimal"}
## Area explicit?
Ecosystem condition indicators designed for ecosystem accounting following the SEEA EA standard needs to be area explicit.
That means they need to be assigned to a spatial position.
This is perhaps not so hard -- if for one variable or indicator you have a single value for all of Norwegian forests, but you don't have actual coordinates for the measurement (for different reasons), then you can still assign that value to all forest areas in Norway.
If Norwegian forests are delineated on a map, then this value is also delineated on a map, i.e. it's area explicit.
:::
:::

Ecosystem condition accounts are designed around ecosystem assets (EA) which are the basal units that are assessed, i.e. the smallest spatial resolution of the final assessment. 
The EAs can be small or large[^1], they are _areas_, not points.

[^1]: They can for example be the same as the ecosystem assessment area (EAA) if there is only one EA (e.g. a single assessments for all forests in Norwegian). They can also be fine scaled, for example a single forest patch.

The data underlying the condition indicators can either have wall-to-wall coverage (withing the EAs), or they can have partial coverage.
Wall-to-wall data typically means satellite or at least remotely sensed data.
These are great, because we get to see the spatial variation in the variable, and we can be sure (with some caveats) that the data are area representative, because it is complete.

However, many important ecological aspects are not detectable in remotely sensed data.
This includes things such as species occurrences, or finer scale structural elements, such as dead wood.
To include these aspect in ecosystem account we must rely on more traditional field data that exists typically as point data or patchy polygon sets (e.g. nature type mapping).
To assign indicator values from these variables to the entire area of the EAs, we must **project** in some way over the areas where we have no data.
This can be done through more complex modelling, such as (species) distribution models.
It can also be a simpler process by saying that these data are area representative and therefore simply assign some central tendency from the data to the EA.

::: {.callout-note appearance="minimal"}
## Methods of assigning indicator values to ETs

1. **Conceptual assignment** -- Assume a variable is relevant to an ecosystem based on theory or previous knowledge (e.g. from habitat requiremnts)
1. **Field based assignment** -- Assign ET in the field when recording the variable
1. **Map based assignment** -- Extract information from maps, such as remotely sensed or modelled wall-to-wall data sets
1. **Landscape ecological indicator** -- Derive variable from spatial patterns in maps

Source: @nybo2023
:::

<br>

::: {.callout-note appearance="minimal"}
## Methods of obtaining wall-to-wall indicator coverage

a. Using wall-to-wall data (e.g. remotely sensed data)
a. Modeled variable from other sets of wall-to-wall environmental variables
a. Modeled (extra- or interpolated) variable from training data
a. Simple projection of some central tendency from area representative data
:::

In Norway we have some nature monitoring programs that are area representative, such as the forest inventory monitoring, ANO, ASO, GRUK,  3Q and TOV-E.
From these we can project indicator values to EAs using method c or d.
However, area representative data have the typical disadvantage that they

- do not adequately cover rare ETs
- do not have the density to allow the calculation of indicator values at smaller scales

Limited by these four methods (a-d) we necessarily discard data where we can not assume area representativity, for example due to non-random or biased sampling.

::: {.callout-note appearance="simple"}
There is one important exception to this, and that concerns indicators that are assumed representative, but only for a subset of the ET.
For example, if the ET is _wetland_, then often indicators are included even if they only attain to one type of wetland, even if this subset is a minor componenet of the entire ET (e.g. helophyttic swamps).
I refer to this as _thematic bias_.
:::

As condition accounts are often limited by data availability and resolution, being able to include data sources from less-than-ideal sampling designs would be very useful.

In a recent project [@nybo2023] I used used a spatially and thematically biased data [@mdir2023] to develop area representative EC indicators for five regions in Norway.
The data set [@mdir2023] contains >100 000 nature type localities sampled non-randomly throughout Norway, typically in pressure areas.
For each polygon there are multiple variables recorded which can be relevant for designing EC indicators.
I selected two relevant variables, [alien species](https://ninanor.github.io/ecosystemCondition/alien-species.html) and [anthropogenic disturbance to soil and vegetation](https://ninanor.github.io/ecosystemCondition/slitasje.html) and used (what I think is) a novel, but simple, stratified aggregatation method based on [homogenous impact areas](https://ninanor.github.io/ecosystemCondition/homogeneous-impact-areas.html) (HIAs) [^1] to project indicator values over a sub-set of ETs (a further development of _Method d_ in the info box above).
These indicators provide information not readily replaced with other more suitable data sources.

[^1]: This approach is somewhat similar to the concept of _homogenous ecosystem areas_ introduced in @jrc2022.

\begin{center}
\begin{LARGE}
Project Aims 
\end{LARGE}
\end{center}

With this project I aim to **write a manuscript** documenting the work already done (see above).
The manuscript will focus on introducing the term **HIAs** and the topic of indicator scalabillity.
I will demonstrate these things using the two indicators _alien species_ and _anthropogenic disturbance to soil and vegetation_ calculated

- at two scales, national/regional and regional/local (comparing municipalities)
- in two ecosystems (with and without ET maps)

[**Time and budget**]{.underline}
: I think I need two weeks (**70 hours**) to draft this manuscript, and another **15 hours** for co-author contributions.
The budget is negotiable.

<br>

[**Potential collaborators**]{.underline}
: Many have already contributed with discussions or even formal review or code review (Matt, Marianne, Zander, Vegar). 
In addition I'm thinking about inviting  people that can provide a link to municipal accounting needs (Trond, Erik).
Finally, the manuscript will be written in markdown and managed over GitHub. 
This is reason enough to include others with an interest in these open science approaches (Jens).

\begin{center}
\begin{LARGE}
Examples 
\end{LARGE}
\end{center}

@fig-ex1 and @fig-ex2 show some example results from the previous work dome on this.

```{r fig-ex1, fig.cap="Alien species indicator extrapolated over Nordre Follo (3020), Trondheim (5001) and Målselv (5418). Note that the maps are not masked by ecosystem type, so the three ecosystem specific indicators are overlaping. The bottom row shows the location of the four human impact sones.", echo=FALSE, out.width='80%'}
knitr::include_graphics("images/ex1.PNG")
```

```{r fig-ex2, fig.cap="Effect of latitude on the alien species indicator values across Norway. The blue line is a loess smoother (span=0.75).", echo=FALSE, out.width='100%'}
knitr::include_graphics("images/ex2.PNG")
```

\newpage

# References {-}
::: {#refs}
:::

### 2023-11-17 project_description.qmd
spatially not area explicit

@@
 subtitle: "Project description"
-format: pdf
+format: 
+  pdf:
+    toc: false
 bibliography: mybibfile_projectDescription.bib
@@
 ::: {.callout-note appearance="minimal"}
-## Area explicit?
-Ecosystem condition indicators designed for ecosystem accounting following the SEEA EA standard needs to be area explicit.
+## Spatially explicit?
+Ecosystem condition indicators designed for ecosystem accounting following the SEEA EA standard needs to be spatially explicit.
 That means they need to be assigned to a spatial position.
 This is perhaps not so hard -- if for one variable or indicator you have a single value for all of Norwegian forests, but you don't have actual coordinates for the measurement (for different reasons), then you can still assign that value to all forest areas in Norway.
-If Norwegian forests are delineated on a map, then this value is also delineated on a map, i.e. it's area explicit.
+If Norwegian forests are delineated on a map, then this value is also delineated on a map, i.e. it's spatially explicit.
 :::

### 2024-01-09 manuscript.Rmd
initiate manuscript doc

---
title: A GIS based workflow for Ecosystem Condition Accounting - an example using spatially biased indicators
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    affiliation: Norwegian Institute for Nature Research
    correspondingauthor: true
    footnote: 1
  - name: Bob Security
    email: bob@example.com
    affiliation: Another University
  - name: Cat Memes
    email: cat@example.com
    affiliation: Another University
    footnote: 2
  - name: Derek Zoolander
    email: derek@example.com
    affiliation: Some Institute of Technology
    footnote: 2
address:
  - code: Some Institute of Technology
    organization: Big Wig University
    addressline: 1 main street
    city: Gotham
    state: State
    postcode: 123456
    country: United States
  - code: Another University
    organization: Department
    addressline: A street 29
    postcode: 2054 NX
    city: Manchester,
    country: The Netherlands
footnote:
  - code: 1
    text: "This is the first author footnote."
  - code: 2
    text: "Another author footnote."
abstract: |
  This is the abstract.

  It consists of two paragraphs.
keywords: 
  - SEEA EA
  - alien species
  - ...
journal: "Ecological Indicators"
date: "`r Sys.Date()`"
classoption: preprint, 3p, authoryear
bibliography: mybibfile.bib
linenumbers: false
numbersections: true
# Use a CSL with `citation_package = "default"`
# csl: https://www.zotero.org/styles/elsevier-harvard
output: 
  rticles::elsevier_article:
    keep_tex: true
    citation_package: natbib
---

# Introduction
SEEA EA principles

The need for local field data

Low data availability

Fast turn around with automated workflows

4 types of indicator

# Material and Methods

In this study we focused on open mire ecosystems in Norway.
For this ecosystem we have an ecosystem delineation map produced using a  
a deep learning model which was trained on field verified ground truth data and and satellite imagery. 
This model had a XX% precision when tested against independent field data.

The reference state defining ecosystems in optimal or pristine condition is defined as one with little or no human influence, with a climate as in the period 1961-1990 and a native species similar as today (no mammoths allowed).

We then identified two _concepts_ that define open mire ecosystems in optimal condition:

- Native species dominate
- No physical damage caused by human activities

To describe these two concepts we use five condition variables from the same data set (ref).
The five variables are:

1. Alien plant cover
1. 7SE
1. 7TK
1. PR..
1. PR..

The data set from a standardised field survey of nature types.
Each locality is scored on a range of variables relevant for describing the state and quality of nature.
The surveys are comissioned with the goal of producing rel spatially

Variable 1 was planned to be developed into a single indicator (i.e. a normalised variable) with the same name.
Variables 2-5 describe very related aspects as so they where planned to be combined into a composite indicator called antropogenic disturbance to soil and vegetation so ADSV or short.

Alien plant cover was assigned to ECT class B1a - Compositional state characteristics
ADSV was assigned to ECT class A1 - physical ...

National indicator and for about 4 municipalities as a case study

Validation/sensitivity analyses (varying HIA resolution and minimum _n_?)

Schematic GIS workflow

Aggregation

# Results
Data availability/coverage

Validation

National/regional indicator - do we have enough data?

Municipal case study

# Discussion

# Conclusion

# Credit authorship contribution statement {-}

# Declaration of Competing Interest {-}

# Acknowledgements {-}

# Data availability {-}

# References {-}

### 2024-01-11 manuscript.Rmd
write methods

@@
 In this study we focused on open mire ecosystems in Norway.
-For this ecosystem we have an ecosystem delineation map produced using a  
-a deep learning model which was trained on field verified ground truth data and and satellite imagery. 
+For this ecosystem we have an ecosystem delineation map produced using remotely sensed data and  
+a deep learning model.
 This model had a XX% precision when tested against independent field data.
 
-The reference state defining ecosystems in optimal or pristine condition is defined as one with little or no human influence, with a climate as in the period 1961-1990 and a native species similar as today (no mammoths allowed).
+The reference state defining ecosystems in optimal or pristine condition is defined as one with little or no human influence, with a climate as in the period 1961-1990 and a native species similar as today (i.ei no mammoths allowed).
 
-We then identified two _concepts_ that define open mire ecosystems in optimal condition:
+We then identified two _ecosystem condition characteristics_ that, along with many other characteristics that we do not specify here, define the typical behavior of open mire ecosystems in the reference condition:
 
-- Native species dominate
-- No physical damage caused by human activities
+- Species composition
+- Vegetation intactness
 
@@
 
-1. Alien plant cover
-1. 7SE
-1. 7TK
-1. PR..
-1. PR..
+1. 7FK (prevalence of alien species)
+1. 7SE (tracks from large vehicles)
+1. 7TK (human caused abration or abration-caused erosion)
+1. PRTK (same as 7TK, but recorded at a higher resolution)
+1. PRSL (same as 7SE, but recorded at a higher resolution)
 
-The data set from a standardised field survey of nature types.
+Varaibles 1-3 are defined in Halvorsen and Bratli 2019.
+Variables 4 and 5 are defined in ...
+
+The data set from a standardised field survey of nature types in Norway (REF).
 Each locality is scored on a range of variables relevant for describing the state and quality of nature.
-The surveys are comissioned with the goal of producing rel spatially
+The surveys are comissioned with the goal of producing data relevant for immediate land-use decisions, and is therefore
+spatially biased, typically towards areas with high human impact or expected impact.
 

### 2024-01-12 manuscript.Rmd
First draft of intro and start Methods

@@
 ---
-title: A GIS based workflow for Ecosystem Condition Accounting - an example using spatially biased indicators
+title: Using spatially biased variables in ecosystem condition accounting with a GIS based workflow
 author:
@@
   This is the abstract.
-
   It consists of two paragraphs.
@@
 numbersections: true
-# Use a CSL with `citation_package = "default"`
 # csl: https://www.zotero.org/styles/elsevier-harvard
@@
     citation_package: natbib
+editor_options: 
+  markdown: 
+    wrap: sentence
 ---
@@
 # Introduction
-SEEA EA principles
 
-The need for local field data
+Some general text about the decline in natural capital?
 
-Low data availability
+Ecosystem condition accounting is the game of compiling relevant data on the status, trends and qualities of ecosystems (i.e. nature) and communicating this in a structured format.
+Its purpose is to make it easier to account for nature in policy by making the environmental costs of certain policies and practices visible to decision makers.
+As natural capital keeps declining all over the world, it is becoming increasingly urgent to make the message clear to decision makers about A statistical standard for ecosystem accounting, including ecosystem condition accounting, was developed by the UN and adopted by the UN Statistical Commission in 2021 and is called SEEA EA (System of Environmental-Economic Accounting - Ecosystem Accounting).
+The standard, or framework, is a set of rules, principles and best practices for compiling Ecosystem accounts, mainly aimed at national accounts.
 
-Fast turn around with automated workflows
+Central to ecosystem condition accounts are variables and indicators.
+These are parameters chosen to reflect the sentral condition characteristics of the ecosystems, and that can be quantified and ideally monitored over time to reflect the status and trends in condition.
+Indicators are (data) variables that are normalised (rescaled) against upper and lower reference values to become bound between the values 0 and 1.
+This normalisation ensures that indicators are more comparable because an indicator value of 1 will mean the same for all indicators, i.e.
+that the variable equals the upper reference value which again reflect the value of the variables under the reference condition.
+Similarly, a value of 0 mean that the variable is the worst possible state.
+The reference condition needs to be defined for each Ecosystem Condition Assessment separately, but SEEA EA gives some suggestion, such as an an ecosystem with no or minimal anthorpogenic disturbance.
 
-4 types of indicator
+A general requirement for indicators in the SEEA EA framework is that they should give an unbiased representation of the condition inside the ecosystem assets (Czucz et al 2021; Table 1 - validity; see also SEEA EA §2.87) Ecosystem assets are defined as "ecological entities [meaning areas] about which information is sought and about which statistics are ultimately compiled (REF).
+This requirement for indicator validity means that spatially biased data are ill suited, especially if sampling intensity varies along gradient of anthropogenic pressures and hence ecosystem condition.
+SEEA EA is spatially explicit, which in practice means that indicators that are in some way sampled (i.e.
+not complete wall-to-wall data like remotely sensed imagery), the values are projected on to areas on the map, so that all areas inside the ecosystem accounting area get assigned value for that indicator.
+There are at least 3 general ways to achieve this complete areal coverage of indicator values:
+
+a.  Using wall-to-wall data (e.g. remotely sensed data)
+b.  Predict values using a model (e.g. by accounting for environmental variation)
+c.  Simple projection of some best estimate, typically a central tendency from area representative data
+
+The need for an unbiased estimation of indicator values is unquestionable, but nonetheless, this requirement puts a large limitation on what types of data one can use.
+Ecosystem condition assessments are generally limited by data availability, and the choice of variables and indicators to include in assessments are more often or not a pragmatic and opportunistic one which is unlikely to reflect the full scope of the ecosystem condition characteristics.
+Note that the same is true for thematic biases.
+For example reflected in the scarcity of data included on insects or soil biota, even though most will agree they represent key ecosystem characteristics.
+Also having data from only one or a subset of nature types inside what is defined as the ecosystem in the assessment, is a typical thematic bias in ecosystem accounting.
+However, in this paper we chose to focus on spatial bias.
+
+Being able to make use of spatially biased data would greatly alleviate data shortage problems in ecosystem condition accounts.
+One way to achieve this is modelling (option b in the list above).
+Models can describe the general associations between the spatially sampled data and the context (e.g.
+the set of environmental variables) where it was samples, and use these arelationships to predict and project indicator values to areas that where not originally sampled.
+Depending on the data that goes into these models, they can be very reliable and make good indicators.
+This is especially true when the ecosystem assets are large (e.g. regions or nations).
+But when they are small, like the scale of a municipality, and when the indicator is more likely to be used as the evidence base in concrete physical land use planning, then the inherent lvel of uncertainty from such models becomes unacceptable.
+
+In this study we explore the potential for using a stratified aggregation technique to make use of spatially biased field data in ecosystem condition accounting.
+We demonstrate this technique using a generic GIS-based workflow for compiling ecosystem condition accounts that can be applied at any spatial scale, and we highlight the opportunities for local use-cases of this workflow by contrasting our findings across four neighboring municipalities in Norway.
+The main question is how much generalisation can we perform on the data we have before the resulting indicator looses its practical value in local governance processes.
+We end by interviewing end-users from the relevant municipalities about the perceived benefits and shortcomings of our condition indicators.
 
@@
 In this study we focused on open mire ecosystems in Norway.
-For this ecosystem we have an ecosystem delineation map produced using remotely sensed data and  
+For this ecosystem we have an ecosystem delineation map produced using remotely sensed data and\
 a deep learning model.
-This model had a XX% precision when tested against independent field data.
+This model, which has a 90.9% precision when tested against independent field data, estimates 12.7% of the area in southern Norway is mire (Bakkestuen REF).
+Mires are ecologically and socially important in Norway simply do to its large extent, and especially due to its role in climate mitigation.
+Examples - how much carbon - (REFS) There has not been a national assessment of the ecosystem condition of mires in Norway, but we recent contributed on a report which presented several new indicators that can be used in future assessment (REF Nybø, and web site).
+This study builds on the work in that report.
 
-The reference state defining ecosystems in optimal or pristine condition is defined as one with little or no human influence, with a climate as in the period 1961-1990 and a native species similar as today (i.ei no mammoths allowed).
-
-We then identified two _ecosystem condition characteristics_ that, along with many other characteristics that we do not specify here, define the typical behavior of open mire ecosystems in the reference condition:
+We started by defining the reference state as one where ecosystems are subject to little or no human influence, with a climate as in the period 1961-1990 and a native species pool similar as today (i.e no mammoths allowed).
+We then identified two *ecosystem condition characteristics* that, along with many other characteristics that we do not specify here, define the typical behavior of open mire ecosystems in the reference condition:
 
@@
 
-To describe these two concepts we use five condition variables from the same data set (ref).
+To describe these two concepts we use five condition variables from the same data set (REF).
 The five variables are:
 
-1. 7FK (prevalence of alien species)
-1. 7SE (tracks from large vehicles)
-1. 7TK (human caused abration or abration-caused erosion)
-1. PRTK (same as 7TK, but recorded at a higher resolution)
-1. PRSL (same as 7SE, but recorded at a higher resolution)
+| Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                    | Reference  |
+|------------|-------------|------------------|------------------|-------------|
+| 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | What fraction of the species composition is made up from aline species                                         |            |
+| 7SE                  | Human caused abration or abration-caused erosion | Unitless, ordinal, non-linear scale from 1 to 4.
+| Measures the frequency of imagined 4 m^2^ quadrats layed over the area that has some sign of abration          |            |
+| PRSL                 | *as above*                                       | Unitless, ordinal, non-linear scale from 1 to 8.
+| Same as 7SE, but recorded at a higher resolution                                                               |            |
+| 7TK                  | Tracks from large vehicles                       | Unitless, ordinal, non-linear scale from 1 to 4.
+| Measures the frequency of imagined 100 m^2^ quadrats layed over the area that has some signs of vehicle tracks |            |
+| PRTK                 | *as above*                                       | Unitless, ordinal, non-linear scale from 1 to 8.
+| Same as 7TK, but recorded at a higher resolution                                                               |            |
+| Infrastructure Index | Infrastructure Index                             | Unitelss linear scale from 0 to xxx                                                     | ...
+| Bakkestuen |
+| HIA                  | Human Impact area                                | Ordinal, non-linear scale from 1 to 4                                                   |                                                                                                                |            |
 
-Varaibles 1-3 are defined in Halvorsen and Bratli 2019.
-Variables 4 and 5 are defined in ...
+: Variables used in this study
 
@@
 Each locality is scored on a range of variables relevant for describing the state and quality of nature.
-The surveys are comissioned with the goal of producing data relevant for immediate land-use decisions, and is therefore
-spatially biased, typically towards areas with high human impact or expected impact.
+The surveys are comissioned with the goal of producing data relevant for immediate land-use decisions, and is therefore spatially biased, typically towards areas with high human impact or expected impact.
 
@@
 
-Alien plant cover was assigned to ECT class B1a - Compositional state characteristics
-ADSV was assigned to ECT class A1 - physical ...
+Alien plant cover was assigned to ECT class B1a - Compositional state characteristics ADSV was assigned to ECT class A1 - physical ...
 
@@
 
-Validation/sensitivity analyses (varying HIA resolution and minimum _n_?)
+Validation/sensitivity analyses (varying HIA resolution and minimum *n*?)
 
@@
 # Results
+
 Data availability/coverage
@@
 
-# Credit authorship contribution statement {-}
+# Credit authorship contribution statement {.unnumbered}
 
-# Declaration of Competing Interest {-}
+# Declaration of Competing Interest {.unnumbered}
 
-# Acknowledgements {-}
+# Acknowledgements {.unnumbered}
 
-# Data availability {-}
+# Data availability {.unnumbered}
 
-# References {-}
+# References {.unnumbered}

### 2024-01-14 R/scrap/bayesian inference.qmd
ways to estimate uncertainty

---
title: "Bayesian inference"
format: html
editor: visual
---

## Bayesian inference

I want to test how I can get best estimates of the population means with a distribution around it to describe the uncertainty. If I just use the sample variance, then with very low sample sizes, like *n =* 1 or 2, I can get very low variance just by chance, making the estimates seem very robust, when in fact it is not.

If I use bayesian inference over a flat prior and using the s.dev. from the national data set (closer to the true sd.dev. than the sub-population sd.dev.), then perhaps I can get what I want?

```{r, warning=F, message=F}
library(Bolstad)
library(tidyverse)
```

Our variable is discrete, 1 to 7.

```{r}
var <- seq(1,7, by=1) 
```

We want to estimate the mean of the population.

The probability density function for the mean (mu) will be normally distributed and truncated at 1 and 7.

Let's generate some example population samples.

```{r}
random1 <- 2
random2 <- c(sample(var, 2, replace = T))
random10 <- c(sample(var, 10, replace = T))

full <- c(sample(var, 1000, replace = T))
```

`Bolstad::normdp` assumes the samples comes from a normal distribution with a known st.dev.

I will take the st.dev from the entire national data set.

```{r}
national.sd <- sd(full)
```

We want to use a uniform prior.

Here's the posterior distribution for mu when we have just *n* = 1

```{r}
pd1 <- normdp(x = random1,
       sigma.x = national.sd,
       n.mu = 50) # default 
```

The plot is truncated at 4. We can see this better if we extract the values manually and make our own plot

```{r}
x1 <- pd1[["param.x"]]
y1 <- pd1[["posterior"]]
df <- data.frame(x1, y1)
df |> 
  ggplot(aes(x = x1, y = y1))+
  geom_area(fill="orange", alpha=.5)+
  xlim(c(0,7))
```

The possible value of mu is restricted. We have to define it.

Mu is actually not discrete. Just the data are.

```{r}
mu <- seq(1, 7, length.out=100)
```

```{r}
pd2 <- normdp(x = random1,
       sigma.x = national.sd,
       n.mu = 50, # not having an effect
       mu = mu,
       mu.prior = rep(1/length(mu), times=length(mu)))

```

OK, but there is practically zero change of mu being 7. We only have one sample. Shouldn't there be a wider spread of probabilities?

Actually, the probability that mu is 7 is calculated to be

```{r}
min(pd2$posterior)
```

So not zero, but still very small. And if the st.dev is 2, then perhaps this is correct.

```{r}
dnorm(7, mean=random1, sd=national.sd)
```

This is also a small number, although not identical.

```{r}
pd10 <- normdp(x = random10,
       sigma.x = national.sd,
       n.mu = 50, # not having an effect
       mu = mu,
       mu.prior = rep(1/length(mu), times=length(mu)))
```

Lets try some more realistic samples.

```{r}
sampleHigh2 <- round(runif(2, 5,7))
sampleHigh5 <- round(runif(5, 5,7))
sampleHigh10 <- round(runif(10, 5,7))
sampleHigh20 <- round(runif(20, 5,7))
```

```{r, output=F}
pd_h2 <- normdp(x = sampleHigh2,
       sigma.x = national.sd,
       mu = mu,
       mu.prior = rep(1/length(mu), times=length(mu)))

pd_h5 <- normdp(x = sampleHigh5,
       sigma.x = national.sd,
       mu = mu,
       mu.prior = rep(1/length(mu), times=length(mu)))

pd_h10 <- normdp(x = sampleHigh10,
       sigma.x = national.sd,
       mu = mu,
       mu.prior = rep(1/length(mu), times=length(mu)))

pd_h20 <- normdp(x = sampleHigh20,
       sigma.x = national.sd,
       mu = mu,
       mu.prior = rep(1/length(mu), times=length(mu)))
```

```{r}
data.frame(x = mu,
                 "n2" = pd_h2$posterior,
                 "n5" = pd_h5$posterior,
                 "n10" = pd_h10$posterior,
                 "n20" = pd_h20$posterior) |>
  pivot_longer(!x, 
               names_to = "n") |>
  ggplot(aes(x = x, y = value, fill = n))+
  geom_area(alpha=.5)+
  theme_bw()

```

This figure is wrong. The data is correct to begin with, but plotting goes wrong.

### 2024-01-15 R/scrap/bayesian inference.qmd
wgt_mean

@@
 
-## Bayesian inference
+## Mean of population
 
@@
 If I just use the sample variance, then with very low sample sizes, like *n =* 1 or 2, I can get very low variance just by chance, making the estimates seem very robust, when in fact it is not.
-
-If I use bayesian inference over a flat prior and using the s.dev. from the national data set (closer to the true sd.dev. than the sub-population sd.dev.), then perhaps I can get what I want?
-
-```{r, warning=F, message=F}
-library(Bolstad)
-library(tidyverse)
-```
 
@@
 var <- seq(1,7, by=1) 
+```
+
+We have a national data set with thousands of samples (*n* \> 1000). From this we want to get the mean for local regions (the data is georeferenced).
+
+If I use bayesian inference over a uniform prior and using the st.dev. from the national data set (closer to the true st.dev. than the sub-population st.dev.), then perhaps I can get what I want?
+
+I will use the Bolstad package for the inference.
+
+```{r, warning=F, message=F}
+library(Bolstad)
+library(tidyverse)
+library(ggridges)
 ```
@@
 
-`Bolstad::normdp` assumes the samples comes from a normal distribution with a known st.dev.
+`Bolstad::normgcp` assumes the samples comes from a normal distribution with a known st.dev.
 
@@
 ```{r}
-pd1 <- normdp(x = random1,
+pd1 <- normgcp(x = random1,
        sigma.x = national.sd,
+       density = "uniform",
        n.mu = 50) # default 
@@
 
-The plot is truncated at 4. We can see this better if we extract the values manually and make our own plot
+The plot goes beyond the possible range of \[1-7\]. We can see this better if we extract the values manually and make our own plot.
 
@@
 
-The possible value of mu is restricted. We have to define it.
+We can perhaps trucate the posterior, like the figure above, or we could use a discrete prior
 
-Mu is actually not discrete. Just the data are.
+`mu` is actually not discrete, just the data are. But with this approach, mu is also discrete, but we can add many steps to make it core or less continous. I use 100 steps below.
 
@@
        sigma.x = national.sd,
-       n.mu = 50, # not having an effect
        mu = mu,
@@
 
-This is also a small number, although not identical.
+This is also a small number, although not identical. I'm not sure I understand the maths here.
+
+Let's try with *n* = 10.
 
@@
                names_to = "n") |>
-  ggplot(aes(x = x, y = value, fill = n))+
-  geom_area(alpha=.5)+
-  theme_bw()
+  mutate(n2 = str_remove(n, "n"),
+         n3 = as.numeric(n2),
+         n = fct_reorder(n2, n3)) |>
+  ggplot(aes(x = x, 
+             y = n, 
+             height = value,
+             fill = n))+
+  geom_density_ridges(alpha=.5,
+                      stat = "identity")+
+  theme_ridges()
 
@@
 
-This figure is wrong. The data is correct to begin with, but plotting goes wrong.
+This is looking good so far.
+
+## Weighted mean
+
+Now for the next bit. We actually need the weighted mean for the population.
+
+Let's use `pd10` to compare against. `pd10` used the `random10` as the population sample.
+
+```{r}
+plot(pd10)
+```
+
+We want to add weights to `random10`. Creating vectors of weights that down weigh the low numbers:
+
+```{r}
+random10_w <- random10
+random10_w2 <- random10^2
+```
+
+I have custumised `Bolstad::normdp` to handle weighted means with a function called `wgt_mean`.
+
+```{r}
+source("../bayes_weightedMean.R")
+```
+
+```{r, output = F}
+dw1 <- wgt_mean(x = random10,
+         weights = random10_w, #notice new argument
+       sigma.x = national.sd,
+       mu = mu,
+       mu.prior = rep(1/length(mu), times=length(mu)))
+
+dw2 <- wgt_mean(x = random10,
+         weights = random10_w2,
+       sigma.x = national.sd,
+       mu = mu,
+       mu.prior = rep(1/length(mu), times=length(mu)))
+```
+
+```{r}
+data.frame(x = mu,
+          "No_weights"    = pd10$posterior,
+          "Proportional" = dw1$posterior,
+          "Exponential"  = dw2$posterior) |>
+  pivot_longer(!x, 
+               names_to = "Weights") |>
+  mutate(Weights = fct_relevel(Weights, 
+                               "No_weights",
+                               "Proportional",
+                               "Exponential")) |>
+    ggplot(aes(x = x, 
+             y = Weights, 
+             height = value,
+             fill = Weights))+
+  geom_density_ridges(alpha=.5,
+                      stat = "identity")+
+  theme_ridges()
+```
+
+Our custom function has succesfully increased the weight on the high values in the sample.
+
+How about if the sample size varies.
+
+```{r, output=F}
+dw3 <- wgt_mean(x = sampleHigh2,
+         weights = sampleHigh2,
+       sigma.x = national.sd,
+       mu = mu,
+       mu.prior = rep(1/length(mu), times=length(mu)))
+
+dw4 <- wgt_mean(x = sampleHigh10,
+         weights = sampleHigh10,
+       sigma.x = national.sd,
+       mu = mu,
+       mu.prior = rep(1/length(mu), times=length(mu)))
+
+dw5 <- wgt_mean(x = sampleHigh20,
+         weights = sampleHigh20,
+       sigma.x = national.sd,
+       mu = mu,
+       mu.prior = rep(1/length(mu), times=length(mu)))
+```
+
+```{r}
+data.frame(x = mu,
+                 "n2" = dw3$posterior,
+                 "n10" = dw4$posterior,
+                 "n20" = dw5$posterior) |>
+  pivot_longer(!x, 
+               names_to = "n") |>
+  mutate(n2 = str_remove(n, "n"),
+         n3 = as.numeric(n2),
+         n = fct_reorder(n2, n3)) |>
+  ggplot(aes(x = x, 
+             y = n, 
+             height = value,
+             fill = n))+
+  geom_density_ridges(alpha=.5,
+                      stat = "identity")+
+  theme_ridges()
+```
+
+It works as predicted.

### 2024-01-15 R/bayes_weightedMean.R

@@ -1,0 +1,57 @@
+wgt_mean <- function (x, weights, sigma.x = NULL, mu = NULL, mu.prior = NULL, n.mu = 50, 
+          ...) 
+{
+  if (n.mu < 3) 
+    stop("Number of prior values of theta must be greater than 2")
+  if (is.null(mu)) {
+    mu = seq(min(x) - sigma.x, max(x) + sigma.x, length = n.mu)
+    mu.prior = rep(1/n.mu, n.mu)
+  }
+  mx = weighted.mean(x, weights)
+  quiet = Bolstad.control(...)$quiet
+  if (is.null(sigma.x)) {
+    sigma.x = sd(x - mx)
+    if (!quiet) {
+      cat(paste("Standard deviation of the residuals :", 
+                signif(sigma.x, 4), "\n", sep = ""))
+    }
+  }
+  else {
+    if (sigma.x > 0) {
+      if (!quiet) {
+        cat(paste("Known standard deviation :", signif(sigma.x, 
+                                                       4), "\n", sep = ""))
+      }
+    }
+    else {
+      stop("The standard deviation must be greater than zero")
+    }
+  }
+  if (any(mu.prior < 0) | any(mu.prior > 1)) 
+    stop("Prior probabilities must be between 0 and 1 inclusive")
+  if (round(sum(mu.prior), 7) != 1) {
+    warning("The prior probabilities did not sum to 1, therefore the prior has been normalized")
+    mu.prior = mu.prior/sum(mu.prior)
+  }
+  n.mu = length(mu)
+  nx = length(x)
+  snx = sigma.x^2/nx
+  likelihood = exp(-0.5 * (mx - mu)^2/snx)
+  posterior = likelihood * mu.prior/sum(likelihood * mu.prior)
+  if (Bolstad.control(...)$plot) {
+    plot(mu, posterior, ylim = c(0, 1.1 * max(posterior, 
+                                              mu.prior)), pch = 20, col = "blue", xlab = expression(mu), 
+         ylab = expression(Probabilty(mu)))
+    points(mu, mu.prior, pch = 20, col = "red")
+    legend("topleft", bty = "n", fill = c("blue", "red"), 
+           legend = c("Posterior", "Prior"), cex = 0.7)
+  }
+  mx = sum(mu * posterior)
+  vx = sum((mu - mx)^2 * posterior)
+  results = list(name = "mu", param.x = mu, prior = mu.prior, 
[…]

### 2024-01-16 manuscript.Rmd
messing with crossrefs

@@
 
+```{r setup, include=F, warning=F, message=F}
+library(tidyverse)
+library(knitr)
+library(bookdown)
+library(readr)
+library(kableExtra)
+options(knitr.table.format = "latex") 
+```
+
 # Introduction
@@
 In this study we focused on open mire ecosystems in Norway.
-For this ecosystem we have an ecosystem delineation map produced using remotely sensed data and\
-a deep learning model.
+For this ecosystem we have an ecosystem delineation map produced using remotely sensed data and a deep learning model (REF Bakkestuen).
 This model, which has a 90.9% precision when tested against independent field data, estimates 12.7% of the area in southern Norway is mire (Bakkestuen REF).
-Mires are ecologically and socially important in Norway simply do to its large extent, and especially due to its role in climate mitigation.
-Examples - how much carbon - (REFS) There has not been a national assessment of the ecosystem condition of mires in Norway, but we recent contributed on a report which presented several new indicators that can be used in future assessment (REF Nybø, and web site).
+Mires are ecologically and socially important in Norway simply do to its large extent, and due to its role in climate mitigation, as mires store a large amount of carbon (REF?).
+There has not been a national assessment of the ecosystem condition of mires in Norway, but we recent contributed on a report which presented several new indicators that can be used in future assessment (REF Nybø, and web site).
 This study builds on the work in that report.
@@
 We started by defining the reference state as one where ecosystems are subject to little or no human influence, with a climate as in the period 1961-1990 and a native species pool similar as today (i.e no mammoths allowed).
-We then identified two *ecosystem condition characteristics* that, along with many other characteristics that we do not specify here, define the typical behavior of open mire ecosystems in the reference condition:
+We then identified two *ecosystem condition characteristics* that, along with many other characteristics that we do not specify here, define the typical behavior of open mire ecosystems in the reference condition: species composition, and; vegetation intactness.
 
--   Species composition
--   Vegetation intactness
-
-To describe these two concepts we use five condition variables from the same data set (REF).
-The five variables are:
+To describe these two concepts we use five condition variables from the same data set (see \@(tab:foo)).
 
 | Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                    | Reference  |
-|------------|-------------|------------------|------------------|-------------|
-| 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | What fraction of the species composition is made up from aline species                                         |            |
+|---------------|---------------|---------------|---------------|---------------|
+| 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                             | Halvorsen  |
 | 7SE                  | Human caused abration or abration-caused erosion | Unitless, ordinal, non-linear scale from 1 to 4.
-| Measures the frequency of imagined 4 m^2^ quadrats layed over the area that has some sign of abration          |            |
-| PRSL                 | *as above*                                       | Unitless, ordinal, non-linear scale from 1 to 8.
-| Same as 7SE, but recorded at a higher resolution                                                               |            |
+| Measures the frequency of imagined 4 m^2^ quadrats layed over the area that has some sign of abration          | ref        |
+| PRSL                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.
+| Same as 7SE, but recorded at a higher resolution                                                               | Mdir       |
 | 7TK                  | Tracks from large vehicles                       | Unitless, ordinal, non-linear scale from 1 to 4.
-| Measures the frequency of imagined 100 m^2^ quadrats layed over the area that has some signs of vehicle tracks |            |
-| PRTK                 | *as above*                                       | Unitless, ordinal, non-linear scale from 1 to 8.
-| Same as 7TK, but recorded at a higher resolution                                                               |            |
-| Infrastructure Index | Infrastructure Index                             | Unitelss linear scale from 0 to xxx                                                     | ...
-| Bakkestuen |
-| HIA                  | Human Impact area                                | Ordinal, non-linear scale from 1 to 4                                                   |                                                                                                                |            |
+| Measures the frequency of imagined 100 m^2^ quadrats layed over the area that has some signs of vehicle tracks | Halvorsen  |
+| PRTK                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.
+| Same as 7TK, but recorded at a higher resolution                                                               | Mdir       |
+| Infrastructure Index | Infrastructure Index                             | Unitelss linear scale from 0 to 13.2                                                    | Unitelss index ranging from from 0 to 13.2                                                                     | Erikstad   |
+| HIA                  | Human Impact area                                | Ordinal, non-linear scale from 1 to 4                                                   | A categorical representation of the infrastructure index                                                       | this paper |
 
-: Variables used in this study
+Table: (\#tab:foo) Variables used in this study
 
-The data set from a standardised field survey of nature types in Norway (REF).
-Each locality is scored on a range of variables relevant for describing the state and quality of nature.
-The surveys are comissioned with the goal of producing data relevant for immediate land-use decisions, and is therefore spatially biased, typically towards areas with high human impact or expected impact.
+Reference to Figure \@(fig:cars-plot)
 
-Variable 1 was planned to be developed into a single indicator (i.e. a normalised variable) with the same name.
-Variables 2-5 describe very related aspects as so they where planned to be combined into a composite indicator called antropogenic disturbance to soil and vegetation so ADSV or short.
+```{r cars-plot, fig.cap="Test caption"}
+plot(cars)
+```
 
-Alien plant cover was assigned to ECT class B1a - Compositional state characteristics ADSV was assigned to ECT class A1 - physical ...
+The data set is from a standardised field survey of nature types in Norway that started in 2018 and which is still ongoing (REF).
+We included data fram 2018 to 2023.
+In this survey, selected nature types are delineatied on a map, and each locality is scored on a range of variables relevant for describing the state and quality of nature.
+The surveys are commissioned with the goal of producing data relevant for immediate land-use decisions, and is therefore spatially biased, typically towards areas with high human impact or expected impact.
+In addition there is a thematic and size bias in the sampling protocol.
+Of all possible mire types, the survey only maps the following:
 
-National indicator and for about 4 municipalities as a case study
+-   Southern ombrotrophic mires \> 2500 m^2^
+
+-   Northern omgrotrophic \> 10.000 m^2^
+
+-   All semi-natural mires (minerotrophic)
+
+-   Calcareous southern fens \>500 m^2^
+
+-   Calcareous northen fens \>1000 m^2^
+
+In the above, *southern* refers to boreonemoral and southboreal zones, and *northern* refers to mid-boreal, north boreal, and alpine zones.
+In addition, the northern fens need to be even more calcareous than the southern fens in order to be surveyed.
+In this paper we assume the survey is representative for the entire mire ecosystem in Norway.
+Although it is possible that smaller or less calcareous mires will score systematically different than the ones that are surveyed we do not think this is the case for our variables at least.
+
+7FK was developed into a single indicator (i.e. a normalised variable) named *Alien plant cover*.
+Variables 2-5 describe very related aspects as so they where planned to be combined into one indicator called antropogenic disturbance to soil and vegetation, or ADSV for short.
+This was done by taking the value from the variable with the lowest value, i.e. worst condition.
+For both indicators, the lower and upper reference values, i.e.
+the worst and best possible condition that the variables can be in, is given by the range of the data where 0 or 1 is the lower limit and 4, 7 or 8 is the upper limit, depending on the variable (see @tbl-variables).
+
+The Alien plant cover was assigned to ECT class B1 - Compositional state characteristics and ADSV was assigned to ECT class A1 - Physical state characteristics (REF).
+
+![Schematic workflow followed in this paper. Sub-index A is calculated using spatially explicit aggregation (i.e .a simple overlay takingthe mean value).](images/workflow.jpg)
+
+We chose three municipalities in Norway to test out the indicator (FIG).
+These municipalities are spread over xxx km north to south, and also differ in the amount of mire area, the total area surveyed, and the prevalence of infrastructure (TABLE)
+
+To
 

### 2024-01-16 manuscript/HIAs.qmd
move to quarto template - crossrefs ok

--- 2024-01-16 manuscript.Rmd
@@
 title: Using spatially biased variables in ecosystem condition accounting with a GIS based workflow
+subtitle: A Short Subtitle
 author:
@@
     email: anders.kolstad@nina.no
-    affiliation: Norwegian Institute for Nature Research
-    correspondingauthor: true
-    footnote: 1
+    affiliations: 
+        - id: some-tech
+          name: Norwegian Institute for Nature Research
+          department: Department of Terrestrial Ecology
+          address: Street Address
+          city: City
+          state: State
+          postal-code: Postal Code
+    attributes:
+        corresponding: true
+    note: This is the first author footnote.
   - name: Bob Security
     email: bob@example.com
-    affiliation: Another University
+    affiliations:
+        - id: another-u
+          name: Another University
+          department: Department Name
+          address: Street Address
+          city: City
+          state: State
+          postal-code: Postal Code
+    note: |
+      Another author footnote, this is a very long footnote and it should be a really long footnote. But this footnote is not yet sufficiently long enough to make two lines of footnote text.
   - name: Cat Memes
     email: cat@example.com
-    affiliation: Another University
-    footnote: 2
+    affiliations:
+        - ref: another-u
+    note: Yet another author footnote.
   - name: Derek Zoolander
     email: derek@example.com
-    affiliation: Some Institute of Technology
-    footnote: 2
-address:
-  - code: Some Institute of Technology
-    organization: Big Wig University
-    addressline: 1 main street
-    city: Gotham
-    state: State
-    postcode: 123456
-    country: United States
-  - code: Another University
-    organization: Department
-    addressline: A street 29
-    postcode: 2054 NX
-    city: Manchester,
-    country: The Netherlands
-footnote:
-  - code: 1
-    text: "This is the first author footnote."
-  - code: 2
-    text: "Another author footnote."
+    affilations:
+        - ref: some-tech
 abstract: |
   This is the abstract.
-  It consists of two paragraphs.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit.
+Nam placerat turpis purus, eu tristique ex tincidunt et.
+Mauris sed augue eget turpis ultrices tincidunt.
+Sed et mi in leo porta egestas.
+Aliquam non laoreet velit.
+Nunc quis ex vitae eros aliquet auctor nec ac libero.
+Duis laoreet sapien eu mi luctus, in bibendum leo molestie.
+Sed hendrerit diam diam, ac dapibus nisl volutpat vitae.
+Aliquam bibendum varius libero, eu efficitur justo rutrum at.
+Sed at tempus elit.
 keywords: 
@@
   - alien species
-  - ...
-journal: "Ecological Indicators"
-date: "`r Sys.Date()`"
-classoption: preprint, 3p, authoryear
-bibliography: mybibfile.bib
-linenumbers: false
-numbersections: true
-# csl: https://www.zotero.org/styles/elsevier-harvard
-output: 
-  rticles::elsevier_article:
-      keep_tex: true
-      citation_package: natbib
-editor_options: 
-  markdown: 
-    wrap: sentence
+date: last-modified
+bibliography: bibliography.bib
+format:
+  elsevier-pdf:
+    keep-tex: true
+    journal:
+      name: Ecological Indicators
+      formatting: preprint
+      model: 3p
+      cite-style: super
 ---
-
-```{r setup, include=F, warning=F, message=F}
-library(tidyverse)
-library(knitr)
-library(bookdown)
-library(readr)
-library(kableExtra)
-options(knitr.table.format = "latex") 
-```
-
 # Introduction
@@
 
-To describe these two concepts we use five condition variables from the same data set (see \@(tab:foo)).
+To describe these two concepts we use five condition variables from the same data set (see @tbl-variables.
 
@@
 
-Table: (\#tab:foo) Variables used in this study
+Table: Variables used in this study {#tbl-variables}
 
-Reference to Figure \@(fig:cars-plot)
+Reference to @fig-carsplot
 
-```{r cars-plot, fig.cap="Test caption"}
+```{r fig-carsplot, fig.cap="Test caption"}
 plot(cars)
@@
 
-![Schematic workflow followed in this paper. Sub-index A is calculated using spatially explicit aggregation (i.e .a simple overlay takingthe mean value).](images/workflow.jpg)
+![Schematic workflow followed in this paper. Sub-index A is calculated using spatially explicit aggregation (i.e .a simple overlay takingthe mean value).](../images/workflow.jpg){#fig-workflow}
 
@@
 
-Schematic GIS workflow
+Schematic GIS workflow, see @fig-workflow
 
@@
 
-# Data availability {.unnumbered}
+# Data availability {-}
 

### 2024-01-16 manuscript/HIAs.tex
move to quarto template - crossrefs ok

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  super,
  preprint,
  3p]{elsarticle}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{241,243,245}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Ecological Indicators}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Using spatially biased variables in ecosystem condition accounting with a GIS based workflow},
  pdfauthor={Anders Lorentzen Kolstad; Bob Security; Cat Memes; Derek Zoolander},
  pdfkeywords={SEEA EA, alien species},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\setlength{\parindent}{6pt}
\begin{document}

\begin{frontmatter}
\title{Using spatially biased variables in ecosystem condition
accounting with a GIS based workflow \\\large{A Short Subtitle} }
\author[1]{Anders Lorentzen Kolstad%
\corref{cor1}%
\fnref{fn1}}
 \ead{anders.kolstad@nina.no} 
\author[2]{Bob Security%
%
\fnref{fn2}}
 \ead{bob@example.com} 
\author[2]{Cat Memes%
%
\fnref{fn3}}
 \ead{cat@example.com} 
\author[]{Derek Zoolander%
%
}
 \ead{derek@example.com}

\affiliation[1]{organization={Norwegian Institute for Nature
Research, Department of Terrestrial Ecology},addressline={Street
Address},city={City},postcode={Postal Code},postcodesep={}}
\affiliation[2]{organization={Another University, Department
Name},addressline={Street Address},city={City},postcode={Postal
Code},postcodesep={}}

\cortext[cor1]{Corresponding author}
\fntext[fn1]{This is the first author footnote.}
\fntext[fn2]{Another author footnote, this is a very long footnote and
it should be a really long footnote. But this footnote is not yet
sufficiently long enough to make two lines of footnote text.}
\fntext[fn3]{Yet another author footnote.}

\begin{abstract}
This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing
elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget
velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris
sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta
egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet
auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo
molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae.
Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at
tempus elit.
\end{abstract}

\begin{keyword}
    SEEA EA \sep 
    alien species
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, sharp corners, breakable]}{\end{tcolorbox}}\fi

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Some general text about the decline in natural capital?

Ecosystem condition accounting is the game of compiling relevant data on
the status, trends and qualities of ecosystems (i.e.~nature) and
communicating this in a structured format. Its purpose is to make it
easier to account for nature in policy by making the environmental costs
of certain policies and practices visible to decision makers. As natural
capital keeps declining all over the world, it is becoming increasingly
urgent to make the message clear to decision makers about A statistical
standard for ecosystem accounting, including ecosystem condition
accounting, was developed by the UN and adopted by the UN Statistical
Commission in 2021 and is called SEEA EA (System of
Environmental-Economic Accounting - Ecosystem Accounting). The standard,
or framework, is a set of rules, principles and best practices for
compiling Ecosystem accounts, mainly aimed at national accounts.

Central to ecosystem condition accounts are variables and indicators.
These are parameters chosen to reflect the sentral condition
characteristics of the ecosystems, and that can be quantified and
ideally monitored over time to reflect the status and trends in
condition. Indicators are (data) variables that are normalised
(rescaled) against upper and lower reference values to become bound
between the values 0 and 1. This normalisation ensures that indicators
are more comparable because an indicator value of 1 will mean the same
for all indicators, i.e.~that the variable equals the upper reference
value which again reflect the value of the variables under the reference
condition. Similarly, a value of 0 mean that the variable is the worst
possible state. The reference condition needs to be defined for each
Ecosystem Condition Assessment separately, but SEEA EA gives some
suggestion, such as an an ecosystem with no or minimal anthorpogenic
disturbance.

A general requirement for indicators in the SEEA EA framework is that
they should give an unbiased representation of the condition inside the
ecosystem assets (Czucz et al 2021; Table 1 - validity; see also SEEA EA
§2.87) Ecosystem assets are defined as ``ecological entities {[}meaning
areas{]} about which information is sought and about which statistics
are ultimately compiled (REF). This requirement for indicator validity
means that spatially biased data are ill suited, especially if sampling
intensity varies along gradient of anthropogenic pressures and hence
ecosystem condition. SEEA EA is spatially explicit, which in practice
means that indicators that are in some way sampled (i.e.~not complete
wall-to-wall data like remotely sensed imagery), the values are
projected on to areas on the map, so that all areas inside the ecosystem
accounting area get assigned value for that indicator. There are at
least 3 general ways to achieve this complete areal coverage of
indicator values:

\begin{enumerate}
\def\labelenumi{\alph{enumi}.}
\tightlist
\item
  Using wall-to-wall data (e.g.~remotely sensed data)
\item
  Predict values using a model (e.g.~by accounting for environmental
  variation)
\item
  Simple projection of some best estimate, typically a central tendency
  from area representative data
\end{enumerate}

The need for an unbiased estimation of indicator values is
unquestionable, but nonetheless, this requirement puts a large
limitation on what types of data one can use. Ecosystem condition
assessments are generally limited by data availability, and the choice
of variables and indicators to include in assessments are more often or
not a pragmatic and opportunistic one which is unlikely to reflect the
full scope of the ecosystem condition characteristics. Note that the
same is true for thematic biases. For example reflected in the scarcity
of data included on insects or soil biota, even though most will agree
they represent key ecosystem characteristics. Also having data from only
one or a subset of nature types inside what is defined as the ecosystem
in the assessment, is a typical thematic bias in ecosystem accounting.
However, in this paper we chose to focus on spatial bias.

Being able to make use of spatially biased data would greatly alleviate
data shortage problems in ecosystem condition accounts. One way to
achieve this is modelling (option b in the list above). Models can
describe the general associations between the spatially sampled data and
the context (e.g.~the set of environmental variables) where it was
samples, and use these arelationships to predict and project indicator
values to areas that where not originally sampled. Depending on the data
that goes into these models, they can be very reliable and make good
indicators. This is especially true when the ecosystem assets are large
(e.g.~regions or nations). But when they are small, like the scale of a
municipality, and when the indicator is more likely to be used as the
evidence base in concrete physical land use planning, then the inherent
lvel of uncertainty from such models becomes unacceptable.

In this study we explore the potential for using a stratified
aggregation technique to make use of spatially biased field data in
ecosystem condition accounting. We demonstrate this technique using a
generic GIS-based workflow for compiling ecosystem condition accounts
that can be applied at any spatial scale, and we highlight the
opportunities for local use-cases of this workflow by contrasting our
findings across four neighboring municipalities in Norway. The main
question is how much generalisation can we perform on the data we have
before the resulting indicator looses its practical value in local
governance processes. We end by interviewing end-users from the relevant
municipalities about the perceived benefits and shortcomings of our
condition indicators.

\hypertarget{material-and-methods}{%
\section{Material and Methods}\label{material-and-methods}}

In this study we focused on open mire ecosystems in Norway. For this
ecosystem we have an ecosystem delineation map produced using remotely
sensed data and a deep learning model (REF Bakkestuen). This model,
which has a 90.9\% precision when tested against independent field data,
estimates 12.7\% of the area in southern Norway is mire (Bakkestuen
REF). Mires are ecologically and socially important in Norway simply do
to its large extent, and due to its role in climate mitigation, as mires
store a large amount of carbon (REF?). There has not been a national
assessment of the ecosystem condition of mires in Norway, but we recent
contributed on a report which presented several new indicators that can
be used in future assessment (REF Nybø, and web site). This study builds
on the work in that report.

We started by defining the reference state as one where ecosystems are
subject to little or no human influence, with a climate as in the period
1961-1990 and a native species pool similar as today (i.e no mammoths
allowed). We then identified two \emph{ecosystem condition
characteristics} that, along with many other characteristics that we do
not specify here, define the typical behavior of open mire ecosystems in
the reference condition: species composition, and; vegetation
intactness.

To describe these two concepts we use five condition variables from the
same data set (see Table~\ref{tbl-variables}.

\hypertarget{tbl-variables}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}@{}}
\caption{\label{tbl-variables}Variables used in this
study}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Variable code
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Variable name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Measurement unit
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Description
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Reference
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Variable code
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Variable name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Measurement unit
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Description
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Reference
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
7FK & Prevalence of alien species & Unitless, ordinal, non-linear scale
from 1 (no alien species) to 7 (only alien species) & The fraction of
the species composition made up from alien species & Halvorsen \\
7SE & Human caused abration or abration-caused erosion & Unitless,
ordinal, non-linear scale from 1 to 4. & Measures the frequency of
imagined 4 m\textsuperscript{2} quadrats layed over the area that has
some sign of abration & ref \\
PRSL & \emph{as above} & Unitless, ordinal, non-linear scale from 0 to
7. & Same as 7SE, but recorded at a higher resolution & Mdir \\
7TK & Tracks from large vehicles & Unitless, ordinal, non-linear scale
from 1 to 4. & Measures the frequency of imagined 100
m\textsuperscript{2} quadrats layed over the area that has some signs of
vehicle tracks & Halvorsen \\
PRTK & \emph{as above} & Unitless, ordinal, non-linear scale from 0 to
7. & Same as 7TK, but recorded at a higher resolution & Mdir \\
Infrastructure Index & Infrastructure Index & Unitelss linear scale from
0 to 13.2 & Unitelss index ranging from from 0 to 13.2 & Erikstad \\
HIA & Human Impact area & Ordinal, non-linear scale from 1 to 4 & A
categorical representation of the infrastructure index & this paper \\
\end{longtable}

Reference to Figure~\ref{fig-carsplot}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{plot}\NormalTok{(cars)}
\end{Highlighting}
\end{Shaded}

\begin{figure}[H]

{\centering \includegraphics{HIAs_files/figure-pdf/fig-carsplot-1.pdf}

}

\caption{\label{fig-carsplot}Test caption}

\end{figure}

The data set is from a standardised field survey of nature types in
Norway that started in 2018 and which is still ongoing (REF). We
included data fram 2018 to 2023. In this survey, selected nature types
are delineatied on a map, and each locality is scored on a range of
variables relevant for describing the state and quality of nature. The
surveys are commissioned with the goal of producing data relevant for
immediate land-use decisions, and is therefore spatially biased,
typically towards areas with high human impact or expected impact. In
addition there is a thematic and size bias in the sampling protocol. Of
all possible mire types, the survey only maps the following:

\begin{itemize}
\item
  Southern ombrotrophic mires \textgreater{} 2500 m\textsuperscript{2}
\item
  Northern omgrotrophic \textgreater{} 10.000 m\textsuperscript{2}
\item
  All semi-natural mires (minerotrophic)
\item
  Calcareous southern fens \textgreater500 m\textsuperscript{2}
\item
  Calcareous northen fens \textgreater1000 m\textsuperscript{2}
\end{itemize}

In the above, \emph{southern} refers to boreonemoral and southboreal
zones, and \emph{northern} refers to mid-boreal, north boreal, and
alpine zones. In addition, the northern fens need to be even more
calcareous than the southern fens in order to be surveyed. In this paper
we assume the survey is representative for the entire mire ecosystem in
Norway. Although it is possible that smaller or less calcareous mires
will score systematically different than the ones that are surveyed we
do not think this is the case for our variables at least.

7FK was developed into a single indicator (i.e.~a normalised variable)
named \emph{Alien plant cover}. Variables 2-5 describe very related
aspects as so they where planned to be combined into one indicator
called antropogenic disturbance to soil and vegetation, or ADSV for
short. This was done by taking the value from the variable with the
lowest value, i.e.~worst condition. For both indicators, the lower and
upper reference values, i.e.~the worst and best possible condition that
the variables can be in, is given by the range of the data where 0 or 1
is the lower limit and 4, 7 or 8 is the upper limit, depending on the
variable (see Table~\ref{tbl-variables}).

The Alien plant cover was assigned to ECT class B1 - Compositional state
characteristics and ADSV was assigned to ECT class A1 - Physical state
characteristics (REF).

\begin{figure}

{\centering \includegraphics{../images/workflow.jpg}

}

\caption{\label{fig-workflow}Schematic workflow followed in this paper.
Sub-index A is calculated using spatially explicit aggregation (i.e .a
simple overlay takingthe mean value).}

\end{figure}

We chose three municipalities in Norway to test out the indicator (FIG).
These municipalities are spread over xxx km north to south, and also
differ in the amount of mire area, the total area surveyed, and the
prevalence of infrastructure (TABLE)

To

Validation/sensitivity analyses (varying HIA resolution and minimum
\emph{n}?)

Schematic GIS workflow, see Figure~\ref{fig-workflow}

Aggregation

\hypertarget{results}{%
\section{Results}\label{results}}

Data availability/coverage

Validation

National/regional indicator - do we have enough data?

Municipal case study

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

\hypertarget{credit-authorship-contribution-statement}{%
\section*{Credit authorship contribution
statement}\label{credit-authorship-contribution-statement}}
\addcontentsline{toc}{section}{Credit authorship contribution statement}

\hypertarget{declaration-of-competing-interest}{%
\section*{Declaration of Competing
Interest}\label{declaration-of-competing-interest}}
\addcontentsline{toc}{section}{Declaration of Competing Interest}

\hypertarget{acknowledgements}{%
\section*{Acknowledgements}\label{acknowledgements}}
\addcontentsline{toc}{section}{Acknowledgements}

\hypertarget{data-availability}{%
\section*{Data availability}\label{data-availability}}
\addcontentsline{toc}{section}{Data availability}

\renewcommand\refname{References}
  \bibliography{bibliography.bib}

\end{document}

### 2024-01-16 manuscript/template.qmd
move to quarto template - crossrefs ok

--- 2024-01-16 manuscript/HIAs.qmd
@@
 ---
-# Introduction
 
-Some general text about the decline in natural capital?
+Please make sure that your manuscript follows the guidelines in the 
+Guide for Authors of the relevant journal. It is not necessary to 
+typeset your manuscript in exactly the same way as an article, 
+unless you are submitting to a camera-ready copy (CRC) journal.
 
-Ecosystem condition accounting is the game of compiling relevant data on the status, trends and qualities of ecosystems (i.e. nature) and communicating this in a structured format.
-Its purpose is to make it easier to account for nature in policy by making the environmental costs of certain policies and practices visible to decision makers.
-As natural capital keeps declining all over the world, it is becoming increasingly urgent to make the message clear to decision makers about A statistical standard for ecosystem accounting, including ecosystem condition accounting, was developed by the UN and adopted by the UN Statistical Commission in 2021 and is called SEEA EA (System of Environmental-Economic Accounting - Ecosystem Accounting).
-The standard, or framework, is a set of rules, principles and best practices for compiling Ecosystem accounts, mainly aimed at national accounts.
+For detailed instructions regarding the elsevier article class, see   <https://www.elsevier.com/authors/policies-and-guidelines/latex-instructions>
 
-Central to ecosystem condition accounts are variables and indicators.
-These are parameters chosen to reflect the sentral condition characteristics of the ecosystems, and that can be quantified and ideally monitored over time to reflect the status and trends in condition.
-Indicators are (data) variables that are normalised (rescaled) against upper and lower reference values to become bound between the values 0 and 1.
-This normalisation ensures that indicators are more comparable because an indicator value of 1 will mean the same for all indicators, i.e.
-that the variable equals the upper reference value which again reflect the value of the variables under the reference condition.
-Similarly, a value of 0 mean that the variable is the worst possible state.
-The reference condition needs to be defined for each Ecosystem Condition Assessment separately, but SEEA EA gives some suggestion, such as an an ecosystem with no or minimal anthorpogenic disturbance.
+# Bibliography styles
 
-A general requirement for indicators in the SEEA EA framework is that they should give an unbiased representation of the condition inside the ecosystem assets (Czucz et al 2021; Table 1 - validity; see also SEEA EA §2.87) Ecosystem assets are defined as "ecological entities [meaning areas] about which information is sought and about which statistics are ultimately compiled (REF).
-This requirement for indicator validity means that spatially biased data are ill suited, especially if sampling intensity varies along gradient of anthropogenic pressures and hence ecosystem condition.
-SEEA EA is spatially explicit, which in practice means that indicators that are in some way sampled (i.e.
-not complete wall-to-wall data like remotely sensed imagery), the values are projected on to areas on the map, so that all areas inside the ecosystem accounting area get assigned value for that indicator.
-There are at least 3 general ways to achieve this complete areal coverage of indicator values:
+Here are two sample references:  @Feynman1963118 @Dirac1953888.
 
-a.  Using wall-to-wall data (e.g. remotely sensed data)
-b.  Predict values using a model (e.g. by accounting for environmental variation)
-c.  Simple projection of some best estimate, typically a central tendency from area representative data
+By default, natbib will be used with the `authoryear` style, set in `classoption` variable in YAML. 
+You can sets extra options with `natbiboptions` variable in YAML header. Example
 
-The need for an unbiased estimation of indicator values is unquestionable, but nonetheless, this requirement puts a large limitation on what types of data one can use.
-Ecosystem condition assessments are generally limited by data availability, and the choice of variables and indicators to include in assessments are more often or not a pragmatic and opportunistic one which is unlikely to reflect the full scope of the ecosystem condition characteristics.
-Note that the same is true for thematic biases.
-For example reflected in the scarcity of data included on insects or soil biota, even though most will agree they represent key ecosystem characteristics.
-Also having data from only one or a subset of nature types inside what is defined as the ecosystem in the assessment, is a typical thematic bias in ecosystem accounting.
-However, in this paper we chose to focus on spatial bias.
-
-Being able to make use of spatially biased data would greatly alleviate data shortage problems in ecosystem condition accounts.
-One way to achieve this is modelling (option b in the list above).
-Models can describe the general associations between the spatially sampled data and the context (e.g.
-the set of environmental variables) where it was samples, and use these arelationships to predict and project indicator values to areas that where not originally sampled.
-Depending on the data that goes into these models, they can be very reliable and make good indicators.
-This is especially true when the ecosystem assets are large (e.g. regions or nations).
-But when they are small, like the scale of a municipality, and when the indicator is more likely to be used as the evidence base in concrete physical land use planning, then the inherent lvel of uncertainty from such models becomes unacceptable.
-
-In this study we explore the potential for using a stratified aggregation technique to make use of spatially biased field data in ecosystem condition accounting.
-We demonstrate this technique using a generic GIS-based workflow for compiling ecosystem condition accounts that can be applied at any spatial scale, and we highlight the opportunities for local use-cases of this workflow by contrasting our findings across four neighboring municipalities in Norway.
-The main question is how much generalisation can we perform on the data we have before the resulting indicator looses its practical value in local governance processes.
-We end by interviewing end-users from the relevant municipalities about the perceived benefits and shortcomings of our condition indicators.
-
-# Material and Methods
-
-In this study we focused on open mire ecosystems in Norway.
-For this ecosystem we have an ecosystem delineation map produced using remotely sensed data and a deep learning model (REF Bakkestuen).
-This model, which has a 90.9% precision when tested against independent field data, estimates 12.7% of the area in southern Norway is mire (Bakkestuen REF).
-Mires are ecologically and socially important in Norway simply do to its large extent, and due to its role in climate mitigation, as mires store a large amount of carbon (REF?).
-There has not been a national assessment of the ecosystem condition of mires in Norway, but we recent contributed on a report which presented several new indicators that can be used in future assessment (REF Nybø, and web site).
-This study builds on the work in that report.
-
-We started by defining the reference state as one where ecosystems are subject to little or no human influence, with a climate as in the period 1961-1990 and a native species pool similar as today (i.e no mammoths allowed).
-We then identified two *ecosystem condition characteristics* that, along with many other characteristics that we do not specify here, define the typical behavior of open mire ecosystems in the reference condition: species composition, and; vegetation intactness.
-
-To describe these two concepts we use five condition variables from the same data set (see @tbl-variables.
-
-| Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                    | Reference  |
-|---------------|---------------|---------------|---------------|---------------|
-| 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                             | Halvorsen  |
-| 7SE                  | Human caused abration or abration-caused erosion | Unitless, ordinal, non-linear scale from 1 to 4.
-| Measures the frequency of imagined 4 m^2^ quadrats layed over the area that has some sign of abration          | ref        |
-| PRSL                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.
-| Same as 7SE, but recorded at a higher resolution                                                               | Mdir       |
-| 7TK                  | Tracks from large vehicles                       | Unitless, ordinal, non-linear scale from 1 to 4.
-| Measures the frequency of imagined 100 m^2^ quadrats layed over the area that has some signs of vehicle tracks | Halvorsen  |
-| PRTK                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.
-| Same as 7TK, but recorded at a higher resolution                                                               | Mdir       |
-| Infrastructure Index | Infrastructure Index                             | Unitelss linear scale from 0 to 13.2                                                    | Unitelss index ranging from from 0 to 13.2                                                                     | Erikstad   |
-| HIA                  | Human Impact area                                | Ordinal, non-linear scale from 1 to 4                                                   | A categorical representation of the infrastructure index                                                       | this paper |
-
-Table: Variables used in this study {#tbl-variables}
-
-Reference to @fig-carsplot
-
-```{r fig-carsplot, fig.cap="Test caption"}
-plot(cars)
+```
+natbiboptions: longnamesfirst,angle,semicolon
 ```
 
-The data set is from a standardised field survey of nature types in Norway that started in 2018 and which is still ongoing (REF).
-We included data fram 2018 to 2023.
-In this survey, selected nature types are delineatied on a map, and each locality is scored on a range of variables relevant for describing the state and quality of nature.
-The surveys are commissioned with the goal of producing data relevant for immediate land-use decisions, and is therefore spatially biased, typically towards areas with high human impact or expected impact.
-In addition there is a thematic and size bias in the sampling protocol.
-Of all possible mire types, the survey only maps the following:
+There are various more specific bibliography styles available at
+<https://support.stmdocs.in/wiki/index.php?title=Model-wise_bibliographic_style_files>. 
+To use one of these, add it in the header using, for example, `biblio-style: model1-num-names`.
 
--   Southern ombrotrophic mires \> 2500 m^2^
+## Using CSL
 
--   Northern omgrotrophic \> 10.000 m^2^
+If `cite-method` is set to `citeproc` in `elsevier_article()`, then pandoc is used for citations instead of `natbib`.
+In this case, the `csl` option is used to format the references.
+By default, this template will provide an appropriate style, but alternative `csl` files are available from <https://www.zotero.org/styles?q=elsevier>.
+These can be downloaded
+and stored locally, or the url can be used as in the example header.
 
--   All semi-natural mires (minerotrophic)
+# Equations
 
--   Calcareous southern fens \>500 m^2^
+Here is an equation:
+$$ 
+  f_{X}(x) = \left(\frac{\alpha}{\beta}\right)
+  \left(\frac{x}{\beta}\right)^{\alpha-1}
+  e^{-\left(\frac{x}{\beta}\right)^{\alpha}}; 
+  \alpha,\beta,x > 0 .
+$$
 
--   Calcareous northen fens \>1000 m^2^
+Inline equations work as well: $\sum_{i = 2}^\infty\{\alpha_i^\beta\}$
 
-In the above, *southern* refers to boreonemoral and southboreal zones, and *northern* refers to mid-boreal, north boreal, and alpine zones.
-In addition, the northern fens need to be even more calcareous than the southern fens in order to be surveyed.
-In this paper we assume the survey is representative for the entire mire ecosystem in Norway.
-Although it is possible that smaller or less calcareous mires will score systematically different than the ones that are surveyed we do not think this is the case for our variables at least.
+# Figures and tables
 
-7FK was developed into a single indicator (i.e. a normalised variable) named *Alien plant cover*.
-Variables 2-5 describe very related aspects as so they where planned to be combined into one indicator called antropogenic disturbance to soil and vegetation, or ADSV for short.
-This was done by taking the value from the variable with the lowest value, i.e. worst condition.
-For both indicators, the lower and upper reference values, i.e.
-the worst and best possible condition that the variables can be in, is given by the range of the data where 0 or 1 is the lower limit and 4, 7 or 8 is the upper limit, depending on the variable (see @tbl-variables).
+@fig-meaningless is generated using an R chunk.
 
-The Alien plant cover was assigned to ECT class B1 - Compositional state characteristics and ADSV was assigned to ECT class A1 - Physical state characteristics (REF).
+```{r}
+#| label: fig-meaningless
+#| fig-cap: A meaningless scatterplot
+#| fig-width: 5
+#| fig-height: 5
+#| fig-align: center
+#| out-width: 50%
+#| echo: false
+plot(runif(25), runif(25))
+```
 
-![Schematic workflow followed in this paper. Sub-index A is calculated using spatially explicit aggregation (i.e .a simple overlay takingthe mean value).](../images/workflow.jpg){#fig-workflow}
+# Tables coming from R
 
-We chose three municipalities in Norway to test out the indicator (FIG).
-These municipalities are spread over xxx km north to south, and also differ in the amount of mire area, the total area surveyed, and the prevalence of infrastructure (TABLE)
+Tables can also be generated using R chunks, as shown in @tbl-simple example.
 
-To
+```{r}
+#| label: tbl-simple
+#| tbl-cap: Caption centered above table
+#| echo: true
+knitr::kable(head(mtcars)[,1:4])
+```
 
-Validation/sensitivity analyses (varying HIA resolution and minimum *n*?)
-
-Schematic GIS workflow, see @fig-workflow
-
-Aggregation
-
-# Results
-
-Data availability/coverage
-
-Validation
-
-National/regional indicator - do we have enough data?
-
-Municipal case study
-
-# Discussion
-
-# Conclusion
-
-# Credit authorship contribution statement {.unnumbered}
-
-# Declaration of Competing Interest {.unnumbered}
-
-# Acknowledgements {.unnumbered}
-
-# Data availability {-}
-
-# References {.unnumbered}
+# References {-}

### 2024-01-16 R/tables.R

@@ -1,0 +1,16 @@
+variables <- read_delim("../../data/variables.csv", 
+                        delim = ";", escape_double = FALSE, locale = locale(encoding = "ISO-8859-1"), 
+                        trim_ws = TRUE)
+
+kbl(variables,
+    booktabs = T) %>%
+  row_spec(0, bold=T) %>%
+  kable_classic() %>%
+  column_spec(1, width = "1cm") %>%
+  column_spec(2, width = "5cm") %>%
+  column_spec(3, width = "20cm") %>%
+  column_spec(4, width = "40cm") %>%
+  column_spec(5, width = "5cm") %>%
+  footnote(general = "Variables PRSL and PRRK are originally on a scale from 0 to 7, but was transposed to lie between 1 and 8 to conform to the other variables.") %>%
+  save_kable("images/tbl_variables.jpg",
+             zoom=3)

### 2024-01-16 manuscript/HIAs.qmd
linenumbers

@@
       cite-style: super
+    include-in-header:
+      text: '\usepackage{lineno}\linenumbers'
 ---

### 2024-01-16 manuscript/HIAs.qmd
use defult citation style (numbers)

--- 2024-01-16 manuscript/HIAs.qmd
@@
       model: 3p
-      cite-style: super
+    include-in-header:
+      text: '\usepackage{lineno}\linenumbers'
+editor: 
+  markdown: 
+    wrap: sentence
 ---
+
 # Introduction
@@
 Some general text about the decline in natural capital?
+
+cite-style: super
 
@@
 
-A general requirement for indicators in the SEEA EA framework is that they should give an unbiased representation of the condition inside the ecosystem assets (Czucz et al 2021; Table 1 - validity; see also SEEA EA §2.87) Ecosystem assets are defined as "ecological entities [meaning areas] about which information is sought and about which statistics are ultimately compiled (REF).
+A general requirement for indicators in the SEEA EA framework is that they should give an unbiased representation of the condition inside the ecosystem assets (Czucz et al 2021; Table 1 - validity; see also SEEA EA §2.87) Ecosystem assets are defined as "ecological entities \[meaning areas\] about which information is sought and about which statistics are ultimately compiled (REF).
 This requirement for indicator validity means that spatially biased data are ill suited, especially if sampling intensity varies along gradient of anthropogenic pressures and hence ecosystem condition.
@@
 
-To describe these two concepts we use five condition variables from the same data set (see @tbl-variables.
+To describe these two concepts we use five condition variables from the same data set ( @tbl-variables).
 
@@
 |---------------|---------------|---------------|---------------|---------------|
-| 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                             | Halvorsen  |
+| 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                             | @rune_halvorsen_dokumentasjon_2019 |
 | 7SE                  | Human caused abration or abration-caused erosion | Unitless, ordinal, non-linear scale from 1 to 4.
@@
 
-Table: Variables used in this study {#tbl-variables}
-
-Reference to @fig-carsplot
-
-```{r fig-carsplot, fig.cap="Test caption"}
-plot(cars)
-```
+: Variables used in this study {#tbl-variables}
 
@@
 
-# Data availability {-}
+# Data availability {.unnumbered}
 

### 2024-01-16 manuscript/HIAs.qmd
add some refs

@@
       model: 3p
+      cite-style: authoryear
     include-in-header:
@@
 
-Some general text about the decline in natural capital?
-
-cite-style: super
-
 Ecosystem condition accounting is the game of compiling relevant data on the status, trends and qualities of ecosystems (i.e. nature) and communicating this in a structured format.
 Its purpose is to make it easier to account for nature in policy by making the environmental costs of certain policies and practices visible to decision makers.
-As natural capital keeps declining all over the world, it is becoming increasingly urgent to make the message clear to decision makers about A statistical standard for ecosystem accounting, including ecosystem condition accounting, was developed by the UN and adopted by the UN Statistical Commission in 2021 and is called SEEA EA (System of Environmental-Economic Accounting - Ecosystem Accounting).
+As natural capital keeps declining all over the world, it is becoming increasingly urgent to make the message clear to decision makers about.
+A statistical standard for ecosystem accounting, including ecosystem condition accounting, was developed by the UN and adopted by the UN Statistical Commission in 2021 and is called SEEA EA (@unitednations2021; System of Environmental-Economic Accounting - Ecosystem Accounting).
 The standard, or framework, is a set of rules, principles and best practices for compiling Ecosystem accounts, mainly aimed at national accounts.
@@
 Central to ecosystem condition accounts are variables and indicators.
-These are parameters chosen to reflect the sentral condition characteristics of the ecosystems, and that can be quantified and ideally monitored over time to reflect the status and trends in condition.
+These are parameters chosen to reflect the central condition characteristics of the ecosystems, and that can be quantified and ideally monitored over time to reflect the status and trends in condition.
 Indicators are (data) variables that are normalised (rescaled) against upper and lower reference values to become bound between the values 0 and 1.
@@
 Similarly, a value of 0 mean that the variable is the worst possible state.
-The reference condition needs to be defined for each Ecosystem Condition Assessment separately, but SEEA EA gives some suggestion, such as an an ecosystem with no or minimal anthorpogenic disturbance.
+The reference condition needs to be defined for each Ecosystem Condition Assessment separately, but SEEA EA gives some suggestion, such as an an ecosystem with no or minimal anthropogenic disturbance.
 
-A general requirement for indicators in the SEEA EA framework is that they should give an unbiased representation of the condition inside the ecosystem assets (Czucz et al 2021; Table 1 - validity; see also SEEA EA §2.87) Ecosystem assets are defined as "ecological entities \[meaning areas\] about which information is sought and about which statistics are ultimately compiled (REF).
+A general requirement for indicators in the SEEA EA framework is that they should give an unbiased representation of the condition inside the ecosystem assets [@czucz_selection_2021, table 1; see also @unitednations2021 §2.87] Ecosystem assets are defined as "ecological entities \[meaning areas\] about which information is sought and about which statistics are ultimately compiled [@unitednations2021].
 This requirement for indicator validity means that spatially biased data are ill suited, especially if sampling intensity varies along gradient of anthropogenic pressures and hence ecosystem condition.
@@
 Models can describe the general associations between the spatially sampled data and the context (e.g.
-the set of environmental variables) where it was samples, and use these arelationships to predict and project indicator values to areas that where not originally sampled.
+the set of environmental variables) where it was samples, and use these relationships to predict and project indicator values to areas that where not originally sampled.
 Depending on the data that goes into these models, they can be very reliable and make good indicators.
 This is especially true when the ecosystem assets are large (e.g. regions or nations).
-But when they are small, like the scale of a municipality, and when the indicator is more likely to be used as the evidence base in concrete physical land use planning, then the inherent lvel of uncertainty from such models becomes unacceptable.
+But when they are small, like the scale of a municipality, and when the indicator is more likely to be used as the evidence base in concrete physical land use planning, then the inherent level of uncertainty from such models becomes unacceptable.
 
@@
 In this study we focused on open mire ecosystems in Norway.
-For this ecosystem we have an ecosystem delineation map produced using remotely sensed data and a deep learning model (REF Bakkestuen).
-This model, which has a 90.9% precision when tested against independent field data, estimates 12.7% of the area in southern Norway is mire (Bakkestuen REF).
-Mires are ecologically and socially important in Norway simply do to its large extent, and due to its role in climate mitigation, as mires store a large amount of carbon (REF?).
-There has not been a national assessment of the ecosystem condition of mires in Norway, but we recent contributed on a report which presented several new indicators that can be used in future assessment (REF Nybø, and web site).
+For this ecosystem we have an ecosystem delineation map produced using remotely sensed data and a deep learning model [@bakkestuen_delineation_2023].
+This model, which has a 90.9% precision when tested against independent field data, estimates 12.7% of the area in southern Norway is mire [@bakkestuen_delineation_2023].
+Mires are ecologically and socially important in Norway simply doe to its large extent, and due to its role in climate mitigation, as mires store a large amount of carbon (REF?).
+There has not been a national assessment of the ecosystem condition of mires in Norway, but we recent contributed on a report which presented several new indicators that can be used in future assessment [@nybo_indikatorer_2023; see also @kolstad2023], and web site).
 This study builds on the work in that report.
@@
 To describe these two concepts we use five condition variables from the same data set ( @tbl-variables).
-
-| Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                    | Reference                          |
-|---------------|---------------|---------------|---------------|---------------|
-| 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                             | @rune_halvorsen_dokumentasjon_2019 |
-| 7SE                  | Human caused abration or abration-caused erosion | Unitless, ordinal, non-linear scale from 1 to 4.
-| Measures the frequency of imagined 4 m^2^ quadrats layed over the area that has some sign of abration          | ref                                |
-| PRSL                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.
-| Same as 7SE, but recorded at a higher resolution                                                               | Mdir                               |
-| 7TK                  | Tracks from large vehicles                       | Unitless, ordinal, non-linear scale from 1 to 4.
-| Measures the frequency of imagined 100 m^2^ quadrats layed over the area that has some signs of vehicle tracks | Halvorsen                          |
-| PRTK                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.
-| Same as 7TK, but recorded at a higher resolution                                                               | Mdir                               |
-| Infrastructure Index | Infrastructure Index                             | Unitelss linear scale from 0 to 13.2                                                    | Unitelss index ranging from from 0 to 13.2                                                                     | Erikstad                           |
-| HIA                  | Human Impact area                                | Ordinal, non-linear scale from 1 to 4                                                   | A categorical representation of the infrastructure index                                                       | this paper                         |
-
-: Variables used in this study {#tbl-variables}
-
 The data set is from a standardised field survey of nature types in Norway that started in 2018 and which is still ongoing (REF).
@@
 
+| Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                    | Reference                          |
+|---------------|---------------|---------------|---------------|---------------|
+| 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                             | @runehalvorsen2019                 |
+| 7SE                  | Human caused abration or abration-caused erosion | Unitless, ordinal, non-linear scale from 1 to 4.
+| Measures the frequency of imagined 4 m^2^ quadrats layed over the area that has some sign of abration          | @runehalvorsen2019                 |
+| PRSL                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.
+| Same as 7SE, but recorded at a higher resolution                                                               | @miljødirektoratet                 |
+| 7TK                  | Tracks from large vehicles                       | Unitless, ordinal, non-linear scale from 1 to 4.
+| Measures the frequency of imagined 100 m^2^ quadrats layed over the area that has some signs of vehicle tracks | @rune_halvorsen_dokumentasjon_2019 |
+| PRTK                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.
+| Same as 7TK, but recorded at a higher resolution                                                               | @miljødirektoratet                 |
+| Infrastructure Index | Infrastructure Index                             | Unitelss linear scale from 0 to 13.2                                                    | Unitelss index ranging from from 0 to 13.2                                                                     | @erikstad_index_2023               |
+| HIA                  | Human Impact area                                | Ordinal, non-linear scale from 1 to 4                                                   | A categorical representation of the infrastructure index                                                       | this paper                         |
+
+: Variables used in this study {#tbl-variables}
+
 7FK was developed into a single indicator (i.e. a normalised variable) named *Alien plant cover*.
-Variables 2-5 describe very related aspects as so they where planned to be combined into one indicator called antropogenic disturbance to soil and vegetation, or ADSV for short.
+Variables 2-5 describe very related aspects and so they where combined into one indicator called antropogenic disturbance to soil and vegetation, or ADSV for short.
 This was done by taking the value from the variable with the lowest value, i.e. worst condition.
@@
 the worst and best possible condition that the variables can be in, is given by the range of the data where 0 or 1 is the lower limit and 4, 7 or 8 is the upper limit, depending on the variable (see @tbl-variables).
+See @fig-workflow for a schematic workflow for the indicator development.
 
-The Alien plant cover was assigned to ECT class B1 - Compositional state characteristics and ADSV was assigned to ECT class A1 - Physical state characteristics (REF).
+The Alien plant cover was assigned to ECT class B1 - Compositional state characteristics, and ADSV was assigned to ECT class A1 - Physical state characteristics [@czucz_common_2021].
 
@@
 Validation/sensitivity analyses (varying HIA resolution and minimum *n*?)
-
-Schematic GIS workflow, see @fig-workflow
 

### 2024-01-18 manuscript/HIAs.qmd
select relevant NTs

@@
     wrap: sentence
+execute:
+  echo: false
+  include: false
+  eval: false
 ---
+
+```{r setup, warning=F, message=F, include=F}
+library(tidyverse)
+library(knitr)
+library(sf)
+
+```
 
@@
 Mires are ecologically and socially important in Norway simply doe to its large extent, and due to its role in climate mitigation, as mires store a large amount of carbon (REF?).
-There has not been a national assessment of the ecosystem condition of mires in Norway, but we recent contributed on a report which presented several new indicators that can be used in future assessment [@nybo_indikatorer_2023; see also @kolstad2023], and web site).
+There has not been a national assessment of the ecosystem condition of mires in Norway, but we recent contributed on a report which presented several new indicators that can be used in future assessment [@nybo_indikatorer_2023; see also @kolstad2023]).
 This study builds on the work in that report.
@@
 To describe these two concepts we use five condition variables from the same data set ( @tbl-variables).
-The data set is from a standardised field survey of nature types in Norway that started in 2018 and which is still ongoing (REF).
-We included data fram 2018 to 2023.
-In this survey, selected nature types are delineatied on a map, and each locality is scored on a range of variables relevant for describing the state and quality of nature.
+The data set is from a standardised field survey of nature types in Norway that started in 2018 and which is still ongoing.
+The data is freely available [@norwegian_environmental_agency_naturtyper_2024].
+See @miljødirektoratet for the sampling protocol (in Norwegian).
+We included data from 2018 to 2023.
+In this survey, selected nature types are delineated on a map, and each locality is scored on a range of variables relevant for describing the state and quality of nature.
 The surveys are commissioned with the goal of producing data relevant for immediate land-use decisions, and is therefore spatially biased, typically towards areas with high human impact or expected impact.
@@
 
--   Northern omgrotrophic \> 10.000 m^2^
+-   Northern ombrotrophic mires \> 10.000 m^2^
 
--   All semi-natural mires (minerotrophic)
+-   All semi-natural mires (minerotrophic fens)
 
@@
 
--   Calcareous northen fens \>1000 m^2^
+-   Calcareous northern fens \>1000 m^2^
 
-In the above, *southern* refers to boreonemoral and southboreal zones, and *northern* refers to mid-boreal, north boreal, and alpine zones.
-In addition, the northern fens need to be even more calcareous than the southern fens in order to be surveyed.
+In the above, *southern* refers to boreonemoral and southboreal zones, and *northern* refers to mid-boreal, north boreal, and alpine zones @asbjørnmoen1998 .
+In addition, the northern fens need to be even more calcareous than the southern fens in order to be surveyed, and the variable 7FK is not recorded in ombrotrophic mires.
 In this paper we assume the survey is representative for the entire mire ecosystem in Norway.
 Although it is possible that smaller or less calcareous mires will score systematically different than the ones that are surveyed we do not think this is the case for our variables at least.
+It may also be that alien plants are more or less common on bogs relative to fens.
+For this we have no assumption.
 
 | Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                    | Reference                          |
-|---------------|---------------|---------------|---------------|---------------|
+|--------------|--------------|--------------|----------------|--------------|
 | 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                             | @runehalvorsen2019                 |
@@
 
-![Schematic workflow followed in this paper. Sub-index A is calculated using spatially explicit aggregation (i.e .a simple overlay takingthe mean value).](../images/workflow.jpg){#fig-workflow}
+![Schematic workflow followed in this paper. Sub-index A is calculated using spatially explicit aggregation (i.e .a simple overlay taking the mean value).](../images/workflow.jpg){#fig-workflow}
 
@@
 
-To
+```{r filepath}
+# Set up conditional file paths
+# The data is too big to be stored on GitHub
+dir <- substr(getwd(), 1,2)
+
+path <- ifelse(dir == "C:", 
+               "R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb",
+               "/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb")
+
+#pData <- ifelse(dir == "C:", 
+#               "P:/41201785_okologisk_tilstand_2022_2023/data/",
+#               "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/")
+```
+
+```{r}
+# I already did some work to identify the relevant nature types
+
+# summary file (https://github.com/NINAnor/ecosystemCondition/blob/main/data/naturetypes/natureType_summary.rds)
+naturetypes_summary <- readRDS("../data/natureType_summary.rds")
+
+# Getting the nature types that have 7FA
+myVars <- "7FA"
+naturetypes_summary2 <- naturetypes_summary %>%
+  rowwise() %>%
+  mutate(keepers = sum(c_across(
+    all_of(myVars))>0, na.rm=T)) %>%
+  filter(keepers >0,
+         Ecosystem == "våtmark") %>%
+  select(Nature_type, NiN_mainType, Year)
+
+naturetypes_summary2$Nature_type
+
+# Getting the nature types that have the ADSV variables
+myVars <- c('7TK', '7SE', 'PRTK', 'PRSL')
+naturetypes_summary3 <- naturetypes_summary %>%
+  rowwise() %>%
+  mutate(keepers = sum(c_across(
+    all_of(myVars))>0, na.rm=T)) %>%
+  filter(keepers >0,
+         Ecosystem == "våtmark") %>%
+  select(Nature_type, NiN_mainType, Year)
+
+naturetypes_summary3$Nature_type
+```
+
+```{r relevantNTs}
+# These are the nature types I'm keeping for 7FA.
+# Note the absence of ombrotriphic types
+alienNTS <- c("Slåttemyr",
+              "Semi-naturlig våteng",
+              "Rik åpen jordvannsmyr i mellomboreal sone",
+              "Rik åpen sørlig jordvannsmyr",
+              "Rik åpen jordvannsmyr i nordboreal og lavalpin sone")
+
+# And these are the nature types that we use for the ADSV indicator
+ advsNTs <- c(# fens
+              "Slåttemyr",
+              "Sørlig slåttemyr",
+              "Rik åpen jordvannsmyr i mellomboreal sone",
+              "Rik åpen sørlig jordvannsmyr",
+              "Semi-naturlig myr",
+              "Rik åpen jordvannsmyr i nordboreal og lavalpin sone",
+              #"Sørlig kaldkilde", # excluded springs
+
+              # bogs (mainly)
+              "Høgereligende og nordlig nedbørsmyr",
+              "Øyblandingsmyr",
+              "Konsentrisk høymyr",
+              "Sørlig nedbørsmyr",
+              "Atlantisk høymyr",
+              "Terrengdekkende myr",
+              "Eksentrisk høymyr",
+              "Kanthøymyr",
+              "Platåhøymyr",
+              "Palsmyr")
+
+```
+
+```{r muniMaps, include=F, eval=F}
+naturetypes <- sf::st_read(dsn = path) %>%
+  # keep only wetlands
+  filter(hoved.kosystem == "våtmark") %>%
+
+  filter(naturtype %in% naturetypes_summary$Nature_type) %>%
+  mutate(hovedøkosystem = case_match(hovedøkosystem,
+                                  "naturligÅpneOmråderILavlandet" ~ "naturligÅpneOmråderUnderSkoggrensa",
+                                  .default = hovedøkosystem),
+         area = st_area(.)) %>%
+  separate_rows(ninBeskrivelsesvariable, sep=",") %>%
+  separate(col = ninBeskrivelsesvariable,
+           into = c("NiN_variable_code", "NiN_variable_value"),
+           sep = "_",
+           remove=F) %>%
+  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) %>%
+  filter(NiN_variable_code %in% myVars)
+
+temp <- naturetypes %>%
+    filter(hoved.kosystem == "våtmark")
+
+```
 

### 2024-01-18 manuscript/HIAs.qmd
try to crop and mask EDM

@@
     wrap: sentence
+# remove this execute part to actually run the analysis
 execute:
@@
 library(sf)
-
+library(tmap)
+library(stars)
+library(terra)
 ```
@@
 
-```{r filepath}
-# Set up conditional file paths
-# The data is too big to be stored on GitHub
-dir <- substr(getwd(), 1,2)
-
-path <- ifelse(dir == "C:", 
-               "R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb",
-               "/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb")
-
-#pData <- ifelse(dir == "C:", 
-#               "P:/41201785_okologisk_tilstand_2022_2023/data/",
-#               "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/")
-```
-
-```{r}
+```{r getRelevantNTs}
 # I already did some work to identify the relevant nature types
@@
 myVars <- "7FA"
-naturetypes_summary2 <- naturetypes_summary %>%
-  rowwise() %>%
-  mutate(keepers = sum(c_across(
-    all_of(myVars))>0, na.rm=T)) %>%
-  filter(keepers >0,
-         Ecosystem == "våtmark") %>%
-  select(Nature_type, NiN_mainType, Year)
 
-naturetypes_summary2$Nature_type
+#naturetypes_summary2 <- naturetypes_summary %>%
+#  rowwise() %>%
+#  mutate(keepers = sum(c_across(
+#    all_of(myVars))>0, na.rm=T)) %>%
+#  filter(keepers >0,
+#         Ecosystem == "våtmark") %>%
+#  select(Nature_type, NiN_mainType, Year)
+#
+#naturetypes_summary2$Nature_type
 
 # Getting the nature types that have the ADSV variables
-myVars <- c('7TK', '7SE', 'PRTK', 'PRSL')
-naturetypes_summary3 <- naturetypes_summary %>%
-  rowwise() %>%
-  mutate(keepers = sum(c_across(
-    all_of(myVars))>0, na.rm=T)) %>%
-  filter(keepers >0,
-         Ecosystem == "våtmark") %>%
-  select(Nature_type, NiN_mainType, Year)
-
-naturetypes_summary3$Nature_type
+myVars2 <- c('7TK', '7SE', 'PRTK', 'PRSL')
+#naturetypes_summary3 <- naturetypes_summary %>%
+#  rowwise() %>%
+#  mutate(keepers = sum(c_across(
+#    all_of(myVars2))>0, na.rm=T)) %>%
+#  filter(keepers >0,
+#         Ecosystem == "våtmark") %>%
+#  select(Nature_type, NiN_mainType, Year)
+#
+#naturetypes_summary3$Nature_type
 ```
@@
 
-```{r muniMaps, include=F, eval=F}
-naturetypes <- sf::st_read(dsn = path) %>%
-  # keep only wetlands
-  filter(hoved.kosystem == "våtmark") %>%
+```{r prepData}
+# Set up conditional file paths
+# The data is too big to be stored on GitHub
+dir <- substr(getwd(), 1,2)
 
-  filter(naturtype %in% naturetypes_summary$Nature_type) %>%
-  mutate(hovedøkosystem = case_match(hovedøkosystem,
-                                  "naturligÅpneOmråderILavlandet" ~ "naturligÅpneOmråderUnderSkoggrensa",
-                                  .default = hovedøkosystem),
-         area = st_area(.)) %>%
-  separate_rows(ninBeskrivelsesvariable, sep=",") %>%
+path <- ifelse(dir == "C:", 
+               "R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb",
+               "/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb")
+
+#pData <- ifelse(dir == "C:", 
+#               "P:/41201785_okologisk_tilstand_2022_2023/data/",
+#               "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/")
+
+# Impart polygon data set
+naturetypes <- sf::st_read(dsn = path)
+# 117k polygons (2023)
+# sometimes this functin struggles with Norwegian letters. Restart R and try again.
+
+naturetypes <- naturetypes |>
+  # keep only wetlands and our selected nature types
+  filter(hovedøkosystem == "våtmark",
+         naturtype %in% c(advsNTs, alienNTS)) |>
+  # calculate the areas (m2) of the polygons
+  mutate(area = SHAPE |> st_area()) |>
+  # the variable codes and values are all in the same column
+  separate_rows(ninBeskrivelsesvariable, sep=",") |>
   separate(col = ninBeskrivelsesvariable,
@@
            sep = "_",
-           remove=F) %>%
-  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) %>%
-  filter(NiN_variable_code %in% myVars)
+           remove=F) |>
+  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) |>
+  filter(NiN_variable_code %in% c(myVars, myVars2)) |>
+  select(id = identifikasjon_lokalId,
+         municipality = kommunenummer,
+         year = kartleggingsår,
+         mosaic = mosaikk,
+         quality = lokalitetskvalitet,
+         biodiversity = naturmangfold,
+         condition = tilstand,
+         natureType = naturtype,
+         variable = NiN_variable_code,
+         value = NiN_variable_value,
+         area)
 
-temp <- naturetypes %>%
-    filter(hoved.kosystem == "våtmark")
+# 13k obs.
+#naturetypes |>
+#  saveRDS("../data/surveyData.RDS")
+#head(naturetypes)
+```
+
+```{r getNorwayOutline}
+outline <- sf::read_sf("../data/outlineOfNorway_EPSG25833.shp")
+```
+
+```{r getMunicipalityOutine}
+path <- paste0("../data/Basisdata_0000_Norge_25833_Kommuner_FGDB.gdb")
+
+# find the correct layer
+#st_layers(path)
+muni <- sf::read_sf(path, options = "ENCODING=UTF8", layer = "kommune")
+#st_crs(muni) #25833
+
+# This returns some multisurfaces as well
+table(st_geometry_type(muni))
+
+# function to make sure that mutipolygons are returned 
+ensure_multipolygons <- function(X) {
+    tmp1 <- tempfile(fileext = ".gpkg")
+    tmp2 <- tempfile(fileext = ".gpkg")
+    st_write(X, tmp1)
+    gdalUtilities::ogr2ogr(tmp1, tmp2, f = "GPKG", nlt = "MULTIPOLYGON")
+    Y <- st_read(tmp2)
+    st_sf(st_drop_geometry(X), geom = st_geometry(Y))
+}
+
+muni <- ensure_multipolygons(muni)
+#> Writing layer `file34a3261c663332' to data source 
+#>   `/tmp/RtmpkT1dSD/file34a3261c663332.gpkg' using driver `GPKG'
+#> Writing 363 features with 13 fields and geometry type Unknown (any).
+#> Reading layer `file34a3261c663332' from data source 
+#>   `/tmp/RtmpkT1dSD/file34a32673f6fef.gpkg' 
+#>   using driver `GPKG'
+#> Simple feature collection with 363 features and 13 fields
+#> Geometry type: MULTIPOLYGON
+#> Dimension:     XY
+#> Bounding box:  xmin: -99551.21 ymin: 6426048 xmax: 1121941 ymax: 7962744
+#> Projected CRS: ETRS89 / UTM zone 33N
+table(st_geometry_type(muni)) #OK
+
+# code to show all municipalities
+#tm_shape(muni)+
+#  tm_polygons(col="kommunenummer")+
+#  tm_layout(legend.outside = T)
+
+# Transform to match the EDM (see below)
+muni <- muni |>
+  st_transform(25832)
+
+# subset of the three taregt municipalities
+muni3 <- muni |>
+  filter(kommunenummer %in% c("3020", #Nordre Follo
+                              "5001", # Trondheim
+                              "5418"  # Målselv
+                              ))
+# To crop EDM, I need the three municipalities seprately. 
+nf <- muni3 |>
+  filter(kommunenummer == "3020")
+tr <- muni3 |>
+  filter(kommunenummer == "5001")
+me <- muni3 |>
+  filter(kommunenummer == "5418")
+
+# Make map to show where the three municipalities are
+#
+#
+
+# I need to intersect nature types with municipality
+# and also to get the dat coverage polygon
+tm_shape(muni3)+
+  tm_polygons()+
+  tm_facets(by = "kommunenummer")+
+  tm_shape(naturetypes)+
+  tm_polygons()
+```
+
+```{r getMireData}
+file <- "P:/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif"
+# This is a big TIF, so I some lazy evalyation on the starts proxy 
+nf_mire  <- stars::read_stars(file, proxy=T) |>
+  st_crop(nf) |>
+  st_as_stars()
+
+# Same result, different syntax:
+#nf_mire  <- stars::read_stars(file, proxy=T)
+#nf_mire <- nf_mire[nf]
+#nf_mire <- st_as_stars(nf_mire)
+
+# This is not working as it puts the cropped raster into the sea
+#tmap_mode("view")
+#tm_shape(st_downsample(nf_mire , 10))+
+#  tm_raster(alpha=.8
+#            )+
+#  tm_shape(outline)+
+#  tm_borders()
+
+# trying terra
+# convert municipal outline to vect via st
+nf2 <- as(nf, "Spatial") |>
+  terra::vect()
+
+# crop and mask (very fast!)
+nf_mire2 <- terra::rast(file) |>
+  crop(nf2) |>
+  terra::mask(nf2)
+
+# Plot to check overlap
+ggplot()+
+  geom_spatraster(data = nf_mire2)+
+  geom_spatvector(data = nf2,
+                  fill = NA)
+# The cropping and masking worked.
+
+# I ilke the stars, sf and tmap combo better, so I return to stars
+nf_mire3 <- nf_mire2 |>
+  st_as_stars()
+
+tmap_mode("plot")
+tm_shape(st_downsample(nf_mire3 , 10))+
+  tm_raster(alpha=.8
+            )+
+  tm_shape(outline)+
+  tm_borders()
+
+# Then I cut out the relevant part of the ecosystem map
+nft <- nf |>
+  terra::vect()
+
+ras_crop <- terra::crop(ras_raw, shp)
+ras_mask <- terra::mask(ras_crop, shp, touches = T)
+
+nf_mire <- st_crop(EDM, st_bbox(nf))
+tr_mire <- st_crop(EDM, tr)
+me_mire <- st_crop(EDM, me)
+
+  tm_shape(nf)+
+  tm_borders(col="red", 
+             lwd=3)+
+tm_shape(st_downsample(nf_mire, 10))+
+  tm_raster(col="myrmodell90pros.tif")
+
+#tm_shape(st_downsample(EDM, 100))+
+  #tm_raster(alpha=.3)+
 

### 2024-01-19 manuscript/HIAs.qmd
kart

@@
 library(terra)
+```
+
+```{r getRelevantNTs}
+# I already did some work to identify the relevant nature types
+
+# summary file (https://github.com/NINAnor/ecosystemCondition/blob/main/data/naturetypes/natureType_summary.rds)
+naturetypes_summary <- readRDS("../data/natureType_summary.rds")
+
+# Getting the nature types that have 7FA
+myVars <- "7FA"
+
+#naturetypes_summary2 <- naturetypes_summary %>%
+#  rowwise() %>%
+#  mutate(keepers = sum(c_across(
+#    all_of(myVars))>0, na.rm=T)) %>%
+#  filter(keepers >0,
+#         Ecosystem == "våtmark") %>%
+#  select(Nature_type, NiN_mainType, Year)
+#
+#naturetypes_summary2$Nature_type
+
+# Getting the nature types that have the ADSV variables
+myVars2 <- c('7TK', '7SE', 'PRTK', 'PRSL')
+#naturetypes_summary3 <- naturetypes_summary %>%
+#  rowwise() %>%
+#  mutate(keepers = sum(c_across(
+#    all_of(myVars2))>0, na.rm=T)) %>%
+#  filter(keepers >0,
+#         Ecosystem == "våtmark") %>%
+#  select(Nature_type, NiN_mainType, Year)
+#
+#naturetypes_summary3$Nature_type
+```
+
+```{r relevantNTs}
+# These are the nature types I'm keeping for 7FA.
+# Note the absence of ombrotriphic types
+alienNTS <- c("Slåttemyr",
+              "Semi-naturlig våteng",
+              "Rik åpen jordvannsmyr i mellomboreal sone",
+              "Rik åpen sørlig jordvannsmyr",
+              "Rik åpen jordvannsmyr i nordboreal og lavalpin sone")
+
+# And these are the nature types that we use for the ADSV indicator
+ advsNTs <- c(# fens
+              "Slåttemyr",
+              "Sørlig slåttemyr",
+              "Rik åpen jordvannsmyr i mellomboreal sone",
+              "Rik åpen sørlig jordvannsmyr",
+              "Semi-naturlig myr",
+              "Rik åpen jordvannsmyr i nordboreal og lavalpin sone",
+              #"Sørlig kaldkilde", # excluded springs
+
+              # bogs (mainly)
+              "Høgereligende og nordlig nedbørsmyr",
+              "Øyblandingsmyr",
+              "Konsentrisk høymyr",
+              "Sørlig nedbørsmyr",
+              "Atlantisk høymyr",
+              "Terrengdekkende myr",
+              "Eksentrisk høymyr",
+              "Kanthøymyr",
+              "Platåhøymyr",
+              "Palsmyr")
+
+```
+
+```{r prepData, eval=F}
+# Set up conditional file paths
+# The data is too big to be stored on GitHub
+dir <- substr(getwd(), 1,2)
+
+path <- ifelse(dir == "C:", 
+               "R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb",
+               "/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb")
+
+#pData <- ifelse(dir == "C:", 
+#               "P:/41201785_okologisk_tilstand_2022_2023/data/",
+#               "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/")
+
+# Impart polygon data set
+naturetypes <- sf::st_read(dsn = path)
+# 117k polygons (2023)
+# sometimes this functin struggles with Norwegian letters. Restart R and try again.
+
+naturetypes <- naturetypes |>
+  # keep only wetlands and our selected nature types
+  filter(hovedøkosystem == "våtmark",
+         naturtype %in% c(advsNTs, alienNTS)) |>
+  # calculate the areas (m2) of the polygons
+  mutate(area = SHAPE |> st_area()) |>
+  # the variable codes and values are all in the same column
+  separate_rows(ninBeskrivelsesvariable, sep=",") |>
+  separate(col = ninBeskrivelsesvariable,
+           into = c("NiN_variable_code", "NiN_variable_value"),
+           sep = "_",
+           remove=F) |>
+  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) |>
+  filter(NiN_variable_code %in% c(myVars, myVars2)) |>
+  select(id = identifikasjon_lokalId,
+         municipality = kommunenummer,
+         year = kartleggingsår,
+         mosaic = mosaikk,
+         quality = lokalitetskvalitet,
+         biodiversity = naturmangfold,
+         condition = tilstand,
+         natureType = naturtype,
+         variable = NiN_variable_code,
+         value = NiN_variable_value,
+         area) |>
+  st_transform(25832)
+
+# 13k obs.
+#naturetypes |>
+#  saveRDS("../data/surveyData.RDS")
+#head(naturetypes)
+```
+
+```{r}
+naturetypes <- readRDS("../data/surveyData.RDS")
+```
+
+```{r getNorwayOutline}
+outline <- sf::read_sf("../data/outlineOfNorway_EPSG25833.shp") |>
+  st_transform(25832)
+```
+
+```{r getMunicipalityOutine}
+path <- paste0("../data/Basisdata_0000_Norge_25833_Kommuner_FGDB.gdb")
+
+# find the correct layer
+#st_layers(path)
+muni <- sf::read_sf(path, layer = "kommune") |>
+  st_transform(25832)
+
+# This returns some multisurfaces as well
+#table(st_geometry_type(muni))
+
+# function to make sure that mutipolygons are returned 
+ensure_multipolygons <- function(X) {
+    tmp1 <- tempfile(fileext = ".gpkg")
+    tmp2 <- tempfile(fileext = ".gpkg")
+    st_write(X, tmp1)
+    gdalUtilities::ogr2ogr(tmp1, tmp2, f = "GPKG", nlt = "MULTIPOLYGON")
+    Y <- st_read(tmp2)
+    st_sf(st_drop_geometry(X), geom = st_geometry(Y))
+}
+
+muni <- ensure_multipolygons(muni)
+#table(st_geometry_type(muni)) #OK
+
+# subset of the three target municipalities
+muni3 <- muni |>
+  filter(kommunenummer %in% c("3020", #Nordre Follo
+                              "5001", # Trondheim
+                              "5418"  # Målselv
+                              )) |>
+  mutate(Municipality = case_when(
+    kommunenummer == "3020" ~ "Nordre Follo",
+    kommunenummer == "5001" ~ "Trondheim",
+    kommunenummer == "5418" ~ "Målselv"
+  ))
+# To crop EDM, I need the three municipalities seprately. 
+nf <- muni3 |>
+  filter(kommunenummer == "3020")
+tr <- muni3 |>
+  filter(kommunenummer == "5001")
+me <- muni3 |>
+  filter(kommunenummer == "5418")
+
+# Make map to show where the three municipalities are
+#
+#
+
+# I need to intersect nature types with municipality
+nature3 <- naturetypes |>
+  st_intersection(muni3)
+
+# and also to get the data coverage polygon
+dk <-   read_sf("../data/dekningskartNF/Naturtyper NiN - Dekningskart.shp") |>
+  rbind(read_sf("../data/dekningskartTR/Naturtyper NiN - Dekningskart.shp")) |>
+  rbind(read_sf("../data/dekningskartME/Naturtyper NiN - Dekningskart.shp")) |>
+  st_transform(st_crs(muni3)) |>
+  st_intersection(muni3) |>
+  st_union()
+
+outline2 <- outline |>
+  st_intersection(muni3)
+
+ocean <- muni3 |>
+  st_difference(outline)
+
+tm_shape(muni3)+
+    tm_borders()+
+  tm_facets(by = "Municipality")+
+  tm_shape(outline2)+
+    tm_fill(col = "lightgreen",
+             alpha=.4)+
+  tm_shape(ocean)+
+    tm_polygons(col = "lightblue",
+                border.col = "black",
+            alpha=1)+
+  #tm_shape(muni3)+
+  #  tm_borders()+
+  tm_shape(dk)+
+    tm_polygons(col = "grey",
+                alpha=.8)+
+  tm_shape(nature3)+
+  tm_polygons(col = "red",
+              border.col = "red",
+              lwd=3)
+
+```
+
+```{r getMireData}
+file <- "P:/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif"
+# This is a big TIF, so I some lazy evalyation on the starts proxy 
+nf_mire  <- stars::read_stars(file, proxy=T) |>
+  st_crop(nf) |>
+  st_as_stars()
+
+# Same result, different syntax:
+#nf_mire  <- stars::read_stars(file, proxy=T)
+#nf_mire <- nf_mire[nf]
+#nf_mire <- st_as_stars(nf_mire)
+
+# This is not working as it puts the cropped raster into the sea
+#tmap_mode("view")
+#tm_shape(st_downsample(nf_mire , 10))+
+#  tm_raster(alpha=.8
+#            )+
+#  tm_shape(outline)+
+#  tm_borders()
+
+# trying terra
+# convert municipal outline to vect via st
+nf2 <- as(nf, "Spatial") |>
+  terra::vect()
+
+# crop and mask (very fast!)
+nf_mire2 <- terra::rast(file) |>
+  crop(nf2) |>
+  terra::mask(nf2)
+
+# Plot to check overlap
+ggplot()+
+  geom_spatraster(data = nf_mire2)+
+  geom_spatvector(data = nf2,
+                  fill = NA)
+# The cropping and masking worked.
+
+# I ilke the stars, sf and tmap combo better, so I return to stars
+nf_mire3 <- nf_mire2 |>
+  st_as_stars()
+
+tmap_mode("plot")
+tm_shape(st_downsample(nf_mire3 , 10))+
+  tm_raster(alpha=.8
+            )+
+  tm_shape(outline)+
+  tm_borders()
+
+# Then I cut out the relevant part of the ecosystem map
+nft <- nf |>
+  terra::vect()
+
+ras_crop <- terra::crop(ras_raw, shp)
+ras_mask <- terra::mask(ras_crop, shp, touches = T)
+
+nf_mire <- st_crop(EDM, st_bbox(nf))
+tr_mire <- st_crop(EDM, tr)
+me_mire <- st_crop(EDM, me)
+
+  tm_shape(nf)+
+  tm_borders(col="red", 
+             lwd=3)+
+tm_shape(st_downsample(nf_mire, 10))+
+  tm_raster(col="myrmodell90pros.tif")
+
+#tm_shape(st_downsample(EDM, 100))+
+  #tm_raster(alpha=.3)+
+
 ```
@@
 | Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                    | Reference                          |
-|--------------|--------------|--------------|----------------|--------------|
+|---------------|---------------|---------------|---------------|---------------|
 | 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                             | @runehalvorsen2019                 |
@@
 
-![Schematic workflow followed in this paper. Sub-index A is calculated using spatially explicit aggregation (i.e .a simple overlay taking the mean value).](../images/workflow.jpg){#fig-workflow}
+![Schematic workflow followed in this paper.
+Sub-index A is calculated using spatially explicit aggregation (i.e .a simple overlay taking the mean value).](../images/workflow.jpg){#fig-workflow alt="Schematic workflow followed in this paper.
+Sub-index A is calculated using spatially explicit aggregation (i.e .a simple overlay taking the mean value)."}
 
@@
 These municipalities are spread over xxx km north to south, and also differ in the amount of mire area, the total area surveyed, and the prevalence of infrastructure (TABLE)
-
-```{r getRelevantNTs}
-# I already did some work to identify the relevant nature types
-
-# summary file (https://github.com/NINAnor/ecosystemCondition/blob/main/data/naturetypes/natureType_summary.rds)
-naturetypes_summary <- readRDS("../data/natureType_summary.rds")
-
-# Getting the nature types that have 7FA
-myVars <- "7FA"
-
-#naturetypes_summary2 <- naturetypes_summary %>%
-#  rowwise() %>%
-#  mutate(keepers = sum(c_across(
-#    all_of(myVars))>0, na.rm=T)) %>%
-#  filter(keepers >0,
-#         Ecosystem == "våtmark") %>%
-#  select(Nature_type, NiN_mainType, Year)
-#
-#naturetypes_summary2$Nature_type
-
-# Getting the nature types that have the ADSV variables
-myVars2 <- c('7TK', '7SE', 'PRTK', 'PRSL')
-#naturetypes_summary3 <- naturetypes_summary %>%
-#  rowwise() %>%
-#  mutate(keepers = sum(c_across(
-#    all_of(myVars2))>0, na.rm=T)) %>%
-#  filter(keepers >0,
-#         Ecosystem == "våtmark") %>%
-#  select(Nature_type, NiN_mainType, Year)
-#
-#naturetypes_summary3$Nature_type
-```
-
-```{r relevantNTs}
-# These are the nature types I'm keeping for 7FA.
-# Note the absence of ombrotriphic types
-alienNTS <- c("Slåttemyr",
-              "Semi-naturlig våteng",
-              "Rik åpen jordvannsmyr i mellomboreal sone",
-              "Rik åpen sørlig jordvannsmyr",
-              "Rik åpen jordvannsmyr i nordboreal og lavalpin sone")
-
-# And these are the nature types that we use for the ADSV indicator
- advsNTs <- c(# fens
-              "Slåttemyr",
-              "Sørlig slåttemyr",
-              "Rik åpen jordvannsmyr i mellomboreal sone",
-              "Rik åpen sørlig jordvannsmyr",
-              "Semi-naturlig myr",
-              "Rik åpen jordvannsmyr i nordboreal og lavalpin sone",
-              #"Sørlig kaldkilde", # excluded springs
-
-              # bogs (mainly)
-              "Høgereligende og nordlig nedbørsmyr",
-              "Øyblandingsmyr",
-              "Konsentrisk høymyr",
-              "Sørlig nedbørsmyr",
-              "Atlantisk høymyr",
-              "Terrengdekkende myr",
-              "Eksentrisk høymyr",
-              "Kanthøymyr",
-              "Platåhøymyr",
-              "Palsmyr")
-
-```
-
-```{r prepData}
-# Set up conditional file paths
-# The data is too big to be stored on GitHub
-dir <- substr(getwd(), 1,2)
-
-path <- ifelse(dir == "C:", 
-               "R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb",
-               "/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb")
-
-#pData <- ifelse(dir == "C:", 
-#               "P:/41201785_okologisk_tilstand_2022_2023/data/",
-#               "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/")
-
-# Impart polygon data set
-naturetypes <- sf::st_read(dsn = path)
-# 117k polygons (2023)
-# sometimes this functin struggles with Norwegian letters. Restart R and try again.
-
-naturetypes <- naturetypes |>
-  # keep only wetlands and our selected nature types
-  filter(hovedøkosystem == "våtmark",
-         naturtype %in% c(advsNTs, alienNTS)) |>
-  # calculate the areas (m2) of the polygons
-  mutate(area = SHAPE |> st_area()) |>
-  # the variable codes and values are all in the same column
-  separate_rows(ninBeskrivelsesvariable, sep=",") |>
-  separate(col = ninBeskrivelsesvariable,
-           into = c("NiN_variable_code", "NiN_variable_value"),
-           sep = "_",
-           remove=F) |>
-  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) |>
-  filter(NiN_variable_code %in% c(myVars, myVars2)) |>
-  select(id = identifikasjon_lokalId,
-         municipality = kommunenummer,
-         year = kartleggingsår,
-         mosaic = mosaikk,
-         quality = lokalitetskvalitet,
-         biodiversity = naturmangfold,
-         condition = tilstand,
-         natureType = naturtype,
-         variable = NiN_variable_code,
-         value = NiN_variable_value,
-         area)
-
-# 13k obs.
-#naturetypes |>
-#  saveRDS("../data/surveyData.RDS")
-#head(naturetypes)
-```
-
-```{r getNorwayOutline}
-outline <- sf::read_sf("../data/outlineOfNorway_EPSG25833.shp")
-```
-
-```{r getMunicipalityOutine}
-path <- paste0("../data/Basisdata_0000_Norge_25833_Kommuner_FGDB.gdb")
-
-# find the correct layer
-#st_layers(path)
-muni <- sf::read_sf(path, options = "ENCODING=UTF8", layer = "kommune")
-#st_crs(muni) #25833
-
-# This returns some multisurfaces as well
-table(st_geometry_type(muni))
-
-# function to make sure that mutipolygons are returned 
-ensure_multipolygons <- function(X) {
-    tmp1 <- tempfile(fileext = ".gpkg")
-    tmp2 <- tempfile(fileext = ".gpkg")
-    st_write(X, tmp1)
-    gdalUtilities::ogr2ogr(tmp1, tmp2, f = "GPKG", nlt = "MULTIPOLYGON")
-    Y <- st_read(tmp2)
-    st_sf(st_drop_geometry(X), geom = st_geometry(Y))
-}
-
-muni <- ensure_multipolygons(muni)
-#> Writing layer `file34a3261c663332' to data source 
-#>   `/tmp/RtmpkT1dSD/file34a3261c663332.gpkg' using driver `GPKG'
-#> Writing 363 features with 13 fields and geometry type Unknown (any).
-#> Reading layer `file34a3261c663332' from data source 
-#>   `/tmp/RtmpkT1dSD/file34a32673f6fef.gpkg' 
-#>   using driver `GPKG'
-#> Simple feature collection with 363 features and 13 fields
-#> Geometry type: MULTIPOLYGON
-#> Dimension:     XY
-#> Bounding box:  xmin: -99551.21 ymin: 6426048 xmax: 1121941 ymax: 7962744
-#> Projected CRS: ETRS89 / UTM zone 33N
-table(st_geometry_type(muni)) #OK
-
-# code to show all municipalities
-#tm_shape(muni)+
-#  tm_polygons(col="kommunenummer")+
-#  tm_layout(legend.outside = T)
-
-# Transform to match the EDM (see below)
-muni <- muni |>
-  st_transform(25832)
-
-# subset of the three taregt municipalities
-muni3 <- muni |>
-  filter(kommunenummer %in% c("3020", #Nordre Follo
-                              "5001", # Trondheim
-                              "5418"  # Målselv
-                              ))
-# To crop EDM, I need the three municipalities seprately. 
-nf <- muni3 |>
-  filter(kommunenummer == "3020")
-tr <- muni3 |>
-  filter(kommunenummer == "5001")
-me <- muni3 |>
-  filter(kommunenummer == "5418")
-
-# Make map to show where the three municipalities are
-#
-#
-
-# I need to intersect nature types with municipality
-# and also to get the dat coverage polygon
-tm_shape(muni3)+
-  tm_polygons()+
-  tm_facets(by = "kommunenummer")+
-  tm_shape(naturetypes)+
-  tm_polygons()
-```
-
-```{r getMireData}
-file <- "P:/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif"
-# This is a big TIF, so I some lazy evalyation on the starts proxy 
-nf_mire  <- stars::read_stars(file, proxy=T) |>
-  st_crop(nf) |>
-  st_as_stars()
-
-# Same result, different syntax:
-#nf_mire  <- stars::read_stars(file, proxy=T)
-#nf_mire <- nf_mire[nf]
-#nf_mire <- st_as_stars(nf_mire)
-
-# This is not working as it puts the cropped raster into the sea
-#tmap_mode("view")
-#tm_shape(st_downsample(nf_mire , 10))+
-#  tm_raster(alpha=.8
-#            )+
-#  tm_shape(outline)+
-#  tm_borders()
-
-# trying terra
-# convert municipal outline to vect via st
-nf2 <- as(nf, "Spatial") |>
-  terra::vect()
-
-# crop and mask (very fast!)
-nf_mire2 <- terra::rast(file) |>
-  crop(nf2) |>
-  terra::mask(nf2)
-
-# Plot to check overlap
-ggplot()+
-  geom_spatraster(data = nf_mire2)+
-  geom_spatvector(data = nf2,
-                  fill = NA)
-# The cropping and masking worked.
-
-# I ilke the stars, sf and tmap combo better, so I return to stars
-nf_mire3 <- nf_mire2 |>
-  st_as_stars()
-
-tmap_mode("plot")
-tm_shape(st_downsample(nf_mire3 , 10))+
-  tm_raster(alpha=.8
-            )+
-  tm_shape(outline)+
-  tm_borders()
-
-# Then I cut out the relevant part of the ecosystem map
-nft <- nf |>
-  terra::vect()
-
-ras_crop <- terra::crop(ras_raw, shp)
-ras_mask <- terra::mask(ras_crop, shp, touches = T)
-
-nf_mire <- st_crop(EDM, st_bbox(nf))
-tr_mire <- st_crop(EDM, tr)
-me_mire <- st_crop(EDM, me)
-
-  tm_shape(nf)+
-  tm_borders(col="red", 
-             lwd=3)+
-tm_shape(st_downsample(nf_mire, 10))+
-  tm_raster(col="myrmodell90pros.tif")
-
-#tm_shape(st_downsample(EDM, 100))+
-  #tm_raster(alpha=.3)+
-
-```
 

### 2024-01-19 manuscript/HIAs.qmd
arrange municipality plot and outline

@@
 # Make map to show where the three municipalities are
-#
-#
+# First, get centroids
+centroids <- muni3 |>
+  st_centroid()
+
+positionMap <- tm_shape(outline)+
+  tm_polygons() +
+  tm_shape(centroids)+
+    tm_text(text = "Municipality",
+      size = 1,
+      xmod = 2,
+      ymod=-1)+
+  tm_symbols(size = 2,
+             shape = 17,
+             col="black")+
+  tm_grid(projection = 4326)+
+  tm_layout(bg.color = "skyblue")
 
@@
 
+# Simplified coastline
 outline2 <- outline |>
@@
 
+# Polygons for the oceans in each municipality
 ocean <- muni3 |>
@@
 
-tm_shape(muni3)+
+# Plot (could perhaps add the EDM here as well)
+muniPlot <- tm_shape(muni3)+
     tm_borders()+
-  tm_facets(by = "Municipality")+
+  tm_facets(by = "Municipality",
+            ncol=2)+
   tm_shape(outline2)+
@@
   tm_shape(ocean)+
-    tm_polygons(col = "lightblue",
-                border.col = "black",
-            alpha=1)+
-  #tm_shape(muni3)+
-  #  tm_borders()+
+    tm_polygons(col = "skyblue",
+                border.col = "black")+
   tm_shape(dk)+
@@
 
+tmap_arrange(positionMap, muniPlot,
+             widths = c(.4, .6))
 ```

### 2024-01-19 manuscript/HIAs.qmd
add figure to ms

@@
 ```{r setup, warning=F, message=F, include=F}
+#| warning: false
+#| message: false
+#| include: false
+#| eval: true
+
 library(tidyverse)
@@
 ```{r}
+# read cached rds file
 naturetypes <- readRDS("../data/surveyData.RDS")
@@
 
-tmap_arrange(positionMap, muniPlot,
+methodsMap <- tmap_arrange(positionMap, muniPlot,
              widths = c(.4, .6))
+
+tmap_save(methodsMap, "../figures/studyLocations.tiff",
+          dpi= 1000,
+          units = "cm",
+          width = 18,
+          height = 10)
+
+saveRDS(methodsMap, "../figures/studyLocation.RDS")
 ```
@@
 
-![Schematic workflow followed in this paper.
-Sub-index A is calculated using spatially explicit aggregation (i.e .a simple overlay taking the mean value).](../images/workflow.jpg){#fig-workflow alt="Schematic workflow followed in this paper.
-Sub-index A is calculated using spatially explicit aggregation (i.e .a simple overlay taking the mean value)."}
+![Schematic workflow followed in this paper. Sub-index A is calculated using spatially explicit aggregation (i.e .a simple overlay taking the mean value).](../images/workflow.jpg){#fig-workflow}
 
-We chose three municipalities in Norway to test out the indicator (FIG).
+We chose three municipalities in Norway to test out the indicator @fig-studyLocation.
 These municipalities are spread over xxx km north to south, and also differ in the amount of mire area, the total area surveyed, and the prevalence of infrastructure (TABLE)
+
+```{r fig-studyLocation}
+#| eval: true
+#| include: true 
+#| out.width: '100%'
+#| fig.cap: 
+#| - "Postion of the three example municipalities in
+#| Norway. For each municipality the map shows ocean in blue and non-ocean 
+#| in green. The survey coverage maps are in grey, and the mapped mires are 
+#| in red, with polygon borderes made extra thich to make them visible, 
+#| but then also exagerating their size."
+
+(temp <- readRDS("../figures/studyLocation.RDS"))
+```
 

### 2024-01-19 manuscript/HIAs.qmd
filling in area stats

@@
   eval: false
+  warning: false
+  message: false
 ---
@@
 
-```{r prepData, eval=F}
+```{r prepData}
+#| eval: false
+
 # Set up conditional file paths
@@
 
-```{r}
+```{r readSurvey}
+#| eval: true
+
 # read cached rds file
@@
 ```{r getNorwayOutline}
+#| eval: true
 outline <- sf::read_sf("../data/outlineOfNorway_EPSG25833.shp") |>
@@
 
-```{r getMunicipalityOutine}
+```{r prepPolygons}
+#| eval: true
+
 path <- paste0("../data/Basisdata_0000_Norge_25833_Kommuner_FGDB.gdb")
@@
 
-# function to make sure that mutipolygons are returned 
+# function to make sure that multipolygons are returned 
 ensure_multipolygons <- function(X) {
@@
   ))
+
 # To crop EDM, I need the three municipalities seprately. 
@@
 
+# I need to intersect nature types with municipality
+nature3 <- naturetypes |>
+  st_intersection(muni3)
+
+# and also to get the data coverage polygon
+dk <-   read_sf("../data/dekningskartNF/Naturtyper NiN - Dekningskart.shp") |>
+  rbind(read_sf("../data/dekningskartTR/Naturtyper NiN - Dekningskart.shp")) |>
+  rbind(read_sf("../data/dekningskartME/Naturtyper NiN - Dekningskart.shp")) |> 
+  st_transform(st_crs(muni3)) |>
+  st_intersection(muni3) |>
+  st_union()
+
+# Simplified coastline / terrestrial area
+terrestrial <- outline |>
+  st_intersection(muni3)
+
+# Polygons for the oceans in each municipality
+ocean <- muni3 |>
+  st_difference(outline)
+
+# claculate stats
+
+# terrestrial area
+terrestrial <- terrestrial |>
+  mutate(area_t = geometry |> st_area())
+
+```
+
+```{r makeMap}
 # Make map to show where the three municipalities are
@@
 
-# I need to intersect nature types with municipality
-nature3 <- naturetypes |>
-  st_intersection(muni3)
-
-# and also to get the data coverage polygon
-dk <-   read_sf("../data/dekningskartNF/Naturtyper NiN - Dekningskart.shp") |>
-  rbind(read_sf("../data/dekningskartTR/Naturtyper NiN - Dekningskart.shp")) |>
-  rbind(read_sf("../data/dekningskartME/Naturtyper NiN - Dekningskart.shp")) |>
-  st_transform(st_crs(muni3)) |>
-  st_intersection(muni3) |>
-  st_union()
-
-# Simplified coastline
-outline2 <- outline |>
-  st_intersection(muni3)
-
-# Polygons for the oceans in each municipality
-ocean <- muni3 |>
-  st_difference(outline)
-
 # Plot (could perhaps add the EDM here as well)
@@
             ncol=2)+
-  tm_shape(outline2)+
+  tm_shape(terrestrial)+
     tm_fill(col = "lightgreen",
@@
 # This is a big TIF, so I some lazy evalyation on the starts proxy 
-nf_mire  <- stars::read_stars(file, proxy=T) |>
-  st_crop(nf) |>
+tr2 <- tr |>
+  st_transform(25833)
+
+mire <- read_stars(file) |>
+  st_downsample(100) |>
+  st_as_stars()
+plot(mire, col="yellow")
+
+tr_mire  <- stars::read_stars(file, proxy=T) |>
+  st_crop(tr) |>
+  st_as_stars()
+#Not working
+
+# What is the EDM is actually utm 33
+tr_mire  <- stars::read_stars(file, proxy=T)
+st_crs(tr_mire) <- 25833
+tr_mire  <- tr_mire |>
+  st_crop(tr2) |>
   st_as_stars()
@@
 # This is not working as it puts the cropped raster into the sea
-#tmap_mode("view")
-#tm_shape(st_downsample(nf_mire , 10))+
-#  tm_raster(alpha=.8
-#            )+
-#  tm_shape(outline)+
-#  tm_borders()
+tmap_mode("view")
+tm_shape(st_downsample(tr_mire , 10))+
+  tm_raster(alpha=.8)+
+  tm_shape(outline)+
+  tm_borders()
 
@@
 
+```{r}
+#| eval: true
+
+# calculate terrestrial area
+terrestrial <- terrestrial |> 
+  mutate(t_area_km = 
+           round(units::drop_units(area_t * 1e-6))) |>
+  as_tibble()
+
+# calculate area of survey coverage maps
+dk2 <-   read_sf("../data/dekningskartNF/Naturtyper NiN - Dekningskart.shp") |>
+  rbind(read_sf("../data/dekningskartTR/Naturtyper NiN - Dekningskart.shp")) |>
+  rbind(read_sf("../data/dekningskartME/Naturtyper NiN - Dekningskart.shp")) |>
+  st_transform(st_crs(muni3)) |>
+  st_intersection(muni3) |>
+  st_cast("POLYGON") |>
+  group_by(Municipality) |>
+  summarise(geometry = st_union(geometry)) |>
+  mutate(dk_area_km = geometry |> st_area(),
+         dk_area_km = round(units::drop_units(dk_area_km * 1e-6)))
+
+muni_tbl <- muni3 |>
+  # calculate total area
+  mutate(area_km = 
+           round(
+             units::drop_units(
+               geom |> st_area() * 1e-6))) |>
+  # make tibble for the coming join
+  as_tibble() |>
+  # paste inn terrestrial area
+  left_join(terrestrial |> select(kommunenummer, t_area_km),
+            keep=F) |>
+  # area of survey
+  left_join(dk2 |> select(Municipality, dk_area_km)) |>
+  mutate(dk_percent = round((dk_area_km/t_area_km) * 100))
+
+```
+
 ```{r fig-studyLocation}
@@
 ```
+
+| Municipality | Total terrestrial area (km2)                                                 | \% of terrestrial area surveyed                                            | \% open mires in relation to total terrestrial area | Total mire area | Fraction mire area included in survey | Number of mire polygons in survey |
+|-----------|-----------|-----------|-----------|-----------|-----------|-----------|
+| Nordre Follo | `r terrestrial |> filter(Municipality == "Nordre Follo") |> pull(t_area_km)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(dk_percent)` |                                                     |                 |                                       |                                   |
+| Trondheim    | `r terrestrial |> filter(Municipality == "Trondheim") |> pull(t_area_km)`    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(dk_percent)`    |                                                     |                 |                                       |                                   |
+| Målselv      | `r terrestrial |> filter(Municipality == "Målselv") |> pull(t_area_km)`      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(dk_percent)`      |                                                     |                 |                                       |                                   |
+
+: Information for the three target municipalities in Norway
 

### 2024-01-19 manuscript/HIAs.qmd
number of polygons

@@
 
-```{r prepData}
+```{r natureTypeData}
 #| eval: false
@@
   st_intersection(muni3)
+
+table(nature3$municipality)
 
@@
 
+# number and area of mire polygons
+nature3_tbl <- nature3 |>
+  group_by(Municipality) |>
+  summarise(n = n()) |>
+  as_tibble()
+
 muni_tbl <- muni3 |>
@@
   left_join(dk2 |> select(Municipality, dk_area_km)) |>
-  mutate(dk_percent = round((dk_area_km/t_area_km) * 100))
+  mutate(dk_percent = round((dk_area_km/t_area_km) * 100)) |>
+  # number of polygons
+  left_join(nature3_tbl |> select(Municipality, n))
 
@@
 |-----------|-----------|-----------|-----------|-----------|-----------|-----------|
-| Nordre Follo | `r terrestrial |> filter(Municipality == "Nordre Follo") |> pull(t_area_km)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(dk_percent)` |                                                     |                 |                                       |                                   |
-| Trondheim    | `r terrestrial |> filter(Municipality == "Trondheim") |> pull(t_area_km)`    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(dk_percent)`    |                                                     |                 |                                       |                                   |
-| Målselv      | `r terrestrial |> filter(Municipality == "Målselv") |> pull(t_area_km)`      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(dk_percent)`      |                                                     |                 |                                       |                                   |
+| Nordre Follo | `r terrestrial |> filter(Municipality == "Nordre Follo") |> pull(t_area_km)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(dk_percent)`  |                                                     |                 |                                       | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(n)` |
+| Trondheim    | `r terrestrial |> filter(Municipality == "Trondheim") |> pull(t_area_km)`    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(dk_percent)`     |                                                     |                 |                                       | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(n)`    |
+| Målselv      | `r terrestrial |> filter(Municipality == "Målselv") |> pull(t_area_km)`      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(dk_percent)`       |                                                     |                 |                                       | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(n)`      |
 

### 2024-01-22 manuscript/HIAs.qmd
crop terra and summary table

--- 2024-01-19 manuscript/HIAs.qmd
@@
 library(terra)
+library(tidyterra)
+
+myCRS <- 25832
 ```
@@
          area) |>
-  st_transform(25832)
+  st_transform(myCRS) # Chosing this to match the EDM (se further down)
 
@@
 #| eval: true
+
 outline <- sf::read_sf("../data/outlineOfNorway_EPSG25833.shp") |>
-  st_transform(25832)
+  st_transform(myCRS)
 ```
@@
 muni <- sf::read_sf(path, layer = "kommune") |>
-  st_transform(25832)
+  st_transform(myCRS)
 
@@
 
+table(nature3$municipality)
+
 # and also to get the data coverage polygon
@@
 
-# claculate stats
-
+# calculate stats
 # terrestrial area
@@
 ```{r makeMap}
+#| eval: false
+
+tmap_mode("plot")
 # Make map to show where the three municipalities are
@@
 
-tmap_save(methodsMap, "../figures/studyLocations.tiff",
-          dpi= 1000,
-          units = "cm",
-          width = 18,
-          height = 10)
-
-saveRDS(methodsMap, "../figures/studyLocation.RDS")
+# tmap_save(methodsMap, "../figures/studyLocations.tiff",
+#           dpi= 1000,
+#           units = "cm",
+#           width = 18,
+#           height = 10)
+# 
+# saveRDS(methodsMap, "../figures/studyLocation.RDS")
 ```
 
-```{r getMireData}
+```{r mireDataTestCrop}
 file <- "P:/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif"
-# This is a big TIF, so I some lazy evalyation on the starts proxy 
-tr2 <- tr |>
-  st_transform(25833)
+# This is a big TIF!
 
-mire <- read_stars(file) |>
+mire_test <- read_stars(file, proxy = T) |>
+  setNames("mire") |>
   st_downsample(100) |>
-  st_as_stars()
-plot(mire, col="yellow")
-
-tr_mire  <- stars::read_stars(file, proxy=T) |>
-  st_crop(tr) |>
-  st_as_stars()
-#Not working
-
-# What is the EDM is actually utm 33
-tr_mire  <- stars::read_stars(file, proxy=T)
-st_crs(tr_mire) <- 25833
-tr_mire  <- tr_mire |>
-  st_crop(tr2) |>
+  mutate(mire = na_if(mire, 0)) |>
   st_as_stars()
 
-# Same result, different syntax:
-#nf_mire  <- stars::read_stars(file, proxy=T)
-#nf_mire <- nf_mire[nf]
-#nf_mire <- st_as_stars(nf_mire)
+mire_test_nf <-  mire_test |>
+  st_crop(nf)
 
-# This is not working as it puts the cropped raster into the sea
-tmap_mode("view")
-tm_shape(st_downsample(tr_mire , 10))+
-  tm_raster(alpha=.8)+
+# The cropping is correct
+tmap_mode("plot")
+tm_shape(mire_test)+
+  tm_raster(palette = "red")+
   tm_shape(outline)+
-  tm_borders()
+  tm_borders()+
+  tm_shape(mire_test_nf)+
+  tm_raster(col="red")
+```
 
+```{r mireNF-error}
+#| eval: false
+# Then we do the same, but without downsampling
+
+mire_nf <- read_stars(file, proxy = T) |>
+  setNames("mire") |>
+  #mutate(mire = na_if(mire, 0)) |>
+  st_crop(nf) |>
+  st_as_stars()
+
+# The cropped raster is for some reason far into the ocean. 
+# The cells have values, so it seems like there is something 
+# that happens after the cropping
+tmap_mode("plot")
+tm_shape(mire_test)+
+  tm_raster(palette = "red")+
+  tm_shape(outline)+
+  tm_borders()+
+  tm_shape(mire_nf)+
+  tm_raster(col="red")
+
+st_crs(mire_nf) <- 25832
+```
+
+```{r mireTerra}
 # trying terra
 # convert municipal outline to vect via st
-nf2 <- as(nf, "Spatial") |>
+nf_vect <- as(nf, "Spatial") |>
   terra::vect()
@@
 # crop and mask (very fast!)
-nf_mire2 <- terra::rast(file) |>
-  crop(nf2) |>
-  terra::mask(nf2)
+mire_nf_terra <- terra::rast(file) |>
+  crop(nf_vect) |>
+  terra::mask(nf_vect)
 
@@
 ggplot()+
-  geom_spatraster(data = nf_mire2)+
-  geom_spatvector(data = nf2,
+  geom_spatraster(data = mire_nf_terra)+
+  geom_spatvector(data = nf_vect,
                   fill = NA)
 # The cropping and masking worked. 
+#tmap_mode("plot")
+#tm_shape(mire_test)+
+#  tm_raster(palette = "red")+
+#  tm_shape(outline)+
+#  tm_borders()+
+#  tm_shape(mire_nf_terra)+
+#  tm_raster(col="red")
 
 # I ilke the stars, sf and tmap combo better, so I return to stars
-nf_mire3 <- nf_mire2 |>
+mire_nf_stars <- mire_nf_terra |>
   st_as_stars()
 
-tmap_mode("plot")
-tm_shape(st_downsample(nf_mire3 , 10))+
-  tm_raster(alpha=.8
-            )+
+tm_shape(mire_test)+
+  tm_raster(palette = "red")+
   tm_shape(outline)+
-  tm_borders()
+  tm_borders()+
+  tm_shape(mire_nf_stars)+
+  tm_raster(col="red")
+# OK
 
-# Then I cut out the relevant part of the ecosystem map
-nft <- nf |>
-  terra::vect()
+ggplot()+
+  geom_stars(data = mire_nf_stars)+
+  coord_equal()+
+  theme_void()+
+  scale_fill_viridis_c()
+```
 
-ras_crop <- terra::crop(ras_raw, shp)
-ras_mask <- terra::mask(ras_crop, shp, touches = T)
+```{r mireArea}
+mireArea <- mire_nf_terra |>
+  global(c("mean", "sum"), na.rm=T) |>
+  add_column("Municipality" = "Nordre Follo") |>
+  mutate(mirePercent = round(mean*100, 1),
+         sum = sum/1e+4) |>
+  dplyr::rename(mire_km2 = sum)
 
-nf_mire <- st_crop(EDM, st_bbox(nf))
-tr_mire <- st_crop(EDM, tr)
-me_mire <- st_crop(EDM, me)
-
-  tm_shape(nf)+
-  tm_borders(col="red", 
-             lwd=3)+
-tm_shape(st_downsample(nf_mire, 10))+
-  tm_raster(col="myrmodell90pros.tif")
-
-#tm_shape(st_downsample(EDM, 100))+
-  #tm_raster(alpha=.3)+
+mire_nf_terra_dk <- mire_nf_terra |>
+  terra::mask(dk2 |> filter(Municipality == "Nordre Follo")) |>
+  global("sum", na.rm=T) |>
+  mutate(mireInSurvey_km2 = sum/1e+4) |>
+  add_column(Municipality = "Nordre Follo")
 
@@
 We chose three municipalities in Norway to test out the indicator @fig-studyLocation.
-These municipalities are spread over xxx km north to south, and also differ in the amount of mire area, the total area surveyed, and the prevalence of infrastructure (TABLE)
+These municipalities are spread over xxx km north to south, and also differ in the amount of mire area, the total area surveyed, and the prevalence of infrastructure (@tbl-variables).
 
 ```{r}
-#| eval: true
-
 # calculate terrestrial area
@@
   left_join(dk2 |> select(Municipality, dk_area_km)) |>
-  mutate(dk_percent = round((dk_area_km/t_area_km) * 100))
+  mutate(dk_percent = round((dk_area_km/t_area_km) * 100)) |>
+  # number of polygons
+  left_join(nature3_tbl |> select(Municipality, n)) |>
+  # total mire area and %
+  left_join(mireArea |> select(Municipality, mirePercent, mire_km2)) |>
+  # % mire inside survey coverage map
+  left_join(mire_nf_terra_dk |> select(Municipality, mireInSurvey_km2)) |>
+  mutate(mireInSurvery_percent = round(mireInSurvey_km2/mire_km2*100, 2))
+
+saveRDS(muni_tbl, "../data/municipality_stats.RDS")
+
+```
+
+```{r}
+#| eval: true
+
+muni_tbl <- readRDS("../data/municipality_stats.RDS")
 
@@
 
-| Municipality | Total terrestrial area (km2)                                                 | \% of terrestrial area surveyed                                            | \% open mires in relation to total terrestrial area | Total mire area | Fraction mire area included in survey | Number of mire polygons in survey |
+| Municipality | Total terrestrial area (km^2^)                                            | \% of terrestrial area surveyed                                            | \% open mires in relation to total terrestrial area                         | Total mire area (km^2^)                                                  | %of mire area inside survey area                                                      | Number of mire polygons in survey                                 |
 |-----------|-----------|-----------|-----------|-----------|-----------|-----------|
-| Nordre Follo | `r terrestrial |> filter(Municipality == "Nordre Follo") |> pull(t_area_km)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(dk_percent)` |                                                     |                 |                                       |                                   |
-| Trondheim    | `r terrestrial |> filter(Municipality == "Trondheim") |> pull(t_area_km)`    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(dk_percent)`    |                                                     |                 |                                       |                                   |
-| Målselv      | `r terrestrial |> filter(Municipality == "Målselv") |> pull(t_area_km)`      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(dk_percent)`      |                                                     |                 |                                       |                                   |
+| Nordre Follo | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(t_area_km)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(dk_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mirePercent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mire_km2)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mireInSurvery_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(n)` |
+| Trondheim    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(t_area_km)`    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(dk_percent)`    |                                                                             |                                                                          |                                                                                       | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(n)`    |
+| Målselv      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(t_area_km)`      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(dk_percent)`      |                                                                             |                                                                          |                                                                                       | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(n)`      |
 

### 2024-01-22 manuscript/HIAs.qmd
vectorize infrastructure index

@@
 myCRS <- 25832
+
+```
+
+```{r paths}
+# Conditional file directory
+dir <- substr(getwd(), 1,2)
+
+mireFile <- "P:/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif"
+
+path_infrastructure <- ifelse(dir == "C:", 
+        "R:/GeoSpatialData/Utility_governmentalServices/Norway_Infrastructure_Index/Original/Infrastrukturindeks_UTM33/infra_tiff.tif",
+        "/data/R/GeoSpatialData/Utility_governmentalServices/Norway_Infrastructure_Index/Original/Infrastrukturindeks_UTM33/infra_tiff.tif")
 ```
@@
 ```{r mireDataTestCrop}
-file <- "P:/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif"
-# This is a big TIF!
 
-mire_test <- read_stars(file, proxy = T) |>
+# TEsting the cropping proceedure using a downsampled starts object 
+mire_test <- read_stars(mireFile, proxy = T) |>
   setNames("mire") |>
@@
 
-# The cropping is correct
-tmap_mode("plot")
-tm_shape(mire_test)+
-  tm_raster(palette = "red")+
-  tm_shape(outline)+
-  tm_borders()+
-  tm_shape(mire_test_nf)+
-  tm_raster(col="red")
+# The cropping is correct:
+#tmap_mode("plot")
+#tm_shape(mire_test)+
+#  tm_raster(palette = "red")+
+#  tm_shape(outline)+
+#  tm_borders()+
+#  tm_shape(mire_test_nf)+
+#  tm_raster(col="red")
 ```
@@
 
-mire_nf <- read_stars(file, proxy = T) |>
+mire_nf <- read_stars(mireFile, proxy = T) |>
   setNames("mire") |>
@@
 # crop and mask (very fast!)
-mire_nf_terra <- terra::rast(file) |>
-  crop(nf_vect) |>
+mire_nf_terra <- terra::rast(mireFile) |>
+  terra::crop(nf_vect) |>
   terra::mask(nf_vect)
@@
 # Plot to check overlap
-ggplot()+
-  geom_spatraster(data = mire_nf_terra)+
-  geom_spatvector(data = nf_vect,
-                  fill = NA)
+#ggplot()+
+#  geom_spatraster(data = mire_nf_terra)+
+#  geom_spatvector(data = nf_vect,
+#                  fill = NA)
 # The cropping and masking worked.
-#tmap_mode("plot")
+
+# I like the stars, sf and tmap combo better, so I return to stars
+mire_nf_stars <- mire_nf_terra |>
+  st_as_stars()
+
 #tm_shape(mire_test)+
@@
 #  tm_borders()+
-#  tm_shape(mire_nf_terra)+
+#  tm_shape(mire_nf_stars)+
 #  tm_raster(col="red")
-
-# I ilke the stars, sf and tmap combo better, so I return to stars
-mire_nf_stars <- mire_nf_terra |>
-  st_as_stars()
-
-tm_shape(mire_test)+
-  tm_raster(palette = "red")+
-  tm_shape(outline)+
-  tm_borders()+
-  tm_shape(mire_nf_stars)+
-  tm_raster(col="red")
 # OK
@@
   scale_fill_viridis_c()
+```
+
+```{r dk2}
+# calculate terrestrial area
+terrestrial <- terrestrial |> 
+  mutate(t_area_km = 
+           round(units::drop_units(area_t * 1e-6))) |>
+  as_tibble()
+
+# calculate area of survey coverage maps
+dk2 <-   read_sf("../data/dekningskartNF/Naturtyper NiN - Dekningskart.shp") |>
+  rbind(read_sf("../data/dekningskartTR/Naturtyper NiN - Dekningskart.shp")) |>
+  rbind(read_sf("../data/dekningskartME/Naturtyper NiN - Dekningskart.shp")) |>
+  st_transform(st_crs(muni3)) |>
+  st_intersection(muni3) |>
+  st_cast("POLYGON") |>
+  group_by(Municipality) |>
+  summarise(geometry = st_union(geometry)) |>
+  mutate(dk_area_km = geometry |> st_area(),
+         dk_area_km = round(units::drop_units(dk_area_km * 1e-6)))
 ```
@@
 
+```
+
+```{r infrastructureIndex}
+# this data is on a 100x100m grid
+infra <- stars::read_stars(path_infrastructure) |>
+  # , which is mor then we need - warp it to 1x1km
+  st_warp(cellsize = c(1000, 1000), 
+                                   crs = st_crs(nf), 
+                                   use_gdal = TRUE,
+                                   method = "average") |>
+  setNames("infrastructureIndex") |>
+  st_transform(myCRS) |>
+   mutate(infrastructureIndex = case_when(
+    infrastructureIndex < 1 ~ 0,
+    infrastructureIndex < 6 ~ 1,
+    infrastructureIndex < 12 ~ 2,
+    infrastructureIndex >= 12 ~ 3)) |>
+  st_crop(outline)
+
+# This step might seem rather stupid.
+We want to vectorize a rather large raster.
+This makes it a quite big data object.
+The reason is that there is no really good way to burn polygon data on to raster grid cells after the disuse of the raster package.
+It was not straight forward then either.
+But calculating intersections between polygons is very fast and easy.
+
+infra <-  eaTools::ea_homogeneous_area(infra, 
+                                groups = infrastructureIndex)
+
+path <- ifelse(dir == "C:", 
+        "P:/41201785_okologisk_tilstand_2022_2023/data/cache/",
+        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/")
+
+saveRDS(infra, paste0(path, "infrastructureIndex_discrete_vectorized.rds"))
+```
+
+```{r muni-tbl}
+# number and area of mire polygons
+nature3_tbl <- nature3 |>
+  group_by(Municipality) |>
+  summarise(n = n()) |>
+  as_tibble()
+
+muni_tbl <- muni3 |>
+  # calculate total area
+  mutate(area_km = 
+           round(
+             units::drop_units(
+               geom |> st_area() * 1e-6))) |>
+  # make tibble for the coming join
+  as_tibble() |>
+  # paste inn terrestrial area
+  left_join(terrestrial |> select(kommunenummer, t_area_km),
+            keep=F) |>
+  # area of survey
+  left_join(dk2 |> select(Municipality, dk_area_km)) |>
+  mutate(dk_percent = round((dk_area_km/t_area_km) * 100)) |>
+  # number of polygons
+  left_join(nature3_tbl |> select(Municipality, n)) |>
+  # total mire area and %
+  left_join(mireArea |> select(Municipality, mirePercent, mire_km2)) |>
+  # % mire inside survey coverage map
+  left_join(mire_nf_terra_dk |> select(Municipality, mireInSurvey_km2)) |>
+  mutate(mireInSurvery_percent = round(mireInSurvey_km2/mire_km2*100, 2))
+
+saveRDS(muni_tbl, "../data/municipality_stats.RDS")
 ```
@@
 
-We chose three municipalities in Norway to test out the indicator @fig-studyLocation.
+We chose three municipalities in Norway to test out the indicators (@fig-studyLocation).
 These municipalities are spread over xxx km north to south, and also differ in the amount of mire area, the total area surveyed, and the prevalence of infrastructure (@tbl-variables).
-
-```{r}
-# calculate terrestrial area
-terrestrial <- terrestrial |> 
-  mutate(t_area_km = 
-           round(units::drop_units(area_t * 1e-6))) |>
-  as_tibble()
-
-# calculate area of survey coverage maps
-dk2 <-   read_sf("../data/dekningskartNF/Naturtyper NiN - Dekningskart.shp") |>
-  rbind(read_sf("../data/dekningskartTR/Naturtyper NiN - Dekningskart.shp")) |>
-  rbind(read_sf("../data/dekningskartME/Naturtyper NiN - Dekningskart.shp")) |>
-  st_transform(st_crs(muni3)) |>
-  st_intersection(muni3) |>
-  st_cast("POLYGON") |>
-  group_by(Municipality) |>
-  summarise(geometry = st_union(geometry)) |>
-  mutate(dk_area_km = geometry |> st_area(),
-         dk_area_km = round(units::drop_units(dk_area_km * 1e-6)))
-
-# number and area of mire polygons
-nature3_tbl <- nature3 |>
-  group_by(Municipality) |>
-  summarise(n = n()) |>
-  as_tibble()
-
-muni_tbl <- muni3 |>
-  # calculate total area
-  mutate(area_km = 
-           round(
-             units::drop_units(
-               geom |> st_area() * 1e-6))) |>
-  # make tibble for the coming join
-  as_tibble() |>
-  # paste inn terrestrial area
-  left_join(terrestrial |> select(kommunenummer, t_area_km),
-            keep=F) |>
-  # area of survey
-  left_join(dk2 |> select(Municipality, dk_area_km)) |>
-  mutate(dk_percent = round((dk_area_km/t_area_km) * 100)) |>
-  # number of polygons
-  left_join(nature3_tbl |> select(Municipality, n)) |>
-  # total mire area and %
-  left_join(mireArea |> select(Municipality, mirePercent, mire_km2)) |>
-  # % mire inside survey coverage map
-  left_join(mire_nf_terra_dk |> select(Municipality, mireInSurvey_km2)) |>
-  mutate(mireInSurvery_percent = round(mireInSurvey_km2/mire_km2*100, 2))
-
-saveRDS(muni_tbl, "../data/municipality_stats.RDS")
-
-```
 
@@
 
-| Municipality | Total terrestrial area (km^2^)                                            | \% of terrestrial area surveyed                                            | \% open mires in relation to total terrestrial area                         | Total mire area (km^2^)                                                  | %of mire area inside survey area                                                      | Number of mire polygons in survey                                 |
-|-----------|-----------|-----------|-----------|-----------|-----------|-----------|
-| Nordre Follo | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(t_area_km)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(dk_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mirePercent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mire_km2)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mireInSurvery_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(n)` |
-| Trondheim    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(t_area_km)`    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(dk_percent)`    |                                                                             |                                                                          |                                                                                       | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(n)`    |
-| Målselv      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(t_area_km)`      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(dk_percent)`      |                                                                             |                                                                          |                                                                                       | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(n)`      |
+| Municipality | Total terrestrial area (km^2^)                                            | \% of terrestrial area surveyed                                            | \% open mires in relation to total terrestrial area                         | Total mire area (km^2^)                                                  | %of mire area inside survey area                                                      | Number of mire polygons in survey                                 | Mean Infrastructure Index value  |
+|-------|----------|----------|----------|----------|----------|----------|----------|
+| Nordre Follo | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(t_area_km)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(dk_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mirePercent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mire_km2)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mireInSurvery_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(n)` |                                  |
+| Trondheim    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(t_area_km)`    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(dk_percent)`    |                                                                             |                                                                          |                                                                                       | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(n)`    |                                  |
+| Målselv      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(t_area_km)`      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(dk_percent)`      |                                                                             |                                                                          |                                                                                       | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(n)`      |                                  |
 

### 2024-01-28 manuscript/HIAs.qmd
clean script and add annotations

@@
     wrap: sentence
-# remove this execute part to actually run the analysis
+# set eval to true to actually run the analysis
 execute:
@@
 
-```{r setup, warning=F, message=F, include=F}
+<!--# The document is set to eval=F in the yaml, and some code chunks are set manually to eval=T.
+When working on the manuscript, you can run all (Ctr+Alt+R).
+This will read inn data cashed locally.
+To also evaluate the more time consuming chunks, change the YAML to eval: true and run all (or render to html):  -->
+
+```{r setup}
 #| warning: false
@@
 library(tidyterra)
+library(ggtext)
 
 myCRS <- 25832
-
 ```
@@
 ```{r paths}
+#| eval: true
+
 # Conditional file directory
@@
 
+# Soem directories
 mireFile <- "P:/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif"
@@
         "/data/R/GeoSpatialData/Utility_governmentalServices/Norway_Infrastructure_Index/Original/Infrastrukturindeks_UTM33/infra_tiff.tif")
+
+path_naturetypes <- ifelse(dir == "C:", 
+               "R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb",
+               "/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb")
+
+path_muni <- paste0("../data/Basisdata_0000_Norge_25833_Kommuner_FGDB.gdb")
+
 ```
@@
 
-# Set up conditional file paths
 # The data is too big to be stored on GitHub
-dir <- substr(getwd(), 1,2)
-
-path <- ifelse(dir == "C:", 
-               "R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb",
-               "/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb")
-
-#pData <- ifelse(dir == "C:", 
-#               "P:/41201785_okologisk_tilstand_2022_2023/data/",
-#               "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/")
 
 # Impart polygon data set
-naturetypes <- sf::st_read(dsn = path)
+naturetypes <- sf::st_read(dsn = path_naturetypes)
 # 117k polygons (2023)
-# sometimes this functin struggles with Norwegian letters. Restart R and try again.
+# sometimes this functin struggles with Norwegian letters. 
+# Restart R and try again.
 
@@
 # 13k obs.
-#naturetypes |>
-#  saveRDS("../data/surveyData.RDS")
+
+naturetypes |>
+  saveRDS("../data/surveyData.RDS")
 #head(naturetypes)
@@
 
-```{r prepPolygons}
+```{r getMunicipalities}
 #| eval: true
 
-path <- paste0("../data/Basisdata_0000_Norge_25833_Kommuner_FGDB.gdb")
+# Preparing the outlines for the three municipalieties
 
 # find the correct layer
-#st_layers(path)
-muni <- sf::read_sf(path, layer = "kommune") |>
+#st_layers(path_muni)
+
+# read inn data and transform
+muni <- sf::read_sf(path_muni, layer = "kommune") |>
   st_transform(myCRS)
@@
 #table(st_geometry_type(muni))
-
 # function to make sure that multipolygons are returned 
@@
   filter(kommunenummer == "5418")
+```
 
-# I need to intersect nature types with municipality
+```{r prepPolygons}
+#| eval: true
+
+# I need to intersect the naturetypes data with the municipalities
 nature3 <- naturetypes |>
@@
 
-# and also to get the data coverage polygon
+# and also to get the data coverage polygon. 
+# These I had to manually download from naturbase.no 
 dk <-   read_sf("../data/dekningskartNF/Naturtyper NiN - Dekningskart.shp") |>
@@
   st_union()
+
+```
+
+```{r prepSomeMoreMunicipalityShapes}
+#| eval: true
 
@@
 
-# calculate stats
-# terrestrial area
+# calculate stats - terrestrial area
 terrestrial <- terrestrial |>
-  mutate(area_t = geometry |> st_area())
-
+  mutate(area_t = geometry |> st_area(),
+         t_area_km = 
+           round(units::drop_units(area_t * 1e-6)))|>
+  as_tibble()
 ```
 
-```{r makeMap}
-#| eval: false
+```{r positionMap}
 
-tmap_mode("plot")
 # Make map to show where the three municipalities are
@@
 
+tmap_mode("plot")
 positionMap <- tm_shape(outline)+
@@
   tm_layout(bg.color = "skyblue")
-
-# Plot (could perhaps add the EDM here as well)
-muniPlot <- tm_shape(muni3)+
-    tm_borders()+
-  tm_facets(by = "Municipality",
-            ncol=2)+
-  tm_shape(terrestrial)+
-    tm_fill(col = "lightgreen",
-             alpha=.4)+
-  tm_shape(ocean)+
-    tm_polygons(col = "skyblue",
-                border.col = "black")+
-  tm_shape(dk)+
-    tm_polygons(col = "grey",
-                alpha=.8)+
-  tm_shape(nature3)+
-  tm_polygons(col = "red",
-              border.col = "red",
-              lwd=3)
-
-methodsMap <- tmap_arrange(positionMap, muniPlot,
-             widths = c(.4, .6))
-
-# tmap_save(methodsMap, "../figures/studyLocations.tiff",
-#           dpi= 1000,
-#           units = "cm",
-#           width = 18,
-#           height = 10)
-# 
-# saveRDS(methodsMap, "../figures/studyLocation.RDS")
 ```
@@
 
-# TEsting the cropping proceedure using a downsampled starts object 
+# Testing the cropping proceedure using a downsampled starts object 
 mire_test <- read_stars(mireFile, proxy = T) |>
@@
 #  tm_raster(col="red")
-```
 
-```{r mireNF-error}
-#| eval: false
 # Then we do the same, but without downsampling
@@
   st_as_stars()
+# The cropped raster is for some reason far into the ocean!
 
-# The cropped raster is for some reason far into the ocean. 
 # The cells have values, so it seems like there is something 
@@
   tm_raster(col="red")
-
-st_crs(mire_nf) <- 25832
 ```
@@
 ```{r mireTerra}
-# trying terra
+
+# trying { terra } instead
+
 # convert municipal outline to vect via st
@@
 
-```{r dk2}
-# calculate terrestrial area
-terrestrial <- terrestrial |> 
-  mutate(t_area_km = 
-           round(units::drop_units(area_t * 1e-6))) |>
-  as_tibble()
-
+```{r dk2-CoverageMaps}
 # calculate area of survey coverage maps
+# values goes into summary table in the ms
 dk2 <-   read_sf("../data/dekningskartNF/Naturtyper NiN - Dekningskart.shp") |>
@@
 ```{r mireArea}
+# calculate area of the mires in each municipality
+# -- Nordre Follo
 mireArea <- mire_nf_terra |>
@@
   mutate(mirePercent = round(mean*100, 1),
-         sum = sum/1e+4) |>
-  dplyr::rename(mire_km2 = sum)
+         mire_km2 = sum/1e+4)
 
+# Caululate the area of mire inside the coverage maps
+# -- Nordre Follo
 mire_nf_terra_dk <- mire_nf_terra |>
@@
     infrastructureIndex >= 12 ~ 3)) |>
+  # takeing away point in the sea
   st_crop(outline)
@@
 
-path <- ifelse(dir == "C:", 
+# path to local cahing folder
+path_temp <- ifelse(dir == "C:", 
         "P:/41201785_okologisk_tilstand_2022_2023/data/cache/",
@@
 
-saveRDS(infra, paste0(path, "infrastructureIndex_discrete_vectorized.rds"))
+saveRDS(infra, paste0(path_temp, "infrastructureIndex_discrete_vectorized.rds"))
+```
+
+```{r readInfraStructureIndex}
+
+# read cached vetorized infrastructure data
+infra <- readRDS(paste0(path, "infrastructureIndex_discrete_vectorized.rds"))
+
+# intersect with the three municipalities
+# and calculate area
+infraMuni3 <- infra |>
+  st_intersection(muni3) |>
+  mutate(area = geometry |> st_area())
+
+# Turn m2 into km2
+# and sum the total area per HIA
+(infraMuni3_tbl <- infraMuni3 |>
+  as.data.frame() |>
+  mutate(area_HIA_km2 = units::drop_units(area)*1e-6) |>
+  group_by(Municipality, infrastructureIndex) |>
+  summarise(total_area_HIAs_km2 = round(sum(area_HIA_km2))))
+
+# Calculate the area weighted mean HIA value per municipality
+infraMuni3_summary <- infraMuni3_tbl |>
+  group_by(Municipality) |>
+  summarise(meanHIA = 
+              round(
+                weighted.mean(
+                  infrastructureIndex, total_area_HIAs_km2), 2))
+
+# Make a plot to check that it has worked 
+infraMuni3_tbl |>
+  ggplot()+
+  geom_bar(aes(x = infrastructureIndex, 
+               y = total_area_HIAs_km2,
+               fill = factor(infrastructureIndex),
+               colour = factor(infrastructureIndex)),
+           stat = "identity",
+           lwd=1.2
+           )+
+  scale_fill_manual(values = RColorBrewer::brewer.pal(4, "YlOrBr"))+
+  scale_color_manual(values = RColorBrewer::brewer.pal(5, "YlOrBr")[-1])+
+  facet_grid(cols = vars(Municipality))+
+  theme_minimal_hgrid()+
+  labs(x = "Human Impact Area (HIA) class",
+       y = "Area (km<sup>2</sup>)")+
+  theme(axis.title.x = element_textbox_simple(
+      width = NULL,
+      padding = margin(4, 4, 4, 4),
+      margin = margin(4, 0, 0, 0),
+      linetype = 1,
+      r = grid::unit(8, "pt"),
+      fill = "azure1"
+    ),
+    axis.title.y = element_textbox_simple(
+      width = NULL,
+      padding = margin(4, 4, 4, 4),
+      margin = margin(4, 0, 0, 0),
+      linetype = 1,
+      orientation = "left-rotated",
+      r = grid::unit(8, "pt"),
+      fill = "azure1"),
+    strip.background = element_blank(),
+    strip.text = element_textbox(
+      size = 12,
+      color = "white", fill = "#5D729D", box.color = "#4A618C",
+      halign = 0.5, linetype = 1, r = unit(5, "pt"), width = unit(1, "npc"),
+      padding = margin(2, 0, 1, 0), margin = margin(3, 3, 3, 3)
+    )
+    )+
+guides(fill= "none", colour = "none")
+```
+
+<!--# Need to update this figure -->
+
+```{r infraMuniMaps}
+infraMuniMap <- tm_shape(muni3)+
+  tm_borders()+
+tm_shape(infraMuni3)+
+  tm_polygons(col= "infrastructureIndex",
+              style="cat",
+              title = "Human Impact\nArea Class")+
+  tm_shape(muni3)+
+  tm_borders(lwd=3, col="black")+
+  tm_facets(by = "Municipality")+
+  tm_layout(legend.show=TRUE)
+
+# Plot
+muniPlot <- tm_shape(muni3)+
+    tm_borders()+
+  tm_facets(by = "Municipality",
+            ncol=3)+
+  tm_shape(terrestrial)+
+    tm_fill(col = "lightgreen",
+             alpha=.4)+
+  tm_shape(ocean)+
+    tm_polygons(col = "skyblue",
+                border.col = "black")+
+  tm_shape(dk)+
+    tm_polygons(col = "grey",
+                alpha=.8)+
+  tm_shape(nature3)+
+  tm_polygons(col = "red",
+              border.col = "red",
+              lwd=3)
+
+(methodsMap <- tmap_arrange(
+                          #positionMap, 
+                           muniPlot,
+                          infraMuniMap,
+             widths = c(.4, .6)))
+
+#cowplot::plot_grid(tmap_grob(muniPlot),
+#                   tmap_grob(infraMuniMap),
+#                   ncol=1,)
+
+# tmap_save(methodsMap, "../figures/studyLocations.tiff",
+#           dpi= 1000,
+#           units = "cm",
+#           width = 18,
+#           height = 10)
+# 
+# saveRDS(methodsMap, "../figures/studyLocation.RDS")
 ```
@@
 ```{r muni-tbl}
-# number and area of mire polygons
+
+# Calulate some more stats for the table with municipality stats
+
+# Number of mire polygons per municipality
+# -- Try count() here instead
 nature3_tbl <- nature3 |>
@@
 
+# Take muni3 and add all sorts of other data to it
+# using left_join.
+# Save local cached file
 muni_tbl <- muni3 |>
@@
   left_join(mire_nf_terra_dk |> select(Municipality, mireInSurvey_km2)) |>
-  mutate(mireInSurvery_percent = round(mireInSurvey_km2/mire_km2*100, 2))
+  mutate(mireInSurvery_percent = round(mireInSurvey_km2/mire_km2*100, 2)) |>
+  left_join(infraMuni3_summary)
 
 saveRDS(muni_tbl, "../data/municipality_stats.RDS")
+```
+
+```{r read-munitbl}
+#| eval: true
+muni_tbl <- readRDS("../data/municipality_stats.RDS")
 ```
@@
 
-To describe these two concepts we use five condition variables from the same data set ( @tbl-variables).
+To describe these two characteristics we use five condition variables from the same data set ( @tbl-variables).
 The data set is from a standardised field survey of nature types in Norway that started in 2018 and which is still ongoing.
@@
 
-```{r}
-#| eval: true
-
-muni_tbl <- readRDS("../data/municipality_stats.RDS")
-
-```
-
 ```{r fig-studyLocation}
@@
 | Municipality | Total terrestrial area (km^2^)                                            | \% of terrestrial area surveyed                                            | \% open mires in relation to total terrestrial area                         | Total mire area (km^2^)                                                  | %of mire area inside survey area                                                      | Number of mire polygons in survey                                 | Mean Infrastructure Index value                                         |
-|-------|----------|----------|----------|----------|----------|----------|----------|
-| Nordre Follo | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(t_area_km)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(dk_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mirePercent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mire_km2)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mireInSurvery_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(n)` |                                  |
-| Trondheim    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(t_area_km)`    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(dk_percent)`    |                                                                             |                                                                          |                                                                                       | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(n)`    |                                  |
-| Målselv      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(t_area_km)`      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(dk_percent)`      |                                                                             |                                                                          |                                                                                       | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(n)`      |                                  |
+|---------|---------|---------|---------|---------|---------|---------|---------|
+| Nordre Follo | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(t_area_km)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(dk_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mirePercent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mire_km2)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mireInSurvery_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(n)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(meanHIA)` |
+| Trondheim    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(t_area_km)`    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(dk_percent)`    |                                                                             |                                                                          |                                                                                       | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(n)`    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(meanHIA)`    |
+| Målselv      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(t_area_km)`      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(dk_percent)`      |                                                                             |                                                                          |                                                                                       | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(n)`      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(meanHIA)`      |
 

### 2024-01-29 manuscript/HIAs.qmd
explicit eval statements

@@
 <!--# The document is set to eval=F in the yaml, and some code chunks are set manually to eval=T.
-When working on the manuscript, you can run all (Ctr+Alt+R).
+When working on the manuscript, you cannot run all with  Ctr+Alt+R as this ignores the YAML.
 This will read inn data cashed locally.
@@
 ```{r getRelevantNTs}
+#| eval: false
+
 # I already did some work to identify the relevant nature types
@@
 ```{r relevantNTs}
+#| eval: false
+
 # These are the nature types I'm keeping for 7FA.
@@
   st_union()
-
 ```
@@
 ```{r positionMap}
+#| eval: false
 
@@
 ```{r mireDataTestCrop}
+#| eval: false
 
@@
 ```{r mireTerra}
+#| eval: false
 
@@
 ```{r dk2-CoverageMaps}
+#| eval: false
+
 # calculate area of survey coverage maps
@@
 ```{r mireArea}
+#| eval: false
+
 # calculate area of the mires in each municipality
@@
 ```{r infrastructureIndex}
+#| eval: false
+
 # this data is on a 100x100m grid
@@
 ```{r readInfraStructureIndex}
+#| eval: false
 
@@
 ```{r infraMuniMaps}
+#| eval: false
+
 infraMuniMap <- tm_shape(muni3)+
@@
 
+<!--# add mean_HIA to the mix -->
+
 ```{r muni-tbl}
+#| eval: false
 

### 2024-01-29 manuscript/HIAs.qmd
styler

@@
 # Conditional file directory
-dir <- substr(getwd(), 1,2)
+dir <- substr(getwd(), 1, 2)
 
@@
   "R:/GeoSpatialData/Utility_governmentalServices/Norway_Infrastructure_Index/Original/Infrastrukturindeks_UTM33/infra_tiff.tif",
-        "/data/R/GeoSpatialData/Utility_governmentalServices/Norway_Infrastructure_Index/Original/Infrastrukturindeks_UTM33/infra_tiff.tif")
+  "/data/R/GeoSpatialData/Utility_governmentalServices/Norway_Infrastructure_Index/Original/Infrastrukturindeks_UTM33/infra_tiff.tif"
+)
 
@@
   "R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb",
-               "/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb")
+  "/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb"
+)
 
 path_muni <- paste0("../data/Basisdata_0000_Norge_25833_Kommuner_FGDB.gdb")
-
 ```
@@
 
-#naturetypes_summary2 <- naturetypes_summary %>%
+# naturetypes_summary2 <- naturetypes_summary %>%
 #  rowwise() %>%
@@
 #
-#naturetypes_summary2$Nature_type
+# naturetypes_summary2$Nature_type
 
 # Getting the nature types that have the ADSV variables
-myVars2 <- c('7TK', '7SE', 'PRTK', 'PRSL')
-#naturetypes_summary3 <- naturetypes_summary %>%
+myVars2 <- c("7TK", "7SE", "PRTK", "PRSL")
+# naturetypes_summary3 <- naturetypes_summary %>%
 #  rowwise() %>%
@@
 #
-#naturetypes_summary3$Nature_type
+# naturetypes_summary3$Nature_type
 ```
@@
 # Note the absence of ombrotriphic types
-alienNTS <- c("Slåttemyr",
+alienNTS <- c(
+  "Slåttemyr",
   "Semi-naturlig våteng",
@@
   "Rik åpen sørlig jordvannsmyr",
-              "Rik åpen jordvannsmyr i nordboreal og lavalpin sone")
+  "Rik åpen jordvannsmyr i nordboreal og lavalpin sone"
+)
 
 # And these are the nature types that we use for the ADSV indicator
- advsNTs <- c(# fens
+advsNTs <- c( # fens
   "Slåttemyr",
@@
   "Rik åpen jordvannsmyr i nordboreal og lavalpin sone",
-              #"Sørlig kaldkilde", # excluded springs
+  # "Sørlig kaldkilde", # excluded springs
 
@@
   "Platåhøymyr",
-              "Palsmyr")
-
+  "Palsmyr"
+)
 ```
@@
   # keep only wetlands and our selected nature types
-  filter(hovedøkosystem == "våtmark",
-         naturtype %in% c(advsNTs, alienNTS)) |>
+  filter(
+    hovedøkosystem == "våtmark",
+    naturtype %in% c(advsNTs, alienNTS)
+  ) |>
   # calculate the areas (m2) of the polygons
@@
   # the variable codes and values are all in the same column
-  separate_rows(ninBeskrivelsesvariable, sep=",") |>
-  separate(col = ninBeskrivelsesvariable,
+  separate_rows(ninBeskrivelsesvariable, sep = ",") |>
+  separate(
+    col = ninBeskrivelsesvariable,
     into = c("NiN_variable_code", "NiN_variable_value"),
     sep = "_",
-           remove=F) |>
+    remove = F
+  ) |>
   mutate(NiN_variable_value = as.numeric(NiN_variable_value)) |>
   filter(NiN_variable_code %in% c(myVars, myVars2)) |>
-  select(id = identifikasjon_lokalId,
+  select(
+    id = identifikasjon_lokalId,
     municipality = kommunenummer,
@@
     value = NiN_variable_value,
-         area) |>
+    area
+  ) |>
   st_transform(myCRS) # Chosing this to match the EDM (se further down)
@@
   saveRDS("../data/surveyData.RDS")
-#head(naturetypes)
+# head(naturetypes)
 ```
@@
 # find the correct layer
-#st_layers(path_muni)
+# st_layers(path_muni)
 
@@
 # This returns some multisurfaces as well
-#table(st_geometry_type(muni))
+# table(st_geometry_type(muni))
 # function to make sure that multipolygons are returned
@@
 muni <- ensure_multipolygons(muni)
-#table(st_geometry_type(muni)) #OK
+# table(st_geometry_type(muni)) #OK
 
@@
 muni3 <- muni |>
-  filter(kommunenummer %in% c("3020", #Nordre Follo
+  filter(kommunenummer %in% c(
+    "3020", # Nordre Follo
     "5001", # Trondheim
@@
 terrestrial <- terrestrial |>
-  mutate(area_t = geometry |> st_area(),
+  mutate(
+    area_t = geometry |> st_area(),
     t_area_km =
-           round(units::drop_units(area_t * 1e-6)))|>
+      round(units::drop_units(area_t * 1e-6))
+  ) |>
   as_tibble()
@@
 tmap_mode("plot")
-positionMap <- tm_shape(outline)+
+positionMap <- tm_shape(outline) +
   tm_polygons() +
-  tm_shape(centroids)+
-    tm_text(text = "Municipality",
+  tm_shape(centroids) +
+  tm_text(
+    text = "Municipality",
     size = 1,
     xmod = 2,
-      ymod=-1)+
-  tm_symbols(size = 2,
+    ymod = -1
+  ) +
+  tm_symbols(
+    size = 2,
     shape = 17,
-             col="black")+
-  tm_grid(projection = 4326)+
+    col = "black"
+  ) +
+  tm_grid(projection = 4326) +
   tm_layout(bg.color = "skyblue")
@@
 # The cropping is correct:
-#tmap_mode("plot")
-#tm_shape(mire_test)+
+# tmap_mode("plot")
+# tm_shape(mire_test)+
 #  tm_raster(palette = "red")+
@@
   setNames("mire") |>
-  #mutate(mire = na_if(mire, 0)) |>
+  # mutate(mire = na_if(mire, 0)) |>
   st_crop(nf) |>
@@
 tmap_mode("plot")
-tm_shape(mire_test)+
-  tm_raster(palette = "red")+
-  tm_shape(outline)+
-  tm_borders()+
-  tm_shape(mire_nf)+
-  tm_raster(col="red")
+tm_shape(mire_test) +
+  tm_raster(palette = "red") +
+  tm_shape(outline) +
+  tm_borders() +
+  tm_shape(mire_nf) +
+  tm_raster(col = "red")
 ```
@@
 # Plot to check overlap
-#ggplot()+
+# ggplot()+
 #  geom_spatraster(data = mire_nf_terra)+
@@
 
-#tm_shape(mire_test)+
+# tm_shape(mire_test)+
 #  tm_raster(palette = "red")+
@@
 
-ggplot()+
-  geom_stars(data = mire_nf_stars)+
-  coord_equal()+
-  theme_void()+
+ggplot() +
+  geom_stars(data = mire_nf_stars) +
+  coord_equal() +
+  theme_void() +
   scale_fill_viridis_c()
@@
   summarise(geometry = st_union(geometry)) |>
-  mutate(dk_area_km = geometry |> st_area(),
-         dk_area_km = round(units::drop_units(dk_area_km * 1e-6)))
+  mutate(
+    dk_area_km = geometry |> st_area(),
+    dk_area_km = round(units::drop_units(dk_area_km * 1e-6))
+  )
 ```
@@
 mireArea <- mire_nf_terra |>
-  global(c("mean", "sum"), na.rm=T) |>
+  global(c("mean", "sum"), na.rm = T) |>
   add_column("Municipality" = "Nordre Follo") |>
-  mutate(mirePercent = round(mean*100, 1),
-         mire_km2 = sum/1e+4)
+  mutate(
+    mirePercent = round(mean * 100, 1),
+    mire_km2 = sum / 1e+4
+  )
 
@@
   terra::mask(dk2 |> filter(Municipality == "Nordre Follo")) |>
-  global("sum", na.rm=T) |>
-  mutate(mireInSurvey_km2 = sum/1e+4) |>
+  global("sum", na.rm = T) |>
+  mutate(mireInSurvey_km2 = sum / 1e+4) |>
   add_column(Municipality = "Nordre Follo")
-
 ```
@@
   # , which is mor then we need - warp it to 1x1km
-  st_warp(cellsize = c(1000, 1000), 
+  st_warp(
+    cellsize = c(1000, 1000),
     crs = st_crs(nf),
     use_gdal = TRUE,
-                                   method = "average") |>
+    method = "average"
+  ) |>
   setNames("infrastructureIndex") |>
@@
     infrastructureIndex < 12 ~ 2,
-    infrastructureIndex >= 12 ~ 3)) |>
+    infrastructureIndex >= 12 ~ 3
+  )) |>
   # takeing away point in the sea
@@
 infra <- eaTools::ea_homogeneous_area(infra,
-                                groups = infrastructureIndex)
+  groups = infrastructureIndex
+)
 
@@
   "P:/41201785_okologisk_tilstand_2022_2023/data/cache/",
-        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/")
+  "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/"
+)
 
@@
   as.data.frame() |>
-  mutate(area_HIA_km2 = units::drop_units(area)*1e-6) |>
+  mutate(area_HIA_km2 = units::drop_units(area) * 1e-6) |>
   group_by(Municipality, infrastructureIndex) |>
@@
   group_by(Municipality) |>
-  summarise(meanHIA = 
+  summarise(
+    meanHIA =
       round(
         weighted.mean(
-                  infrastructureIndex, total_area_HIAs_km2), 2))
+          infrastructureIndex, total_area_HIAs_km2
+        ), 2
+      )
+  )
 
@@
 infraMuni3_tbl |>
-  ggplot()+
-  geom_bar(aes(x = infrastructureIndex, 
+  ggplot() +
+  geom_bar(
+    aes(
+      x = infrastructureIndex,
       y = total_area_HIAs_km2,
       fill = factor(infrastructureIndex),
-               colour = factor(infrastructureIndex)),
+      colour = factor(infrastructureIndex)
+    ),
     stat = "identity",
-           lwd=1.2
-           )+
-  scale_fill_manual(values = RColorBrewer::brewer.pal(4, "YlOrBr"))+
-  scale_color_manual(values = RColorBrewer::brewer.pal(5, "YlOrBr")[-1])+
-  facet_grid(cols = vars(Municipality))+
-  theme_minimal_hgrid()+
-  labs(x = "Human Impact Area (HIA) class",
-       y = "Area (km<sup>2</sup>)")+
-  theme(axis.title.x = element_textbox_simple(
+    lwd = 1.2
+  ) +
+  scale_fill_manual(values = RColorBrewer::brewer.pal(4, "YlOrBr")) +
+  scale_color_manual(values = RColorBrewer::brewer.pal(5, "YlOrBr")[-1]) +
+  facet_grid(cols = vars(Municipality)) +
+  theme_minimal_hgrid() +
+  labs(
+    x = "Human Impact Area (HIA) class",
+    y = "Area (km<sup>2</sup>)"
+  ) +
+  theme(
+    axis.title.x = element_textbox_simple(
       width = NULL,
@@
       r = grid::unit(8, "pt"),
-      fill = "azure1"),
+      fill = "azure1"
+    ),
     strip.background = element_blank(),
@@
     )
-    )+
-guides(fill= "none", colour = "none")
+  ) +
+  guides(fill = "none", colour = "none")
 ```
@@
 
-infraMuniMap <- tm_shape(muni3)+
-  tm_borders()+
-tm_shape(infraMuni3)+
-  tm_polygons(col= "infrastructureIndex",
-              style="cat",
-              title = "Human Impact\nArea Class")+
-  tm_shape(muni3)+
-  tm_borders(lwd=3, col="black")+
-  tm_facets(by = "Municipality")+
-  tm_layout(legend.show=TRUE)
+infraMuniMap <- tm_shape(muni3) +
+  tm_borders() +
+  tm_shape(infraMuni3) +
+  tm_polygons(
+    col = "infrastructureIndex",
+    style = "cat",
+    title = "Human Impact\nArea Class"
+  ) +
+  tm_shape(muni3) +
+  tm_borders(lwd = 3, col = "black") +
+  tm_facets(by = "Municipality") +
+  tm_layout(legend.show = TRUE)
 
 # Plot
-muniPlot <- tm_shape(muni3)+
-    tm_borders()+
-  tm_facets(by = "Municipality",
-            ncol=3)+
-  tm_shape(terrestrial)+
-    tm_fill(col = "lightgreen",
-             alpha=.4)+
-  tm_shape(ocean)+
-    tm_polygons(col = "skyblue",
-                border.col = "black")+
-  tm_shape(dk)+
-    tm_polygons(col = "grey",
-                alpha=.8)+
-  tm_shape(nature3)+
-  tm_polygons(col = "red",
+muniPlot <- tm_shape(muni3) +
+  tm_borders() +
+  tm_facets(
+    by = "Municipality",
+    ncol = 3
+  ) +
+  tm_shape(terrestrial) +
+  tm_fill(
+    col = "lightgreen",
+    alpha = .4
+  ) +
+  tm_shape(ocean) +
+  tm_polygons(
+    col = "skyblue",
+    border.col = "black"
+  ) +
+  tm_shape(dk) +
+  tm_polygons(
+    col = "grey",
+    alpha = .8
+  ) +
+  tm_shape(nature3) +
+  tm_polygons(
+    col = "red",
     border.col = "red",
-              lwd=3)
+    lwd = 3
+  )
 
 (methodsMap <- tmap_arrange(
-                          #positionMap, 
+  # positionMap,
   muniPlot,
   infraMuniMap,
-             widths = c(.4, .6)))
+  widths = c(.4, .6)
+))
 
-#cowplot::plot_grid(tmap_grob(muniPlot),
+# cowplot::plot_grid(tmap_grob(muniPlot),
 #                   tmap_grob(infraMuniMap),
@@
   # calculate total area
-  mutate(area_km = 
+  mutate(
+    area_km =
       round(
         units::drop_units(
-               geom |> st_area() * 1e-6))) |>
+          geom |> st_area() * 1e-6
+        )
+      )
+  ) |>
   # make tibble for the coming join
@@
   left_join(terrestrial |> select(kommunenummer, t_area_km),
-            keep=F) |>
+    keep = F
+  ) |>
   # area of survey
   left_join(dk2 |> select(Municipality, dk_area_km)) |>
-  mutate(dk_percent = round((dk_area_km/t_area_km) * 100)) |>
+  mutate(dk_percent = round((dk_area_km / t_area_km) * 100)) |>
   # number of polygons
@@
   left_join(mire_nf_terra_dk |> select(Municipality, mireInSurvey_km2)) |>
-  mutate(mireInSurvery_percent = round(mireInSurvey_km2/mire_km2*100, 2)) |>
+  mutate(mireInSurvery_percent = round(mireInSurvey_km2 / mire_km2 * 100, 2)) |>
   left_join(infraMuni3_summary)
@@
 | Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                    | Reference                          |
-|---------------|---------------|---------------|---------------|---------------|
+|-------------|-------------|---------------|--------------------|-------------|
 | 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                             | @runehalvorsen2019                 |

### 2024-01-29 manuscript/HIAs.qmd
fix methods map with infrastructure map

@@
     wrap: sentence
-# set eval to true to actually run the analysis
 execute:
@@
 
-<!--# The document is set to eval=F in the yaml, and some code chunks are set manually to eval=T.
-When working on the manuscript, you cannot run all with  Ctr+Alt+R as this ignores the YAML.
-This will read inn data cashed locally.
-To also evaluate the more time consuming chunks, change the YAML to eval: true and run all (or render to html):  -->
+<!--# The document is set to eval=F in the yaml.
+This only affect the rendering, and not the interactive session.
+Eval is also set explicitly for each chunk.
+When eval is set to false it means the job is quite time consuming and I instead run it once and cache it locally.
+-->
 
@@
 library(ggtext)
+library(cowplot)
 
@@
 
-# Soem directories
+# Some directories
 mireFile <- "P:/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif"
@@
 path_muni <- paste0("../data/Basisdata_0000_Norge_25833_Kommuner_FGDB.gdb")
+
+# path to local caching folder
+path_temp <- ifelse(dir == "C:",
+  "P:/41201785_okologisk_tilstand_2022_2023/data/cache/",
+  "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/"
+)
 ```
@@
 
-# read cached rds file
+# read cached rds file. This is not so big.
 naturetypes <- readRDS("../data/surveyData.RDS")
@@
 
-table(nature3$municipality)
+nature3 |>
+  as_tibble() |>
+  count(municipality,
+    sort = TRUE,
+    name = "Number of polygons")
 
@@
       round(units::drop_units(area_t * 1e-6))
-  ) |>
-  as_tibble()
+  ) 
 ```
@@
 
-tmap_mode("plot")
-positionMap <- tm_shape(outline) +
+(positionMap <- tm_shape(outline) +
   tm_polygons() +
@@
   tm_grid(projection = 4326) +
-  tm_layout(bg.color = "skyblue")
+  tm_layout(
+    bg.color = "skyblue",
+    outer.margins = c(0.01, .02, .02, .02)))
 ```
@@
 
-# tm_shape(mire_test)+
-#  tm_raster(palette = "red")+
-#  tm_shape(outline)+
-#  tm_borders()+
-#  tm_shape(mire_nf_stars)+
-#  tm_raster(col="red")
-# OK
-
 ggplot() +
@@
 
-# Caululate the area of mire inside the coverage maps
+# Calculate the area of mire inside the coverage maps
 # -- Nordre Follo
@@
 
-# path to local cahing folder
-path_temp <- ifelse(dir == "C:",
-  "P:/41201785_okologisk_tilstand_2022_2023/data/cache/",
-  "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/"
-)
-
 saveRDS(infra, paste0(path_temp, "infrastructureIndex_discrete_vectorized.rds"))
@@
 # read cached vetorized infrastructure data
-infra <- readRDS(paste0(path, "infrastructureIndex_discrete_vectorized.rds"))
+infra <- readRDS(paste0(path_temp, "infrastructureIndex_discrete_vectorized.rds"))
 
@@
   scale_color_manual(values = RColorBrewer::brewer.pal(5, "YlOrBr")[-1]) +
-  facet_grid(cols = vars(Municipality)) +
   theme_minimal_hgrid() +
@@
   ) +
-  guides(fill = "none", colour = "none")
+  guides(fill = "none", colour = "none") +
+  scale_y_log10() +
+  facet_grid(cols = vars(Municipality)) 
 ```
@@
 
+# Infratructure in each municipality
 infraMuniMap <- tm_shape(muni3) +
@@
   tm_facets(by = "Municipality") +
-  tm_layout(legend.show = TRUE)
+  tm_layout(
+    legend.show = FALSE,
+    panel.label.height = 0)
 
-# Plot
+# A figure with just the legend
+infraMuniMap_l <- tm_shape(muni3) +
+  tm_borders() +
+  tm_shape(infraMuni3) +
+  tm_polygons(
+    col = "infrastructureIndex",
+    style = "cat",
+    title = "Human Impact\nArea Class"
+  ) +
+  tm_layout(legend.only = TRUE,
+            legend.position = c("left", "bottom"),
+            legend.outside = TRUE)
+
+# Syrvey coverage and other colour overlayed the municipalities
 muniPlot <- tm_shape(muni3) +
@@
 (methodsMap <- tmap_arrange(
-  # positionMap,
+  positionMap,
   muniPlot,
+  infraMuniMap_l,
   infraMuniMap,
-  widths = c(.4, .6)
+  widths = c(.35, .65),
+  heights = c(.6, .4),
+  outer.margins = NULL
 ))
-
-# cowplot::plot_grid(tmap_grob(muniPlot),
-#                   tmap_grob(infraMuniMap),
-#                   ncol=1,)
 

### 2024-01-29 manuscript/HIAs.qmd
calculate distance NF to Målselv

@@
     text = "Municipality",
-    size = 1,
+    size = .8,
     xmod = 2,
@@
   tm_symbols(
-    size = 2,
+    size = 1.5,
     shape = 17,
@@
     outer.margins = c(0.01, .02, .02, .02)))
+```
+
+```{r distanceBetweenMunis}
+#| eval: true
+
+# what is the distance between Nordre Follo and Målselv
+(latlong <- centroids |>
+  st_transform(4326) |>
+st_coordinates() |>
+  as_tibble())
+
+round(max(latlong$Y)-min(latlong$Y), 1)
+
 ```
@@
 
-<!--# add mean_HIA to the mix -->
-
 ```{r muni-tbl}
@@
   # paste inn terrestrial area
-  left_join(terrestrial |> select(kommunenummer, t_area_km),
+  left_join(terrestrial |>
+              as_tibble() |> 
+              select(kommunenummer, t_area_km),
     keep = F
@@
 We chose three municipalities in Norway to test out the indicators (@fig-studyLocation).
-These municipalities are spread over xxx km north to south, and also differ in the amount of mire area, the total area surveyed, and the prevalence of infrastructure (@tbl-variables).
+These municipalities are spread over `r round(max(latlong$Y)-min(latlong$Y), 1)` degrees latitude, and also differ in the amount of mire area, the total area surveyed, and the prevalence of infrastructure (@tbl-variables).
 

### 2024-01-29 manuscript/HIAs.qmd
calculate km distance

@@
   as_tibble())
+
+km_distance <- centroids |>
+  st_distance() |>
+  max() |>
+  set_units("km") |>
+  drop_units() |>
+  round()
 
@@
 We chose three municipalities in Norway to test out the indicators (@fig-studyLocation).
-These municipalities are spread over `r round(max(latlong$Y)-min(latlong$Y), 1)` degrees latitude, and also differ in the amount of mire area, the total area surveyed, and the prevalence of infrastructure (@tbl-variables).
+These municipalities are spread over `r round(max(latlong$Y)-min(latlong$Y), 1)` degrees latitude, or `r km_distance` km, and also differ in the amount of mire area, the total area surveyed, and the prevalence of infrastructure (@tbl-variables).
 

### 2024-01-29 manuscript/HIAs.qmd
make variables continous

@@
 library(cowplot)
+library(units)
 
@@
 naturetypes <- readRDS("../data/surveyData.RDS")
+```
+
+```{r convertToPercent}
+#| eval: true
+
+# I now want to take the variable and normalise it before I can then combine them
+# despite them being on different scales
+# I will first normalise by converting into %.
+# Remeber the ordinal categories represnts frequency ranges
+
+# The data is strongly right skewed, so simply taking the center value of each
+# bin will not work.
+naturetypes %>% 
+  ggplot()+
+  theme_bw() +
+  geom_histogram(aes(x = value),
+                 binwidth = 1) +
+  facet_wrap(.~variable,
+             scales = "free")
+
+# I will use th lower bound for each bin.
+# The exception in when the variable is 1, because then the lower bound
+# is 0, same as when the variable is 0.
+# For these I will set manually a slightly higher value.
+
+naturetypes <- naturetypes %>%
+  mutate(value = case_when(
+    # selecting the variable that follow the same 4 step scale
+    variable %in% c("7TK", "7SE", "7FA") ~ 
+      case_match(value,
+             0 ~ 0,
+             1 ~ mean(c(0, 1/16))*100,
+             2 ~ 1/16*100,
+             3 ~ 50), # note that it is not possible to get a value of 1
+    # selecting the eight step variables
+    variable %in% c('PRTK', 'PRSL') ~ 
+      case_match(value,
+             0 ~ 0,
+             1 ~ 1.5,
+             2 ~ 3,
+             3 ~ 6.25,
+             4 ~ 12.5,
+             5 ~ 25,
+             6 ~ 50,
+             7 ~ 75)
+    )
+  )
+
+naturetypes %>% 
+  ggplot()+
+  theme_bw() +
+  geom_histogram(aes(x = value),
+                 binwidth = 1,
+                 color = "orange",
+                 fill = "orange") +
+  facet_wrap(.~variable,
+             scales = "free")
+
+```
+
+```{r}
+#| eval: true
+# Now I make the data wide, and remove 7TK and 7SE if PRTK or PRSL are present,
+# respectively
+
+naturetypes_wide <- naturetypes |>
+  # Column names starting with a number is problematic, so adding a prefix
+  mutate(variable = paste0("var_", variable)) |>
+  pivot_wider(
+    names_from = "variable",
+    values_from = "value",
+    id_cols = "id") |>
+    as_tibble()
+
+head(naturetypes_wide, 10)
+```
+
+```{r}
+#| eval: true
+# First I will combine 7TK and PRTK, and also 7SE and PRSL.
+
+naturetypes_wide <- naturetypes_wide %>%
+  mutate(TK = if_else(
+    is.na(var_PRTK), var_7TK, var_PRTK),
+    SE = if_else(
+      is.na(var_PRSL), var_7SE, var_PRSL)
+    )
+
+plot_grid(
+naturetypes_wide %>%
+  as_tibble() |>
+  count(SE,
+        name = "sum") |>
+  ggplot(
+    aes(
+      x = factor(SE),
+      y = sum))+
+  geom_bar(stat="identity",
+           fill="grey",
+           colour = "black")+
+  theme_bw(base_size = 12)+
+  labs(x = "7SE or PRSL score",
+       y = "Number of localities"),
+
+naturetypes_wide %>%
+  as_tibble() |>
+  count(TK,
+        name = "sum") |>
+  ggplot(
+    aes(
+      x = factor(TK),
+      y = sum))+
+  geom_bar(stat="identity",
+           fill="grey",
+           colour = "black")+
+  theme_bw(base_size = 12)+
+  labs(x = "7TK or PRTK score",
+       y = "Number of localities")
+)
+
+# The NA's represnets localities where just one of the two variables 
+# (then thinking 7SE and PRSL as the same variable)
+# is recorded. To combine these into one metric ADSV, I could take the 
+# one with the highest value (worst-rule) og the sum.
+# Sum is problematic is not all locations ahve two values to sum together.
+# But the other option is problematic since I think field workers often
+# tend to split the effects over two variables is they have that option.
+# And if we have 50% vehicle damaeg and 50% hiking damage, that is no doubt
+# worst than just having 50% of either. So I will use the sum, despite its issues.
+```
+
+```{r}
+#| eval: true
+
+# Taking he sum of 7SE and 7TK (incl the PR.. variabels)
+naturetypes_wide <- naturetypes_wide |>
+  rowwise() |>
+  mutate(ADSV = sum(c(SE, TK), na.rm=TRUE))
+
+naturetypes_wide %>%
+  as_tibble() |>
+  count(ADSV,
+        name = "sum") |>
+  ggplot(
+    aes(
+      x = (ADSV),
+      y = sum))+
+  geom_bar(stat="identity",
+           fill="grey",
+           colour = "black")+
+  theme_bw(base_size = 12)+
+  labs(x = "Summed ADVS score",
+       y = "Number of localities") +
+    scale_x_continuous(
+    labels = scales::label_number(accuracy = 1))
+
+```
+
+```{r}
+#| eval: true
+# Now I will copy these ADVS-values into the sf object again
+naturetypes <- naturetypes |>
+  left_join(naturetypes_wide |> select(id, var_7FA, ADSV), by = "id")
 ```
@@
 
-<!--# Need to update this figure -->
-
 ```{r infraMuniMaps}
@@
 | Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                    | Reference                          |
-|-------------|-------------|---------------|--------------------|-------------|
+|--------------|--------------|--------------|----------------|--------------|
 | 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                             | @runehalvorsen2019                 |

### 2024-01-30 manuscript/HIAs.qmd
validate

@@
 
+```{r rescale}
+#| eval: true
+
+# Now I rescale the now continous variables using reference and threshold values
+# I will use the same reference levels/values for all of Norway:
+
+upper <- 0
+lower <- 100
+threshold <- 10
+
+eaTools::ea_normalise(data = naturetypes,
+                      vector = "ADSV",
+                      upper_reference_level = lower,
+                      lower_reference_level = upper,
+                      break_point = threshold,
+                      plot=T,
+                      reverse = T
+                      )
+
+# There is no point yet making this a time series
+# I will assign all the indicator value to the same time (2018-2022)
+
+# same for 7FK
+eaTools::ea_normalise(data = naturetypes,
+                      vector = "var_7FA",
+                      upper_reference_level = lower,
+                      lower_reference_level = upper,
+                      break_point = threshold,
+                      plot=T,
+                      reverse = T
+                      )
+# The variables are really coarse
+```
+
+```{r}
+#| eval: true
+
+# Adding scaled indicator values to the dataset
+naturetypes$i_ADSV <- eaTools::ea_normalise(data = naturetypes,
+                      vector = "ADSV",
+                      upper_reference_level = lower,
+                      lower_reference_level = upper,
+                      break_point = threshold,
+                      reverse = T
+                      )
+
+naturetypes$i_alien <- eaTools::ea_normalise(data = naturetypes,
+                      vector = "var_7FA",
+                      upper_reference_level = lower,
+                      lower_reference_level = upper,
+                      break_point = threshold,
+                      reverse = T
+                      )
+```
+
 ```{r getNorwayOutline}
@@
 
+```{r HIA-validate}
+#| eval: false
+
+# Now I want to see if the indicator values depend on the HIA is a predictable
+# way to justify the stratification
+infra |>
+  as_tibble() |>
+  count(infrastructureIndex)
+
+corrCheck <- st_intersection(naturetypes, infra)
+corrCheck |>
+  as_tibble() |>
+  mutate(i_ADSV_fct = floor(round(i_ADSV*10,2))/10,
+         i_alien_fct = floor(round(i_alien*10,2))/10) |>
+  count(i_ADSV_fct, infrastructureIndex) |>
+  View()
+
+corrCheck |>
+  mutate(i_ADSV_fct = floor(round(i_ADSV*10,2))/10,
+         i_alien_fct = floor(round(i_alien*10,2))/10) |>
+  pivot_longer(cols = c(i_ADSV_fct, i_alien_fct),
+               values_to = "indicatorValue",
+               names_to = "indicator") |>
+  filter(!is.na(indicatorValue)) |>
+  ggplot(aes(
+    x = factor(infrastructureIndex), 
+    fill = factor(round(indicatorValue,2))))+
+  geom_bar(
+    #position="fill"
+    )+
+  theme_bw(base_size = 12)+
+  guides(fill = guide_legend("Scaled indicator values"))+
+  ylab("Fraction of data points")+
+  xlab("HIA")+
+  scale_fill_brewer(palette = "RdYlGn")+
+  facet_grid(~indicator)
+```
+
+```{r}
+#| eval: false
+
+# There is very little data from HIA 3, so I combine it with HIA 2
+corrCheck |>
+  mutate(i_ADSV_fct = floor(round(i_ADSV*10,2))/10,
+         i_alien_fct = floor(round(i_alien*10,2))/10,
+         infrastructureIndex = case_when(
+           infrastructureIndex == 3 ~ 2,
+           .default = infrastructureIndex
+         )) |>
+  pivot_longer(cols = c(i_ADSV_fct, i_alien_fct),
+               values_to = "indicatorValue",
+               names_to = "indicator") |>
+  filter(!is.na(indicatorValue)) |>
+  ggplot(aes(
+    x = factor(infrastructureIndex), 
+    fill = factor(round(indicatorValue,2))))+
+  geom_bar(
+    position="fill"
+    )+
+  theme_bw(base_size = 12)+
+  guides(fill = guide_legend("Scaled indicator values"))+
+  ylab("Fraction of data points")+
+  xlab("HIA")+
+  scale_fill_brewer(palette = "RdYlGn")+
+  facet_grid(~indicator)
+```
+
+```{r}
+# That did not made it better
+# Is there a year effect?
+
+corrCheck |>
+  mutate(i_ADSV_fct = floor(round(i_ADSV*10,2))/10,
+         i_alien_fct = floor(round(i_alien*10,2))/10,
+         infrastructureIndex = case_when(
+           infrastructureIndex == 3 ~ 2,
+           .default = infrastructureIndex
+         )) |>
+  pivot_longer(cols = c(i_ADSV_fct, i_alien_fct),
+               values_to = "indicatorValue",
+               names_to = "indicator") |>
+  filter(!is.na(indicatorValue)) |>
+  ggplot(aes(
+    x = factor(infrastructureIndex), 
+    fill = factor(round(indicatorValue,2))))+
+  geom_bar(
+    position="fill"
+    )+
+  theme_bw(base_size = 12)+
+  guides(fill = guide_legend("Scaled indicator values"))+
+  ylab("Fraction of data points")+
+  xlab("HIA")+
+  scale_fill_brewer(palette = "RdYlGn")+
+  facet_grid(indicator~year)
+```
+
+```{r}
+# Next, make a violing/box plot
+
+corrCheck |>
+  mutate(i_ADSV_fct = floor(round(i_ADSV*10,2))/10,
+         i_alien_fct = floor(round(i_alien*10,2))/10,
+         infrastructureIndex = case_when(
+           infrastructureIndex == 3 ~ 2,
+           .default = infrastructureIndex
+         )) |>
+  pivot_longer(cols = c(i_ADSV_fct, i_alien_fct),
+               values_to = "indicatorValue",
+               names_to = "indicator") |>
+  filter(!is.na(indicatorValue)) |>
+  ggplot(aes(
+    x = factor(infrastructureIndex), 
+    y = indicatorValue))+
+  geom_boxplot(
+    )+
+  theme_bw(base_size = 12)+
+  guides(fill = guide_legend("Scaled indicator values"))+
+  ylab("Fraction of data points")+
+  xlab("HIA")+
+  scale_fill_brewer(palette = "RdYlGn")+
+  facet_grid(indicator~year)
+```
+
+```{r}
+#| eval: false
+
+# All the 1's might make it hard to see the pattern.
+# Try removing the 1's (the areas in best condition)
+
+corrCheck |>
+  mutate(i_ADSV_fct = floor(round(i_ADSV*10,2))/10,
+         i_alien_fct = floor(round(i_alien*10,2))/10,
+         infrastructureIndex = case_when(
+           infrastructureIndex == 3 ~ 2,
+           .default = infrastructureIndex
+           )) |>
+  pivot_longer(cols = c(i_ADSV_fct, i_alien_fct),
+               values_to = "indicatorValue",
+               names_to = "indicator") |>
+  mutate(indicatorValue = na_if(indicatorValue, 1)) |>
+  filter(!is.na(indicatorValue)) |>
+  ggplot(aes(
+    x = factor(infrastructureIndex), 
+    fill = factor(round(indicatorValue,2))))+
+  geom_bar(
+    position="fill"
+    )+
+  theme_bw(base_size = 12)+
+  guides(fill = guide_legend("Scaled indicator values"))+
+  ylab("Fraction of data points")+
+  xlab("HIA")+
+  scale_fill_brewer(palette = "RdYlGn")+
+  facet_grid(~indicator)
+
+```
+
 ```{r infraMuniMaps}

### 2024-01-31 manuscript/HIAs.qmd
validate fail

--- 2024-01-29 manuscript/HIAs.qmd
@@
 infra <- readRDS(paste0(path_temp, "infrastructureIndex_discrete_vectorized.rds"))
+
+# Calulculate area
+infra <- infra |>
+  mutate(area = geometry |> st_area(),
+         area_km = area |> set_units("km2"))
+
+# show the summered area per HIA
+infra |>
+  as_tibble() |>
+  group_by(infrastructureIndex) |>
+  summarise(area_km = sum(area_km)) |>
+  ggplot(aes(x = infrastructureIndex,
+             y = area_km)) +
+  geom_col()
 
@@
 
+```{r HIA-validate}
+#| eval: false
+
+# Now I want to see if the indicator values depend on the HIA is a predictable
+# way to justify the stratification
+corrCheck <- st_intersection(naturetypes, infra)
+
+corrCheck |>
+  as_tibble() |>
+  mutate(i_ADSV_fct = floor(round(i_ADSV*10,2))/10,
+         i_alien_fct = floor(round(i_alien*10,2))/10) |>
+  count(i_ADSV_fct, infrastructureIndex) |>
+  View()
+
+# A first look
+corrCheck |>
+  mutate(i_ADSV_fct = floor(round(i_ADSV*10,2))/10,
+         i_alien_fct = floor(round(i_alien*10,2))/10) |>
+  pivot_longer(cols = c(i_ADSV_fct, i_alien_fct),
+               values_to = "indicatorValue",
+               names_to = "indicator") |>
+  filter(!is.na(indicatorValue)) |>
+  ggplot(aes(
+    x = factor(infrastructureIndex), 
+    fill = factor(round(indicatorValue,2))))+
+  geom_bar(
+    position="fill"
+    )+
+  theme_bw(base_size = 12)+
+  guides(fill = guide_legend("Scaled indicator values"))+
+  ylab("Fraction of data points")+
+  xlab("HIA")+
+  scale_fill_brewer(palette = "RdYlGn")+
+  facet_grid(indicator~year)
+```
+
+```{r}
+#| eval: false
+
+# Using a colour gradient emphasises the first colour (dark green).
+# Lets try discrete colours, and merge som classes to simplify
+corrCheck |>
+  pivot_longer(cols = c(i_ADSV, i_alien),
+               values_to = "indicatorValue",
+               names_to = "indicator") |>
+  filter(!is.na(indicatorValue)) |>
+  mutate(condition = case_when(
+    indicatorValue <0.6 ~"Poor",
+    indicatorValue <0.8 ~"Intermediate",
+    .default = "Good"
+  )) |>
+  ggplot(aes(
+    x = factor(infrastructureIndex), 
+    fill = condition))+
+  geom_bar(
+    position="fill"
+    )+
+  theme_bw(base_size = 12)+
+  guides(fill = guide_legend("Scaled indicator values"))+
+  ylab("Fraction of data points")+
+  xlab("HIA")+
+  scale_y_log10() +
+  scale_fill_viridis_d(alpha = .8) +
+  facet_grid(~indicator)
+
+```
+
 ```{r infraMuniMaps}

### 2024-01-31 manuscript/HIAs.qmd
try different validation plots

@@
 
-```{r positionMap}
-#| eval: false
-
+```{r centroids}
+#| eval: true
 # Make map to show where the three municipalities are
@@
   st_centroid()
+```
+
+```{r positionMap}
+#| eval: false
 
@@
   round()
-
-round(max(latlong$Y)-min(latlong$Y), 1)
-
+#round(max(latlong$Y)-min(latlong$Y), 1)
 ```
@@
 ```{r readInfraStructureIndex}
-#| eval: false
+#| eval: true
 
@@
 infra <- readRDS(paste0(path_temp, "infrastructureIndex_discrete_vectorized.rds"))
+
+```
+
+```{r}
+#| eval: false
 
@@
 
-corrCheck |>
-  as_tibble() |>
-  mutate(i_ADSV_fct = floor(round(i_ADSV*10,2))/10,
-         i_alien_fct = floor(round(i_alien*10,2))/10) |>
-  count(i_ADSV_fct, infrastructureIndex) |>
-  View()
+#corrCheck |>
+#  as_tibble() |>
+#  mutate(i_ADSV_fct = floor(round(i_ADSV*10,2))/10,
+#         i_alien_fct = floor(round(i_alien*10,2))/10) |>
+#  count(i_ADSV_fct, infrastructureIndex) |>
+#  View()
 
@@
 ```
+```{r}
+#| eval: false
+
+# Try looking across all years
+# A first look
+corrCheck |>
+  mutate(i_ADSV_fct = floor(round(i_ADSV*10,2))/10,
+         i_alien_fct = floor(round(i_alien*10,2))/10) |>
+  pivot_longer(cols = c(i_ADSV_fct, i_alien_fct),
+               values_to = "indicatorValue",
+               names_to = "indicator") |>
+  filter(!is.na(indicatorValue)) |>
+  ggplot(aes(
+    x = factor(infrastructureIndex), 
+    fill = factor(round(indicatorValue,2))))+
+  geom_bar(
+    position="fill"
+    )+
+  theme_bw(base_size = 12)+
+  guides(fill = guide_legend("Scaled indicator values"))+
+  ylab("Fraction of data points")+
+  xlab("HIA")+
+  scale_fill_brewer(palette = "RdYlGn")+
+  facet_grid(~indicator)
+```
 
@@
     indicatorValue <0.8 ~"Intermediate",
-    .default = "Good"
-  )) |>
+    indicatorValue <0.91 ~"Good",
+    .default = "Perfect"
+  ),
+  condition = fct_reorder(condition, indicatorValue)
+  ) |>
+  ggplot(aes(
+    x = factor(infrastructureIndex), 
+    fill = condition))+
+  geom_bar(
+    position="fill"
+    )+
+  theme_bw(base_size = 12)+
+  guides(fill = guide_legend("Scaled indicator values"))+
+  ylab("Fraction of data points")+
+  xlab("HIA")+
+  #scale_y_log10() +
+  scale_fill_viridis_d(alpha = .8) +
+  facet_grid(~indicator)
+
+```
+```{r}
+#| eval: false
+
+# The figure changes a lot depending on which value gets the darker colour
+corrCheck |>
+  pivot_longer(cols = c(i_ADSV, i_alien),
+               values_to = "indicatorValue",
+               names_to = "indicator") |>
+  filter(!is.na(indicatorValue)) |>
+  mutate(condition = case_when(
+    indicatorValue <0.6 ~"Poor",
+    indicatorValue <0.8 ~"Intermediate",
+    indicatorValue <0.91 ~"Good",
+    .default = "Perfect"
+  ),
+  condition = fct_reorder(condition, -indicatorValue)
+  ) |>
+  ggplot(aes(
+    x = factor(infrastructureIndex), 
+    fill = condition))+
+  geom_bar(
+    position="fill"
+    )+
+  theme_bw(base_size = 12)+
+  guides(fill = guide_legend("Scaled indicator values"))+
+  ylab("Fraction of data points")+
+  xlab("HIA")+
+  #scale_y_log10() +
+  scale_fill_viridis_d(alpha = .8) +
+  facet_grid(~indicator)
+```
+
+```{r}
+#| eval: false
+
+# If I look at this on a log scale, it also changes and becomes less convincing
+corrCheck |>
+  pivot_longer(cols = c(i_ADSV, i_alien),
+               values_to = "indicatorValue",
+               names_to = "indicator") |>
+  filter(!is.na(indicatorValue)) |>
+  mutate(condition = case_when(
+    indicatorValue <0.6 ~"Poor",
+    indicatorValue <0.8 ~"Intermediate",
+    indicatorValue <0.91 ~"Good",
+    .default = "Perfect"
+  ),
+  condition = fct_reorder(condition, indicatorValue)
+  ) |>
   ggplot(aes(
@@
 
+```
+
+```{r}
+#| eval: false
+
+# HIAs 1 and 2 are quite similar, so perhaps I can combine them
+corrCheck |>
+  pivot_longer(cols = c(i_ADSV, i_alien),
+               values_to = "indicatorValue",
+               names_to = "indicator") |>
+  filter(!is.na(indicatorValue)) |>
+  mutate(condition = case_when(
+    indicatorValue <0.6 ~"Poor",
+    indicatorValue <0.8 ~"Intermediate",
+    indicatorValue <0.91 ~"Good",
+    .default = "Perfect"
+  ),
+  condition = fct_reorder(condition, indicatorValue),
+  HIA = case_when(
+    infrastructureIndex %in% c(1,2) ~ "1 & 2",
+    .default = factor(infrastructureIndex)
+  )
+  ) |>
+  ggplot(aes(
+    x = HIA, 
+    fill = condition))+
+  geom_bar(
+    position="fill"
+    )+
+  theme_bw(base_size = 12)+
+  guides(fill = guide_legend("Scaled indicator values"))+
+  ylab("Fraction of data points")+
+  xlab("HIA")+
+  scale_y_log10() +
+  scale_fill_viridis_d(alpha = .8) +
+  facet_grid(~indicator)
 ```

### 2024-01-31 manuscript/HIAs.qmd
include ditching

--- 2024-01-31 manuscript/HIAs.qmd
@@
 ```{r getRelevantNTs}
-#| eval: false
+#| eval: true
 
@@
 
-# Getting the nature types that have 7FA
-myVars <- "7FA"
-
-# naturetypes_summary2 <- naturetypes_summary %>%
-#  rowwise() %>%
-#  mutate(keepers = sum(c_across(
-#    all_of(myVars))>0, na.rm=T)) %>%
-#  filter(keepers >0,
-#         Ecosystem == "våtmark") %>%
-#  select(Nature_type, NiN_mainType, Year)
-#
-# naturetypes_summary2$Nature_type
-
-# Getting the nature types that have the ADSV variables
-myVars2 <- c("7TK", "7SE", "PRTK", "PRSL")
-# naturetypes_summary3 <- naturetypes_summary %>%
-#  rowwise() %>%
-#  mutate(keepers = sum(c_across(
-#    all_of(myVars2))>0, na.rm=T)) %>%
-#  filter(keepers >0,
-#         Ecosystem == "våtmark") %>%
-#  select(Nature_type, NiN_mainType, Year)
-#
-# naturetypes_summary3$Nature_type
-```
-
-```{r relevantNTs}
-#| eval: false
-
-# These are the nature types I'm keeping for 7FA.
-# Note the absence of ombrotriphic types
-alienNTS <- c(
-  "Slåttemyr",
-  "Semi-naturlig våteng",
-  "Rik åpen jordvannsmyr i mellomboreal sone",
-  "Rik åpen sørlig jordvannsmyr",
-  "Rik åpen jordvannsmyr i nordboreal og lavalpin sone"
-)
-
-# And these are the nature types that we use for the ADSV indicator
-advsNTs <- c( # fens
-  "Slåttemyr",
-  "Sørlig slåttemyr",
-  "Rik åpen jordvannsmyr i mellomboreal sone",
-  "Rik åpen sørlig jordvannsmyr",
-  "Semi-naturlig myr",
-  "Rik åpen jordvannsmyr i nordboreal og lavalpin sone",
-  # "Sørlig kaldkilde", # excluded springs
-
-  # bogs (mainly)
-  "Høgereligende og nordlig nedbørsmyr",
-  "Øyblandingsmyr",
-  "Konsentrisk høymyr",
-  "Sørlig nedbørsmyr",
-  "Atlantisk høymyr",
-  "Terrengdekkende myr",
-  "Eksentrisk høymyr",
-  "Kanthøymyr",
-  "Platåhøymyr",
-  "Palsmyr"
-)
+myVars <- c("7TK", "7SE", "PRTK", "PRSL", "7FA", "7GR-GI")
+nts <- naturetypes_summary %>%
+  rowwise() %>%
+  mutate(keepers = sum(c_across(
+    all_of(myVars))>0, na.rm=T)) |>
+  filter(keepers >0,
+         Ecosystem == "våtmark") |>
+  pull(Nature_type)
 ```
@@
 naturetypes <- sf::st_read(dsn = path_naturetypes)
-# 117k polygons (2023)
-# sometimes this functin struggles with Norwegian letters.
+# 117k polygons (2022)
+# sometimes this function struggles with Norwegian letters.
 # Restart R and try again.
@@
 naturetypes <- naturetypes |>
-  # keep only wetlands and our selected nature types
+  # keep only wetlands
   filter(
     hovedøkosystem == "våtmark",
-    naturtype %in% c(advsNTs, alienNTS)
+    naturtype %in% nts,
+    naturtype != "Kalkrik helofyttsump"
   ) |>
@@
   mutate(NiN_variable_value = as.numeric(NiN_variable_value)) |>
-  filter(NiN_variable_code %in% c(myVars, myVars2)) |>
+  filter(NiN_variable_code %in% myVars) |>
   select(
@@
   ) |>
-  st_transform(myCRS) # Chosing this to match the EDM (se further down)
-
-# 13k obs.
+  st_transform(myCRS) # Choosing this to match the EDM (se further down)
+# 19k obs.
+
+naturetypes |>
+  as_tibble() |>
+  count(natureType, sort=T) |>
+  mutate(natureType = fct_reorder(natureType, n)) |>
+  ggplot(aes(x = natureType, y = n))+
+  geom_col()+
+  coord_flip()
 
@@
              6 ~ 50,
-             7 ~ 75)
+             7 ~ 75),
+    .default = value
     )
@@
 naturetypes_wide <- naturetypes |>
+  filter(variable %in% c("7TK", "7SE", "PRTK", "PRSL")) |>
   # Column names starting with a number is problematic, so adding a prefix
@@
 
-# Taking he sum of 7SE and 7TK (incl the PR.. variabels)
+# Taking the sum of 7SE and 7TK (incl the PR.. variables)
 naturetypes_wide <- naturetypes_wide |>
@@
 #| eval: true
-# Now I will copy these ADVS-values into the sf object again
+
+# Now I will copy these ADVS-values into the sf object again, keeping things in wide format
 naturetypes <- naturetypes |>
-  left_join(naturetypes_wide |> select(id, var_7FA, ADSV), by = "id")
+  pivot_wider(#id_cols="id",
+              names_from = "variable",
+              values_from = "value") |>
+  left_join(naturetypes_wide |> select(id, ADSV), by = "id") |>
+  select(!c("7TK", "7SE", "PRSL", "PRTK")) 
+  # >
+  #pivot_longer(cols = c(ADSV, "7FA", "7GR-GI"),
+  #             names_to = "variable",
+  #             values_drop_na = T)
 ```
@@
 
-# Now I rescale the now continous variables using reference and threshold values
-# I will use the same reference levels/values for all of Norway:
+# Now I rescale the now continuous variables using reference and threshold values
+# I will use the same reference levels/values for all of Norway for ADSV and alien species:
 
@@
 threshold <- 10
+
+# For 7GR-GI I use this
+upper2 <- 1
+lower2 <- 5
+threshold2 <- 2.5 # = observable effect. Value 3 indicates a shit to a new type (grunntype)
 
@@
 eaTools::ea_normalise(data = naturetypes,
-                      vector = "var_7FA",
+                      vector = "7FA",
                       upper_reference_level = lower,
@@
 # The variables are really coarse
+
+eaTools::ea_normalise(data = naturetypes,
+                      vector = "7GR-GI",
+                      upper_reference_level = lower2,
+                      lower_reference_level = upper2,
+                      break_point = threshold2,
+                      plot=T,
+                      reverse = T
+                      )
+
 ```
@@
 naturetypes$i_alien <- eaTools::ea_normalise(data = naturetypes,
-                      vector = "var_7FA",
+                      vector = "7FA",
                       upper_reference_level = lower,
@@
                       break_point = threshold,
+                      reverse = T
+                      )
+
+naturetypes$i_ditch <- eaTools::ea_normalise(data = naturetypes,
+                      vector = "7GR-GI",
+                      upper_reference_level = lower2,
+                      lower_reference_level = upper2,
+                      break_point = threshold2,
                       reverse = T
@@
   mutate(i_ADSV_fct = floor(round(i_ADSV*10,2))/10,
-         i_alien_fct = floor(round(i_alien*10,2))/10) |>
-  pivot_longer(cols = c(i_ADSV_fct, i_alien_fct),
+         i_alien_fct = floor(round(i_alien*10,2))/10,
+         i_ditch_fct = floor(round(i_ditch*10,2))/10) |>
+  pivot_longer(cols = c(i_ADSV_fct, i_alien_fct, i_ditch_fct),
                values_to = "indicatorValue",
-               names_to = "indicator") |>
-  filter(!is.na(indicatorValue)) |>
+               names_to = "indicator",
+               values_drop_na =T) |>
   ggplot(aes(
     x = factor(infrastructureIndex), 
-    fill = factor(round(indicatorValue,2))))+
+    fill = factor(indicatorValue)))+
   geom_bar(
@@
 ```
+```{r}
+#| eval: false
+
+# The year effect is constant, so ignoring that
+corrCheck |>
+  mutate(i_ADSV_fct = floor(round(i_ADSV*10,2))/10,
+         i_alien_fct = floor(round(i_alien*10,2))/10,
+         i_ditch_fct = floor(round(i_ditch*10,2))/10) |>
+  pivot_longer(cols = c(i_ADSV_fct, i_alien_fct, i_ditch_fct),
+               values_to = "indicatorValue",
+               names_to = "indicator",
+               values_drop_na =T) |>
+  ggplot(aes(
+    x = factor(infrastructureIndex), 
+    fill = factor(indicatorValue)))+
+  geom_bar(
+    position="fill"
+    )+
+  theme_bw(base_size = 12)+
+  guides(fill = guide_legend("Scaled indicator values"))+
+  ylab("Fraction of data points")+
+  xlab("HIA")+
+  scale_fill_brewer(palette = "RdYlGn")+
+  facet_grid(~indicator)
+```
 
@@
 # Lets try discrete colours, and merge som classes to simplify
+corrCheck |>
+  pivot_longer(cols = c(i_ADSV, i_alien, i_ditch),
+               values_to = "indicatorValue",
+               names_to = "indicator",
+               values_drop_na = T) |>
+  mutate(condition = case_when(
+    indicatorValue <0.6 ~"Poor",
+    indicatorValue <0.8 ~"Intermediate",
+    indicatorValue <0.91 ~"Good",
+    .default = "Perfect"
+  ),
+  condition = fct_reorder(condition, indicatorValue)
+  ) |>
+  ggplot(aes(
+    x = factor(infrastructureIndex), 
+    fill = condition))+
+  geom_bar(
+    position="fill"
+    )+
+  theme_bw(base_size = 12)+
+  guides(fill = guide_legend("Scaled indicator values"))+
+  ylab("Fraction of data points")+
+  xlab("HIA")+
+  scale_fill_viridis_d(alpha = .8) +
+  facet_grid(~indicator)
+
+```
+```{r}
+#| eval: false
+
+# The figure changes a lot depending on which value gets the darker colour
+corrCheck |>
+  pivot_longer(cols = c(i_ADSV, i_alien, i_ditch),
+               values_to = "indicatorValue",
+               names_to = "indicator",
+               values_drop_na = T) |>
+  mutate(condition = case_when(
+    indicatorValue <0.6 ~"Poor",
+    indicatorValue <0.8 ~"Intermediate",
+    indicatorValue <0.91 ~"Good",
+    .default = "Perfect"
+  ),
+  condition = fct_reorder(condition, -indicatorValue)
+  ) |>
+  ggplot(aes(
+    x = factor(infrastructureIndex), 
+    fill = condition))+
+  geom_bar(
+    position="fill"
+    )+
+  theme_bw(base_size = 12)+
+  guides(fill = guide_legend("Scaled indicator values"))+
+  ylab("Fraction of data points")+
+  xlab("HIA")+
+  scale_fill_viridis_d(alpha = .8) +
+  facet_grid(~indicator)
+```
+
+```{r}
+#| eval: false
+
+# If I look at this on a log scale, it also changes and becomes less convincing for ADSV and alien
+corrCheck |>
+  pivot_longer(cols = c(i_ADSV, i_alien, i_ditch),
+               values_to = "indicatorValue",
+               names_to = "indicator",
+               values_drop_na = T) |>
+  mutate(condition = case_when(
+    indicatorValue <0.6 ~"Poor",
+    indicatorValue <0.8 ~"Intermediate",
+    indicatorValue <0.91 ~"Good",
+    .default = "Perfect"
+  ),
+  condition = fct_reorder(condition, -indicatorValue)
+  ) |>
+  ggplot(aes(
+    x = factor(infrastructureIndex), 
+    fill = condition))+
+  geom_bar(
+    position="fill"
+    )+
+  theme_bw(base_size = 12)+
+  guides(fill = guide_legend("Scaled indicator values"))+
+  ylab("Fraction of data points")+
+  xlab("HIA")+
+  scale_y_log10()+
+  scale_fill_viridis_d(alpha = .8) +
+  facet_grid(~indicator)
+
+# This is not correct I think as it evens out differences that are clearly there.
+# i_ditch is showin less of a pattern for example
+```
+
+```{r}
+#| eval: false
+
+# HIAs 1 and 2 are quite similar, so perhaps I can combine them
 corrCheck |>
@@
     indicatorValue <0.8 ~"Intermediate",
-    .default = "Good"
-  )) |>
+    indicatorValue <0.91 ~"Good",
+    .default = "Perfect"
+  ),
+  condition = fct_reorder(condition, indicatorValue),
+  HIA = case_when(
+    infrastructureIndex %in% c(1,2) ~ "1 & 2",
+    .default = factor(infrastructureIndex)
+  )
+  ) |>
   ggplot(aes(
-    x = factor(infrastructureIndex), 
+    x = HIA, 
     fill = condition))+
@@
   facet_grid(~indicator)
-
 ```

### 2024-01-31 manuscript/HIAs.qmd
methods for adding ditching

@@
 ```
+
 ```{r}
@@
 # This is not correct I think as it evens out differences that are clearly there.
-# i_ditch is showin less of a pattern for example
+# i_ditch is shown less of a pattern for example
 ```
@@
 
-# HIAs 1 and 2 are quite similar, so perhaps I can combine them
+# HIAs 1 and 2 are quite similar for i_alien and i_ADSV, so perhaps I can combine them
 corrCheck |>
@@
 
-In this study we focused on open mire ecosystems in Norway.
-For this ecosystem we have an ecosystem delineation map produced using remotely sensed data and a deep learning model [@bakkestuen_delineation_2023].
+This study makes use of a data set from from a standardised field survey of nature types in Norway that started in 2018 and which is still ongoing [@norwegian_environmental_agency_naturtyper_2024].
+We included data from 2018 to 2022.
+In this survey, selected nature types are delineated on a map, and each locality is scored on a range of variables relevant for describing the state and quality of nature.
+The surveys are commissioned with the goal of producing data relevant for immediate land-use decisions, and is therefore spatially biased, typically towards areas with high human impact or expected impact.
+In addition there is a thematic and size bias in the sampling protocol.
+For examples, for the forest ecosystem, rare, endangered or calcareous forest types are delineated, whereas more common or ordinary forest types are not.
+For that reason we we focused on open mire ecosystems in Norway where the thematic bias is less severe.
+Of all possible mire types, the survey only maps the following:
+
+-   Southern ombrotrophic mires \> 2500 m^2^
+
+-   Northern ombrotrophic mires \> 10.000 m^2^
+
+-   All semi-natural mires (minerotrophic fens)
+
+-   Calcareous southern fens \>500 m^2^
+
+-   Calcareous northern fens \>1000 m^2^
+
+In the above, *southern* refers to boreonemoral and southboreal zones, and *northern* refers to mid-boreal, north boreal, and alpine zones @asbjørnmoen1998 .
+In addition, the northern fens need to be even more calcareous than the southern fens in order to be surveyed.
+
+<!--# and the variable 7FK is not recorded in ombrotrophic mires -->
+
+In this paper we assume the survey is representative for the entire mire ecosystem in Norway.
+Although it is possible that smaller or less calcareous mires will score systematically different than the ones that are surveyed we do not think this is so much the case for our variables.
+It may also be that alien plants are more or less common on bogs relative to fens.
+For this we have no assumption.
+
+For the mire ecosystem we have an ecosystem delineation map produced using remotely sensed data and a deep learning model [@bakkestuen_delineation_2023].
 This model, which has a 90.9% precision when tested against independent field data, estimates 12.7% of the area in southern Norway is mire [@bakkestuen_delineation_2023].
@@
 
-To describe these two characteristics we use five condition variables from the same data set ( @tbl-variables).
-The data set is from a standardised field survey of nature types in Norway that started in 2018 and which is still ongoing.
-The data is freely available [@norwegian_environmental_agency_naturtyper_2024].
+To describe these two characteristics we use six condition variables from the same data set ( @tbl-variables).
+
 See @miljødirektoratet for the sampling protocol (in Norwegian).
-We included data from 2018 to 2023.
-In this survey, selected nature types are delineated on a map, and each locality is scored on a range of variables relevant for describing the state and quality of nature.
-The surveys are commissioned with the goal of producing data relevant for immediate land-use decisions, and is therefore spatially biased, typically towards areas with high human impact or expected impact.
-In addition there is a thematic and size bias in the sampling protocol.
-Of all possible mire types, the survey only maps the following:
-
--   Southern ombrotrophic mires \> 2500 m^2^
-
--   Northern ombrotrophic mires \> 10.000 m^2^
-
--   All semi-natural mires (minerotrophic fens)
-
--   Calcareous southern fens \>500 m^2^
-
--   Calcareous northern fens \>1000 m^2^
-
-In the above, *southern* refers to boreonemoral and southboreal zones, and *northern* refers to mid-boreal, north boreal, and alpine zones @asbjørnmoen1998 .
-In addition, the northern fens need to be even more calcareous than the southern fens in order to be surveyed, and the variable 7FK is not recorded in ombrotrophic mires.
-In this paper we assume the survey is representative for the entire mire ecosystem in Norway.
-Although it is possible that smaller or less calcareous mires will score systematically different than the ones that are surveyed we do not think this is the case for our variables at least.
-It may also be that alien plants are more or less common on bogs relative to fens.
-For this we have no assumption.
-
-| Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                    | Reference                          |
-|---------------|---------------|---------------|---------------|---------------|
-| 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                             | @runehalvorsen2019                 |
-| 7SE                  | Human caused abration or abration-caused erosion | Unitless, ordinal, non-linear scale from 1 to 4.
+
+| id  | Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                                                                        | Reference                          |
+|--|--------------|--------------|--------------|--------------|--------------|
+| 1   | 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                                                                                 | @runehalvorsen2019                 |
+| 2   | 7SE                  | Human caused abration or abration-caused erosion | Unitless, ordinal, non-linear scale from 1 to 4.
 | Measures the frequency of imagined 4 m^2^ quadrats layed over the area that has some sign of abration                                                              | @runehalvorsen2019                 |
-| PRSL                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.
+| 3   | PRSL                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.
 | Same as 7SE, but recorded at a higher resolution                                                                                                                   | @miljødirektoratet                 |
-| 7TK                  | Tracks from large vehicles                       | Unitless, ordinal, non-linear scale from 1 to 4.
+| 4   | 7TK                  | Tracks from large vehicles                       | Unitless, ordinal, non-linear scale from 1 to 4.
 | Measures the frequency of imagined 100 m^2^ quadrats layed over the area that has some signs of vehicle tracks                                                     | @rune_halvorsen_dokumentasjon_2019 |
-| PRTK                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.
+| 5   | PRTK                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.
 | Same as 7TK, but recorded at a higher resolution                                                                                                                   | @miljødirektoratet                 |
-| Infrastructure Index | Infrastructure Index                             | Unitelss linear scale from 0 to 13.2                                                    | Unitelss index ranging from from 0 to 13.2                                                                     | @erikstad_index_2023               |
-| HIA                  | Human Impact area                                | Ordinal, non-linear scale from 1 to 4                                                   | A categorical representation of the infrastructure index                                                       | this paper                         |
+| 6   | 7GR-GI               | Ditching intensity                               | Unitless, ordinal scale from 1 to 5                                                     | Describes the effect that drainage ditches is estimate to have on the species composition and environmental variables ones the system reached its new equilibrium  | @miljødirektoratet                 |
+| 7   | Infrastructure Index | Infrastructure Index                             | Unitelss linear scale from 0 to 13.2                                                    | Unitelss index ranging from from 0 to 13.2                                                                                                                         | @erikstad_index_2023               |
+| 8   | HIA                  | Human Impact area                                | Ordinal, non-linear scale from 1 to 4                                                   | A categorical representation of the infrastructure index                                                                                                           | this paper                         |
 
@@
 
+Variables 1 -5 were 7FK were originally recorded along frequency ranges.
+We the lower limit for each frequency range to convert them into percentages.
+This was done because the data was strongly right skewed .
 7FK was developed into a single indicator (i.e. a normalised variable) named *Alien plant cover*.
 Variables 2-5 describe very related aspects and so they where combined into one indicator called antropogenic disturbance to soil and vegetation, or ADSV for short.
-This was done by taking the value from the variable with the lowest value, i.e. worst condition.
-For both indicators, the lower and upper reference values, i.e.
-the worst and best possible condition that the variables can be in, is given by the range of the data where 0 or 1 is the lower limit and 4, 7 or 8 is the upper limit, depending on the variable (see @tbl-variables).
+This was done by summing the two variables after they had been converted to percentages.
+This was not a perfect solution, especially since some localities only had one of the variables recorded, but we chose this, rather than for example using a *worst rule* principle, in order to better separate the localities in terms of their indicator values.
+For both indicators (alien species and ADSV), the lower and upper reference values, i.e. the worst and best possible condition that the variables can be in, is defined as 100% and 0%, respectively.
+The threshold for *good ecosystem condition* was defined as 10%, and this was mapped to the value 0.6 on the rescaled indicators.
+Variable 7GR-GI was rescaled into an indicator named *ditching intensity* by using the lower and upper reference values 1 and 5, respectively, and a threshold value of 2.5.
+A variable value of 1 indicates an intact mire, and a value of 5 indicates a mire transitioning away from a wetland.
+A value of 2 indicator observable change within the range expected for the same mapping unit, and a value of 3 indicates a mire transitioning into a neighboring (ecologically speaking) mapping unit.
+
+The .
 See @fig-workflow for a schematic workflow for the indicator development.

### 2024-02-01 manuscript/HIAs.qmd
something with the table

@@
 | id  | Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                                                                       | Reference                          |
-|--|--------------|--------------|--------------|--------------|--------------|
+|-----|----------------------|--------------------------------------------------|-----------------------------------------------------------------------------------------|-------------------------------------------------------------------------------------------------------------------------------------------------------------------|------------------------------------|
 | 1   | 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                                                                                | @runehalvorsen2019                 |
@@
 | Municipality | Total terrestrial area (km^2^)                                            | \% of terrestrial area surveyed                                            | \% open mires in relation to total terrestrial area                         | Total mire area (km^2^)                                                  | %of mire area inside survey area                                                      | Number of mire polygons in survey                                 | Mean Infrastructure Index value                                         |
-|---------|---------|---------|---------|---------|---------|---------|---------|
+|--------------|---------------------------------------------------------------------------|----------------------------------------------------------------------------|-----------------------------------------------------------------------------|--------------------------------------------------------------------------|---------------------------------------------------------------------------------------|-------------------------------------------------------------------|-------------------------------------------------------------------------|
 | Nordre Follo | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(t_area_km)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(dk_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mirePercent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mire_km2)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mireInSurvery_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(n)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(meanHIA)` |

### 2024-02-21 manuscript/HIAs.qmd
change municipalities

@@
 # Some directories
-mireFile <- "P:/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif"
-
+
+# Ecosystem delineation map
+path_mire <- "P:/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif"
+
+# infrastructure index
 path_infrastructure <- ifelse(dir == "C:",
@@
 
-path_naturetypes <- ifelse(dir == "C:",
-  "R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb",
-  "/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb"
-)
-
-path_muni <- paste0("../data/Basisdata_0000_Norge_25833_Kommuner_FGDB.gdb")
+# field survey
+# # downloaded from https://kartkatalog.geonorge.no/metadata/naturtyper-miljoedirektoratets-instruks/eb48dd19-03da-41e1-afd9-7ebc3079265c
+path_naturetypes <- "../data/survey.gdb"
+
+# municipality outline
+path_muni <- "../data/Basisdata_0000_Norge_25833_Kommuner_FGDB.gdb"
 
@@
 
-```{r getRelevantNTs}
-#| eval: true
+```{r import}
+#| cache: true
 
 # I already did some work to identify the relevant nature types
-
 # summary file (https://github.com/NINAnor/ecosystemCondition/blob/main/data/naturetypes/natureType_summary.rds)
 naturetypes_summary <- readRDS("../data/natureType_summary.rds")
+
+# Survey data
+# The data is too big to be stored on GitHub
+# Impart polygon data set
+st_layers(path_naturetypes)
+naturetypes <- sf::st_read(dsn = path_naturetypes, layer = "naturtyper_nin_omr")
+# 142k polygons (2023)
+# 
+# Impart survey coverage map
+coverage <- sf::st_read(dsn = path_naturetypes, layer = "naturtyper_nin_dekning") |>
+  st_transform(myCRS)
+
+# Outline of norway (coastline)
+outline <- sf::read_sf("../data/outlineOfNorway_EPSG25833.shp") |>
+  st_transform(myCRS)
+
+# Municipalities
+# find the correct layer
+st_layers(path_muni)
+# read inn data and transform
+muni <- sf::read_sf(path_muni, layer = "kommune") |>
+  st_transform(myCRS)
+
+# Mire data
+mire_terra <- terra::rast(path_mire)
+```
+
+```{r getRelevantNTs}
+#| eval: true
 
@@
 ```{r natureTypeData}
-#| eval: false
-
-# The data is too big to be stored on GitHub
-
-# Impart polygon data set
-naturetypes <- sf::st_read(dsn = path_naturetypes)
-# 117k polygons (2022)
-# sometimes this function struggles with Norwegian letters.
-# Restart R and try again.
+#| eval: true
+#| cache: true
+
+# Clean the survey data
 
@@
 # 19k obs.
-
+```
+
+```{r}
+#| eval: false
+# Plot to show what the most common nature types in the data set are
 naturetypes |>
@@
   coord_flip()
-
-naturetypes |>
-  saveRDS("../data/surveyData.RDS")
-# head(naturetypes)
-```
-
-```{r readSurvey}
-#| eval: true
-
-# read cached rds file. This is not so big.
-naturetypes <- readRDS("../data/surveyData.RDS")
 ```
@@
 #| eval: true
-
-# I now want to take the variable and normalise it before I can then combine them
-# despite them being on different scales
-# I will first normalise by converting into %.
-# Remeber the ordinal categories represnts frequency ranges
-
+# I now want to take the variables and normalise them before I can then combine
+# them despite them being on different scales.
+# I will first normalise by converting into % (not for 7GR-GI).
+# Remember the ordinal categories represents frequency ranges
 # The data is strongly right skewed, so simply taking the center value of each
-# bin will not work.
+# bin will not work:
 naturetypes %>%
-  ggplot()+
+  ggplot() +
   theme_bw() +
   geom_histogram(aes(x = value),
-                 binwidth = 1) +
-  facet_wrap(.~variable,
-             scales = "free")
-
-# I will use th lower bound for each bin.
+    binwidth = 1
+  ) +
+  facet_wrap(. ~ variable,
+    scales = "free"
+  )
+
+# I will use the lower bound for each bin instead.
 # The exception in when the variable is 1, because then the lower bound
@@
   mutate(value = case_when(
-    # selecting the variable that follow the same 4 step scale
+    # selecting the variables that follow the same 4 step scale
     variable %in% c("7TK", "7SE", "7FA") ~
-      case_match(value,
+      case_match(
+        value,
         0 ~ 0,
-             1 ~ mean(c(0, 1/16))*100,
-             2 ~ 1/16*100,
-             3 ~ 50), # note that it is not possible to get a value of 1
+        1 ~ mean(c(0, 1 / 16)) * 100,
+        2 ~ 1 / 16 * 100,
+        3 ~ 50
+      ), # note that it is not possible to get a value of 1
     # selecting the eight step variables
-    variable %in% c('PRTK', 'PRSL') ~ 
-      case_match(value,
+    variable %in% c("PRTK", "PRSL") ~
+      case_match(
+        value,
         0 ~ 0,
@@
         6 ~ 50,
-             7 ~ 75),
+        7 ~ 75
+      ),
     .default = value
-    )
-  )
+  ))
 
 naturetypes %>%
-  ggplot()+
+  filter(variable != "7GR-GI") |>
+  ggplot() +
   theme_bw() +
@@
     color = "orange",
-                 fill = "orange") +
-  facet_wrap(.~variable,
-             scales = "free")
+    fill = "orange"
+  ) +
+  xlab("%") +
+  facet_wrap(. ~ variable,
+    scales = "free"
+  )
 
@@
 # First I will combine 7TK and PRTK, and also 7SE and PRSL.
-
 naturetypes_wide <- naturetypes_wide %>%
-  mutate(TK = if_else(
-    is.na(var_PRTK), var_7TK, var_PRTK),
+  mutate(
+    TK = if_else(
+      is.na(var_PRTK), var_7TK, var_PRTK
+    ),
     SE = if_else(
-      is.na(var_PRSL), var_7SE, var_PRSL)
+      is.na(var_PRSL), var_7SE, var_PRSL
+    )
   )
@@
     count(SE,
-        name = "sum") |>
+      name = "sum"
+    ) |>
     ggplot(
@@
         x = factor(SE),
-      y = sum))+
-  geom_bar(stat="identity",
-           fill="grey",
-           colour = "black")+
-  theme_bw(base_size = 12)+
-  labs(x = "7SE or PRSL score",
-       y = "Number of localities"),
-
+        y = sum
+      )
+    ) +
+    geom_bar(
+      stat = "identity",
+      fill = "grey",
+      colour = "black"
+    ) +
+    theme_bw(base_size = 12) +
+    labs(
+      x = "7SE or PRSL score",
+      y = "Number of localities"
+    ),
   naturetypes_wide %>%
@@
     count(TK,
-        name = "sum") |>
+      name = "sum"
+    ) |>
     ggplot(
@@
         x = factor(TK),
-      y = sum))+
-  geom_bar(stat="identity",
-           fill="grey",
-           colour = "black")+
-  theme_bw(base_size = 12)+
-  labs(x = "7TK or PRTK score",
-       y = "Number of localities")
-)
-
-# The NA's represnets localities where just one of the two variables 
+        y = sum
+      )
+    ) +
+    geom_bar(
+      stat = "identity",
+      fill = "grey",
+      colour = "black"
+    ) +
+    theme_bw(base_size = 12) +
+    labs(
+      x = "7TK or PRTK score",
+      y = "Number of localities"
+    )
+)
+
+# The NA's represents localities where just one of the two variables
 # (then thinking 7SE and PRSL as the same variable)
@@
 # one with the highest value (worst-rule) og the sum.
-# Sum is problematic is not all locations ahve two values to sum together.
+# Sum is problematic is not all locations have two values to sum together.
 # But the other option is problematic since I think field workers often
 # tend to split the effects over two variables is they have that option.
-# And if we have 50% vehicle damaeg and 50% hiking damage, that is no doubt
+# And if we have 50% vehicle damage and 50% hiking damage, that is no doubt
 # worst than just having 50% of either. So I will use the sum, despite its issues.
@@
 #| eval: true
-
 # Taking the sum of 7SE and 7TK (incl the PR.. variables)
@@
   rowwise() |>
-  mutate(ADSV = sum(c(SE, TK), na.rm=TRUE))
+  mutate(ADSV = sum(c(SE, TK), na.rm = TRUE))
 
@@
   count(ADSV,
-        name = "sum") |>
+    name = "sum"
+  ) |>
   ggplot(
     aes(
-      x = (ADSV),
-      y = sum))+
-  geom_bar(stat="identity",
-           fill="grey",
-           colour = "black")+
-  theme_bw(base_size = 12)+
-  labs(x = "Summed ADVS score",
-       y = "Number of localities") +
+      x = ADSV,
+      y = sum
+    )
+  ) +
+  geom_bar(
+    stat = "identity",
+    fill = "grey",
+    colour = "black"
+  ) +
+  theme_bw(base_size = 12) +
+  labs(
+    x = "Summed ADVS score",
+    y = "Number of localities"
+  ) +
   scale_x_continuous(
-    labels = scales::label_number(accuracy = 1))
+    labels = scales::label_number(accuracy = 1)
+  )
 
@@
 naturetypes <- naturetypes |>
-  pivot_wider(#id_cols="id",
+  pivot_wider(
     names_from = "variable",
-              values_from = "value") |>
+    values_from = "value"
+  ) |>
   left_join(naturetypes_wide |> select(id, ADSV), by = "id") |>
@@
   #             values_drop_na = T)
+head(naturetypes)
 ```
@@
 #| eval: true
-
 # Adding scaled indicator values to the dataset
-naturetypes$i_ADSV <- eaTools::ea_normalise(data = naturetypes,
+naturetypes$i_ADSV <- eaTools::ea_normalise(
+  data = naturetypes,
   vector = "ADSV",
@@
 
-naturetypes$i_alien <- eaTools::ea_normalise(data = naturetypes,
+naturetypes$i_alien <- eaTools::ea_normalise(
+  data = naturetypes,
   vector = "7FA",
@@
 
-naturetypes$i_ditch <- eaTools::ea_normalise(data = naturetypes,
+naturetypes$i_ditch <- eaTools::ea_normalise(
+  data = naturetypes,
   vector = "7GR-GI",
@@
 
-```{r getNorwayOutline}
-#| eval: true
-
-outline <- sf::read_sf("../data/outlineOfNorway_EPSG25833.shp") |>
-  st_transform(myCRS)
-```
-
 ```{r getMunicipalities}
@@
 
-# find the correct layer
-# st_layers(path_muni)
-
-# read inn data and transform
-muni <- sf::read_sf(path_muni, layer = "kommune") |>
-  st_transform(myCRS)
-
-# This returns some multisurfaces as well
+# The data contains some multisurfaces as well
 # table(st_geometry_type(muni))
@@
     "3020", # Nordre Follo
-    "5001", # Trondheim
-    "5418" # Målselv
+    "3451", # Nord-Aurdal
+    "3446" # Gran
   )) |>
@@
     kommunenummer == "3020" ~ "Nordre Follo",
-    kommunenummer == "5001" ~ "Trondheim",
-    kommunenummer == "5418" ~ "Målselv"
+    kommunenummer == "3451" ~ "Nord-Aurdal",
+    kommunenummer == "3446" ~ "Gran"
   ))
@@
   filter(kommunenummer == "3020")
-tr <- muni3 |>
-  filter(kommunenummer == "5001")
-me <- muni3 |>
-  filter(kommunenummer == "5418")
+na <- muni3 |>
+  filter(kommunenummer == "3451")
+gr <- muni3 |>
+  filter(kommunenummer == "3446")
 ```
@@
     name = "Number of polygons")
+# There where some polygons that spanned municipal borders. 
+# It's not a problem
 
 # and also to get the data coverage polygon.
-# These I had to manually download from naturbase.no
-dk <- read_sf("../data/dekningskartNF/Naturtyper NiN - Dekningskart.shp") |>
-  rbind(read_sf("../data/dekningskartTR/Naturtyper NiN - Dekningskart.shp")) |>
-  rbind(read_sf("../data/dekningskartME/Naturtyper NiN - Dekningskart.shp")) |>
-  st_transform(st_crs(muni3)) |>
-  st_intersection(muni3) |>
-  st_union()
+coverage3 <- coverage |>
+  st_intersection(muni3)
+
+## These I had to manually download from naturbase.no
+#dk <- read_sf("../data/dekningskartNF/Naturtyper NiN - Dekningskart.shp") |>
+#  rbind(read_sf("../data/dekningskartTR/Naturtyper NiN - Dekningskart.shp")) |>
+#  rbind(read_sf("../data/dekningskartME/Naturtyper NiN - Dekningskart.shp")) |>
+#  st_transform(st_crs(muni3)) |>
+#  st_intersection(muni3) |>
+#  st_union()
 ```
@@
 
-(positionMap <- tm_shape(outline) +
+inc <- 1e5
+myBbox <- st_bbox(centroids)
+myBbox[1:2] <- myBbox[1:2]-inc 
+myBbox[3:4] <- myBbox[3:4]+inc
+
+(positionMap <- 
+  tm_shape(outline,
+           bbox = myBbox) +
   tm_polygons() +
@@
     text = "Municipality",
+    just= "left",
     size = .8,
-    xmod = 2,
-    ymod = -1
+    xmod = 1,
+    ymod = 0
   ) +
@@
     bg.color = "skyblue",
-    outer.margins = c(0.01, .02, .02, .02)))
+    outer.margins = c(0.01, .02, .02, .02))+
+  tm_compass()+
+  tm_scale_bar()+
+    tm_basemap(leaflet::providers$Stamen.Watercolor)
+)
+
 ```

### 2024-02-21 manuscript/HIAs.qmd
fix methods map

@@
 library(units)
+library(rnaturalearth)
+library(rnaturalearthdata)
 
@@
 mire_terra <- terra::rast(path_mire)
+
+# Infrastructure index (read proxy)
+infra <- stars::read_stars(path_infrastructure)
+
 ```
@@
 ```{r positionMap}
-#| eval: false
-
-inc <- 1e5
+#| eval: true
+
+world <- ne_countries(scale = "medium", returnclass = "sf") |>
+  st_transform(myCRS) |>
+  filter(admin %in% c("Norway", "Sweden")) |>
+  st_make_valid()
+
+inc <- 200000
 myBbox <- st_bbox(centroids)
@@
 (positionMap <- 
-  tm_shape(outline,
+  tm_shape(world,
            bbox = myBbox) +
     tm_polygons() +
+  tm_shape(muni3) +
+    tm_polygons(col = "green") +
   tm_shape(centroids) +
@@
   ) +
-  tm_symbols(
-    size = 1.5,
-    shape = 17,
-    col = "black"
-  ) +
   tm_grid(projection = 4326) +
@@
   tm_compass()+
-  tm_scale_bar()+
-    tm_basemap(leaflet::providers$Stamen.Watercolor)
+  tm_scale_bar()
 )
@@
 
-# what is the distance between Nordre Follo and Målselv
-(latlong <- centroids |>
-  st_transform(4326) |>
-st_coordinates() |>
-  as_tibble())
-
-km_distance <- centroids |>
+# what is the distance between Nordre Follo and Nord-Aurdal
+(km_distance <- centroids |>
   st_distance() |>
@@
   drop_units() |>
-  round()
-#round(max(latlong$Y)-min(latlong$Y), 1)
-```
-
-```{r mireDataTestCrop}
-#| eval: false
-
-# Testing the cropping proceedure using a downsampled starts object
-mire_test <- read_stars(mireFile, proxy = T) |>
-  setNames("mire") |>
-  st_downsample(100) |>
-  mutate(mire = na_if(mire, 0)) |>
-  st_as_stars()
-
-mire_test_nf <- mire_test |>
-  st_crop(nf)
-
-# The cropping is correct:
-# tmap_mode("plot")
-# tm_shape(mire_test)+
-#  tm_raster(palette = "red")+
-#  tm_shape(outline)+
-#  tm_borders()+
-#  tm_shape(mire_test_nf)+
-#  tm_raster(col="red")
-
-# Then we do the same, but without downsampling
-
-mire_nf <- read_stars(mireFile, proxy = T) |>
-  setNames("mire") |>
-  # mutate(mire = na_if(mire, 0)) |>
-  st_crop(nf) |>
-  st_as_stars()
-# The cropped raster is for some reason far into the ocean!
-
-# The cells have values, so it seems like there is something
-# that happens after the cropping
-tmap_mode("plot")
-tm_shape(mire_test) +
-  tm_raster(palette = "red") +
-  tm_shape(outline) +
-  tm_borders() +
-  tm_shape(mire_nf) +
-  tm_raster(col = "red")
+  round())
 ```
@@
 ```{r mireTerra}
-#| eval: false
-
+#| eval: true
+
+# I first tried to import and crop the mire data using starts, but that failed (se pre 21 feb 2023)
 # trying { terra } instead
@@
   terra::vect()
+gr_vect <- as(gr, "Spatial") |>
+  terra::vect()
+na_vect <- as(na, "Spatial") |>
+  terra::vect()
 
 # crop and mask (very fast!)
-mire_nf_terra <- terra::rast(mireFile) |>
+mire_terra_nf <- mire_terra |>
   terra::crop(nf_vect) |>
   terra::mask(nf_vect)
+
+mire_terra_gr <- mire_terra |>
+  terra::crop(gr_vect) |>
+  terra::mask(gr_vect)
+
+mire_terra_na <- mire_terra |>
+  terra::crop(na_vect) |>
+  terra::mask(na_vect)
 
@@
 # I like the stars, sf and tmap combo better, so I return to stars
-mire_nf_stars <- mire_nf_terra |>
+mire_stars_nf <- mire_terra_nf |>
   st_as_stars()
-
-ggplot() +
-  geom_stars(data = mire_nf_stars) +
-  coord_equal() +
-  theme_void() +
-  scale_fill_viridis_c()
+mire_stars_gr <- mire_terra_gr |>
+  st_as_stars()
+mire_stars_na <- mire_terra_na |>
+  st_as_stars()
+
+par(mfrow=c(3,1))
+plot(mire_stars_nf)
+plot(mire_stars_gr)
+plot(mire_stars_na)
+
 ```
@@
 # values goes into summary table in the ms
-dk2 <- read_sf("../data/dekningskartNF/Naturtyper NiN - Dekningskart.shp") |>
-  rbind(read_sf("../data/dekningskartTR/Naturtyper NiN - Dekningskart.shp")) |>
-  rbind(read_sf("../data/dekningskartME/Naturtyper NiN - Dekningskart.shp")) |>
-  st_transform(st_crs(muni3)) |>
-  st_intersection(muni3) |>
-  st_cast("POLYGON") |>
+dk2 <- coverage3 |>
   group_by(Municipality) |>
-  summarise(geometry = st_union(geometry)) |>
+  summarise(SHAPE = st_union(SHAPE)) |>
   mutate(
-    dk_area_km = geometry |> st_area(),
+    dk_area_km = SHAPE |> st_area(),
     dk_area_km = round(units::drop_units(dk_area_km * 1e-6))
@@
 # -- Nordre Follo
-mireArea <- mire_nf_terra |>
+(mireArea <- mire_terra_nf |>
   global(c("mean", "sum"), na.rm = T) |>
@@
     mire_km2 = sum / 1e+4
-  )
+  ))
+
+# -- Gran
+mireArea2 <- mire_terra_gr |>
+  global(c("mean", "sum"), na.rm = T) |>
+  add_column("Municipality" = "Gran") |>
+  mutate(
+    mirePercent = round(mean * 100, 1),
+    mire_km2 = sum / 1e+4
+  )
+
+# -- Nord-Aurdal
+mireArea3 <- mire_terra_na |>
+  global(c("mean", "sum"), na.rm = T) |>
+  add_column("Municipality" = "Nord-Aurdal") |>
+  mutate(
+    mirePercent = round(mean * 100, 1),
+    mire_km2 = sum / 1e+4
+  )
+
+mireArea <- mireArea |>
+  rbind(mireArea2, mireArea3)
 
@@
 # -- Nordre Follo
-mire_nf_terra_dk <- mire_nf_terra |>
+mire_in_dk <- mire_terra_nf |>
   terra::mask(dk2 |> filter(Municipality == "Nordre Follo")) |>
@@
   add_column(Municipality = "Nordre Follo")
+
+mire_in_dk2 <- mire_terra_gr |>
+  terra::mask(dk2 |> filter(Municipality == "Gran")) |>
+  global("sum", na.rm = T) |>
+  mutate(mireInSurvey_km2 = sum / 1e+4) |>
+  add_column(Municipality = "Gran")
+mire_in_dk3 <- mire_terra_na |>
+  terra::mask(dk2 |> filter(Municipality == "Nord-Aurdal")) |>
+  global("sum", na.rm = T) |>
+  mutate(mireInSurvey_km2 = sum / 1e+4) |>
+  add_column(Municipality = "Nord-Aurdal")
+
+mire_in_dk <- mire_in_dk |>
+  rbind(mire_in_dk2, mire_in_dk3)
 ```
@@
 #| eval: false
-
 # this data is on a 100x100m grid
-infra <- stars::read_stars(path_infrastructure) |>
-  # , which is mor then we need - warp it to 1x1km
+infra <- infra |>
+  # , which is more then we need - warp it to 1x1km
   st_warp(
@@
   )) |>
-  # takeing away point in the sea
+  # taking away point in the sea
   st_crop(outline)
 
-# This step might seem rather stupid.
-We want to vectorize a rather large raster.
-This makes it a quite big data object.
-The reason is that there is no really good way to burn polygon data on to raster grid cells after the disuse of the raster package.
-It was not straight forward then either.
-But calculating intersections between polygons is very fast and easy.
+# This step might seem rather stupid. We want to vectorize a rather large 
+# raster. This makes it a quite big data object. The reason is that there is no 
+# really good way to burn polygon data on to raster grid cells after the disuse 
+# of the raster package. It was not straight forward then either. But 
+# calculating intersections between polygons is very fast and easy.
 
@@
 
-```{r readInfraStructureIndex}
+```{r InfraStructureIndex}
 #| eval: true
@@
 ```{r}
-#| eval: false
-
-# Calulculate area
+#| eval: true
+
+# Calculate area
 infra <- infra |>
-  mutate(area = geometry |> st_area(),
-         area_km = area |> set_units("km2"))
+  mutate(
+    area = geometry |> st_area(),
+    area_km = area |> set_units("km2")
+  )
 
@@
   summarise(area_km = sum(area_km)) |>
-  ggplot(aes(x = infrastructureIndex,
-             y = area_km)) +
+  ggplot(aes(
+    x = infrastructureIndex,
+    y = area_km
+  )) +
   geom_col()
@@
 ```
-
-```{r HIA-validate}
-#| eval: false
-
+```{r corrCheck}
+#| eval: true
 # Now I want to see if the indicator values depend on the HIA is a predictable
@@
 
-#corrCheck |>
+# corrCheck |>
 #  as_tibble() |>
@@
 
+```
+
+```{r HIA-validate}
+#| eval: false
 # A first look
 corrCheck |>
-  mutate(i_ADSV_fct = floor(round(i_ADSV*10,2))/10,
-         i_alien_fct = floor(round(i_alien*10,2))/10,
-         i_ditch_fct = floor(round(i_ditch*10,2))/10) |>
-  pivot_longer(cols = c(i_ADSV_fct, i_alien_fct, i_ditch_fct),
+  mutate(
+    i_ADSV_fct = floor(round(i_ADSV * 10, 2)) / 10,
+    i_alien_fct = floor(round(i_alien * 10, 2)) / 10,
+    i_ditch_fct = floor(round(i_ditch * 10, 2)) / 10
+  ) |>
+  pivot_longer(
+    cols = c(i_ADSV_fct, i_alien_fct, i_ditch_fct),
     values_to = "indicatorValue",
     names_to = "indicator",
-               values_drop_na =T) |>
+    values_drop_na = T
+  ) |>
   ggplot(aes(
     x = factor(infrastructureIndex),
-    fill = factor(indicatorValue)))+
+    fill = factor(indicatorValue)
+  )) +
   geom_bar(
-    position="fill"
-    )+
-  theme_bw(base_size = 12)+
-  guides(fill = guide_legend("Scaled indicator values"))+
-  ylab("Fraction of data points")+
-  xlab("HIA")+
-  scale_fill_brewer(palette = "RdYlGn")+
-  facet_grid(indicator~year)
+    position = "fill"
+  ) +
+  theme_bw(base_size = 12) +
+  guides(fill = guide_legend("Scaled indicator values")) +
+  ylab("Fraction of data points") +
+  xlab("HIA") +
+  scale_fill_brewer(palette = "RdYlGn") +
+  facet_grid(indicator ~ year)
 ```
@@
 #| eval: false
-
 # The year effect is constant, so ignoring that
 corrCheck |>
-  mutate(i_ADSV_fct = floor(round(i_ADSV*10,2))/10,
-         i_alien_fct = floor(round(i_alien*10,2))/10,
-         i_ditch_fct = floor(round(i_ditch*10,2))/10) |>
-  pivot_longer(cols = c(i_ADSV_fct, i_alien_fct, i_ditch_fct),
+  mutate(
+    i_ADSV_fct = floor(round(i_ADSV * 10, 2)) / 10,
+    i_alien_fct = floor(round(i_alien * 10, 2)) / 10,
+    i_ditch_fct = floor(round(i_ditch * 10, 2)) / 10
+  ) |>
+  pivot_longer(
+    cols = c(i_ADSV_fct, i_alien_fct, i_ditch_fct),
     values_to = "indicatorValue",
     names_to = "indicator",
-               values_drop_na =T) |>
+    values_drop_na = T
+  ) |>
   ggplot(aes(
     x = factor(infrastructureIndex),
-    fill = factor(indicatorValue)))+
+    fill = factor(indicatorValue)
+  )) +
   geom_bar(
-    position="fill"
-    )+
-  theme_bw(base_size = 12)+
-  guides(fill = guide_legend("Scaled indicator values"))+
-  ylab("Fraction of data points")+
-  xlab("HIA")+
-  scale_fill_brewer(palette = "RdYlGn")+
+    position = "fill"
+  ) +
+  theme_bw(base_size = 12) +
+  guides(fill = guide_legend("Scaled indicator values")) +
+  ylab("Fraction of data points") +
+  xlab("HIA") +
+  scale_fill_brewer(
+    palette = "RdYlGn") +
   facet_grid(~indicator)
+```
+
+```{r}
+#| eval: true
+# Using a color gradient emphasizes the first color (dark green).
+# Lets try discrete colors, and merge som classes to simplify
+(validationPlot <- corrCheck |>
+  #rename(ADSV = i_ADSV) |>
+  pivot_longer(
+    cols = c(i_ADSV, i_alien, i_ditch),
+    values_to = "indicatorValue",
+    names_to = "indicator",
+    values_drop_na = T
+  ) |>
+  mutate(
+    condition = case_when(
+      indicatorValue < 0.6 ~ "<0.6",
+      indicatorValue < 0.8 ~ "0.6 to 0.8",
+      indicatorValue < 0.91 ~ "0.8 to 0.9",
+      .default = "0.9 to 1"
+    ),
+    condition = fct_reorder(condition, indicatorValue),
+    indicator = case_when(
+      indicator == "i_ADSV" ~ "ADSV",
+      indicator == "i_alien" ~ "Alien species",
+      indicator == "i_ditch" ~ "Trenching"
+    )
+  ) |>
+  ggplot(aes(
+    x = factor(infrastructureIndex),
+    fill = condition
+  )) +
+  geom_bar(
+    position = "fill"
+  ) +
+  theme_bw(base_size = 12) +
+  guides(fill = guide_legend("Indicator values")) +
+  ylab("Fraction of data points") +
+  xlab("Human Impact Areas") +
+  scale_fill_brewer(palette = "RdYlGn") +
+  facet_grid(~indicator)
+)
 ```
@@
 #| eval: false
-
-# Using a colour gradient emphasises the first colour (dark green).
-# Lets try discrete colours, and merge som classes to simplify
+# The figure changes a lot depending on which value gets the darker colour
 corrCheck |>
-  pivot_longer(cols = c(i_ADSV, i_alien, i_ditch),
+  pivot_longer(
+    cols = c(i_ADSV, i_alien, i_ditch),
     values_to = "indicatorValue",
     names_to = "indicator",
-               values_drop_na = T) |>
-  mutate(condition = case_when(
-    indicatorValue <0.6 ~"Poor",
-    indicatorValue <0.8 ~"Intermediate",
-    indicatorValue <0.91 ~"Good",
-    .default = "Perfect"
-  ),
-  condition = fct_reorder(condition, indicatorValue)
-  ) |>
-  ggplot(aes(
-    x = factor(infrastructureIndex), 
-    fill = condition))+
-  geom_bar(
-    position="fill"
-    )+
-  theme_bw(base_size = 12)+
-  guides(fill = guide_legend("Scaled indicator values"))+
-  ylab("Fraction of data points")+
-  xlab("HIA")+
-  scale_fill_viridis_d(alpha = .8) +
-  facet_grid(~indicator)
-
-```
-
-```{r}
-#| eval: false
-
-# The figure changes a lot depending on which value gets the darker colour
-corrCheck |>
-  pivot_longer(cols = c(i_ADSV, i_alien, i_ditch),
-               values_to = "indicatorValue",
-               names_to = "indicator",
-               values_drop_na = T) |>
-  mutate(condition = case_when(
-    indicatorValue <0.6 ~"Poor",
-    indicatorValue <0.8 ~"Intermediate",
-    indicatorValue <0.91 ~"Good",
+    values_drop_na = T
+  ) |>
+  mutate(
+    condition = case_when(
+      indicatorValue < 0.6 ~ "Poor",
+      indicatorValue < 0.8 ~ "Intermediate",
+      indicatorValue < 0.91 ~ "Good",
       .default = "Perfect"
@@
     x = factor(infrastructureIndex),
-    fill = condition))+
+    fill = condition
+  )) +
   geom_bar(
-    position="fill"
-    )+
-  theme_bw(base_size = 12)+
-  guides(fill = guide_legend("Scaled indicator values"))+
-  ylab("Fraction of data points")+
-  xlab("HIA")+
+    position = "fill"
+  ) +
+  theme_bw(base_size = 12) +
+  guides(fill = guide_legend("Scaled indicator values")) +
+  ylab("Fraction of data points") +
+  xlab("HIA") +
   scale_fill_viridis_d(alpha = .8) +
@@
 #| eval: false
-
 # If I look at this on a log scale, it also changes and becomes less convincing for ADSV and alien
 corrCheck |>
-  pivot_longer(cols = c(i_ADSV, i_alien, i_ditch),
+  pivot_longer(
+    cols = c(i_ADSV, i_alien, i_ditch),
     values_to = "indicatorValue",
     names_to = "indicator",
-               values_drop_na = T) |>
-  mutate(condition = case_when(
-    indicatorValue <0.6 ~"Poor",
-    indicatorValue <0.8 ~"Intermediate",
-    indicatorValue <0.91 ~"Good",
+    values_drop_na = T
+  ) |>
+  mutate(
+    condition = case_when(
+      indicatorValue < 0.6 ~ "Poor",
+      indicatorValue < 0.8 ~ "Intermediate",
+      indicatorValue < 0.91 ~ "Good",
       .default = "Perfect"
@@
     x = factor(infrastructureIndex),
-    fill = condition))+
+    fill = condition
+  )) +
   geom_bar(
-    position="fill"
-    )+
-  theme_bw(base_size = 12)+
-  guides(fill = guide_legend("Scaled indicator values"))+
-  ylab("Fraction of data points")+
-  xlab("HIA")+
-  scale_y_log10()+
+    position = "fill"
+  ) +
+  theme_bw(base_size = 12) +
+  guides(fill = guide_legend("Scaled indicator values")) +
+  ylab("Fraction of data points") +
+  xlab("HIA") +
+  scale_y_log10() +
   scale_fill_viridis_d(alpha = .8) +
@@
     style = "cat",
-    title = "Human Impact\nArea Class"
+    title = "Human\nImpact\nArea"
   ) +
@@
 
-# Syrvey coverage and other colour overlayed the municipalities
+# Survey coverage and other colour overlayed the municipalities
 muniPlot <- tm_shape(muni3) +
@@
   ) +
-  tm_shape(dk) +
+  tm_shape(dk2) +
   tm_polygons(

### 2024-02-22 manuscript/HIAs.qmd
geom_taxt inside bars

@@
 ```
-
 ```{r}
-#| eval: true
-# Using a color gradient emphasizes the first color (dark green).
-# Lets try discrete colors, and merge som classes to simplify
-(validationPlot <- corrCheck |>
-  #rename(ADSV = i_ADSV) |>
+temp <- corrCheck |>
   pivot_longer(
@@
   ) |>
+  as_tibble() |>
+  group_by(indicator, infrastructureIndex, condition) |>
+  summarise(n = n()) |>
+  ungroup() |>
+  group_by(indicator, infrastructureIndex) |>
+  mutate(lab = round(n/sum(n)*100),
+         lab = case_when(
+           lab < 5 ~ NA,
+           .default = paste0(lab, "%")
+         ))
+
+nrow(temp)
+
+temp |>
+ggplot(aes(
+    x = infrastructureIndex,
+    y = n,
+    fill = condition
+  )) +
+  geom_bar(
+    position = "fill",
+    stat = "identity"
+  ) +
+  geom_text(aes(label = lab),
+    position = position_fill(vjust = 0.5),
+    color= "black", vjust = 0.5, size = 4) +
+  theme_minimal(base_size = 15) +
+  theme(
+    panel.grid = element_blank(), 
+    axis.text.x = element_text(margin = margin(t = -10)),
+    axis.text.y = element_blank(),
+    strip.text = element_textbox(
+      size = 12,
+      halign = 0.5, linetype = 1, r = unit(5, "pt"), width = unit(1, "npc"),
+      padding = margin(2, 0, 1, 0), margin = margin(3, 3, 3, 3))) +
+  guides(fill = guide_legend("Indicator values")) +
+  ylab("Fraction of data points") +
+  xlab("Human Impact Areas") +
+  scale_fill_manual(values = c("#E85437","#FBBF00", "#B5DF73", "#009000")) +
+  facet_grid(~indicator)
+
+```
+
+```{r}
+#| eval: true
+# Using a color gradient emphasizes the first color (dark green).
+# Lets try discrete colors, and merge some classes to simplify
+(validationPlot <- corrCheck |>
+  #rename(ADSV = i_ADSV) |>
+  pivot_longer(
+    cols = c(i_ADSV, i_alien, i_ditch),
+    values_to = "indicatorValue",
+    names_to = "indicator",
+    values_drop_na = T
+  ) |>
+  mutate(
+    condition = case_when(
+      indicatorValue < 0.6 ~ "<0.6",
+      indicatorValue < 0.8 ~ "0.6 to 0.8",
+      indicatorValue < 0.91 ~ "0.8 to 0.9",
+      .default = "0.9 to 1"
+    ),
+    condition = fct_reorder(condition, indicatorValue),
+    indicator = case_when(
+      indicator == "i_ADSV" ~ "ADSV",
+      indicator == "i_alien" ~ "Alien species",
+      indicator == "i_ditch" ~ "Trenching"
+    )
+  ) |>
   ggplot(aes(
@@
   ) +
-  theme_bw(base_size = 12) +
+  theme_minimal(base_size = 12) +
+  theme(
+    panel.grid = element_blank(), 
+    axis.text.x = element_markdown(),
+    axis.text.y = element_blank()) +
   guides(fill = guide_legend("Indicator values")) +
@@
   heights = c(.6, .4),
-  outer.margins = NULL
+  outer.margins = .01
 ))

### 2024-02-22 manuscript/HIAs.qmd
methods map no country

@@
   xlab("Human Impact Areas") +
-  scale_fill_manual(values = c("#E85437","#FBBF00", "#B5DF73", "#009000")) +
+  scale_fill_manual(values = c("#E85437","#FBAF00", "#B5DF73", "#009000")) +
   facet_grid(~indicator)
@@
 # Infratructure in each municipality
-infraMuniMap <- tm_shape(muni3) +
+(infraMuniMap <- tm_shape(muni3) +
   tm_borders() +
@@
     style = "cat",
-    title = "Human Impact\nArea Class"
-  ) +
+    title = "Human Impact Areas"
+  ) +
+  tm_layout(
+    legend.show = F,
+    panel.label.height = 0) +
   tm_shape(muni3) +
   tm_borders(lwd = 3, col = "black") +
-  tm_facets(by = "Municipality") +
-  tm_layout(
-    legend.show = FALSE,
-    panel.label.height = 0)
+  tm_facets(by = "Municipality"))
 
@@
             legend.position = c("left", "bottom"),
-            legend.outside = TRUE)
+            legend.outside = F)
+
+empty <- tm_shape(muni3) +
+  tm_borders(col="white") +
+  tm_layout(frame = F)
 
@@
 (methodsMap <- tmap_arrange(
-  positionMap,
   muniPlot,
+  empty,
+  infraMuniMap,
   infraMuniMap_l,
-  infraMuniMap,
-  widths = c(.35, .65),
+  ncol = 2,
+  widths = c(.8, .2),
   heights = c(.6, .4),
-  outer.margins = .01
+  outer.margins = NULL
 ))

### 2024-02-22 manuscript/HIAs.qmd

--- 2024-02-21 manuscript/HIAs.qmd
@@
 ```{r import}
+#| eval: true
 #| cache: true
@@
 
-# Mire data
-mire_terra <- terra::rast(path_mire)
-
 # Infrastructure index (read proxy)
@@
 
+```
+
+```{r terraLoad}
+#| eval: true
+# Mire data
+# Spat rasters cannot be cached
+mire_terra <- terra::rast(path_mire)
 ```
@@
 ```{r mireTerra}
-#| eval: true
-
-# I first tried to import and crop the mire data using starts, but that failed (se pre 21 feb 2023)
+#| eval: false
+
+# I first tried to import and crop the mire data using stars, but that failed (see pre 21 feb 2023)
 # trying { terra } instead
@@
 # The cropping and masking worked.
-
-# I like the stars, sf and tmap combo better, so I return to stars
+# 
+# # I like the stars, sf and tmap combo better, so I return to stars
 mire_stars_nf <- mire_terra_nf |>
@@
 
+saveRDS(mire_stars_nf, "manual_cache/mire_stars_nf.RDS")
+saveRDS(mire_stars_gr, "manual_cache/mire_stars_gr.RDS")
+saveRDS(mire_stars_na, "manual_cache/mire_stars_na.RDS")
+
+```
+
+```{r terraToSTars}
+#| eval: true
+mire_stars_nf <- readRDS("manual_cache/mire_stars_nf.RDS")
+mire_stars_gr <- readRDS("manual_cache/mire_stars_gr.RDS")
+mire_stars_na <- readRDS("manual_cache/mire_stars_na.RDS")
 ```
@@
 ```{r dk2-CoverageMaps}
-#| eval: false
+#| eval: true
+#| cache: true
 
@@
 ```{r mireArea}
-#| eval: false
+#| eval: true
+#| cache: true
 
@@
 ```
+
 ```{r corrCheck}
 #| eval: true
+#| cache: true
 # Now I want to see if the indicator values depend on the HIA is a predictable
@@
 
-```{r}
+```{r test 4}
 #| eval: false
@@
 
-```{r}
+```{r realValidationPlot}
 #| eval: true
 # Using a color gradient emphasizes the first color (dark green).
-# Lets try discrete colors, and merge som classes to simplify
+# Lets try discrete colors, and merge some classes to simplify
 (validationPlot <- corrCheck |>
-  #rename(ADSV = i_ADSV) |>
   pivot_longer(
@@
   ) |>
+  as_tibble() |>
+  group_by(indicator, infrastructureIndex, condition) |>
+  summarise(n = n()) |>
+  ungroup() |>
+  group_by(indicator, infrastructureIndex) |>
+  mutate(lab = round(n/sum(n)*100),
+         lab = case_when(
+           lab < 5 ~ NA,
+           .default = paste0(lab, "%")
+         )) |>
   ggplot(aes(
-    x = factor(infrastructureIndex),
+    x = infrastructureIndex,
+    y = n,
     fill = condition
@@
   geom_bar(
-    position = "fill"
-  ) +
-  theme_bw(base_size = 12) +
+    position = "fill",
+    stat = "identity"
+  ) +
+  geom_text(aes(label = lab),
+    position = position_fill(vjust = 0.5),
+    color= "black", vjust = 0.5, size = 4) +
+  theme_minimal(base_size = 15) +
+  theme(
+    panel.grid = element_blank(), 
+    axis.text.x = element_text(margin = margin(t = -10)),
+    axis.text.y = element_blank(),
+    axis.title.y = element_blank(),
+    strip.text = element_textbox(
+      size = 12,
+      halign = 0.5, linetype = 1, r = unit(5, "pt"), width = unit(1, "npc"),
+      padding = margin(2, 0, 1, 0), margin = margin(3, 3, 3, 3))) +
   guides(fill = guide_legend("Indicator values")) +
-  ylab("Fraction of data points") +
   xlab("Human Impact Areas") +
-  scale_fill_brewer(palette = "RdYlGn") +
+  scale_fill_manual(values = c("#E85437","#FBAF00", "#B5DF73", "#009000")) +
   facet_grid(~indicator)
@@
 
-```{r}
+```{r test3}
 #| eval: false
@@
 
-```{r}
+```{r test2}
 #| eval: false
@@
 
-```{r}
+```{r test}
 #| eval: false
@@
 ```{r infraMuniMaps}
-#| eval: false
+#| eval: true
+#| cache: true
 
 # Infratructure in each municipality
-infraMuniMap <- tm_shape(muni3) +
+(infraMuniMap <- tm_shape(muni3) +
   tm_borders() +
@@
 (methodsMap <- tmap_arrange(
-  positionMap,
   muniPlot,
+  empty,
+  infraMuniMap,
   infraMuniMap_l,
-  infraMuniMap,
-  widths = c(.35, .65),
+  ncol = 2,
+  widths = c(.8, .2),
   heights = c(.6, .4),
@@
 
-# Calulate some more stats for the table with municipality stats
+# Calculate some more stats for the table with municipality stats
 
@@
   # paste inn terrestrial area
-  left_join(terrestrial |>
+  left_join(
+    terrestrial |>
       as_tibble() |>
@@
   # % mire inside survey coverage map
-  left_join(mire_nf_terra_dk |> select(Municipality, mireInSurvey_km2)) |>
+  left_join(mire_in_dk |> select(Municipality, mireInSurvey_km2)) |>
   mutate(mireInSurvery_percent = round(mireInSurvey_km2 / mire_km2 * 100, 2)) |>
@@
 | id  | Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                                                                       | Reference                          |
-|-----|----------------------|--------------------------------------------------|-----------------------------------------------------------------------------------------|-------------------------------------------------------------------------------------------------------------------------------------------------------------------|------------------------------------|
+|------------|------------|------------|------------|----------------|------------|
 | 1   | 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                                                                                | @runehalvorsen2019                 |
@@
 We chose three municipalities in Norway to test out the indicators (@fig-studyLocation).
-These municipalities are spread over `r round(max(latlong$Y)-min(latlong$Y), 1)` degrees latitude, or `r km_distance` km, and also differ in the amount of mire area, the total area surveyed, and the prevalence of infrastructure (@tbl-variables).
+The differ in several aspect, such as the amount of mire area, the total area surveyed, and the prevalence of infrastructure (@tbl-variables).
 
@@
 
-(temp <- readRDS("../figures/studyLocation.RDS"))
+#(temp <- readRDS("../figures/studyLocation.RDS"))
+positionMap
+```
+
+```{r fig-studyLocation2}
+#| eval: true
+#| include: true
+#| out.width: '100%'
+#| fig.cap:
+#| - "Map of the three focal municiaplities. For each municipality the map shows 
+#| ocean in blue and non-ocean
+#| in green. The survey coverage maps are in grey, and the mapped mires are
+#| in red, with polygon borderes made extra thich to make them visible,
+#| but then also exagerating their size."
+
+#(temp <- readRDS("../figures/studyLocation.RDS"))
+methodsMap
 ```
@@
 | Municipality  | Total terrestrial area (km^2^)                                            | \% of terrestrial area surveyed                                            | \% open mires in relation to total terrestrial area                         | Total mire area (km^2^)                                                  | %of mire area inside survey area                                                      | Number of mire polygons in survey                                 | Mean Infrastructure Index value                                         |
-|--------------|---------------------------------------------------------------------------|----------------------------------------------------------------------------|-----------------------------------------------------------------------------|--------------------------------------------------------------------------|---------------------------------------------------------------------------------------|-------------------------------------------------------------------|-------------------------------------------------------------------------|
+|---------|---------|---------|---------|---------|---------|---------|---------|
 | Nordre Follo  | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(t_area_km)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(dk_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mirePercent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mire_km2)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mireInSurvery_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(n)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(meanHIA)` |
-| Trondheim    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(t_area_km)`    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(dk_percent)`    |                                                                             |                                                                          |                                                                                       | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(n)`    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(meanHIA)`    |
-| Målselv      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(t_area_km)`      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(dk_percent)`      |                                                                             |                                                                          |                                                                                       | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(n)`      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(meanHIA)`      |
+| `Gran`        | `r muni_tbl |> filter(Municipality == "Gran") |> pull(t_area_km)`         | `r muni_tbl |> filter(Municipality == "Gran") |> pull(dk_percent)`         | `r muni_tbl |> filter(Municipality == "Gran") |> pull(mirePercent)`         | `r muni_tbl |> filter(Municipality == "Gran") |> pull(mire_km2)`         | `r muni_tbl |> filter(Municipality == "Gran") |> pull(mireInSurvery_percent)`         | `r muni_tbl |> filter(Municipality == "Gran") |> pull(n)`         | `r muni_tbl |> filter(Municipality == "Gran") |> pull(meanHIA)`         |
+| `Nord-Aurdal` | `r muni_tbl |> filter(Municipality == "Nord-Aurdal") |> pull(t_area_km)`  | `r muni_tbl |> filter(Municipality == "Nord-Aurdal") |> pull(dk_percent)`  | `r muni_tbl |> filter(Municipality == "Nord-Aurdal") |> pull(mirePercent)`  | `r muni_tbl |> filter(Municipality == "Nord-Aurdal") |> pull(mire_km2)`  | `r muni_tbl |> filter(Municipality == "Nord-Aurdal") |> pull(mireInSurvery_percent)`  | `r muni_tbl |> filter(Municipality == "Nord-Aurdal") |> pull(n)`  | `r muni_tbl |> filter(Municipality == "Nord-Aurdal") |> pull(meanHIA)`  |
 
@@
 
+```{r fig-validation}
+#| eval: true
+#| include: true
+#| out.width: '100%'
+#| fig.cap:
+#| - "Proportion of localities with different scaled indicator values for three 
+#| ecosystem condition indicators along a Human Impact Gradient"
+
+validationPlot
+```
+
 National/regional indicator - do we have enough data?

### 2024-02-22 manuscript/HIAs.qmd
test bolstad function

@@
 mire_stars_na <- readRDS("manual_cache/mire_stars_na.RDS")
+mire_terra_nf <- rast(mire_stars_nf)
+mire_terra_gr <- rast(mire_stars_gr)
+mire_terra_na <- rast(mire_stars_na)
 ```
@@
 ```{r muni-tbl}
-#| eval: false
+#| eval: true
+#| cache: true
 
@@
   left_join(infraMuni3_summary)
-
-saveRDS(muni_tbl, "../data/municipality_stats.RDS")
 ```
@@
 #| eval: true
-muni_tbl <- readRDS("../data/municipality_stats.RDS")
+#muni_tbl <- readRDS("../data/municipality_stats.RDS")
+```
+
+```{r}
+#| eval: true
+
+# Now I need to spatially (horizontally) aggregate the indicator values for each 
+# HIA. Rather than taking the arithmetic mean, I will use a Bayesian updating 
+# approach. The point estimate (central tendency) will probably be the same
+# more or less, with the two approaches, but with the updating approach I can
+# get a honest measure for the uncertainty even with very small sample sizes.
+# For this I need a 'true' value for the variation in the indicator
+# I will use the entire national data set to determine this number
+
+national_sd_ADSV <- sd(naturetypes$i_ADSV, na.rm=T)
+national_sd_alien <- sd(naturetypes$i_alien, na.rm=T)
+national_sd_ditch <- sd(naturetypes$i_ditch, na.rm=T)
+
+barplot(c(national_sd_ADSV, national_sd_alien, national_sd_ditch),
+        names.arg = c("ADSV", "Alien", "Trenching"),
+        ylab="SD")
+
+# The figure shows that the Trenching indicator is more spatially variable
+```
+
+```{r wgt-mean-fn}
+#| eval: true
+
+# This function is modified from Bolstad::normdp
+# It updates a flat prior based on a sample of values (indicator values)
+# and weight (polygon area) and returns a distribution for the weighted mean
+# that has a gaussian distribution. It assumes the sampled population is also
+# Gaussian. The variance is estimated from the full sample of indicator values.
+
+wgt_mean <- function(x, weights, 
+                     sigma.x = NULL, 
+                     mu = seq(0, 1, length.out=1000), 
+                     mu.prior = rep(1/length(mu), times=length(mu)), 
+                     stat = "mean",
+                     ...) {
+
+  mx <- weighted.mean(x, weights)
+  if (round(sum(mu.prior), 7) != 1) {
+    warning("The prior probabilities did not sum to 1, therefore the prior has been normalized")
+    mu.prior <- mu.prior / sum(mu.prior)
+  }
+  n.mu <- length(mu)
+  nx <- length(x)
+  snx <- sigma.x^2 / nx
+  likelihood <- exp(-0.5 * (mx - mu)^2 / snx)
+  posterior <- likelihood * mu.prior / sum(likelihood * mu.prior)
+  mx <- sum(mu * posterior)
+  vx <- sum((mu - mx)^2 * posterior)
+
+  # draw 1k samples from the posterior to calculate the quantiles and sd from
+  sample <- sample(mu, size = 1000, prob = posterior, replace = T)
+  lower <- quantile(sample, probs = 0.025)
+  upper <- quantile(sample, probs = 0.975)
+  # we can assume that the distrubutin for the mean is gaussian, so we take a 
+  # symetrical sd from here and use it, along with the mean, to recreate a
+  # normal distribution later which we can sample from
+  sdx <- sd(sample)
+  results <- list(
+    name = "mu", 
+    param.x = mu, 
+    prior = mu.prior,
+    likelihood = likelihood, 
+    posterior = posterior, 
+    weighted_mean = mx,
+    var = vx
+  )
+  if(stat == "mean") return(mx)
+  if(stat == "lower") return(lower)
+  if(stat == "upper") return(upper)
+  if(stat == "sd") return(sdx)
+}
+
+```
+
+```{r test-function}
+#| eval: false
+test_sd <- wgt_mean(x = nature3$i_ADSV[!is.na(nature3$i_ADSV)], 
+                 weights = drop_units(nature3$area[!is.na(nature3$i_ADSV)]), 
+                 sigma.x = national_sd_ADSV,
+                 stat = "sd")
+test_mean <- wgt_mean(x = nature3$i_ADSV[!is.na(nature3$i_ADSV)], 
+                 weights = drop_units(nature3$area[!is.na(nature3$i_ADSV)]), 
+                 sigma.x = national_sd_ADSV,
+                 stat = "mean")
+
+hist(rnorm(1000, test_mean, test_sd))
+
+```
+
+```{r meanPerHIA}
+#| eval: true
+#| cache: true
+
+mu <- seq(0, 1, length.out=1000)
+# Intersect nature3 with the HIA
+temp <- nature3 |>
+  st_intersection(infraMuni3) |>
+  pivot_longer(cols = c(
+    i_ADSV,
+    i_alien,
+    i_ditch),
+    names_to = "indicator",
+    values_to = "indicatorValue") |>
+  filter(!is.na(indicatorValue)) |>
+  group_by(indicator, Municipality, infrastructureIndex) |>
+  summarise(mean )
 ```
@@
 | id  | Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                                                                       | Reference                          |
-|------------|------------|------------|------------|----------------|------------|
+|----------|----------|----------|----------|-----------------------|----------|
 | 1   | 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                                                                                | @runehalvorsen2019                 |

### 2024-02-22 manuscript/HIAs.qmd
indicator mean and sd per HIA

@@
 
-mu <- seq(0, 1, length.out=1000)
 # Intersect nature3 with the HIA
-temp <- nature3 |>
+stats_tbl <- nature3 |>
   st_intersection(infraMuni3) |>
@@
   filter(!is.na(indicatorValue)) |>
+  mutate(area = drop_units(area)) |>
   group_by(indicator, Municipality, infrastructureIndex) |>
-  summarise(mean )
+  summarise(sd = wgt_mean(indicatorValue,
+      weights = area,
+      sigma.x = case_when(
+        indicator == "i_alien" ~ national_sd_alien,
+        indicator == "i_ditch" ~ national_sd_ditch,
+        indicator == "i_ADSV" ~ national_sd_ADSV,
+        .default = NULL
+      ),
+      stat = "sd"),
+    mean = wgt_mean(indicatorValue,
+      weights = area,
+      sigma.x = case_when(
+        indicator == "i_alien" ~ national_sd_alien,
+        indicator == "i_ditch" ~ national_sd_ditch,
+        indicator == "i_ADSV" ~ national_sd_ADSV,
+        .default = NULL
+      ),
+      stat = "mean"),
+    n = n())
+stats_tbl
+```
+
+```{r forets plot}
+# Forest plot
+# https://stackoverflow.com/questions/58657802/forest-plot-with-subgroups-in-ggplot2
 ```

### 2024-02-22 manuscript/HIAs.qmd
forest plot v1

@@
 # https://stackoverflow.com/questions/58657802/forest-plot-with-subgroups-in-ggplot2
+# 
+# #define colours for dots and bars
+barCOLS = c("grey80","grey60", "grey40", "grey20")
+dotCOLS = c("grey90","grey70", "grey50", "grey40")
+
+stats_tbl |>
+  ggplot(aes(x=Municipality, 
+             y=mean, 
+             ymin=mean-sd, 
+             ymax=mean+sd,
+             col=factor(infrastructureIndex),
+             fill=factor(infrastructureIndex))) + 
+#specify position here
+  geom_linerange(
+    size=5,
+    position=position_dodge(width = .9)) +
+  #geom_hline(yintercept=1, lty=2) +
+#specify position here too
+  geom_point(
+    size=3, 
+    shape=21, 
+    colour="white", 
+    stroke = 0.5,
+    position=position_dodge(width = .9)) +
+  scale_fill_manual(values=barCOLS)+
+  scale_color_manual(values=dotCOLS)+
+  scale_x_discrete(name="") +
+  scale_y_continuous(name="Indicator values", limits = c(0, 1)) +
+  coord_flip() +
+  theme_bw() +
+  labs(fill = "HIA",
+       col = "HIA") +
+  facet_wrap(.~indicator,
+             ncol=1)
 ```

### 2024-02-23 manuscript/HIAs.qmd
sample size in forest plot

--- 2024-02-22 manuscript/HIAs.qmd
@@
 # https://stackoverflow.com/questions/58657802/forest-plot-with-subgroups-in-ggplot2
+# 
+# #define colours for dots and bars
+dotCOLS = c("grey90","grey70", "grey50", "grey40")
+
+barCOLS <- tmaptools::get_brewer_pal("Set1", n = 4)
+#barCOLS = c("grey80","grey60", "grey40", "grey20")
+
+stats_tbl |>
+  mutate(
+    indicator = case_when(
+      indicator == "i_ADSV" ~ "ADSV",
+      indicator == "i_alien" ~ "Alien species",
+      indicator == "i_ditch" ~ "Trenching"
+    )
+  ) |>
+  ggplot(aes(x=infrastructureIndex, 
+             y=mean, 
+             ymin=mean-sd, 
+             ymax=mean+sd,
+             col=factor(infrastructureIndex),
+             fill=factor(infrastructureIndex))) + 
+#specify position here
+  geom_linerange(
+    size=10,
+    #position=position_dodge(width = .9)
+    ) +
+  #geom_hline(yintercept=1, lty=2) +
+#specify position here too
+  geom_point(
+    size=3, 
+    shape=21, 
+    colour="white", 
+    stroke = 0.5,
+    #position=position_dodge(width = .9)
+    ) +
+  geom_text(
+    aes(y = mean-sd, label = n),
+    nudge_y = -0.1,
+    show.legend = F
+  ) +
+  scale_fill_manual(values=dotCOLS)+
+  scale_color_manual(values=barCOLS)+
+  scale_x_discrete(name="") +
+  scale_y_continuous(name="Indicator values", limits = c(0, 1)) +
+  coord_flip() +
+  theme_bw() +
+  labs(fill = "HIA",
+       col = "HIA") +
+  facet_grid(indicator ~ Municipality)
 ```

### 2024-02-23 manuscript/HIAs.qmd
rowwise

--- 2024-02-22 manuscript/HIAs.qmd
@@
 library(tmap)
+libarary(tmaptools)
 library(stars)
@@
 # https://stackoverflow.com/questions/58657802/forest-plot-with-subgroups-in-ggplot2
+# 
+# #define colours for dots and bars
+dotCOLS = c("grey90","grey70", "grey50", "grey40")
+barCOLS <- tmaptools::get_brewer_pal("Set1", n = 4)
+
+stats_tbl |>
+  mutate(
+    indicator = case_when(
+      indicator == "i_ADSV" ~ "ADSV",
+      indicator == "i_alien" ~ "Alien species",
+      indicator == "i_ditch" ~ "Trenching"
+    )
+  ) |>
+  rowwise() |>
+  mutate(
+    low = quantile(rnorm(200, mean, sd), probs = 0.025),
+    high = quantile(rnorm(200, mean, sd), probs = 0.975),
+    high = ifelse(high > 1, 1, high)
+  ) |>
+  ggplot(aes(x=infrastructureIndex, 
+             y=mean, 
+             ymin=low,
+             ymax=high,
+             col=factor(infrastructureIndex),
+             fill=factor(infrastructureIndex))) + 
+  geom_linerange(
+    size=10,
+  geom_point(
+    size=3, 
+    shape=21, 
+    colour="white", 
+    stroke = 0.5,
+    ) +
+  geom_text(
+    aes(y = low, label = n),
+    nudge_y = -0.1,
+    show.legend = F
+  ) +
+  scale_fill_manual(values=dotCOLS)+
+  scale_color_manual(values=barCOLS)+
+  scale_x_discrete(name="") +
+  scale_y_continuous(name="Indicator values", limits = c(-0.1, 1)) +
+  coord_flip() +
+  theme_bw() +
+  labs(fill = "HIA",
+       col = "HIA") +
+  facet_grid(indicator ~ Municipality)
+```
+
+```{r}
+# Next I need to 
+#  - copy the weighted means and SDs over to the ecosystem delineation map polygons 
+#  - show intermittent resulting map with ggmagnify
+#  - sample from these distributions (ignore missing HIAs)
+```
+
+```{r}
+
 ```

### 2024-02-23 manuscript/HIAs.qmd
EAA means and CI

@@
 library(tmap)
-libarary(tmaptools)
+library(tmaptools)
 library(stars)
@@
 library(rnaturalearthdata)
+library(ggmagnify)
 
@@
   guides(fill = guide_legend("Indicator values")) +
-  xlab("Human Impact Areas") +
+  xlab("Homogeneous Impact Areas") +
   scale_fill_manual(values = c("#E85437","#FBAF00", "#B5DF73", "#009000")) +
@@
 )
+
+ggsave("../images/validation-plot.jpg",
+       plot=validationPlot,
+       width = 9,
+  height = 5)
 ```
@@
     style = "cat",
-    title = "Human Impact Areas"
+    title = "Homogeneous Impact Areas"
   ) +
@@
     style = "cat",
-    title = "Human\nImpact\nArea"
+    title = "Homogeneous\nImpact Area"
   ) +
@@
 
-# tmap_save(methodsMap, "../figures/studyLocations.tiff",
+#tmap_save(methodsMap, "../images/studyLocations.tiff",
 #           dpi= 1000,
@@
 #           height = 10)
+#tmap_save(methodsMap, "../images/studyLocations.jpg",
+#           units = "cm",
+#           width = 18,
+#           height = 10)
+
 #
-# saveRDS(methodsMap, "../figures/studyLocation.RDS")
+#saveRDS(methodsMap, "../figures/studyLocation.RDS")
 ```
@@
 }
-
-```
-
-```{r test-function}
-#| eval: false
-test_sd <- wgt_mean(x = nature3$i_ADSV[!is.na(nature3$i_ADSV)], 
-                 weights = drop_units(nature3$area[!is.na(nature3$i_ADSV)]), 
-                 sigma.x = national_sd_ADSV,
-                 stat = "sd")
-test_mean <- wgt_mean(x = nature3$i_ADSV[!is.na(nature3$i_ADSV)], 
-                 weights = drop_units(nature3$area[!is.na(nature3$i_ADSV)]), 
-                 sigma.x = national_sd_ADSV,
-                 stat = "mean")
-
-hist(rnorm(1000, test_mean, test_sd))
 
@@
 
-stats_tbl |>
+(forest_plot <- stats_tbl |>
   mutate(
@@
   geom_linerange(
-    size=10,
+    size=10) +
   geom_point(
@@
   facet_grid(indicator ~ Municipality)
+)
+
+#ggsave("../images/forest-plot.jpg",
+#       plot=forest_plot)
 ```
@@
 # Next I need to 
-#  - copy the weighted means and SDs over to the ecosystem delineation map polygons 
+#  - copy the weighted means and SDs over to  HIA map and then over to the 
+#    ecosystem delineation map polygons. 
 #  - show intermittent resulting map with ggmagnify
@@
 ```{r}
-
+#| eval: true
+#| cache: true
+
+# Copy the weighted means and SDs over to HIA map and then over to the 
+# ecosystem delineation map polygons. 
+spread_na <- mire_stars_na |>
+  st_as_sf(merge=T) |>
+  filter(Myr153 == 1) |>
+  st_intersection(infraMuni3 |> select(infrastructureIndex)) |>
+  mutate(area = geometry |> st_area()) |>
+  left_join(stats_tbl |> 
+              as_tibble() |>
+              filter(Municipality == "Nord-Aurdal") |>
+              select(mean, sd, indicator, n, infrastructureIndex, Municipality),
+            by = "infrastructureIndex")
+
+spread_nf <- mire_stars_nf |>
+  st_as_sf(merge=T) |>
+  filter(Myr153 == 1) |>
+  st_intersection(infraMuni3 |> select(infrastructureIndex)) |>
+  mutate(area = geometry |> st_area()) |>
+  left_join(stats_tbl |> 
+              as_tibble() |>
+              filter(Municipality == "Nordre Follo") |>
+              select(mean, sd, indicator, n, infrastructureIndex, Municipality),
+            by = "infrastructureIndex")
+
+spread_gr <- mire_stars_gr |>
+  st_as_sf(merge=T) |>
+  filter(Myr153 == 1) |>
+  st_intersection(infraMuni3 |> select(infrastructureIndex)) |>
+  mutate(area = geometry |> st_area()) |>
+  left_join(stats_tbl |> 
+              as_tibble() |>
+              filter(Municipality == "Gran") |>
+              select(mean, sd, indicator, n, infrastructureIndex, Municipality),
+            by = "infrastructureIndex")
+
+```
+
+```{r example-spread}
+#| eval: true
+#| cache: true
+
+# A plot of Nord-Aurdal with the mean indicator values per HIA spread over the 
+# EDM
+from <- c(xmin = 510000, xmax = 515000, ymin = 6741000, ymax = 6746000)
+to <-   c(xmin = 495000, xmax = 520000, ymin = 6765000, ymax = 6790000)
+myCols <- c(
+  #"#E85437",
+  "#FBAF00", 
+  "#B5DF73", 
+  "#009000"
+  )
+
+spread_na_map <- spread_na |>
+  filter(indicator == "i_ditch") |>
+  mutate(Trenching = factor(round(mean, 2))) |>
+  ggplot() +
+  geom_sf(aes(fill=Trenching,
+              color = Trenching)) +
+  geom_sf(data = na,
+          alpha=0) +
+  scale_fill_manual(values = myCols) +
+  scale_color_manual(values = myCols) +
+  coord_sf(
+    datum = st_crs(myCRS),
+    xlim = c(494174.8 , 537114.7 ), 
+    ylim = c(6737092 , 6789676)) +
+  ggmagnify::geom_magnify(from = from, to = to, 
+                          expand = 0,
+                          shadow =T,
+                          corners = 0.1) +
+  theme_bw()
+
+ggsave("../images/spread-na.tiff",
+       plot = spread_na_map)
+ggsave("../images/spread-na-small.tiff",
+       plot = spread_na_map,
+       dpi=150)
+ggsave("../images/spread-na-small.jpg",
+       plot = spread_na_map)
+```
+
+```{r EAA}
+#| eval: true
+#| cache: true
+
+combineALl <- rbind(
+  spread_nf,
+  spread_gr,
+  spread_na
+) |>
+  as_tibble() |>
+  drop_na() |>
+  group_by(Municipality, indicator) |>
+  rowwise() |>
+  mutate(
+    i_sample = list(sample(rnorm(10, mean, sd)))) |>
+  unnest(cols=i_sample) |>
+  summarise(mean = mean(i_sample),
+            lower = quantile(i_sample, probs = c(0.025)),
+            upper = quantile(i_sample, probs = c(0.975)))
+
+# add area weights to the above and make forest plot
+
+#combineALl
+#tail(combineALl)
+#combineALl$i_sample[2]
 ```
@@
 In this study we explore the potential for using a stratified aggregation technique to make use of spatially biased field data in ecosystem condition accounting.
-We demonstrate this technique using a generic GIS-based workflow for compiling ecosystem condition accounts that can be applied at any spatial scale, and we highlight the opportunities for local use-cases of this workflow by contrasting our findings across four neighboring municipalities in Norway.
+We demonstrate this technique using a generic GIS-based workflow for compiling ecosystem condition accounts that can be applied at any spatial scale, and we highlight the opportunities for local use-cases of this workflow by contrasting our findings across three neighboring municipalities in Norway (@fig-studyLocation).
 The main question is how much generalisation can we perform on the data we have before the resulting indicator looses its practical value in local governance processes.
 We end by interviewing end-users from the relevant municipalities about the perceived benefits and shortcomings of our condition indicators.
+
+```{r fig-studyLocation}
+#| eval: true
+#| include: true
+#| out.width: '100%'
+#| fig.cap:
+#| - "Postion of the three example municipalities in
+#| Norway. "
+
+positionMap
+```
 
@@
 We then identified two *ecosystem condition characteristics* that, along with many other characteristics that we do not specify here, define the typical behavior of open mire ecosystems in the reference condition: species composition, and; vegetation intactness.
-
 To describe these two characteristics we use six condition variables from the same data set ( @tbl-variables).
-
 See @miljødirektoratet for the sampling protocol (in Norwegian).
 
+```{r fig-studyLocation2}
+#| eval: true
+#| include: true
+#| out.width: '80%'
+#| fig.cap:
+#| - "Map of the three focal municiaplities. For each municipality the map shows 
+#| ocean in blue and non-ocean
+#| in green. The survey coverage maps are in grey, and the mapped mires are
+#| in red, with polygon borderes made extra thich to make them visible,
+#| but then also exagerating their size."
+
+knitr::include_graphics("../images/studyLocations.jpg")
+```
+
+<!--# Three characteristics now? -->
+
 | id  | Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                                                                       | Reference                          |
-|----------|----------|----------|----------|-----------------------|----------|
+|-------|---------|-----------|--------------|----------------------|----------|
 | 1   | 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                                                                                | @runehalvorsen2019                 |
@@
 A value of 2 indicator observable change within the range expected for the same mapping unit, and a value of 3 indicates a mire transitioning into a neighboring (ecologically speaking) mapping unit.
-
-The .
 See @fig-workflow for a schematic workflow for the indicator development.
@@
 
-```{r fig-studyLocation}
-#| eval: true
-#| include: true
-#| out.width: '100%'
-#| fig.cap:
-#| - "Postion of the three example municipalities in
-#| Norway. For each municipality the map shows ocean in blue and non-ocean
-#| in green. The survey coverage maps are in grey, and the mapped mires are
-#| in red, with polygon borderes made extra thich to make them visible,
-#| but then also exagerating their size."
-
-#(temp <- readRDS("../figures/studyLocation.RDS"))
-positionMap
-```
-
-```{r fig-studyLocation2}
-#| eval: true
-#| include: true
-#| out.width: '100%'
-#| fig.cap:
-#| - "Map of the three focal municiaplities. For each municipality the map shows 
-#| ocean in blue and non-ocean
-#| in green. The survey coverage maps are in grey, and the mapped mires are
-#| in red, with polygon borderes made extra thich to make them visible,
-#| but then also exagerating their size."
-
-#(temp <- readRDS("../figures/studyLocation.RDS"))
-methodsMap
-```
+{{< pagebreak >}}
 
@@
 #| include: true
-#| out.width: '100%'
+#| out.width: '90%'
 #| fig.cap:
 #| - "Proportion of localities with different scaled indicator values for three 
-#| ecosystem condition indicators along a Human Impact Gradient"
-
-validationPlot
-```
-
-National/regional indicator - do we have enough data?
-
-Municipal case study
+#| ecosystem condition indicators along a gradient of infrastructure
+#| densities."
+
+knitr::include_graphics("../images/validation-plot.jpg")
+```
+
+```{r fig-forest}
+#| eval: true
+#| include: true
+#| out.width: '80%'
+#| fig.cap:
+#| - "Indicator values (circles = mean; bars = 95% credible intervals) for three
+#| mire ecosystem condition indicators in three municipalities. The indicator values 
+#| are calculated uniquely for each Homogeneous Impact Area (HIA) in each 
+#| municipality. The numbers to the left of each bar is the sample size, i.e. 
+#| the number of surveyed mires."
+
+knitr::include_graphics("../images/forest-plot.jpg")
+```
+
+```{r fig-spread-example}
+#| eval: true
+#| include: true
+#| out.width: '80%'
+#| fig.cap:
+#| - "An indicator for mire trenching shown for Nord-Aurdal municipality. 
+#| Individual mire polygons are colored by the mean indicator values for the
+#| homogenous impact area where it lies. Colors are chosen to best reflect 
+#| categorical differences and exagerates the absolute difference between 
+#| areas. The inset it just a visual aid."
+
+knitr::include_graphics("../images/spread-na-small.jpg")
+```
 
@@
 
-# References {.unnumbered}
+{{< pagebreak >}} \# References {.unnumbered}

### 2024-02-26 manuscript/HIAs.qmd

@@
 
-ggsave("../images/spread-na.tiff",
-       plot = spread_na_map)
-ggsave("../images/spread-na-small.tiff",
-       plot = spread_na_map,
-       dpi=150)
-ggsave("../images/spread-na-small.jpg",
-       plot = spread_na_map)
+#ggsave("../images/spread-na.tiff",
+#       plot = spread_na_map)
+#ggsave("../images/spread-na-small.tiff",
+#       plot = spread_na_map,
+#       dpi=150)
+#ggsave("../images/spread-na-small.jpg",
+#       plot = spread_na_map)
 ```
@@
 
-combineALl <- rbind(
+# Now I sample indicator values from the EDM and create a new distribution for 
+# indicator values in the EEA (the municipalities). First I sample the individual
+# distributions for each mire polygon with n defined by the polygon area.
+# The distribution for the polygon areas is strongly right skewed, meaning
+# some polygons will contribute much more to the EAA value than others.
+# I will keep this design in-line with SEEA EA guidelines for area weighting, 
+# but it's worth noting that this could be solved in other ways.
+# 
+combineAll <- rbind(
   spread_nf,
@@
   mutate(
-    i_sample = list(sample(rnorm(10, mean, sd)))) |>
-  unnest(cols=i_sample) |>
-  summarise(mean = mean(i_sample),
-            lower = quantile(i_sample, probs = c(0.025)),
-            upper = quantile(i_sample, probs = c(0.975)))
-
-# add area weights to the above and make forest plot
-
-#combineALl
-#tail(combineALl)
-#combineALl$i_sample[2]
+    i_sample = list(sample(rnorm(n, mean, sd)))) |>
+  select(-geometry) |>
+  group_by(Municipality, indicator) |>
+  reframe(i_sample = 
+            sample(
+              unlist(i_sample), 
+              size = 1000,
+              replace = TRUE)) |>
+  mutate(i_sample = case_when(
+    i_sample < 0 ~ 0,
+    i_sample >1  ~ 1,
+    .default = i_sample
+  ))
+```
+
+```{r EEA-plot}
+#| eval: true
+#| cache: true
+library(ggridges)
+
+(ridgepPlot <- combineAll |>
+  mutate(
+    indicator = case_when(
+      indicator == "i_ADSV" ~ "ADSV",
+      indicator == "i_alien" ~ "Alien species",
+      indicator == "i_ditch" ~ "Trenching"
+    )) |>
+  ggplot(
+    aes(
+      x = i_sample, 
+      y = Municipality,
+      fill = after_stat(x))
+  ) +
+  geom_density_ridges_gradient(
+    bandwidth=.05,
+    quantile_lines = TRUE, 
+    quantiles = c(0.025, .5, 0.975)
+  ) +
+  #scale_fill_viridis_c(option = "D") +
+  scale_fill_distiller(palette = "RdYlGn",
+    direction = 1) +
+  theme_bw(base_size = 15) +
+  guides(fill = "none") +
+  geom_vline(xintercept = .6,
+    size=1.2,
+    lty=2) +
+  theme(axis.text.x = element_text(angle=45, hjust=1, vjust=1))+
+  labs(y = "", x = "Indicator values")+
+  xlim(c(0,1)) +
+  facet_wrap(.~indicator)
+)
+
+#ggsave(plot = ridgepPlot,
+#       "../images/ridgePlot.jpg",
+#       width=10,
+#       height=5)
+```
+
+```{r EEA-tbl}
+#| eval: true
+(EEA_tbl <- combineAll |>
+  mutate(
+    indicator = case_when(
+      indicator == "i_ADSV" ~ "ADSV",
+      indicator == "i_alien" ~ "Alien species",
+      indicator == "i_ditch" ~ "Trenching"
+    )) |>
+  group_by(Municipality, indicator) |>
+  summarise(mean = round(mean(i_sample), 2),
+            median = round(median(i_sample), 2),
+            percentile_025 = round(quantile(i_sample, probs = 0.025), 2),
+            percentile_975 = round(quantile(i_sample, probs = 0.975), 2))
+)
+
+# saveRDS(EEA_tbl, "../output/EEA-table.RDS")
 ```
@@
 
-![Schematic workflow followed in this paper. Sub-index A is calculated using spatially explicit aggregation (i.e .a simple overlay taking the mean value).](../images/workflow.jpg){#fig-workflow}
+```{r fig-workflow}
+#| eval: true
+#| include: true
+#| out.width: '90%'
+#| fig.cap:
+#| - "Schematic workflow followed in this paper. Sub-index A is calculated using 
+#| spatially explicit aggregation."
+
+knitr::include_graphics("../images/workflow.jpg")
+```
 
@@
 
+```{r fig-EEA-ridgePlot}
+#| eval: true
+#| include: true
+#| out.width: '80%'
+#| fig.cap:
+#| - "Distributions for three ecosystem condition indicators in the Norwegian 
+#| Municiaplities. The color gradient reflects the same as the x-axis. 
+#| The dotted vertical line represnet the threshold for what is considered 
+#| reduced ecosystem condtion (< 0.6). "
+
+knitr::include_graphics("../images/ridgePlot.jpg")
+```
+
 # Discussion

### 2024-02-27 manuscript/HIAs.qmd
minor clean

@@
     email: anders.kolstad@nina.no
+    orchid: https://orcid.org/0000-0002-9623-9491
     affiliations: 
-        - id: some-tech
+        - id: nina
           name: Norwegian Institute for Nature Research
           department: Department of Terrestrial Ecology
-          address: Street Address
-          city: City
-          state: State
-          postal-code: Postal Code
+          address: Pb 5685 Torgarden
+          city: Trondheim
+          postal-code: 7485
     attributes:
         corresponding: true
-    note: This is the first author footnote.
-  - name: Bob Security
-    email: bob@example.com
+  - name: Matthew Grainger
+    email: matthew.grainger@nina.no
+    orchid: https://orcid.org/0000-0001-8426-6495
     affiliations:
-        - id: another-u
-          name: Another University
-          department: Department Name
-          address: Street Address
-          city: City
-          state: State
-          postal-code: Postal Code
-    note: |
-      Another author footnote, this is a very long footnote and it should be a really long footnote. But this footnote is not yet sufficiently long enough to make two lines of footnote text.
-  - name: Cat Memes
-    email: cat@example.com
-    affiliations:
-        - ref: another-u
-    note: Yet another author footnote.
-  - name: Derek Zoolander
-    email: derek@example.com
-    affilations:
-        - ref: some-tech
+        - ref: nina
 abstract: |
@@
 keywords: 
+  - alien species
+  - disturbance
+  - indicator
+  - ecosystem condition
+  - wetlands
+  - mire
+  - peatlands
+  - ecosystem accounting
   - SEEA EA
-  - alien species
 date: last-modified
@@
 Eval is also set explicitly for each chunk.
-When eval is set to false it means the job is quite time consuming and I instead run it once and cache it locally.
+Larger operations are cached.
+Remember to clean the cache occatinally and re-run.
 -->
@@
 library(ggmagnify)
+library(ggridges)
 
@@
     all_of(myVars))>0, na.rm=T)) |>
-  filter(keepers >0,
-         Ecosystem == "våtmark") |>
+  filter(
+    keepers >0,
+    Ecosystem == "våtmark"
+    ) |>
   pull(Nature_type)
@@
   select(!c("7TK", "7SE", "PRSL", "PRTK"))
-  # >
-  #pivot_longer(cols = c(ADSV, "7FA", "7GR-GI"),
-  #             names_to = "variable",
-  #             values_drop_na = T)
+
 head(naturetypes)
@@
 lower2 <- 5
-threshold2 <- 2.5 # = observable effect. Value 3 indicates a shit to a new type (grunntype)
+threshold2 <- 2.5 # = observable effect. Value 3 indicates a shift to a new type (grunntype)
 
@@
 # Adding scaled indicator values to the dataset
+# Same code as above, but with plot=F.
 naturetypes$i_ADSV <- eaTools::ea_normalise(
@@
 
-# The data contains some multisurfaces as well
+# The data contains some multisurfaces 
 # table(st_geometry_type(muni))
-# function to make sure that multipolygons are returned
+# Here is a function to make sure that multipolygons are returned
 ensure_multipolygons <- function(X) {
@@
   st_intersection(muni3)
-
-## These I had to manually download from naturbase.no
-#dk <- read_sf("../data/dekningskartNF/Naturtyper NiN - Dekningskart.shp") |>
-#  rbind(read_sf("../data/dekningskartTR/Naturtyper NiN - Dekningskart.shp")) |>
-#  rbind(read_sf("../data/dekningskartME/Naturtyper NiN - Dekningskart.shp")) |>
-#  st_transform(st_crs(muni3)) |>
-#  st_intersection(muni3) |>
-#  st_union()
 ```
@@
 
-```{r centroids}
+```{r positionMap}
 #| eval: true
 # Make map to show where the three municipalities are
-# First, get centroids
-centroids <- muni3 |>
-  st_centroid()
-```
-
-```{r positionMap}
-#| eval: true
-
 world <- ne_countries(scale = "medium", returnclass = "sf") |>
@@
 # what is the distance between Nordre Follo and Nord-Aurdal
+# First, get centroids
+centroids <- muni3 |>
+  st_centroid()
+
 (km_distance <- centroids |>
@@
 
-# I first tried to import and crop the mire data using stars, but that failed (see pre 21 feb 2023)
-# trying { terra } instead
+# I first tried to import and crop the mire data using stars, 
+# but that failed (see pre 21 feb 2023).
+# Trying { terra } instead
 
@@
 # The cropping and masking worked.
-# 
+
 # # I like the stars, sf and tmap combo better, so I return to stars
@@
 # Make a plot to check that it has worked
-infraMuni3_tbl |>
+(infra_dist_plot <- infraMuni3_tbl |>
   ggplot() +
@@
   labs(
-    x = "Human Impact Area (HIA) class",
+    x = "Homogeneous Impact Areas",
     y = "Area (km<sup>2</sup>)"
@@
   guides(fill = "none", colour = "none") +
-  scale_y_log10() +
+  #scale_y_log10() +
   facet_grid(cols = vars(Municipality))
+)
+
+#ggsave(plot = infra_dist_plot,
+#       "../images/infra-dist-plot.jpg")
 ```
@@
 corrCheck <- st_intersection(naturetypes, infra)
-
-# corrCheck |>
-#  as_tibble() |>
-#  mutate(i_ADSV_fct = floor(round(i_ADSV*10,2))/10,
-#         i_alien_fct = floor(round(i_alien*10,2))/10) |>
-#  count(i_ADSV_fct, infrastructureIndex) |>
-#  View()
-
 ```
@@
   facet_grid(indicator ~ year)
-```
-
-```{r test 4}
-#| eval: false
-# The year effect is constant, so ignoring that
-corrCheck |>
-  mutate(
-    i_ADSV_fct = floor(round(i_ADSV * 10, 2)) / 10,
-    i_alien_fct = floor(round(i_alien * 10, 2)) / 10,
-    i_ditch_fct = floor(round(i_ditch * 10, 2)) / 10
-  ) |>
-  pivot_longer(
-    cols = c(i_ADSV_fct, i_alien_fct, i_ditch_fct),
-    values_to = "indicatorValue",
-    names_to = "indicator",
-    values_drop_na = T
-  ) |>
-  ggplot(aes(
-    x = factor(infrastructureIndex),
-    fill = factor(indicatorValue)
-  )) +
-  geom_bar(
-    position = "fill"
-  ) +
-  theme_bw(base_size = 12) +
-  guides(fill = guide_legend("Scaled indicator values")) +
-  ylab("Fraction of data points") +
-  xlab("HIA") +
-  scale_fill_brewer(
-    palette = "RdYlGn") +
-  facet_grid(~indicator)
+
+# After this I also tried chaning the color gradient, 
+# and the number of categories. I tries discrete colors and log-transformation.
+# These variants are interpretted slightly differetly by the brain.
+# See versions pre 27.02.2023
 ```
@@
 
-```{r test3}
-#| eval: false
-# The figure changes a lot depending on which value gets the darker colour
-corrCheck |>
-  pivot_longer(
-    cols = c(i_ADSV, i_alien, i_ditch),
-    values_to = "indicatorValue",
-    names_to = "indicator",
-    values_drop_na = T
-  ) |>
-  mutate(
-    condition = case_when(
-      indicatorValue < 0.6 ~ "Poor",
-      indicatorValue < 0.8 ~ "Intermediate",
-      indicatorValue < 0.91 ~ "Good",
-      .default = "Perfect"
-    ),
-    condition = fct_reorder(condition, -indicatorValue)
-  ) |>
-  ggplot(aes(
-    x = factor(infrastructureIndex),
-    fill = condition
-  )) +
-  geom_bar(
-    position = "fill"
-  ) +
-  theme_bw(base_size = 12) +
-  guides(fill = guide_legend("Scaled indicator values")) +
-  ylab("Fraction of data points") +
-  xlab("HIA") +
-  scale_fill_viridis_d(alpha = .8) +
-  facet_grid(~indicator)
-```
-
-```{r test2}
-#| eval: false
-# If I look at this on a log scale, it also changes and becomes less convincing for ADSV and alien
-corrCheck |>
-  pivot_longer(
-    cols = c(i_ADSV, i_alien, i_ditch),
-    values_to = "indicatorValue",
-    names_to = "indicator",
-    values_drop_na = T
-  ) |>
-  mutate(
-    condition = case_when(
-      indicatorValue < 0.6 ~ "Poor",
-      indicatorValue < 0.8 ~ "Intermediate",
-      indicatorValue < 0.91 ~ "Good",
-      .default = "Perfect"
-    ),
-    condition = fct_reorder(condition, -indicatorValue)
-  ) |>
-  ggplot(aes(
-    x = factor(infrastructureIndex),
-    fill = condition
-  )) +
-  geom_bar(
-    position = "fill"
-  ) +
-  theme_bw(base_size = 12) +
-  guides(fill = guide_legend("Scaled indicator values")) +
-  ylab("Fraction of data points") +
-  xlab("HIA") +
-  scale_y_log10() +
-  scale_fill_viridis_d(alpha = .8) +
-  facet_grid(~indicator)
-
-# This is not correct I think as it evens out differences that are clearly there.
-# i_ditch is shown less of a pattern for example
-```
-
-```{r test}
-#| eval: false
-
-# HIAs 1 and 2 are quite similar for i_alien and i_ADSV, so perhaps I can combine them
-corrCheck |>
-  pivot_longer(cols = c(i_ADSV, i_alien),
-               values_to = "indicatorValue",
-               names_to = "indicator") |>
-  filter(!is.na(indicatorValue)) |>
-  mutate(condition = case_when(
-    indicatorValue <0.6 ~"Poor",
-    indicatorValue <0.8 ~"Intermediate",
-    indicatorValue <0.91 ~"Good",
-    .default = "Perfect"
-  ),
-  condition = fct_reorder(condition, indicatorValue),
-  HIA = case_when(
-    infrastructureIndex %in% c(1,2) ~ "1 & 2",
-    .default = factor(infrastructureIndex)
-  )
-  ) |>
-  ggplot(aes(
-    x = HIA, 
-    fill = condition))+
-  geom_bar(
-    position="fill"
-    )+
-  theme_bw(base_size = 12)+
-  guides(fill = guide_legend("Scaled indicator values"))+
-  ylab("Fraction of data points")+
-  xlab("HIA")+
-  scale_y_log10() +
-  scale_fill_viridis_d(alpha = .8) +
-  facet_grid(~indicator)
-```
-
 ```{r infraMuniMaps}
@@
 #           height = 10)
-
 #
@@
 # Number of mire polygons per municipality
-# -- Try count() here instead
 nature3_tbl <- nature3 |>
@@
 # using left_join.
-# Save local cached file
 muni_tbl <- muni3 |>
@@
   left_join(infraMuni3_summary)
-```
-
-```{r read-munitbl}
-#| eval: true
-#muni_tbl <- readRDS("../data/municipality_stats.RDS")
 ```
@@
 ```{r forets plot}
+#| eval: true
+#| cache: true
+
 # Forest plot
-# https://stackoverflow.com/questions/58657802/forest-plot-with-subgroups-in-ggplot2
-# 
+
 # #define colours for dots and bars
@@
 
-```{r}
+```{r notes}
 # Next I need to 
@@
 
-spread_na_map <- spread_na |>
+(spread_na_map <- spread_na |>
   filter(indicator == "i_ditch") |>
@@
   theme_bw()
-
+)
 #ggsave("../images/spread-na.tiff",
@@
 # Now I sample indicator values from the EDM and create a new distribution for 
-# indicator values in the EEA (the municipalities). First I sample the individual
+# indicator values in the EAAs (the municipalities). First I sample the individual
 # distributions for each mire polygon with n defined by the polygon area.
@@
 # some polygons will contribute much more to the EAA value than others.
-# I will keep this design in-line with SEEA EA guidelines for area weighting, 
+# I will keep this design, in-line with SEEA EA guidelines for area weighting, 
 # but it's worth noting that this could be solved in other ways.
@@
   mutate(
+    # this draws one sample pre m2 from a normal distribution:
     i_sample = list(sample(rnorm(n, mean, sd)))) |>
@@
   reframe(i_sample = 
+  # this samples randomly from i_sample, ie large polygons are more likely
+  # to contibute:
             sample(
@@
               replace = TRUE)) |>
+  # truncation, since the tails from the normal distribution can go beyond 0 and 1
   mutate(i_sample = case_when(
@@
 #| cache: true
-library(ggridges)
-
 (ridgepPlot <- combineAll |>
@@
 
+#EEA_tbl |>
+#  kable()
 # saveRDS(EEA_tbl, "../output/EEA-table.RDS")
@@
 | id  | Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                                                                       | Reference                          |
-|-------|---------|-----------|--------------|----------------------|----------|
+|----------|----------|----------|----------|-----------------------|----------|
 | 1   | 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                                                                                | @runehalvorsen2019                 |

### 2024-02-27 manuscript/HIAs.qmd
add EEA table

@@
 
-#EEA_tbl |>
-#  kable()
+library(kableExtra)
+
 # saveRDS(EEA_tbl, "../output/EEA-table.RDS")
+```
+
+```{r}
+(EEA_tbl_out <- EEA_tbl |>
+  ungroup() |>
+  mutate(value = 
+    paste0(
+      format(median, 2), 
+      " [",
+      format(percentile_025,2),
+      " - ",
+      format(percentile_975,2),
+      "] ")) |>
+  rename(Indicator = indicator) |>
+  select(
+    -mean,
+    -median,
+    -percentile_025,
+    -percentile_975,
+    -Municipality
+  )|>
+  kbl(table.attr = "style = \"color: black;\"",
+      align = "lr") |>
+  kable_classic("striped",
+    full_width = F) |>
+  row_spec(0, bold=T) %>%
+    pack_rows("Gran", 1, 3, 
+      label_row_css = "background-color: #cef598; color: #000000;") |>
+  pack_rows("Nord-Aurdal", 4, 6, 
+      label_row_css = "background-color: #cef598; color: #000000;") |>
+  pack_rows("Nordre Follo", 7, 9, 
+      label_row_css = "background-color: #cef598; color: #000000;")
+)
 ```

### 2024-02-28 manuscript/Appendix A.qmd
Appendix A

---
title: Appendix A
subtitle: For the article 'Using spatially biased variables in ecosystem condition accounting with a GIS based workflow'
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    orchid: https://orcid.org/0000-0002-9623-9491
    affiliations: 
        - id: nina
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Pb 5685 Torgarden
          city: Trondheim
          postal-code: 7485
  - name: Matthew Grainger
    email: matthew.grainger@nina.no
    orchid: https://orcid.org/0000-0001-8426-6495
    affiliations:
        - ref: nina
date: last-modified
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Indicators
      formatting: preprint
      model: 3p
      cite-style: authoryear
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
editor: 
  markdown: 
    wrap: sentence
crossref:
  fig-title: |
    Supplementary Figure
  fig-prefix: Fig S.
  tbl-title: |
    Supplementary Table
  tbl-prefix: Fig S.
execute:
  echo: false
---

```{r fig-studyLocations}
#| fig.cap: "Position of the three example municiaplieies in Norway."
#| out.width: '50%'

knitr::include_graphics("../images/positionMap.jpg")
```

{{< pagebreak >}}
```{r tbl-EEA}
#| tbl-cap: "Idicator values for three ecosystem condition indicators in three 
#| Norwegian municipalities. Values are medians with 2.5 and 97.5 percentiles
#| for the distribution of possible values."

(tbl <- readRDS("../output/EEA-tbl.RDS"))
```

### 2024-02-28 manuscript/HIAs.qmd
some example figs

--- 2024-02-27 manuscript/HIAs.qmd
@@
 ---
-title: Using spatially biased variables in ecosystem condition accounting with a GIS based workflow
+title: Horizontal aggregation of spatially biased ecosystem condition indicators – a GIS based workflow
 subtitle: A Short Subtitle
@@
 
-<!--# The document is set to eval=F in the yaml.
-This only affect the rendering, and not the interactive session.
-Eval is also set explicitly for each chunk.
-Larger operations are cached.
-Remember to clean the cache occatinally and re-run.
--->
-
 ```{r setup}
-#| warning: false
-#| message: false
-#| include: false
 #| eval: true
@@
 library(ggridges)
+library(eaTools) #https://ninanor.github.io/eaTools/ version 0.0.0.9000
+library(ggpubr)
+library(kableExtra)
 
@@
 # The data is too big to be stored on GitHub
-# Impart polygon data set
+# Import polygon data set
 st_layers(path_naturetypes)
@@
 # 142k polygons (2023)
-# 
+
 # Impart survey coverage map
@@
 # (then thinking 7SE and PRSL as the same variable)
-# is recorded. To combine these into one metric ADSV, I could take the
-# one with the highest value (worst-rule) og the sum.
-# Sum is problematic is not all locations have two values to sum together.
+# is recorded.
+
+# To combine these into one metric, ADSV, I could take the
+# one with the highest value (worst-rule) or the sum.
+# Sum is problematic as not all locations have two values to sum together.
 # But the other option is problematic since I think field workers often
@@
 
-eaTools::ea_normalise(data = naturetypes,
+scale1 <- eaTools::ea_normalise(data = naturetypes,
   vector = "ADSV",
@@
   reverse = T
-  )
+  ) +
+  labs(x = "ADVS (converted to %)") +
+  ylim(0,1)
 
@@
 # same for 7FK
-eaTools::ea_normalise(data = naturetypes,
+scale2 <- eaTools::ea_normalise(data = naturetypes,
   vector = "7FA",
@@
   reverse = T
-  )
+  ) +
+  labs(x = "7FA (converted to %)",
+       y = "") +
+  ylim(0,1)
 # The variables are really coarse
 
-eaTools::ea_normalise(data = naturetypes,
+scale3 <- eaTools::ea_normalise(data = naturetypes,
   vector = "7GR-GI",
@@
   reverse = T
-  )
-
+  )+
+  labs(x = "7GR-GI (original units)",
+       y = "") +
+  ylim(0,1)
+
+(scaling_plot <- ggarrange(scale1,
+          scale2,
+          scale3,
+          ncol=3)
+)
+
+#ggsave(plot = scaling_plot,
+#  "../images/scaling-plot.jpg",
+#  width=8,
+#  height=5)
 ```
@@
   st_make_valid()
+
+# get centroids
+centroids <- muni3 |>
+  st_centroid()
 
@@
 
+tmap_save(tm = positionMap,
+       "../images/positionMap.jpg")
 ```
@@
 # what is the distance between Nordre Follo and Nord-Aurdal
-# First, get centroids
-centroids <- muni3 |>
-  st_centroid()
-
 (km_distance <- centroids |>
@@
 
-ggsave("../images/validation-plot.jpg",
-       plot=validationPlot,
-       width = 9,
-  height = 5)
+#ggsave("../images/validation-plot.jpg",
+#  plot=validationPlot,
+#  width = 9,
+#  height = 5)
 ```
@@
 # Infratructure in each municipality
-(infraMuniMap <- tm_shape(muni3) +
+infraMuniMap <- tm_shape(muni3) +
   tm_borders() +
@@
   tm_borders(lwd = 3, col = "black") +
-  tm_facets(by = "Municipality"))
+  tm_facets(by = "Municipality")
 
@@
 
-```{r forets plot}
+```{r forets-plot}
 #| eval: true
@@
 #       plot=forest_plot)
+```
+
+```{r forest_plotExample}
+#| eval: true
+#| cache:true
+# Here is the same plot but only for trenching
+
+# #define colours for dots and bars
+dotCOLS = c("grey90","grey70", "grey50", "grey40")
+barCOLS <- tmaptools::get_brewer_pal("Set1", n = 4)
+
+(forest_plot_ex <- stats_tbl |>
+  filter(indicator == "i_ditch",
+         Municipality == "Nord-Aurdal") |>
+  rowwise() |>
+  mutate(
+    low = quantile(rnorm(200, mean, sd), probs = 0.025),
+    high = quantile(rnorm(200, mean, sd), probs = 0.975),
+    high = ifelse(high > 1, 1, high)
+  ) |>
+  ggplot(aes(x=infrastructureIndex, 
+             y=mean, 
+             ymin=low,
+             ymax=high,
+             col=factor(infrastructureIndex),
+             fill=factor(infrastructureIndex))) + 
+  geom_linerange(
+    size=10) +
+  geom_point(
+    size=3, 
+    shape=21, 
+    colour="white", 
+    stroke = 0.5,
+    ) +
+  scale_fill_manual(values=dotCOLS)+
+  scale_color_manual(values=barCOLS)+
+  scale_x_discrete(name="") +
+  scale_y_continuous(name="Indicator values", limits = c(.7, 1)) +
+  coord_flip() +
+  theme_bw() +
+  labs(fill = "HIA",
+       col = "HIA")
+)
+
+#ggsave(plot = forest_plot_ex,
+#       "../images/forest-plot-ex.jpg",
+#       width=3,
+#       height = 3)
 ```
@@
 
+```{r EEA-plot2}
+#| eval:true
+#| cache: true
+# Same as above, but for trenching only
+
+(ridgepPlot2 <- combineAll |>
+  filter(indicator == "i_ditch",
+     Municipality == "Nord-Aurdal") |>
+  select(indicator, i_sample, Municipality) |>  
+  ggplot(
+    aes(
+      x = i_sample)
+    ) +
+  geom_density(
+    fill = "lightgreen",
+    alpha=.5
+  ) +
+  theme_bw(base_size = 15) +
+  guides(fill = "none") +
+  theme(
+        axis.ticks.y = element_blank(),
+        axis.text.y = element_blank())+
+  labs(y = "", x = "Indicator values")+
+  xlim(c(0.7,1))
+)
+
+#ggsave(plot = ridgepPlot2,
+#       "../images/ridgePlot-ex.jpg",
+#       width=6,
+#       height=4)
+```
+
 ```{r EEA-tbl}
@@
 )
-
-#EEA_tbl |>
-#  kable()
 # saveRDS(EEA_tbl, "../output/EEA-table.RDS")
+```
+
+```{r}
+(EEA_tbl_out <- EEA_tbl |>
+  ungroup() |>
+  mutate("Indicator value" = 
+    paste0(
+      format(median, 2), 
+      " [",
+      format(percentile_025,2),
+      " - ",
+      format(percentile_975,2),
+      "] ")) |>
+  rename(Indicator = indicator) |>
+  select(
+    -mean,
+    -median,
+    -percentile_025,
+    -percentile_975,
+    -Municipality
+  )|>
+  kbl(table.attr = "style = \"color: black;\"",
+      align = "lr") |>
+  kable_classic("striped",
+    full_width = F) |>
+  row_spec(0, bold=T) %>%
+    pack_rows("Gran", 1, 3, 
+      label_row_css = "background-color: #cef598; color: #000000;") |>
+  pack_rows("Nord-Aurdal", 4, 6, 
+      label_row_css = "background-color: #cef598; color: #000000;") |>
+  pack_rows("Nordre Follo", 7, 9, 
+      label_row_css = "background-color: #cef598; color: #000000;")
+)
+
+#saveRDS(EEA_tbl_out, "../output/EEA-tbl.RDS")
+```
+
+```{r indicator-map-zoom}
+#| eval: true
+#| cache: true
+
+# Just an example figure to show what the spatial indicator data looks like
+from <- c(xmin = 510000, xmax = 510600, ymin = 6747200, ymax = 6747800)
+to <-   c(xmin = 495000, xmax = 520000, ymin = 6765000, ymax = 6790000)
+myCols <- c(
+  #"#E85437",
+  "#FBAF00", 
+  "#B5DF73", 
+  "#009000"
+  )
+
+(indicator_magnify <- nature3 |>
+  select(i_ditch) |>
+  drop_na() |>
+  mutate(Trenching = factor(format(round(i_ditch, 2), 2))) |>
+  ggplot() +
+  geom_sf(data = na,
+          alpha=0) +
+  geom_sf(aes(fill=Trenching,
+              color = Trenching)) +
+  scale_color_manual(values = RColorBrewer::brewer.pal(5, "Set2")) +
+  scale_fill_manual(values = RColorBrewer::brewer.pal(5, "Set2")) +
+  coord_sf(
+    datum = st_crs(myCRS),
+    xlim = c(494174.8 , 537114.7 ), 
+    ylim = c(6737092 , 6789676)) +
+  ggmagnify::geom_magnify(from = from, to = to, 
+                          expand = 0,
+                          shadow =T,
+                          corners = 0.1) +
+  theme_bw()
+)
+
+#ggsave(plot = indicator_magnify,
+#       "../images/indicator-magnify.jpg")
 ```
@@
 
-```{r fig-studyLocation}
-#| eval: true
-#| include: true
-#| out.width: '100%'
-#| fig.cap:
-#| - "Postion of the three example municipalities in
-#| Norway. "
-
-positionMap
-```
-
 # Material and Methods
@@
 This study makes use of a data set from from a standardised field survey of nature types in Norway that started in 2018 and which is still ongoing [@norwegian_environmental_agency_naturtyper_2024].
-We included data from 2018 to 2022.
-In this survey, selected nature types are delineated on a map, and each locality is scored on a range of variables relevant for describing the state and quality of nature.
+In this survey, selected nature types are delineated on a map (over 140 000 polygons at the end of 2023), and each locality is scored on a range of variables relevant for describing the state and quality of nature (REF INSTRUKS).
 The surveys are commissioned with the goal of producing data relevant for immediate land-use decisions, and is therefore spatially biased, typically towards areas with high human impact or expected impact.
@@
 
-In the above, *southern* refers to boreonemoral and southboreal zones, and *northern* refers to mid-boreal, north boreal, and alpine zones @asbjørnmoen1998 .
+In the above, *southern* refers to boreonemoral and southboreal zones, and *northern* refers to mid-boreal, north boreal, and alpine zones @asbjørnmoen1998.
 In addition, the northern fens need to be even more calcareous than the southern fens in order to be surveyed.
-
-<!--# and the variable 7FK is not recorded in ombrotrophic mires -->
+We included data from 2018 to 2023.
 
@@
 Although it is possible that smaller or less calcareous mires will score systematically different than the ones that are surveyed we do not think this is so much the case for our variables.
-It may also be that alien plants are more or less common on bogs relative to fens.
-For this we have no assumption.
-
-For the mire ecosystem we have an ecosystem delineation map produced using remotely sensed data and a deep learning model [@bakkestuen_delineation_2023].
+However, we do assume that alien plants are slightly less common on bogs relative to fens, and therefore that this variable will biased since that variable for only recorded in fens.
+
+From the survey data set we identified six relevnt variables (variable 1-6 in @tbl-variables) which we attribute to three different ecosystem condition characteristics that describe the typical behavior of open mire ecosystems in the reference condition @fig-workflow.
+
+Variables 1 -5 were were originally recorded along binned frequency ranges.
+because the data was strongly right skewed we used the lower limit for each frequency range to convert them into percentages.
+Variables 2-5 describe very related aspects, and are attributed to the same ecosystem condition characteristic (vegetation intactness) and so they where combined into a single indicator called antropogenic disturbance to soil and vegetation, or ADSV for short.
+This was done by summing the two variables after they had been converted to percentages.
+This was not a perfect solution, especially since some localities only had one of the variables recorded, but we chose this, rather than for example using a *worst rule* principle, in order to better separate the localities in terms of their indicator values.
+
+To turn the variables into indicators we scaled them using three normative reference values each: an upper (best possible condition), a lower (worst possible condition), and a threshold value that defines the breaking point between good and poor condition.
+The reference values are numerical representation of the reference condition.
+We define the reference condition as one where ecosystems are subject to little or no human influence, with a climate as in the period 1961-1990 and a native species pool similar as today.
+
+For the indicators Alien species and ADSV, the lower and upper reference values were defined as 100% and 0%, respectively.
+The threshold for *good ecosystem condition* was defined as 10%, which was then mapped to the value 0.6 on the rescaled indicators [@fig-scaling].
+The variable 7GR-GI was rescaled into the indicator named *trenchin* by using the lower and upper reference values 1 and 5, respectively, and a threshold value of 2.5.
+A variable value of 1 indicates an intact mire, and a value of 5 indicates a mire transitioning away from a wetland.
+A value of 2 indicates observable change within the range expected for the same mapping unit, and a value of 3 indicates a mire transitioning into a neighboring (ecologically speaking) mapping unit.
+See @fig-workflow for a schematic workflow for the indicator development.
+
+*Alien species* was assigned to ECT class B1 - Compositional state characteristics and *ADSV* and *Trenching* were both assigned to ECT class A1 - Physical state characteristics [@czucz_common_2021].
+
+We used an ecosystem delineation map for open mires in southern Norway, produced using remotely sensed data and a deep learning model [@bakkestuen_delineation_2023].
 This model, which has a 90.9% precision when tested against independent field data, estimates 12.7% of the area in southern Norway is mire [@bakkestuen_delineation_2023].
-Mires are ecologically and socially important in Norway simply doe to its large extent, and due to its role in climate mitigation, as mires store a large amount of carbon (REF?).
-There has not been a national assessment of the ecosystem condition of mires in Norway, but we recent contributed on a report which presented several new indicators that can be used in future assessment [@nybo_indikatorer_2023; see also @kolstad2023]).
-This study builds on the work in that report.
-
-We started by defining the reference state as one where ecosystems are subject to little or no human influence, with a climate as in the period 1961-1990 and a native species pool similar as today (i.e no mammoths allowed).
-We then identified two *ecosystem condition characteristics* that, along with many other characteristics that we do not specify here, define the typical behavior of open mire ecosystems in the reference condition: species composition, and; vegetation intactness.
-To describe these two characteristics we use six condition variables from the same data set ( @tbl-variables).
-See @miljødirektoratet for the sampling protocol (in Norwegian).
+Mires are ecologically and socially important in Norway simply doe to its large extent, and due to its role in climate mitigation as mires store a large amount of carbon (REF?).
+There has not been a national assessment of the ecosystem condition of mires in Norway, but the authors recent contributed on a report which presented several new indicators that can be used in future assessments for this ecosystem [@nybo_indikatorer_2023; see also @kolstad2023]).
+The current study builds on the work in that report.
 
@@
 
-<!--# Three characteristics now? -->
+{{< pagebreak >}}
 
 | id  | Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                                                                       | Reference                          |
-|----------|----------|----------|----------|-----------------------|----------|
+|------------|------------|------------|------------|------------|------------|
 | 1   | 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                                                                                | @runehalvorsen2019                 |
@@
 | 6   | 7GR-GI               | Ditching intensity                               | Unitless, ordinal scale from 1 to 5                                                     | Describes the effect that drainage ditches is estimate to have on the species composition and environmental variables ones the system reached its new equilibrium | @miljødirektoratet                 |
-| 7   | Infrastructure Index | Infrastructure Index                             | Unitelss linear scale from 0 to 13.2                                                    | Unitelss index ranging from from 0 to 13.2                                                                                                                        | @erikstad_index_2023               |
+| 7   | Infrastructure Index | Infrastructure Index                             | Unitless linear scale from 0 to 13.2                                                    | Unitelss index ranging from from 0 to 13.2                                                                                                                        | @erikstad_index_2023               |
 | 8   | HIA                  | Human Impact area                                | Ordinal, non-linear scale from 1 to 4                                                   | A categorical representation of the infrastructure index                                                                                                          | this paper                         |
@@
 : Variables used in this study {#tbl-variables}
-
-Variables 1 -5 were 7FK were originally recorded along frequency ranges.
-We the lower limit for each frequency range to convert them into percentages.
-This was done because the data was strongly right skewed .
-7FK was developed into a single indicator (i.e. a normalised variable) named *Alien plant cover*.
-Variables 2-5 describe very related aspects and so they where combined into one indicator called antropogenic disturbance to soil and vegetation, or ADSV for short.
-This was done by summing the two variables after they had been converted to percentages.
-This was not a perfect solution, especially since some localities only had one of the variables recorded, but we chose this, rather than for example using a *worst rule* principle, in order to better separate the localities in terms of their indicator values.
-For both indicators (alien species and ADSV), the lower and upper reference values, i.e. the worst and best possible condition that the variables can be in, is defined as 100% and 0%, respectively.
-The threshold for *good ecosystem condition* was defined as 10%, and this was mapped to the value 0.6 on the rescaled indicators.
-Variable 7GR-GI was rescaled into an indicator named *ditching intensity* by using the lower and upper reference values 1 and 5, respectively, and a threshold value of 2.5.
-A variable value of 1 indicates an intact mire, and a value of 5 indicates a mire transitioning away from a wetland.
-A value of 2 indicator observable change within the range expected for the same mapping unit, and a value of 3 indicates a mire transitioning into a neighboring (ecologically speaking) mapping unit.
-See @fig-workflow for a schematic workflow for the indicator development.
-
-The Alien plant cover was assigned to ECT class B1 - Compositional state characteristics, and ADSV was assigned to ECT class A1 - Physical state characteristics [@czucz_common_2021].
 
@@
 
-We chose three municipalities in Norway to test out the indicators (@fig-studyLocation).
-The differ in several aspect, such as the amount of mire area, the total area surveyed, and the prevalence of infrastructure (@tbl-variables).
+We chose three municipalities in Norway to test out the indicators.
+THe municipalities differ in several aspects, such as the amount of mire area, the total area surveyed, and the prevalence of infrastructure (@tbl-munis, @fig-studyLocation2, Figure S1).
 
@@
 
-: Information for the three target municipalities in Norway
-
-Validation/sensitivity analyses (varying HIA resolution and minimum *n*?)
-
-Aggregation
+: Information for the three target municipalities in Norway {#tbl-munis}
+
+Because the nature type survey data is spatially biased, we cannot simply assume that they are area representative.
+In an attempt to overcome this issue we divided the area of Norway into four non-overlapping Homogeneous Impact Areas (HIAs) based on an infrastructure index [@tbl-variables, @erikstad_index_2023].
+This index measures the frequency of different infrastructure types inside 500 m radius circles around each 10×10 m pixel.
+We categorised this continuous variable into four classes (0–3) using the following value ranges: 0 = \<1; 1 ≤ 1 ≥ 6; 6 ≤ 2 ≥12; 3 ≥ 12.
+We plotted this categorical map over an area in Norway that we are familiar with and confirmed a sensible seperation of areas from areas with no roads of major infrastructure (HIA = 0) to urban areas (HIA = 3).
+We then aggregated the data to 1×1 km pixels to ease computations.
+
+We then took the nature type polygons with the indicator values and intersected with the HIA map and a map of municipal outlines in Norway.
+For each HIA and municipality combination we then created a probability distribution for the indicator values using Bayesian updating of a uniform prior between 0 and 1.
+The resulting distribution will be
 
@@
 Data availability/coverage
+
+```{r fig-scaling}
+#| eval: true
+#| include: true
+#| out.width: '70%'
+#| fig.cap:
+#| - "Custom re-scaling of variables to indicators. Unique threshold values are 
+#| assignes to each variable and this value is normalised to become 0.6 on the 
+#| indicator scale (y-axis). The variables also differ in the number of possible 
+#| values they can take, hence the varying number of points for the three panes."
+
+knitr::include_graphics("../images/scaling-plot.jpg")
+```
 

### 2024-02-29 manuscript/Appendix A.qmd
workflow example

@@
 format:
-  elsevier-pdf:
-    keep-tex: true
-    journal:
-      name: Ecological Indicators
-      formatting: preprint
-      model: 3p
-      cite-style: authoryear
-    include-in-header:
-      text: '\usepackage{lineno}\linenumbers'
+  pdf
 editor: 
@@
   fig-title: |
-    Supplementary Figure
+    Figure S
+  tbl-prefix: Fig S.
   fig-prefix: Fig S.
   tbl-title: |
-    Supplementary Table
-  tbl-prefix: Fig S.
+    Table S
 execute:
@@
 
+```{r fig-studyLocations}
+#| fig.cap: "Barplot showing the total area from each homogeneous impact area 
+#| in three Norwegian municipalities"
+#| out.width: '50%'
+
+knitr::include_graphics("../images/infra-dist-plot.jpg")
+```
+
+```{r fig-scaling}
+#| eval: true
+#| include: true
+#| out.width: '70%'
+#| fig.cap:
+#| - "Custom re-scaling of variables to indicators. Unique threshold values are 
+#| assignes to each variable and this value is normalised to become 0.6 on the 
+#| indicator scale (y-axis). The variables also differ in the number of possible 
+#| values they can take, hence the varying number of points for the three panes."
+
+knitr::include_graphics("../images/scaling-plot.jpg")
+```
+
 {{< pagebreak >}}
 ```{r tbl-EEA}
-#| tbl-cap: "Idicator values for three ecosystem condition indicators in three 
+#| tbl-cap: "Indicator values for three ecosystem condition indicators in three 
 #| Norwegian municipalities. Values are medians with 2.5 and 97.5 percentiles

### 2024-02-29 manuscript/HIAs.qmd
workflow example

@@
 
+```{r infra-NA}
+# Infratructure in Nord-Aurdal only
+(infraMuniMap_NA <- 
+  tm_shape(muni3 |> filter(Municipality == "Nord-Aurdal")) +
+  tm_borders() +
+  tm_shape(infraMuni3) +
+  tm_polygons(
+    col = "infrastructureIndex",
+    style = "cat",
+    title = "Homogeneous\nImpact\nAreas"
+  ) +
+  tm_layout(
+    legend.show = T,
+    legend.position = c("left", "top"),
+    legend.text.size = 1.2) +
+  tm_shape(muni3) +
+  tm_borders(lwd = 3, col = "black")
+)
+
+tmap_save(tm = infraMuniMap_NA,
+       "../images/HIA-NA.jpg",
+       width=4,
+       height=4)
+```
+
 ```{r muni-tbl}
@@
 
-Data availability/coverage
-
-```{r fig-scaling}
+```{r fig-workflowExample}
 #| eval: true
 #| include: true
-#| out.width: '70%'
-#| fig.cap:
-#| - "Custom re-scaling of variables to indicators. Unique threshold values are 
-#| assignes to each variable and this value is normalised to become 0.6 on the 
-#| indicator scale (y-axis). The variables also differ in the number of possible 
-#| values they can take, hence the varying number of points for the three panes."
-
-knitr::include_graphics("../images/scaling-plot.jpg")
-```
-
-Validation
+#| out.width: '100%'
+#| fig.cap: |
+#| - "Simplified schematic showing the workflow (Figures a-E) for horizontal aggregation of a spatially biased ecosystem condition indicator.
+Figure A shows a spatially explicit indicator.
+The colors indicator different indicator values (highlighted in the inset).
+The outline is the ecosystem accounting area (EEA).
+Figure B shows the distributio of four homogeneous impact areas (HIAs) inside the EEA.
+The indicator values in Figure A are used to update a uniform prior to produce a posterior probability distributio for the mean indicator value for each HIA (Figure C).
+The bayesian updating is informed by the standard deviation for the indicator in the much bigger national data set.
+Unlike an arithmetic approach where the variance in the mean indicator value per HIA is unreligable for very small sample sizes, the posterior distribution gives a religable measure for the indicator uncertainty even with a singe observation.
+The distributions in Figure C are assumed normally distributed.
+In Figure D the mean and SD from the posterior distributions are transferred to individual polygons in an ecosystem delineation map, for each HIA seperately.
+For each ecosystem occurence (i.e.
+ecosystem assets), we draw one random value for each square meter from a normal distribution with the mean and SD that is associated with that polygon.
+In Figure E we randomly sample 1000 values from the entire vector of sample in the previous step, and we get a area weighted probability distribution for the mean indicator value for the EEA."
+
+knitr::include_graphics("../images/aggregation-workflow.jpg")
+
+```
 

### 2024-02-29 manuscript/Appendix A.qmd
methods draft complete

@@
 title: Appendix A
-subtitle: For the article 'Using spatially biased variables in ecosystem condition accounting with a GIS based workflow'
+subtitle: For the article 'Horizontal aggregation of spatially biased ecosystem condition indicators – a GIS based workflow'
 author:
@@
         - ref: nina
+  - name: Marianne Evju
+    email: marianne.evju@nina.no
+    orchid: https://orcid.org/0000-0001-7338-5376
+    affiliations:
+        - id: nina2
+          name: Norwegian Institute for Nature Research
+          department: NINA Oslo
+          address: Sognsveien 68
+          city: Oslo
+          postal-code: NO-0855
 date: last-modified
@@
 
-```{r fig-studyLocations}
-#| fig.cap: "Position of the three example municiaplieies in Norway."
-#| out.width: '50%'
+```{r fig-workflow}
+#| eval: true
+#| include: true
+#| out.width: '90%'
+#| fig.cap:
+#| - "Schematic workflow followed in this paper. ADSV = Anthropogenic Disturbance to Soil and Vegetation."
 
-knitr::include_graphics("../images/positionMap.jpg")
-```
-
-```{r fig-studyLocations}
-#| fig.cap: "Barplot showing the total area from each homogeneous impact area 
-#| in three Norwegian municipalities"
-#| out.width: '50%'
-
-knitr::include_graphics("../images/infra-dist-plot.jpg")
+knitr::include_graphics("../images/workflow.jpg")
 ```
@@
 
+```{r fig-muniPosition}
+#| fig.cap: "Position of the three example municiaplieies in Norway."
+#| out.width: '50%'
+
+knitr::include_graphics("../images/positionMap.jpg")
+```
+
+```{r fig-muniHIAs}
+#| fig.cap: "Barplot showing the total area from each homogeneous impact area 
+#| in three Norwegian municipalities"
+#| out.width: '50%'
+
+knitr::include_graphics("../images/infra-dist-plot.jpg")
+```
+
 {{< pagebreak >}}

### 2024-02-29 manuscript/Manuscript.qmd

--- 2024-02-29 manuscript/HIAs.qmd
@@
         - ref: nina
+  - name: Marianne Evju
+    email: marianne.evju@nina.no
+    orchid: https://orcid.org/0000-0001-7338-5376
+    affiliations:
+        - id: nina2
+          name: Norwegian Institute for Nature Research
+          department: NINA Oslo
+          address: Sognsveien 68
+          city: Oslo
+          postal-code: NO-0855
 abstract: |
@@
   left_join(mire_in_dk |> select(Municipality, mireInSurvey_km2)) |>
-  mutate(mireInSurvery_percent = round(mireInSurvey_km2 / mire_km2 * 100, 2)) |>
-  left_join(infraMuni3_summary)
+  mutate(mireInSurvery_percent = round(mireInSurvey_km2 / mire_km2 * 100, 1)) |>
+  left_join(infraMuni3_summary) |>
+  mutate(mire_km2 = round(mire_km2, 1),
+         meanHIA = round(meanHIA, 1))
 ```
@@
 
-#ggsave("../images/forest-plot.jpg",
-#       plot=forest_plot)
+ggsave("../images/forest-plot.jpg",
+       plot=forest_plot,
+       width = 8,
+       height = 6)
 ```
@@
 #| eval: true
-#| cache:true
+#| cache: true
+
 # Here is the same plot but only for trenching
@@
 ```{r EEA-plot2}
-#| eval:true
+#| eval: true
 #| cache: true
+
 # Same as above, but for trenching only
@@
 In this study we explore the potential for using a stratified aggregation technique to make use of spatially biased field data in ecosystem condition accounting.
-We demonstrate this technique using a generic GIS-based workflow for compiling ecosystem condition accounts that can be applied at any spatial scale, and we highlight the opportunities for local use-cases of this workflow by contrasting our findings across three neighboring municipalities in Norway (@fig-studyLocation).
+We demonstrate this technique using a generic GIS-based workflow for compiling ecosystem condition accounts that can be applied at any spatial scale, and we highlight the opportunities for local use-cases of this workflow by contrasting our findings across three neighboring municipalities in Norway.
 The main question is how much generalisation can we perform on the data we have before the resulting indicator looses its practical value in local governance processes.
@@
 
-# Material and Methods
+# Material and Methods {#sec-methods}
 
 This study makes use of a data set from from a standardised field survey of nature types in Norway that started in 2018 and which is still ongoing [@norwegian_environmental_agency_naturtyper_2024].
-In this survey, selected nature types are delineated on a map (over 140 000 polygons at the end of 2023), and each locality is scored on a range of variables relevant for describing the state and quality of nature (REF INSTRUKS).
+In this survey, selected nature types are delineated on a map (over 140 000 polygons at the end of 2023), and each locality is scored on a range of variables relevant for describing the state and quality of nature [@norwegian_environmental_agency_protocol].
 The surveys are commissioned with the goal of producing data relevant for immediate land-use decisions, and is therefore spatially biased, typically towards areas with high human impact or expected impact.
@@
 
-In the above, *southern* refers to boreonemoral and southboreal zones, and *northern* refers to mid-boreal, north boreal, and alpine zones @asbjørnmoen1998.
+In the above, *southern* refers to boreonemoral and southboreal zones, and *northern* refers to mid-boreal, north boreal, and alpine zones [@asbjørnmoen1998].
 In addition, the northern fens need to be even more calcareous than the southern fens in order to be surveyed.
 We included data from 2018 to 2023.
-
 In this paper we assume the survey is representative for the entire mire ecosystem in Norway.
-Although it is possible that smaller or less calcareous mires will score systematically different than the ones that are surveyed we do not think this is so much the case for our variables.
-However, we do assume that alien plants are slightly less common on bogs relative to fens, and therefore that this variable will biased since that variable for only recorded in fens.
-
-From the survey data set we identified six relevnt variables (variable 1-6 in @tbl-variables) which we attribute to three different ecosystem condition characteristics that describe the typical behavior of open mire ecosystems in the reference condition @fig-workflow.
-
-Variables 1 -5 were were originally recorded along binned frequency ranges.
-because the data was strongly right skewed we used the lower limit for each frequency range to convert them into percentages.
-Variables 2-5 describe very related aspects, and are attributed to the same ecosystem condition characteristic (vegetation intactness) and so they where combined into a single indicator called antropogenic disturbance to soil and vegetation, or ADSV for short.
+Although it is possible that smaller or less calcareous mires will score systematically different than the ones that are surveyed we do not think this is so much the case for our variables (see below).
+However, we do assume that alien plants are slightly less common on bogs relative to fens, and therefore that this variable will be biased since that variable for only recorded in fens.
+The other variables are recorded in all delineated mires.
+
+From the survey data set we identified six relevant variables (variable 1-6 in @tbl-variables) which we attribute to three different ecosystem condition characteristics that describe the typical behavior of open mire ecosystems in the reference condition (Figure S1).
+Variables 1 -5 were were originally recorded along binned frequency ranges @norwegian_environmental_agency_protocol .
+Because the data was strongly right skewed we used the lower limit for each frequency range to convert them into percentages.
+Variables 2-5 describe very related aspects, and are attributed to the same ecosystem condition characteristic (vegetation intactness) and so they where combined into a single indicator called antropogenic disturbance to soil and vegetation (ADSV; Figure S1).
 This was done by summing the two variables after they had been converted to percentages.
 This was not a perfect solution, especially since some localities only had one of the variables recorded, but we chose this, rather than for example using a *worst rule* principle, in order to better separate the localities in terms of their indicator values.
-
-To turn the variables into indicators we scaled them using three normative reference values each: an upper (best possible condition), a lower (worst possible condition), and a threshold value that defines the breaking point between good and poor condition.
+Variable 6 is different from the preceding variables in @tbl-variables in that it includes an estimation of future effects that the observed trenches are projected to have on mire vegetation, function or structure over time.
+This is not a favorable trait in a metric used to evaluate the ecosystem condition as it is today, yet we include it here nonetheless because there is a general shortage of data on mire hydrology, which is a fundamental part of mire integrity.
+
+To turn the remaining three variables into indicators we scaled them using three normative reference values each: an upper (best possible condition), a lower (worst possible condition), and a threshold value that defines the breaking point between good and poor condition.
 The reference values are numerical representation of the reference condition.
@@
 For the indicators Alien species and ADSV, the lower and upper reference values were defined as 100% and 0%, respectively.
-The threshold for *good ecosystem condition* was defined as 10%, which was then mapped to the value 0.6 on the rescaled indicators [@fig-scaling].
-The variable 7GR-GI was rescaled into the indicator named *trenchin* by using the lower and upper reference values 1 and 5, respectively, and a threshold value of 2.5.
+The threshold for *good ecosystem condition* was defined as 10%, which was then mapped to the value 0.6 on the rescaled indicators, thus creating a non-linear rescaling of the variable (Figure S2).
+The variable 7GR-GI was rescaled into the indicator named *Trenching* by using the lower and upper reference values 1 and 5, respectively, and a threshold value of 2.5 (Figure S2).
 A variable value of 1 indicates an intact mire, and a value of 5 indicates a mire transitioning away from a wetland.
 A value of 2 indicates observable change within the range expected for the same mapping unit, and a value of 3 indicates a mire transitioning into a neighboring (ecologically speaking) mapping unit.
-See @fig-workflow for a schematic workflow for the indicator development.
-
-*Alien species* was assigned to ECT class B1 - Compositional state characteristics and *ADSV* and *Trenching* were both assigned to ECT class A1 - Physical state characteristics [@czucz_common_2021].
+See Figure S1 and @fig-workflowExample for a schematic workflow for the indicator development.
+
+According to the SEEA ECT, *Alien species* can be attributed to ECT class B1 - Compositional state characteristics and *ADSV* and *Trenching* attributed to ECT class A1 - Physical state characteristics [@czucz_common_2021].
 
@@
 This model, which has a 90.9% precision when tested against independent field data, estimates 12.7% of the area in southern Norway is mire [@bakkestuen_delineation_2023].
-Mires are ecologically and socially important in Norway simply doe to its large extent, and due to its role in climate mitigation as mires store a large amount of carbon (REF?).
-There has not been a national assessment of the ecosystem condition of mires in Norway, but the authors recent contributed on a report which presented several new indicators that can be used in future assessments for this ecosystem [@nybo_indikatorer_2023; see also @kolstad2023]).
+Mires are ecologically and socially important in Norway simply doe to its large extent, and due to its role in climate mitigation as mires store a large amount of carbon.
+There has not been a national assessment of the ecosystem condition of mires in Norway, but the authors recent contributed on a report which presented several new indicators that can be used in future national assessments for this ecosystem [@nybo_indikatorer_2023; see also @kolstad2023]).
 The current study builds on the work in that report.
+
+We chose three municipalities in Norway to test out the indicators.
+The municipalities differ in several aspects, such as the amount of mire area, the total area surveyed, and the prevalence of infrastructure (@tbl-munis, @fig-studyLocation2, Figures S3, S4).
 
@@
 #| fig.cap:
-#| - "Map of the three focal municiaplities. For each municipality the map shows 
-#| ocean in blue and non-ocean
-#| in green. The survey coverage maps are in grey, and the mapped mires are
+#| - "Map of the three focal municiaplities. For each municipality the maps
+#| in the top row shows ocean in blue and non-ocean in green. 
+#| The survey coverage maps are in grey, and the mapped mires are
 #| in red, with polygon borderes made extra thich to make them visible,
-#| but then also exagerating their size."
+#| but then also exagerating their size.
+#| The bottom row shows the delineation for homogeneous impact areas which
+#| is an ordinal gradient from 0–3 with increasing presence of human 
+#| instrastructure."
 
@@
 | id  | Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                                                                       | Reference            |
-|------------|------------|------------|------------|------------|------------|
+|--------|------------|------------|------------|-----------------|----------------|
 | 1   | 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                                                                                | @runehalvorsen2019   |
@@
 | 4   | 7TK                  | Tracks from large vehicles                       | Unitless, ordinal, non-linear scale from 1 to 4.
-| Measures the frequency of imagined 100 m^2^ quadrats layed over the area that has some signs of vehicle tracks                                                    | @rune_halvorsen_dokumentasjon_2019 |
+| Measures the frequency of imagined 100 m^2^ quadrats layed over the area that has some signs of vehicle tracks                                                    | @runehalvorsen2019   |
 | 5   | PRTK                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.
 | Same as 7TK, but recorded at a higher resolution                                                                                                                  | @miljødirektoratet   |
-| 6   | 7GR-GI               | Ditching intensity                               | Unitless, ordinal scale from 1 to 5                                                     | Describes the effect that drainage ditches is estimate to have on the species composition and environmental variables ones the system reached its new equilibrium | @miljødirektoratet                 |
+| 6   | 7GR-GI               | Trenching intensity                              | Unitless, ordinal scale from 1 to 5                                                     | Describes the effect that drainage ditches is estimate to have on the species composition and environmental variables ones the system reached its new equilibrium | @miljødirektoratet   |
 | 7   | Infrastructure Index | Infrastructure Index                             | Unitless linear scale from 0 to 13.2                                                    | Unitelss index ranging from from 0 to 13.2                                                                                                                        | @erikstad_index_2023 |
-| 8   | HIA                  | Human Impact area                                | Ordinal, non-linear scale from 1 to 4                                                   | A categorical representation of the infrastructure index                                                                                                          | this paper                         |
+| 8   | HIA                  | Homogeneous Impact Area                          | Ordinal, non-linear scale from 1 to 4                                                   | A categorical representation of the infrastructure index                                                                                                          | this paper           |
 
 : Variables used in this study {#tbl-variables}
-
-```{r fig-workflow}
-#| eval: true
-#| include: true
-#| out.width: '90%'
-#| fig.cap:
-#| - "Schematic workflow followed in this paper. Sub-index A is calculated using 
-#| spatially explicit aggregation."
-
-knitr::include_graphics("../images/workflow.jpg")
-```
-
-We chose three municipalities in Norway to test out the indicators.
-THe municipalities differ in several aspects, such as the amount of mire area, the total area surveyed, and the prevalence of infrastructure (@tbl-munis, @fig-studyLocation2, Figure S1).
 
@@
 
-Because the nature type survey data is spatially biased, we cannot simply assume that they are area representative.
-In an attempt to overcome this issue we divided the area of Norway into four non-overlapping Homogeneous Impact Areas (HIAs) based on an infrastructure index [@tbl-variables, @erikstad_index_2023].
-This index measures the frequency of different infrastructure types inside 500 m radius circles around each 10×10 m pixel.
-We categorised this continuous variable into four classes (0–3) using the following value ranges: 0 = \<1; 1 ≤ 1 ≥ 6; 6 ≤ 2 ≥12; 3 ≥ 12.
-We plotted this categorical map over an area in Norway that we are familiar with and confirmed a sensible seperation of areas from areas with no roads of major infrastructure (HIA = 0) to urban areas (HIA = 3).
-We then aggregated the data to 1×1 km pixels to ease computations.
+Because the nature type survey data is spatially biased, we cannot assume that they are area representative.
+In an attempt to overcome this issue we divided the area of Norway into four non-overlapping Homogeneous Impact Areas (HIAs) based on an infrastructure index for the year 2022 [@tbl-variables, @erikstad_index_2023].
+This index is a continuous variable that represnets the frequency of different infrastructure types inside 500 m radius circles around each 100×100 m pixel.
+We then categorised this continuous variable into four classes (0–3) using the following value ranges: 0 = \<1; 1 ≤ 1 \> 6; 6 ≤ 2 \>12; 3 ≥ 12.
+The name HIAs is not a perfect representation of the information found in the infrastructure index, but we introduce this name here as a general term which is aligned with the concept of Homogeneous Ecosystem Areas, *sensu* @vallecillo_eu-wide_2022.
+We plotted the HIA map over an area in Norway that we are familiar with and confirmed a sensible separation of areas from areas with no roads of major infrastructure (HIA = 0) to urban areas (HIA = 3).
+We then aggregated the data to 1×1 km pixels to ease computations, and vectorized it.
 
 We then took the nature type polygons with the indicator values and intersected with the HIA map and a map of municipal outlines in Norway.
-For each HIA and municipality combination we then created a probability distribution for the indicator values using Bayesian updating of a uniform prior between 0 and 1.
-The resulting distribution will be
+The relationship between the HIA classes and indicator values was examined visually.
+For each HIA and municipality combination we then created a probability distribution for the area weighted mean indicator values using Bayesian updating of a uniform prior between 0 and 1, informed by the standard deviation (SD) of the indicator values in the national data set [@fig-workflowExample C; Appendix C, line XXX].
+The resulting distributions are assumed normally distributed, and we therefore simply carried the mean and SD from the posterior distributions over to individual polygons in the ecosystem delineation map, for each HIA class separately, and for each polygon we then sampled random numbers from a normal distribution with this same mean and SD.
+The number of m^3^ dictated the number of samples for each polygon, thus ensuring that large polygons end up counting more towards the indicator value in the entire municipality.
+We then drew 1000 random values from this much large vector of possible indicator values and created a probability distribution for the indicators for the entire municipalities (i.e.
+the ecosystem accounting areas).
+When there were no indicator values for a given HIA class, we ignored that class also in the municipal estimate.
 
 # Results
+
+The three indicators showed some association with the HIAs when we lookad at data from all of Norway with `r nrow(naturetypes)` individual mires.
+The indicators mostly showed a worsening of condition with increasing presence of human infrastructure (@fig-validation).
+This relationship was much stringer for the indicator Trenching, and weakest for the indicator Alien species.
+Alien species also had the highest indicator values overall, with most mirer having no alien species recorded.
+For Trenching, 57% of mires in HIA-3 had some trenches, whereas in HIA-0 this number was 10%.
+However, only in Gran municipality was there a significant difference in the Trenching indicator between HIAs, with HIA-2 having worse condition than HIA-1 (@fig-forest).
+Conversely, the ADSV indicator in Gran showed worse condition in HIA-1 compared to HIA-2.
+Nordre-Follo had considerably fewer data points compared to Gran, and especially to Nord-Aurdal who had the most data points, and this paucity of data is reflected in the wide credible intervals for all three indicators in Nordre Follo.
+The credible intervals are widest for the Trenching indicators, and this is because the national data set showed more variation in the underlying variable 7GR-GI, and this information is informing the Bayesian updating process (see @sec-methods).
+None of the three municipalities had data for all four HIAs.
+Therefore, when we transferred the mean and SD for each HIA over to the ecosystem delineation map (@fig-spread-example), some mire polygons did not get assigned any data, and were therefore dropped from any further analyses.
+At the municipality level (i.e. the ecosystem accounting area level) the three municipalities show some differentiation.
+Nordre-Follo had the highest (the best) indicator values for ADSV, but the lowest for Trenching, which was also the only instance of an indicator crossing the threshold from good to deteriorated condition (@fig-EEA-ridgePlot; Table S1).
 
@@
 #| out.width: '100%'
-#| fig.cap: |
-#| - "Simplified schematic showing the workflow (Figures a-E) for horizontal aggregation of a spatially biased ecosystem condition indicator.
+#| fig.cap:
+#| - "Simplified schematic showing the workflow (Figures A-E) for horizontal aggregation of a spatially biased ecosystem condition indicator.
 Figure A shows a spatially explicit indicator.
@@
 The outline is the ecosystem accounting area (EEA).
-Figure B shows the distributio of four homogeneous impact areas (HIAs) inside the EEA.
-The indicator values in Figure A are used to update a uniform prior to produce a posterior probability distributio for the mean indicator value for each HIA (Figure C).
-The bayesian updating is informed by the standard deviation for the indicator in the much bigger national data set.
-Unlike an arithmetic approach where the variance in the mean indicator value per HIA is unreligable for very small sample sizes, the posterior distribution gives a religable measure for the indicator uncertainty even with a singe observation.
+Figure B shows the location of four homogeneous impact areas (HIAs) inside the EEA.
+The indicator values in Figure A are used, in combination with the HIA map in Figure B, to update a uniform prior and produce a posterior probability distribution for the mean indicator value for each HIA (Figure C; here simplified to only show the mean (circles) and 95% credible intervals (colored bands)).
+The bayesian updating is informed by the standard deviation for the indicator in a much bigger national data set.
+Unlike an arithmetic approach, the width of the posterior distribution responds both to variation in the data and to the sample size, giving  a religable measure for the uncertainty around the indicator even with a singe observation.
 The distributions in Figure C are assumed normally distributed.
+Note that because in this example there were no indicator data for HIA-3, indicator values are only aggregated for HIAs 0–2.
 In Figure D the mean and SD from the posterior distributions are transferred to individual polygons in an ecosystem delineation map, for each HIA seperately.
 For each ecosystem occurence (i.e.
-ecosystem assets), we draw one random value for each square meter from a normal distribution with the mean and SD that is associated with that polygon.
-In Figure E we randomly sample 1000 values from the entire vector of sample in the previous step, and we get a area weighted probability distribution for the mean indicator value for the EEA."
+ecosystem assets) in Figure D, we draw one random value for each square meter of area from a normal distribution with the mean and SD that is associated with that polygon.
+In Figure E we randomly sample 1000 values from the entire vector of samples in the previous step, and we get an area weighted probability distribution for the mean indicator value for the EEA."
 
@@
 #| fig.cap:
-#| - "Proportion of localities with different scaled indicator values for three 
+#| - "Proportion of localities with different indicator values for three 
 #| ecosystem condition indicators along a gradient of infrastructure
@@
 #| - "Indicator values (circles = mean; bars = 95% credible intervals) for three
-#| mire ecosystem condition indicators in three municipalities. The indicator values 
+#| mire ecosystem condition indicators in three Norwegian municipalities. The indicator values 
 #| are calculated uniquely for each Homogeneous Impact Area (HIA) in each 
@@
 #| categorical differences and exagerates the absolute difference between 
-#| areas. The inset it just a visual aid."
+#| areas. The inset it just a visual aid. Coordinate reference system is 
+#| EPSG 25832. Axis are in meters."
 
@@
 #| - "Distributions for three ecosystem condition indicators in the Norwegian 
-#| Municiaplities. The color gradient reflects the same as the x-axis. 
+#| municiaplities. The color gradient reflects the value of the x-axis. 
 #| The dotted vertical line represnet the threshold for what is considered 
-#| reduced ecosystem condtion (< 0.6). "
+#| reduced ecosystem condtion (< 0.6). Vertical lines under the density curves
+#| are 2.5%, 50% (the median) and 97.5% percentiles."
 
@@
 
-# Credit authorship contribution statement {.unnumbered}
+# CRediT authorship contribution statement {.unnumbered}
+
+**Anders L. Kolstad**: Conceptualization, Methodology, Software, Validation, Formal analysis, Data Curation, Writing - Original Draft, Visualization, Project administration, Funding acquisition.
+**Matthew Grainger**: ...
+**Marianne Evju**: ...
 
@@
 
+Kwaku Peprah Adjei Hanno Sandvik
+
 # Data availability {.unnumbered}
 
-{{< pagebreak >}} \# References {.unnumbered}
+This manuscript is written in Quarto, and the source file (Appendix B) also contains code for all the analyses underlying this study, including data exploration and cleaning.
+For a rendered version of the source file, with all code and calculation visible, see Appendix C. The source files and the data are also located on GitHub (https://github.com/anders-kolstad/HIAs).
+The exception is the large nature type survey data which was downloaded locally, but which is freely available (see @norwegian_environmental_agency_naturtyper_2024).
+
+{{< pagebreak >}}
+
+# References {.unnumbered}
+
+::: {#refs}
+:::

### 2024-03-02 manuscript/Manuscript.qmd
Some typos

@@
 title: Horizontal aggregation of spatially biased ecosystem condition indicators – a GIS based workflow
-subtitle: A Short Subtitle
 author:
@@
 
-# Survey coverage and other colour overlayed the municipalities
+# Survey coverage and other colour overlaid the municipalities
 muniPlot <- tm_shape(muni3) +
@@
 
-Ecosystem condition accounting is the game of compiling relevant data on the status, trends and qualities of ecosystems (i.e. nature) and communicating this in a structured format.
+Ecosystem condition accounting is the process of compiling relevant data on the status, trends and qualities of ecosystems (i.e. nature) and communicating this in a structured format.
 Its purpose is to make it easier to account for nature in policy by making the environmental costs of certain policies and practices visible to decision makers.
 As natural capital keeps declining all over the world, it is becoming increasingly urgent to make the message clear to decision makers about.
+<!--# About what? -->
+
 A statistical standard for ecosystem accounting, including ecosystem condition accounting, was developed by the UN and adopted by the UN Statistical Commission in 2021 and is called SEEA EA (@unitednations2021; System of Environmental-Economic Accounting - Ecosystem Accounting).
@@
 
-Central to ecosystem condition accounts are variables and indicators.
+Central to ecosystem condition accounts are *variables* and *indicators*.
 These are parameters chosen to reflect the central condition characteristics of the ecosystems, and that can be quantified and ideally monitored over time to reflect the status and trends in condition.
@@
 that the variable equals the upper reference value which again reflect the value of the variables under the reference condition.
-Similarly, a value of 0 mean that the variable is the worst possible state.
+Similarly, a value of 0 means that the variable is in the worst possible state.
 The reference condition needs to be defined for each Ecosystem Condition Assessment separately, but SEEA EA gives some suggestion, such as an an ecosystem with no or minimal anthropogenic disturbance.
 
-A general requirement for indicators in the SEEA EA framework is that they should give an unbiased representation of the condition inside the ecosystem assets [@czucz_selection_2021, table 1; see also @unitednations2021 §2.87] Ecosystem assets are defined as "ecological entities \[meaning areas\] about which information is sought and about which statistics are ultimately compiled [@unitednations2021].
+A general requirement for indicators in the SEEA EA framework is that they should give an unbiased representation of the condition inside the ecosystem assets [@czucz_selection_2021, table 1 <!--# refering to table 1 in Czucz 2021?
+-->; see also @unitednations2021 §2.87] Ecosystem assets are defined as "ecological entities \[meaning areas\] about which information is sought and about which statistics are ultimately compiled [@unitednations2021].
 This requirement for indicator validity means that spatially biased data are ill suited, especially if sampling intensity varies along gradient of anthropogenic pressures and hence ecosystem condition.
 SEEA EA is spatially explicit, which in practice means that indicators that are in some way sampled (i.e.
-not complete wall-to-wall data like remotely sensed imagery), the values are projected on to areas on the map, so that all areas inside the ecosystem accounting area get assigned value for that indicator.
+not complete wall-to-wall data like remotely sensed imagery), the values are projected on to areas on the map, so that all areas inside the ecosystem accounting area get assigned a value for that indicator.
 There are at least 3 general ways to achieve this complete areal coverage of indicator values:
@@
 Models can describe the general associations between the spatially sampled data and the context (e.g.
-the set of environmental variables) where it was samples, and use these relationships to predict and project indicator values to areas that where not originally sampled.
+the set of environmental variables) where it was sampled, and use these relationships to predict and project indicator values to areas that where not originally sampled.
 Depending on the data that goes into these models, they can be very reliable and make good indicators.
@@
 In this study we explore the potential for using a stratified aggregation technique to make use of spatially biased field data in ecosystem condition accounting.
-We demonstrate this technique using a generic GIS-based workflow for compiling ecosystem condition accounts that can be applied at any spatial scale, and we highlight the opportunities for local use-cases of this workflow by contrasting our findings across three neighboring municipalities in Norway.
+We demonstrate this technique using a generic GIS-based workflow for compiling ecosystem condition accounts that can be applied at any spatial scale, and we highlight the opportunities for local use-cases of this workflow by contrasting our findings across three neighbouring municipalities in Norway.
 The main question is how much generalisation can we perform on the data we have before the resulting indicator looses its practical value in local governance processes.
@@
 Because the data was strongly right skewed we used the lower limit for each frequency range to convert them into percentages.
-Variables 2-5 describe very related aspects, and are attributed to the same ecosystem condition characteristic (vegetation intactness) and so they where combined into a single indicator called antropogenic disturbance to soil and vegetation (ADSV; Figure S1).
+Variables 2-5 describe very related aspects, and are attributed to the same ecosystem condition characteristic (vegetation intactness) and so they where combined into a single indicator called anthropogenic disturbance to soil and vegetation (ADSV; Figure S1).
 This was done by summing the two variables after they had been converted to percentages.
@@
 Variable 6 is different from the preceding variables in @tbl-variables in that it includes an estimation of future effects that the observed trenches are projected to have on mire vegetation, function or structure over time.
-This is not a favorable trait in a metric used to evaluate the ecosystem condition as it is today, yet we include it here nonetheless because there is a general shortage of data on mire hydrology, which is a fundamental part of mire integrity.
+This is not a favourable trait in a metric used to evaluate the ecosystem condition as it is today, yet we include it here nonetheless because there is a general shortage of data on mire hydrology, which is a fundamental part of mire integrity.
 
@@
 A variable value of 1 indicates an intact mire, and a value of 5 indicates a mire transitioning away from a wetland.
-A value of 2 indicates observable change within the range expected for the same mapping unit, and a value of 3 indicates a mire transitioning into a neighboring (ecologically speaking) mapping unit.
+A value of 2 indicates observable change within the range expected for the same mapping unit, and a value of 3 indicates a mire transitioning into a neighbouring (ecologically speaking) mapping unit.
 See Figure S1 and @fig-workflowExample for a schematic workflow for the indicator development.
@@
 This model, which has a 90.9% precision when tested against independent field data, estimates 12.7% of the area in southern Norway is mire [@bakkestuen_delineation_2023].
-Mires are ecologically and socially important in Norway simply doe to its large extent, and due to its role in climate mitigation as mires store a large amount of carbon.
-There has not been a national assessment of the ecosystem condition of mires in Norway, but the authors recent contributed on a report which presented several new indicators that can be used in future national assessments for this ecosystem [@nybo_indikatorer_2023; see also @kolstad2023]).
+Mires are ecologically and socially important in Norway simply due to its large extent, and due to its role in climate mitigation as mires store a large amount of carbon.
+There has not been a national assessment of the ecosystem condition of mires in Norway, but the authors recently contributed to a report which presented several new indicators that can be used in future national assessments for this ecosystem [@nybo_indikatorer_2023; see also @kolstad2023]).
 The current study builds on the work in that report.
@@
 | id  | Variable code        | Variable name                                    | Measurement unit                                                                         | Description                                                                                                                                                       | Reference            |
-|--------|------------|------------|------------|-----------------|----------------|
-| 1   | 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                                                                                | @runehalvorsen2019   |
-| 2   | 7SE                  | Human caused abration or abration-caused erosion | Unitless, ordinal, non-linear scale from 1 to 4.
-| Measures the frequency of imagined 4 m^2^ quadrats layed over the area that has some sign of abration                                                             | @runehalvorsen2019   |
-| 3   | PRSL                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.
+|------------|------------|------------|------------|------------|------------|
+| 1   | 7FK                  | Prevalence of alien species                      | Unit-less, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                                                                                | @runehalvorsen2019   |
+| 2   | 7SE                  | Human caused abration or abration-caused erosion | Unit-less, ordinal, non-linear scale from 1 to 4.
+| Measures the frequency of imagined 4 m^2^ quadrats laid over the area that has some sign of abration                                                              | @runehalvorsen2019   |
+| 3   | PRSL                 | *as above*                                       | Unit-less, ordinal, non-linear scale from 0 to 7.
 | Same as 7SE, but recorded at a higher resolution                                                                                                                  | @miljødirektoratet   |
-| 4   | 7TK                  | Tracks from large vehicles                       | Unitless, ordinal, non-linear scale from 1 to 4.
-| Measures the frequency of imagined 100 m^2^ quadrats layed over the area that has some signs of vehicle tracks                                                    | @runehalvorsen2019   |
-| 5   | PRTK                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.
+| 4   | 7TK                  | Tracks from large vehicles                       | Unit-less, ordinal, non-linear scale from 1 to 4.
+| Measures the frequency of imagined 100 m^2^ quadrats laid over the area that has some signs of vehicle tracks                                                     | @runehalvorsen2019   |
+| 5   | PRTK                 | *as above*                                       | Unit-less, ordinal, non-linear scale from 0 to 7.
 | Same as 7TK, but recorded at a higher resolution                                                                                                                  | @miljødirektoratet   |
-| 6   | 7GR-GI               | Trenching intensity                              | Unitless, ordinal scale from 1 to 5                                                     | Describes the effect that drainage ditches is estimate to have on the species composition and environmental variables ones the system reached its new equilibrium | @miljødirektoratet   |
-| 7   | Infrastructure Index | Infrastructure Index                             | Unitless linear scale from 0 to 13.2                                                    | Unitelss index ranging from from 0 to 13.2                                                                                                                        | @erikstad_index_2023 |
+| 6   | 7GR-GI               | Trenching intensity                              | Unit-less, ordinal scale from 1 to 5                                                     | Describes the effect that drainage ditches is estimate to have on the species composition and environmental variables ones the system reached its new equilibrium | @miljødirektoratet   |
+| 7   | Infrastructure Index | Infrastructure Index                             | Unit-less linear scale from 0 to 13.2                                                    | Unit-less index ranging from from 0 to 13.2                                                                                                                       | @erikstad_index_2023 |
 | 8   | HIA                  | Homogeneous Impact Area                          | Ordinal, non-linear scale from 1 to 4                                                    | A categorical representation of the infrastructure index                                                                                                          | this paper           |
@@
 
-Because the nature type survey data is spatially biased, we cannot assume that they are area representative.
+Because the nature type survey data is spatially biased, we cannot assume that they are area-representative.
 In an attempt to overcome this issue we divided the area of Norway into four non-overlapping Homogeneous Impact Areas (HIAs) based on an infrastructure index for the year 2022 [@tbl-variables, @erikstad_index_2023].
-This index is a continuous variable that represnets the frequency of different infrastructure types inside 500 m radius circles around each 100×100 m pixel.
+This index is a continuous variable that represents the frequency of different infrastructure types inside 500 m radius circles around each 100×100 m pixel.
 We then categorised this continuous variable into four classes (0–3) using the following value ranges: 0 = \<1; 1 ≤ 1 \> 6; 6 ≤ 2 \>12; 3 ≥ 12.
@@
 For each HIA and municipality combination we then created a probability distribution for the area weighted mean indicator values using Bayesian updating of a uniform prior between 0 and 1, informed by the standard deviation (SD) of the indicator values in the national data set [@fig-workflowExample C; Appendix C, line XXX].
-The resulting distributions are assumed normally distributed, and we therefore simply carried the mean and SD from the posterior distributions over to individual polygons in the ecosystem delineation map, for each HIA class separately, and for each polygon we then sampled random numbers from a normal distribution with this same mean and SD.
+The resulting distributions are assumed to be normally distributed, and we therefore simply carried the mean and SD from the posterior distributions over to individual polygons in the ecosystem delineation map, for each HIA class separately, and for each polygon we then sampled random numbers from a normal distribution with this same mean and SD.
 The number of m^3^ dictated the number of samples for each polygon, thus ensuring that large polygons end up counting more towards the indicator value in the entire municipality.
@@
 
-The three indicators showed some association with the HIAs when we lookad at data from all of Norway with `r nrow(naturetypes)` individual mires.
+The three indicators showed some association with the HIAs when we looked at data from all of Norway with `r nrow(naturetypes)` individual mires.
 The indicators mostly showed a worsening of condition with increasing presence of human infrastructure (@fig-validation).
-This relationship was much stringer for the indicator Trenching, and weakest for the indicator Alien species.
-Alien species also had the highest indicator values overall, with most mirer having no alien species recorded.
+This relationship was much stronger for the indicator Trenching, and weakest for the indicator Alien species.
+Alien species also had the highest indicator values overall, with most mires having no alien species recorded.
 For Trenching, 57% of mires in HIA-3 had some trenches, whereas in HIA-0 this number was 10%.
@@
 #| municiaplities. The color gradient reflects the value of the x-axis. 
-#| The dotted vertical line represnet the threshold for what is considered 
+#| The dotted vertical line represent the threshold for what is considered 
 #| reduced ecosystem condtion (< 0.6). Vertical lines under the density curves

### 2024-03-18 R/presentation.qmd
add presentation

---
title: "Presenation"
format: revealjs
editor: visual
include-in-header: 
  text: |
    <style>
      .custom-small table {
        font-size: .5em
      }
    </style>
---

## Horizontal aggregation of spatially biased ecosystem condition indicators – a GIS based workflow

Anders L. Kolstad

Mathew Grainger

Marianne Evju

<br> *1st draft* - aimed at Ecosystem Indicators

## Question

-   Is there a way to make use of spatially biased data in ecosystem condition indicators?\

::: {.callout-note appearance="minimal"}
Ecosystem condition indicators should be wall-to-wall. For municipalities with little available data, the indicators values could be inferred from projections based on a model. If municipalities are somehow judged based on their indicator scores, it is not ideal if this is based on an assumption that all municipality are the same.
:::

-   Can we make local wall-to-wall indicator maps from sporadic data without relying on data from other municipalities?

## Workflow

```{r}
#| fig.cap: ""
knitr::include_graphics("images/aggregation-workflow.jpg")
```

## Variables

::: custom-small
| id  | Variable code        | Variable name                                    |
|---------------|---------------|-------------------------------------------|
| 1   | 7FK                  | Prevalence of alien species                      |
| 2   | 7SE                  | Human caused abration or abration-caused erosion |
| 3   | PRSL                 | *as above*                                       |
| 4   | 7TK                  | Tracks from large vehicles                       |
| 5   | PRTK                 | *as above*                                       |
| 6   | 7GR-GI               | Trenching intensity                              |
| 7   | Infrastructure Index | Infrastructure Index                             |
| 8   | HIA                  | Homogeneous Impact Area                          |
:::

## Workflow 2

![](images/workflow-01.jpg)

## Pilot municipalities

![](images/positionMap.jpg)

## Pilot municipalities

![](images/studyLocations.jpg)

## Scaling![](images/scaling-plot.jpg)

## Trenching variable

![](images/indicator-magnify.jpg)

## Validation

![](images/validation-plot.jpg)

## Generalise

![](images/forest-plot.jpg)

## Spread

![](images/spread-na-small.jpg)

## Aggregate

![](images/ridgePlot.jpg)

### 2024-04-08 manuscript/Manuscript.qmd
update methods section

@@
 
-# same for 7FK
+# same for 7FA
 scale2 <- eaTools::ea_normalise(data = naturetypes,
@@
 In this survey, selected nature types are delineated on a map (over 140 000 polygons at the end of 2023), and each locality is scored on a range of variables relevant for describing the state and quality of nature [@norwegian_environmental_agency_protocol].
-The surveys are commissioned with the goal of producing data relevant for immediate land-use decisions, and is therefore spatially biased, typically towards areas with high human impact or expected impact.
+The surveys are commissioned with the goal of producing data relevant for immediate land-use decisions, and are therefore spatially biased, typically towards areas with high human impact or expected impact.
 In addition there is a thematic and size bias in the sampling protocol.
-For examples, for the forest ecosystem, rare, endangered or calcareous forest types are delineated, whereas more common or ordinary forest types are not.
-For that reason we we focused on open mire ecosystems in Norway where the thematic bias is less severe.
-Of all possible mire types, the survey only maps the following:
-
--   Southern ombrotrophic mires \> 2500 m^2^
+For example, for the forest ecosystem, rare, endangered or calcareous forest types are delineated, whereas more common or ordinary forest types are not.
+In this study we focused on open mire ecosystems in Norway where the thematic bias is less severe.
+The survey maps the following mire types which we use in our analyses:
+
+-   Southern ombrotrophic mires (bogs) \> 2500 m^2^
 
@@
 
--   All semi-natural mires (minerotrophic fens)
-
--   Calcareous southern fens \>500 m^2^
+-   All semi-natural mires \> 1000 m^2^
+
+-   Calcareous southern fens (minerotrophic mires) \>500 m^2^
 
@@
 From the survey data set we identified six relevant variables (variable 1-6 in @tbl-variables) which we attribute to three different ecosystem condition characteristics that describe the typical behavior of open mire ecosystems in the reference condition (Figure S1).
-Variables 1 -5 were were originally recorded along binned frequency ranges @norwegian_environmental_agency_protocol .
+Variable 1, 7FA, represents the abundance of alien species, mainly plants.\
+Variables 2-5 (7SE, PRSL, 7TK and PRTK) describe very related aspects, and are attributed to the same ecosystem condition characteristic (vegetation intactness) and so they where combined into a single indicator called anthropogenic disturbance to soil and vegetation (ADSV; Figure S1).
+These variables were were originally recorded along binned frequency ranges @norwegian_environmental_agency_protocol .
 Because the data was strongly right skewed we used the lower limit for each frequency range to convert them into percentages.
-Variables 2-5 describe very related aspects, and are attributed to the same ecosystem condition characteristic (vegetation intactness) and so they where combined into a single indicator called anthropogenic disturbance to soil and vegetation (ADSV; Figure S1).
-This was done by summing the two variables after they had been converted to percentages.
+This was done by summing the four variables after they had been converted to percentages.
 This was not a perfect solution, especially since some localities only had one of the variables recorded, but we chose this, rather than for example using a *worst rule* principle, in order to better separate the localities in terms of their indicator values.
-Variable 6 is different from the preceding variables in @tbl-variables in that it includes an estimation of future effects that the observed trenches are projected to have on mire vegetation, function or structure over time.
+Variable 6, 7GR-GI, is different from the preceding variables in @tbl-variables in that it includes an estimation of future effects that the observed trenches are projected to have on mire vegetation, function or structure over time.
 This is not a favourable trait in a metric used to evaluate the ecosystem condition as it is today, yet we include it here nonetheless because there is a general shortage of data on mire hydrology, which is a fundamental part of mire integrity.
 
-To turn the remaining three variables into indicators we scaled them using three normative reference values each: an upper (best possible condition), a lower (worst possible condition), and a threshold value that defines the breaking point between good and poor condition.
-The reference values are numerical representation of the reference condition.
-We define the reference condition as one where ecosystems are subject to little or no human influence, with a climate as in the period 1961-1990 and a native species pool similar as today.
-
-For the indicators Alien species and ADSV, the lower and upper reference values were defined as 100% and 0%, respectively.
+To turn variables into indicators we scaled them using three normative reference values each: an upper (best possible condition), a lower (worst possible condition), and a threshold value that defines the breaking point between good and poor condition [@jakobsson2020].
+The reference values makes up a numerical representation of the reference condition.
+We define the reference condition as one where ecosystems are subject to little or no human influence, with a climate as in the period 1961-1990 and a native species pool similar as today [@jakobsson2021].
+
+The variable 7TK was rescaled into the indicator named *Alien species*, 7GR-GI was rescaled into the indicator *Trenching*, and ADSV was rescaled and kept the name *ADSV*.
+For Alien species and ADSV, the lower and upper reference values were defined as 100% and 0%, respectively.
 The threshold for *good ecosystem condition* was defined as 10%, which was then mapped to the value 0.6 on the rescaled indicators, thus creating a non-linear rescaling of the variable (Figure S2).
-The variable 7GR-GI was rescaled into the indicator named *Trenching* by using the lower and upper reference values 1 and 5, respectively, and a threshold value of 2.5 (Figure S2).
+For *Trenching* we used the lower and upper reference values 1 and 5, respectively, and a threshold value of 2.5 (Figure S2).
 A variable value of 1 indicates an intact mire, and a value of 5 indicates a mire transitioning away from a wetland.
@@
 This model, which has a 90.9% precision when tested against independent field data, estimates 12.7% of the area in southern Norway is mire [@bakkestuen_delineation_2023].
-Mires are ecologically and socially important in Norway simply due to its large extent, and due to its role in climate mitigation as mires store a large amount of carbon.
+Mires are ecologically and socially important in Norway simply due to their large extent, and due to their role in climate mitigation as mires store a large amount of carbon.
 There has not been a national assessment of the ecosystem condition of mires in Norway, but the authors recently contributed to a report which presented several new indicators that can be used in future national assessments for this ecosystem [@nybo_indikatorer_2023; see also @kolstad2023]).
@@
 
-We chose three municipalities in Norway to test out the indicators.
+We used the entire national data set for assessing the relationship between the HIA levels and the indicator values.
+We then sub-setted the national data set and chose three municipalities in south Norway to test out the indicators.
 The municipalities differ in several aspects, such as the amount of mire area, the total area surveyed, and the prevalence of infrastructure (@tbl-munis, @fig-studyLocation2, Figures S3, S4).
@@
 #| - "Map of the three focal municiaplities. For each municipality the maps
-#| in the top row shows ocean in blue and non-ocean in green. 
+#| in the top row show ocean in blue and non-ocean in green. 
 #| The survey coverage maps are in grey, and the mapped mires are
-#| in red, with polygon borderes made extra thich to make them visible,
-#| but then also exagerating their size.
+#| in red, with polygon borders made extra thich to make them visible,
+#| but then also exaggerating their size.
 #| The bottom row shows the delineation for homogeneous impact areas which
@@
 |------------|------------|------------|------------|------------|------------|
-| 1   | 7FK                  | Prevalence of alien species                      | Unit-less, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                                                                                | @runehalvorsen2019   |
+| 1   | 7FA                  | Prevalence of alien species                      | Unit-less, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                                                                                | @runehalvorsen2019   |
 | 2   | 7SE                  | Human caused abration or abration-caused erosion | Unit-less, ordinal, non-linear scale from 1 to 4.
@@
 
-Because the nature type survey data is spatially biased, we cannot assume that they are area-representative.
-In an attempt to overcome this issue we divided the area of Norway into four non-overlapping Homogeneous Impact Areas (HIAs) based on an infrastructure index for the year 2022 [@tbl-variables, @erikstad_index_2023].
-This index is a continuous variable that represents the frequency of different infrastructure types inside 500 m radius circles around each 100×100 m pixel.
-We then categorised this continuous variable into four classes (0–3) using the following value ranges: 0 = \<1; 1 ≤ 1 \> 6; 6 ≤ 2 \>12; 3 ≥ 12.
+Because the nature type survey data are spatially biased, we cannot assume that they are area-representative.
+In an attempt to overcome this issue we divided the area of Norway into four non-overlapping Homogeneous Impact Areas (HIAs) based on an infrastructure index for the year 2022 (@tbl-variables).
+This index is a continuous variable that represents the frequency of different infrastructure types inside 500 m radius circles around each 100×100 m pixel \[ @erikstad2023\].
+We then categorised this continuous variable into four classes (0–3) using the following value ranges: 0 = \<1; 1 ≤ 1 \> 6; 6 ≤ 2 \>12; 3 ≤ 12.
+
 The name HIAs is not a perfect representation of the information found in the infrastructure index, but we introduce this name here as a general term which is aligned with the concept of Homogeneous Ecosystem Areas, *sensu* @vallecillo_eu-wide_2022.
@@
 The relationship between the HIA classes and indicator values was examined visually.
-For each HIA and municipality combination we then created a probability distribution for the area weighted mean indicator values using Bayesian updating of a uniform prior between 0 and 1, informed by the standard deviation (SD) of the indicator values in the national data set [@fig-workflowExample C; Appendix C, line XXX].
-The resulting distributions are assumed to be normally distributed, and we therefore simply carried the mean and SD from the posterior distributions over to individual polygons in the ecosystem delineation map, for each HIA class separately, and for each polygon we then sampled random numbers from a normal distribution with this same mean and SD.
-The number of m^3^ dictated the number of samples for each polygon, thus ensuring that large polygons end up counting more towards the indicator value in the entire municipality.
-We then drew 1000 random values from this much large vector of possible indicator values and created a probability distribution for the indicators for the entire municipalities (i.e.
+For each HIA and municipality combination we then created a probability distribution for the area weighted mean of the indicator values using Bayesian updating of a uniform prior between 0 and 1, informed by the standard deviation (SD) of the indicator values in the national data set [@fig-workflowExample C; Appendix C, line XXX].
+The resulting distributions are assumed to be normally distributed, and we therefore simply carried the mean and SD from the posterior distributions over to individual polygons in the ecosystem delineation map, for each HIA class separately.
+For each polygon we then sampled random numbers from a normal distribution with this same mean and SD.
+The number of m^2^ dictated the number of samples for each polygon, thus ensuring that large polygons ended up counting more towards the indicator value in the given municipality.
+We then drew 1000 random values from this vector of possible indicator values and created a probability distribution for the indicators for the entire municipalities (i.e.
 the ecosystem accounting areas).

### 2024-04-08 manuscript/Manuscript.qmd

@@
 #| fig.cap:
-#| - "Map of the three focal municiaplities. For each municipality the maps
+#| - "Map of the three focal municipalities. For each municipality the maps
 #| in the top row show ocean in blue and non-ocean in green. 
@@
 In an attempt to overcome this issue we divided the area of Norway into four non-overlapping Homogeneous Impact Areas (HIAs) based on an infrastructure index for the year 2022 (@tbl-variables).
-This index is a continuous variable that represents the frequency of different infrastructure types inside 500 m radius circles around each 100×100 m pixel \[ @erikstad2023\].
+This index is a continuous variable that represents the frequency of different infrastructure types inside 500 m radius circles around each 100×100 m pixel [@erikstad2023].
 We then categorised this continuous variable into four classes (0–3) using the following value ranges: 0 = \<1; 1 ≤ 1 \> 6; 6 ≤ 2 \>12; 3 ≤ 12.
@@
 
+## Indicator validty and trends in Norway
+
 The three indicators showed some association with the HIAs when we looked at data from all of Norway with `r nrow(naturetypes)` individual mires.
@@
 For Trenching, 57% of mires in HIA-3 had some trenches, whereas in HIA-0 this number was 10%.
-However, only in Gran municipality was there a significant difference in the Trenching indicator between HIAs, with HIA-2 having worse condition than HIA-1 (@fig-forest).
+
+## The three focal municipalities
+
+Only in Gran municipality was there a significant difference in the Trenching indicator between HIAs, with HIA-2 having worse condition than HIA-1 (@fig-forest).
 Conversely, the ADSV indicator in Gran showed worse condition in HIA-1 compared to HIA-2.
 Nordre-Follo had considerably fewer data points compared to Gran, and especially to Nord-Aurdal who had the most data points, and this paucity of data is reflected in the wide credible intervals for all three indicators in Nordre Follo.
-The credible intervals are widest for the Trenching indicators, and this is because the national data set showed more variation in the underlying variable 7GR-GI, and this information is informing the Bayesian updating process (see @sec-methods).
-None of the three municipalities had data for all four HIAs.
-Therefore, when we transferred the mean and SD for each HIA over to the ecosystem delineation map (@fig-spread-example), some mire polygons did not get assigned any data, and were therefore dropped from any further analyses.
-At the municipality level (i.e. the ecosystem accounting area level) the three municipalities show some differentiation.
-Nordre-Follo had the highest (the best) indicator values for ADSV, but the lowest for Trenching, which was also the only instance of an indicator crossing the threshold from good to deteriorated condition (@fig-EEA-ridgePlot; Table S1).
+The credible intervals are widest for the Trenching indicator, and this is because the national data set showed more variation in the underlying variable 7GR-GI compared to the other variables 1-5 (@tbl-variables), and this information is informing the Bayesian updating process (see @sec-methods).
+Besides Nordre Follo, which do not contain any HIA-0 areas (i.e. relatovely wild areas), the three municipalities had areas in all HIA classes.
+Yet, none of the three municipalities had survey data from all HIAs.
+Therefore, when we transferred the mean and SD for each HIA (@fig-forest) over to the ecosystem delineation map (@fig-spread-example), some mire polygons in the ecosystem delineation map did not get assigned any data, and were therefore not included (i.e.
+did not have any influence on) in the following aggregation to ecosystem accounting areas (EAA).
+At the municipality level (i.e. the EAA level) the three municipalities show some differentiation.
+Nordre Follo had the highest (the best) indicator values for ADSV, but the lowest for Trenching, which was also the only instance of an indicator crossing the threshold from good to deteriorated condition (@fig-EEA-ridgePlot; Table S1).
 
@@
 #| ecosystem condition indicators along a gradient of infrastructure
-#| densities."
+#| densities. The data is from a national vegetation type survey in Norway."
 

### 2024-04-09 manuscript/Manuscript.qmd

@@
 Similarly, a value of 0 means that the variable is in the worst possible state.
-The reference condition needs to be defined for each Ecosystem Condition Assessment separately, but SEEA EA gives some suggestion, such as an an ecosystem with no or minimal anthropogenic disturbance.
-
-A general requirement for indicators in the SEEA EA framework is that they should give an unbiased representation of the condition inside the ecosystem assets [@czucz_selection_2021, table 1 <!--# refering to table 1 in Czucz 2021?
--->; see also @unitednations2021 §2.87] Ecosystem assets are defined as "ecological entities \[meaning areas\] about which information is sought and about which statistics are ultimately compiled [@unitednations2021].
-This requirement for indicator validity means that spatially biased data are ill suited, especially if sampling intensity varies along gradient of anthropogenic pressures and hence ecosystem condition.
-SEEA EA is spatially explicit, which in practice means that indicators that are in some way sampled (i.e.
-not complete wall-to-wall data like remotely sensed imagery), the values are projected on to areas on the map, so that all areas inside the ecosystem accounting area get assigned a value for that indicator.
-There are at least 3 general ways to achieve this complete areal coverage of indicator values:
-
-a.  Using wall-to-wall data (e.g. remotely sensed data)
-b.  Predict values using a model (e.g. by accounting for environmental variation)
-c.  Simple projection of some best estimate, typically a central tendency from area representative data
+This could mean that the species or ecosystem function in question is completely lost fro the ecosystem.
+The reference condition needs to be defined for each Ecosystem Condition Assessment separately, but SEEA EA gives some suggestions, such as an an ecosystem with no or minimal anthropogenic disturbance.
+
+The SEEA EA framework recommends that ecosystem condition indicators give an unbiased representation of the condition [table 1 in @czucz_selection_2021; see also @unitednations2021 §2.87], and that condition indicators are recorded uniquely for each ecosystem asset (EA) so that the condition in the ecosystem accounting area (EAA) can be inferred from an area weighted mean based on the relative sizes of the EAs [@unitednations2021 §5.54].
+Ecosystem assets are defined as "ecological entities" (meaning areas) about which information is sought and about which statistics are ultimately compiled [@unitednations2021].
+This recommendation for representativity in the indicators means that spatially biased data are ill suited, especially if sampling intensity varies along gradient of anthropogenic pressures and hence ecosystem condition.
+This recommendation for spatial resolution also means that when a condition metric is sampled coarsely, for example in an opportunistic and spatially biased field campaign, then it is very likely that you will not have a enough data points to to religably estimate an indicator value for each EA.
+This problem does not exist for complete wall-to-wall data like remotely sensed imagery where each EA will get its own unbiased indicator value.
+
+There are at least 2 ways to achieve this same complete areal coverage, that is relatively unbiased, from originally point-based indicator values:
+
+a.  Predict (project) values from a statistical model that accounts for the effect of environmental variation on the variables
+b.  Simple projection of some best estimate, for example a central tendency from a stratified selection of area representative data
 
 The need for an unbiased estimation of indicator values is unquestionable, but nonetheless, this requirement puts a large limitation on what types of data one can use.
-Ecosystem condition assessments are generally limited by data availability, and the choice of variables and indicators to include in assessments are more often or not a pragmatic and opportunistic one which is unlikely to reflect the full scope of the ecosystem condition characteristics.
+Ecosystem condition assessments are generally limited by data availability, and the choice of variables and indicators to include in assessments are more often than not a pragmatic and opportunistic one which is unlikely to reflect the full scope of the ecosystem condition characteristics.
 Note that the same is true for thematic biases.
@@
 We used an ecosystem delineation map for open mires in southern Norway, produced using remotely sensed data and a deep learning model [@bakkestuen_delineation_2023].
-This model, which has a 90.9% precision when tested against independent field data, estimates 12.7% of the area in southern Norway is mire [@bakkestuen_delineation_2023].
+This model estimates 12.7% of the area in southern Norway is mire [@bakkestuen_delineation_2023].
 Mires are ecologically and socially important in Norway simply due to their large extent, and due to their role in climate mitigation as mires store a large amount of carbon.
@@
 We then categorised this continuous variable into four classes (0–3) using the following value ranges: 0 = \<1; 1 ≤ 1 \> 6; 6 ≤ 2 \>12; 3 ≤ 12.
+These ranges were chosen as they produced a sensible and relatively linear area classification when visualised along an urban-wilderness gradient, whilst at the same time delineating sufficient areas for each range across Norway.
 

### 2024-04-09 manuscript/Manuscript.qmd

@@
 dotCOLS = c("grey90","grey70", "grey50", "grey40")
-barCOLS <- tmaptools::get_brewer_pal("Set1", n = 4)
+#barCOLS <- tmaptools::get_brewer_pal("Set1", n = 4)
+barCOLS <- c("#fff7bd", "#fecf66", "#f88b22", "#cc4c02")
 
@@
              ymin=low,
-             ymax=high,
-             col=factor(infrastructureIndex),
-             fill=factor(infrastructureIndex))) + 
+             ymax=high
+             )) + 
   geom_linerange(
+    aes(colour=factor(infrastructureIndex)),
     size=10) +
   geom_point(
+    aes(fill=factor(infrastructureIndex)),
     size=3, 
@@
   ) +
-  scale_fill_manual(values=dotCOLS)+
+  scale_fill_manual(values=rev(dotCOLS))+
   scale_color_manual(values=barCOLS)+
@@
 
-ggsave("../images/forest-plot.jpg",
-       plot=forest_plot,
-       width = 8,
-       height = 6)
+#ggsave("../images/forest-plot.jpg",
+#       plot=forest_plot,
+#       width = 8,
+#       height = 6)
 ```
@@
 # Here is the same plot but only for trenching
-
-# #define colours for dots and bars
-dotCOLS = c("grey90","grey70", "grey50", "grey40")
-barCOLS <- tmaptools::get_brewer_pal("Set1", n = 4)
 
@@
     size=10) +
+  geom_linerange(
+    size=10,
+    lwd=1,
+    colour="black") +
   geom_point(
@@
     ) +
-  scale_fill_manual(values=dotCOLS)+
+  scale_fill_manual(values=rev(dotCOLS))+
   scale_color_manual(values=barCOLS)+
@@
 
-#ggsave(plot = forest_plot_ex,
-#       "../images/forest-plot-ex.jpg",
-#       width=3,
-#       height = 3)
+ggsave(plot = forest_plot_ex,
+       "../images/forest-plot-ex.jpg",
+       width=3,
+       height = 3)
 ```

### 2024-04-09 manuscript/Appendix A.qmd
typo
some comments from david

@@
 title: Appendix A
-subtitle: For the article 'Horizontal aggregation of spatially biased ecosystem condition indicators – a GIS based workflow'
+subtitle: For the article 'Making use of spatially biased variables in ecosystem condition accounting – a GIS based workflow'
 author:
@@
 ```{r fig-muniPosition}
-#| fig.cap: "Position of the three example municiaplieies in Norway."
+#| fig.cap: "Position of the three example municipalities in Norway."
 #| out.width: '50%'

### 2024-04-09 manuscript/Manuscript.qmd

@@
 ---
-title: Horizontal aggregation of spatially biased ecosystem condition indicators – a GIS based workflow
+title: Making use of spatially biased variables in ecosystem condition accounting – a GIS based workflow
 author:
@@
 This recommendation for representativity in the indicators means that spatially biased data are ill suited, especially if sampling intensity varies along gradient of anthropogenic pressures and hence ecosystem condition.
-This recommendation for spatial resolution also means that when a condition metric is sampled coarsely, for example in an opportunistic and spatially biased field campaign, then it is very likely that you will not have a enough data points to to religably estimate an indicator value for each EA.
+This recommendation for the spatial resolution also means that when a condition metric is sampled coarsely, for example in an opportunistic and spatially biased field campaign, then it is very likely that you will not have a enough data points to to religably estimate an indicator value for each EA.
 This problem does not exist for complete wall-to-wall data like remotely sensed imagery where each EA will get its own unbiased indicator value.
 
-There are at least 2 ways to achieve this same complete areal coverage, that is relatively unbiased, from originally point-based indicator values:
+There are at least 3 ways to achieve this same complete areal coverage (i.e. spatialisation) that is relatively unbiased, from originally point-based indicator values:
 
 a.  Predict (project) values from a statistical model that accounts for the effect of environmental variation on the variables
-b.  Simple projection of some best estimate, for example a central tendency from a stratified selection of area representative data
+b.  Take a central tendency (e.g. the mean) from the sample population and project it to all EAs of a given ecosystem type.
+c.  Take a central tendency from stratified sections of the sample polulation and projection them to the EAs, adhering to the corresponding strata.
+
+This challenge if spatial generalisation is not unique to ecosystem condition accounts, and it is much discussed in the ecosystem service accounting community of practice via the related term *value transfer*[@unitednations2022; @ncaves; @grammatikopoulou2023; @barton2023].
 
@@
 For example, for the forest ecosystem, rare, endangered or calcareous forest types are delineated, whereas more common or ordinary forest types are not.
-In this study we focused on open mire ecosystems in Norway where the thematic bias is less severe.
+In this study we focused on open mire ecosystems in Norway where the thematic bias is less severe (the spatial bias being presumably equal for all ecosystems).
 The survey maps the following mire types which we use in our analyses:
@@
 These ranges were chosen as they produced a sensible and relatively linear area classification when visualised along an urban-wilderness gradient, whilst at the same time delineating sufficient areas for each range across Norway.
+We then aggregated the data to 1×1 km pixels to ease computations, and vectorized it.
 
 The name HIAs is not a perfect representation of the information found in the infrastructure index, but we introduce this name here as a general term which is aligned with the concept of Homogeneous Ecosystem Areas, *sensu* @vallecillo_eu-wide_2022.
-We plotted the HIA map over an area in Norway that we are familiar with and confirmed a sensible separation of areas from areas with no roads of major infrastructure (HIA = 0) to urban areas (HIA = 3).
-We then aggregated the data to 1×1 km pixels to ease computations, and vectorized it.
 
@@
 For each HIA and municipality combination we then created a probability distribution for the area weighted mean of the indicator values using Bayesian updating of a uniform prior between 0 and 1, informed by the standard deviation (SD) of the indicator values in the national data set [@fig-workflowExample C; Appendix C, line XXX].
+We refer to this step as the first of two *aggregation* steps (Fig. S1).
 The resulting distributions are assumed to be normally distributed, and we therefore simply carried the mean and SD from the posterior distributions over to individual polygons in the ecosystem delineation map, for each HIA class separately.
+This step is referred to as *spreading* the data (Fig. S1).
 For each polygon we then sampled random numbers from a normal distribution with this same mean and SD.
 The number of m^2^ dictated the number of samples for each polygon, thus ensuring that large polygons ended up counting more towards the indicator value in the given municipality.
-We then drew 1000 random values from this vector of possible indicator values and created a probability distribution for the indicators for the entire municipalities (i.e.
-the ecosystem accounting areas).
+We then drew 1000 random values from this vector of possible indicator values and created a probability distribution for the indicators for each municipality (the second aggregation step; Fig. S1).
 When there were no indicator values for a given HIA class, we ignored that class also in the municipal estimate.

### 2024-04-11 manuscript/Manuscript.qmd
draft two paras

@@
 | id  | Variable code        | Variable name                                    | Measurement unit                                                                         | Description                                                                                                                                                       | Reference            |
-|------------|------------|------------|------------|------------|------------|
+|----------|----------|----------|----------|------------------------|----------|
 | 1   | 7FA                  | Prevalence of alien species                      | Unit-less, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                                                                                | @runehalvorsen2019   |
@@
 
+In this paper we have demonstrated a GIS-based workflow for making use of spatially biased field data to produce representative ecosystem condition indicators at a local municipal scale.
+This approach is valuable especially for local assessments were we do not want to data from outside the EAA inform or influence our estimates for condition indicators inside the EAA.
+The method relies on stratifying the field data by Homogeneous Impact Areas (HIAs).
+Our HIA was an ordinal gradient from low to high prevalence of infrastructure, and we found a relationship between increasing infrastructure and indicator values ranging from weak in the case of the Alien species indicator, to strong, for the indicator for Trenching in mires (the ADSV indicator falling somewhere in the middle).
+This implies that the stratification was warranted in the latter case (Trenching), since field data from urban or sub-urban mires could not be said to represent the status of mires in more remote wilderness areas.
+In the case for Alien species, stratifying the data can be predicted to have less directional effect on average, but we did find a difference between HIAs 1 and 2 in Nord-Aurdal municipality, indicating that the impact from alien species in these two strata of varying infrastructure differs (@fig-forest).
+The drawback that we see from using this stratification technique when the HIA variable is not a good predictor for variation in the indicator metric, is that we end up with less data points, and therefore higher uncertainties around our indicator estimates, than we otherwise would.
+However, we do not see this as a major problem.
+Also, since our HIA areas are somewhat cohesive, forming a clumped or aggregated pattern, stratifying by it will on average lead to more local data being used for each local indicator estimate, with less geographic extrapolation.
+Therefor, we see little risk in using the stratification process also on indicators where the relationship between the indicators and the HIA are weak, although we stress that not accounting for potentially other, more relevant predictors of indicator values, would leave the indicator as spatially biased as the original variable.
+
+We have also demonstrated how Bayesian updating and resampling can be used instead of traditional arithmetic aggregation of indicators, from EAs to EAAs, to produce probability density functions for indicator values.
+A major benefit of this approach in our example was that it allowed us to present reliable uncertainty ranges for our indicator estimates even with very low sampling sizes, thus making it possible to estimate indicator values for a much greater area than if we were forced to set a minimum sample size threshold.
+An indicator estimate is of little value unless it is accompanied by a measure of uncertainty.
+The variance is commonly used, usually calcualated from the variation between spatially distant sampling points.
+However, the variance is not reliable for low sample sizes.
+For example, even if the true numerical population has a large variance, it is not impossible to sample two or three numbers that are very close to each other, and this sample would then display a low variance which is not reflective of the true variance.
+Knowing this, an often used approach is to set a sample size limit, and refrain from estimating indicator values unless there are more that a given number of data points.
+In our test municipality Nordre Follo, we had only 1-3 data points for each HIA strata, yet we were able to use these data regardless because the Bayesian updating produces a reliable, and relatively broad, unceratinty range for the estimate.
+This is an advantage, especially for local ecosystem condition assessments where data scarcity is likely to be more of an issue than in for example national assessments.
+
 # Conclusion

### 2024-04-12 manuscript/Manuscript.qmd
discussion para 3 and 4

@@
 | id  | Variable code        | Variable name                                    | Measurement unit                                                                         | Description                                                                                                                                                       | Reference            |
-|----------|----------|----------|----------|------------------------|----------|
+|-----|----------------------|--------------------------------------------------|------------------------------------------------------------------------------------------|-------------------------------------------------------------------------------------------------------------------------------------------------------------------|----------------------|
 | 1   | 7FA                  | Prevalence of alien species                      | Unit-less, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                                                                                | @runehalvorsen2019   |
@@
 | Municipality  | Total terrestrial area (km^2^)                                            | \% of terrestrial area surveyed                                            | \% open mires in relation to total terrestrial area                         | Total mire area (km^2^)                                                  | %of mire area inside survey area                                                      | Number of mire polygons in survey                                 | Mean Infrastructure Index value                                         |
-|---------|---------|---------|---------|---------|---------|---------|---------|
+|---------------|---------------------------------------------------------------------------|----------------------------------------------------------------------------|-----------------------------------------------------------------------------|--------------------------------------------------------------------------|---------------------------------------------------------------------------------------|-------------------------------------------------------------------|-------------------------------------------------------------------------|
 | Nordre Follo  | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(t_area_km)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(dk_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mirePercent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mire_km2)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mireInSurvery_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(n)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(meanHIA)` |
@@
 
+The UN standard for ecosystem accounting has been tailed towards national accounts.
+Similarly, reporting of Ecosystem Accounts to Eurostat, mandatory for all EU and EEA countries from 2026, is required for each major ecosystem type, but with to requirements for sub-national resolution.
+Assessments at national scales can make use of data sets, and hence indicators, that local assessments can not.
+For example national forest inventories or other national area representative field-based nature monitoring provide enough data points for a reliable estimates of indicator values only at large spatial scales.
+Local ecosystem accounts istead rely more on remote sensing or field measurements.
+Remote sensing holds great promise for delivering valuable information to ecosystem accounts across all scales, yet there are some ecosystem characteristics that can not easily be sensed remotes.
+This include for example species identities, and hence indicators on topics such as alien species or occurrence of threatened or certain keystone species.
+Fieldwork will therefore still be vital for collection information for ecosystem accounts also in the future.
+To reduce costs, municipalities can stratify their field sampling according to the HIAs for efficient field efforts at local scales.
+For example, Gran ad Nord-Aurdal have not recorded any field data from their most urban areas (HIA 3), and could make these areas a priority in upcoming field campaigns in order to get more complete spatial indicator coverage.
+
+We used a national data set of nature type surveys that follow a shared protocol.
+There can be other data set, that may not be compiled nationally, but which could be relevant.
+In some cases, a local ecosystem condition account is mostly interested in describing and monitoring the status within its own EAA, but great benefits may be had if accounts can also be compared between EAA, say between municipalities.
+This kind of information can give grounds for a more regional assessment, for example looking at the effectys of policies on ecosystem condition trajectories.
+One way to accomplish level of synthesis is to design and use indicators that are common across EAAs, or to compile and publish existing data in a standardised format.
+In Norway, EIAs are frequently being produced at scales, and with information that is highly relevant for local ECAs, yet the data is not standardised or available as a common resource.
+In contrast, the Biodiversity Net Gain program in England enforces developers to record and compile data on ecosystem extent and condition at very fine resolution in order to document they effect on nature and ensure a positive net gain from their activities.
+The data is made publicly available in a national data base (?) and could be used to design indicators that are informative at a municipal scale, larger than the project level scale at which they were recorded.
+The other major difference between Biodiversity Net Gain and the Norwegian EIA is the way the BNG and the work to estimate the biodiversity metric, is initiated with a wall-to-wall delineation of nature types, and relevnt ecosystem condition variables are then recorded explicitly in these, whereas
+
 # Conclusion

### 2024-04-12 manuscript/Manuscript.qmd
conclusion

--- 2024-04-09 manuscript/Manuscript.qmd
@@
 | id  | Variable code        | Variable name                                    | Measurement unit                                                                         | Description                                                                                                                                                       | Reference            |
-|------------|------------|------------|------------|------------|------------|
+|-----------|-----------|-----------|-----------|----------------|-----------|
 | 1   | 7FA                  | Prevalence of alien species                      | Unit-less, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                                                                                | @runehalvorsen2019   |
@@
 The credible intervals are widest for the Trenching indicator, and this is because the national data set showed more variation in the underlying variable 7GR-GI compared to the other variables 1-5 (@tbl-variables), and this information is informing the Bayesian updating process (see @sec-methods).
-Besides Nordre Follo, which do not contain any HIA-0 areas (i.e. relatovely wild areas), the three municipalities had areas in all HIA classes.
+Besides Nordre Follo, which do not contain any HIA-0 areas (i.e. relatively wild areas), the three municipalities had areas in all HIA classes.
 Yet, none of the three municipalities had survey data from all HIAs.
 Therefore, when we transferred the mean and SD for each HIA (@fig-forest) over to the ecosystem delineation map (@fig-spread-example), some mire polygons in the ecosystem delineation map did not get assigned any data, and were therefore not included (i.e.
-did not have any influence on) in the following aggregation to ecosystem accounting areas (EAA).
+did not have any influence on) the following aggregation to ecosystem accounting areas (EAA).
 At the municipality level (i.e. the EAA level) the three municipalities show some differentiation.
@@
 
+## The stratified GIS-based approach using HIAs
+
+In this paper we have demonstrated a generalisable GIS-based workflow for making use of spatially biased field data to produce representative ecosystem condition indicators at a local municipal scale (@fig-workflowExample).
+This approach is valuable especially for local assessments were we do not want to use data from outside the EAA inform or influence our estimates for condition indicators inside the EAA.
+The method relies on stratifying the field data by Homogeneous Impact Areas (HIAs).
+Our HIA was an ordinal gradient from low to high prevalence of infrastructure, and we found a relationship between increasing infrastructure and indicator values ranging from weak in the case of the Alien species indicator, to strong, for the indicator for Trenching in mires (the ADSV indicator falling somewhere in the middle).
+This implies that the stratification was warranted in the latter case (Trenching), i.e that field data from urban or sub-urban mires could not be said to represent the status of mires in more remote wilderness areas.
+In the case for Alien species, stratifying the variable by the HIAs can be predicted to have had a less directional effect on average, but we did find a difference between HIAs 1 and 2 in Nord-Aurdal municipality, indicating that the impact from alien species differs between these two strata of varying infrastructure (@fig-forest).
+The drawback that we see from using this stratification technique when the HIA variable is not a good predictor for variation in the indicator metric, is that we end up with less data points, and therefore higher uncertainties around our indicator estimates, than we otherwise would.
+However, we do not see this as a major problem.
+Also, since our HIA areas are somewhat cohesive, forming a clumped or aggregated pattern, stratifying by it will on average lead to more local data being used for each local indicator estimate, with less geographic extrapolation.
+Therefor, we see little risk in using the stratification process also on indicators where the relationship between the indicators and the HIA are weak, although we stress that not accounting for other, potentially more relevant predictors of indicator values, would leave the indicator as spatially biased as the original variable.
+
+## The benefits of using Bayesian updating
+
+We have also demonstrated how Bayesian updating and re-sampling can be used instead of traditional arithmetic aggregation of indicators, from EAs to EAAs, to produce probability density distributions for indicator values.
+A major benefit of this approach in our example was that it allowed us to present reliable uncertainty ranges for our indicator estimates even with very low sampling sizes, thus making it possible to estimate indicator values for a much greater areas than if we were forced to set a minimum sample size threshold.
+An indicator estimate is of little value unless it is accompanied by a measure of uncertainty.
+The variance is commonly used, usually calculated from the variation between spatially distant sampling points.
+However, the variance is not reliable for low sample sizes.
+For example, even if a numerical population has a large variance, it is possible to sample, by chance alone, two or three numbers that are very close to each other, and this sample would then display a low variance which is not reflective of the true variance.
+Knowing this, an often used approach is to set a sample size limit, and refrain from estimating indicator values unless there are more than a given number of data points.
+In our test municipality Nordre Follo, we had only 1-3 data points for each HIA strata, yet we were able to use these data regardless because the Bayesian updating produces a reliable, and relatively broad, uncertainty range for the estimate.
+This is an advantage, especially for local ecosystem condition assessments where data scarcity is likely to be more of an issue than in for example national assessments.
+
+## How to get more data for local Ecosystem Condition Accounts
+
+The UN standard for ecosystem accounting has been tailed towards national accounts [@unitednations2021].
+Similarly, reporting of Ecosystem Accounts to Eurostat, expected to become mandatory for all EU and EEA countries from 2026, is unlikely to require sub-national resolution.
+Assessments at national scales can make use of data sets, and hence indicators, that local assessments can not.
+For example national forest inventories or other national area representative field-based nature monitoring provide enough data points for a reliable estimates of indicator values only at large spatial scales.
+Local ecosystem accounts instead rely more on remote sensing or field measurements.
+Remote sensing holds great promise for delivering valuable information to ecosystem accounts across all scales, yet there are some ecosystem characteristics that can not easily be sensed remotey.
+This include for example species identities, and hence indicators on topics such as alien species or occurrence of threatened or certain keystone species.
+Fieldwork will therefore still be vital for collection information for ecosystem accounts also in the future.
+To reduce costs, municipalities can stratify their field sampling according to the HIAs for efficient field efforts at local scales.
+For example, Gran and Nord-Aurdal have not recorded any field data from their most urban areas (HIA 3), and could make these areas a priority in upcoming field campaigns in order to get more complete spatial indicator coverage.
+
+We used a national data set of nature type surveys that follow a shared protocol.
+There can be other data set, that may not be compiled nationally, but which could be relevant.
+In some cases, a local ecosystem condition account is mostly interested in describing and monitoring the status within its own EAA, but great benefits may be had if accounts can also be compared between EAA, say between municipalities.
+This kind of information can give grounds for a more regional assessment, for example looking at the effects of policies on ecosystem condition trajectories.
+To accomplish this level of synthesis it helps to be able to use indicators that are common across EAAs.
+This all depends on the right variables being collected in a similar way, for example across different projects or monitoring frameworks.
+In Norway, ecosystem impact assessments (EIA) are frequently being produced in conjunction with planned development projects.
+This data could be made very valuable for local ecosystem condition accounts, but currently there are several think stopping this.
+Firstly, the data collection is mainly on the occurrence of specific nature type or species of interest, and true condition parameters are not part of the information required by the authorities.
+In contrast, other nations have implemented EIA or EIA equivalent systems that have much greater synergistic properties towards ecosystem accounting, such as the Biodiversity Net Gain program in England\[\^1\].
+This program enforces developers to record and compile data on ecosystem extent and condition at very fine resolution in order to document they effect on nature and ensure a positive net gain from their activities.
+Relevant ecosystem condition variables are recorded for each delineated ecosytsen areas, providing an explicit link between ecosystem type and condition.
+Second, there is no standardised reporting for EIA in Norway, and no mandatory publication pathway, and therefore no way to obtain a unified data set of variables recorded across different EIAs.
+If data is made publicly available in a national data base they could be used to design indicators that are informative at a multiple scales, also larger than the project level scale at which they were recorded.
+
+\[\^1\]: <https://www.gov.uk/guidance/understanding-biodiversity-net-gain> \[accessed 12.04.2024\]
+
+## Indicator values in the three pilot municipalities
+
+The indicator values for our three indicators across the three municipalities showed in general a good condition in the mire ecosystems (@fig-EEA-ridgePlot).
+The exception was for variable Trenching in Nordre Follo, where the indicator value was 0.25 which is well below the threshold for good ecosystem condition (0.6).
+This estimate is based on just 4–5 data points (i.e. mires), and this is also reflected in the wide credible intervals \[0.04 – 0.50\].
+Nordre Follo has surveyed more of its total land area, but has less mire area to begin with (@tbl-munis) , and that is the reason for this low sample size.
+We believe this estimate is correct, in light of the wide credible intervals, and that it could reflect the true prevalence of trenching in Nordre Follo.
+Future work could test this assumption by simulating the effect of varying the sample size for an EAA where all the mires are sampled.
+
+Mire condition in Norway has not been assessed at a national scale.
+Still, we know that they are subject to multiple threats, especially land conversion into agriculture, forestry or development, and we see this reflected in the Norwegian Red List for Nature types where many ombrotrophic bog types, and rich southern minerotrophic fens are listed [@artsdatabanken2018].
+Land conversion of mires is often initiated with trenching to lower the water table, but note that our Trenching indicator doen not document the effect of trenching on old mires that are today transitioned into another ecosystem.
+This is in line with the UN standard for ecosystem accounting, where these effects should be captured by the extent accounts [@unitednations2021].
+Therefore, the negative consequences of trenching on all Norwegian mires, old and current, must be higher than what our indicator is made to show, most likely considerably higher.
+
+Alien species risk is not considered a threat to any of the Norwegian mire types , and our results confirm this, at least for the present [@fig-EEA-ridgePlot].
+Motorised vehicle traffic and other form of human transportation are also threats listed for some mire types in the red list assessment, and we see some reduction in the ADSV indicator, most so in Gran and in areas closer to infrastructure [@fig-forest].
+Still, the indicator values are consistently above the threshold value [@fig-EEA-ridgePlot].
+
 # Conclusion
+
+In conclusion, we have demonstrated a new GIS based workflow for constructing ecosystem condition indicators from spatially biased variables using a method of stratified aggregation by an ordinal gradient of Homogeneous Impact Areas (HIA).
+In addition, we show how Bayesian updating can help produce reliable uncertainty estimates for condition indicators that suffer from low sampling sizes where traditional variances would be impossible.
+Our workflow can also help guide local field efforts to in the future get better and less resource demanding local ecosystem condition accounts.
 

### 2024-04-12 manuscript/Manuscript.qmd
acknowledgements

@@
 abstract: |
-  This is the abstract.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit.
-Nam placerat turpis purus, eu tristique ex tincidunt et.
-Mauris sed augue eget turpis ultrices tincidunt.
-Sed et mi in leo porta egestas.
-Aliquam non laoreet velit.
-Nunc quis ex vitae eros aliquet auctor nec ac libero.
-Duis laoreet sapien eu mi luctus, in bibendum leo molestie.
-Sed hendrerit diam diam, ac dapibus nisl volutpat vitae.
-Aliquam bibendum varius libero, eu efficitur justo rutrum at.
-Sed at tempus elit.
+  Coming soon...
 keywords: 
@@
 
+We have no competing interests to declare.
+
 # Acknowledgements {.unnumbered}
 
-Kwaku Peprah Adjei Hanno Sandvik
+We wich to thank Kwaku Peprah Adjei for statistical help, and Hanno Sandvik for code review.
 

### 2024-04-15 manuscript/Manuscript.qmd
abstract

@@
 abstract: |
-  Coming soon...
+  Ecosystem Condition Accounts (ECA) rely on spatially representative indicators for condition.
+All ECAs are subject to data constraints in some way.
+Therefore, being able to make use of spatially biased data sets would be very valuable.
+For national ECAs, modelling aproaches can in some cases be used to control for sampling bias.
+For local ECAs however, this is often not an option as it involved spatial extrapolation using data from outside the ecosystem accounting area.
+In this study we develop three ecosystem condition indicators from the same spatially biased data set from nature type mapping of Norwegian mires.
+These indicators are Alien species, Trenching and Antropogenic Disturbance to Soil and Vegetation (ADSV).
+We test our approach in three municipalities in south Norway.
+A spatial variable was created with four steps of increasing of infrastructure prevalence that we refer to as Homogeneous Impact Areas (HIAs).
+The spatially aggregated indicator values were calculated explicity for each HIA (ie.
+a stratification).
+To facilitate religable estimation of indicator uncertainties, even in cases with very low sample sizes, we use a Bayesian updating method and produce probability distributions for the area-weighted mean indicator values in each HIA seperately.
+Then we use area-weighted resampling and produce similar propabiliy distributions for each municipality.
+With this Bayesian updating and resampling approach, small sample sizes can be compensated for by correspondingly large uncertainty ranges, as long as the full dataset is large enough to estimate the true population standard deviation.
+This paper demonstrates the use of a GIS-based workflow to control for some of the most problematic biases in an oppurtunistic field survey.
+The workflow can be used at any scale.
+By calculating indicator values for unique spatial strata, local governemnts or others can target their data aquisition towards strata with low sample sizes, and thus achieve higher cost effectiveness and ultuimately better spatial indicator coverage.
 keywords: 
@@
 
-Ecosystem condition accounting is the process of compiling relevant data on the status, trends and qualities of ecosystems (i.e. nature) and communicating this in a structured format.
+Ecosystem condition accounting is the process of compiling relevant data on the status, trends and qualities of ecosystems (i.e. nature) and communicating this in a structured format [@comte2022].
 Its purpose is to make it easier to account for nature in policy by making the environmental costs of certain policies and practices visible to decision makers.
-As natural capital keeps declining all over the world, it is becoming increasingly urgent to make the message clear to decision makers about.
-<!--# About what? -->
-
 A statistical standard for ecosystem accounting, including ecosystem condition accounting, was developed by the UN and adopted by the UN Statistical Commission in 2021 and is called SEEA EA (@unitednations2021; System of Environmental-Economic Accounting - Ecosystem Accounting).
@@
 Central to ecosystem condition accounts are *variables* and *indicators*.
-These are parameters chosen to reflect the central condition characteristics of the ecosystems, and that can be quantified and ideally monitored over time to reflect the status and trends in condition.
-Indicators are (data) variables that are normalised (rescaled) against upper and lower reference values to become bound between the values 0 and 1.
+These are metrics chosen to reflect the central condition characteristics of the ecosystems.
+These metrics are quantified and ideally monitored over time to reflect the status and trends in condition.
+Indicators are variables that are normalised (rescaled) against upper and lower reference values to become bound between the values 0 and 1.
 This normalisation ensures that indicators are more comparable because an indicator value of 1 will mean the same for all indicators, i.e.
 that the variable equals the upper reference value which again reflect the value of the variables under the reference condition.
-Similarly, a value of 0 means that the variable is in the worst possible state.
-This could mean that the species or ecosystem function in question is completely lost fro the ecosystem.
-The reference condition needs to be defined for each Ecosystem Condition Assessment separately, but SEEA EA gives some suggestions, such as an an ecosystem with no or minimal anthropogenic disturbance.
-
-The SEEA EA framework recommends that ecosystem condition indicators give an unbiased representation of the condition [table 1 in @czucz_selection_2021; see also @unitednations2021 §2.87], and that condition indicators are recorded uniquely for each ecosystem asset (EA) so that the condition in the ecosystem accounting area (EAA) can be inferred from an area weighted mean based on the relative sizes of the EAs [@unitednations2021 §5.54].
-Ecosystem assets are defined as "ecological entities" (meaning areas) about which information is sought and about which statistics are ultimately compiled [@unitednations2021].
-This recommendation for representativity in the indicators means that spatially biased data are ill suited, especially if sampling intensity varies along gradient of anthropogenic pressures and hence ecosystem condition.
-This recommendation for the spatial resolution also means that when a condition metric is sampled coarsely, for example in an opportunistic and spatially biased field campaign, then it is very likely that you will not have a enough data points to to religably estimate an indicator value for each EA.
-This problem does not exist for complete wall-to-wall data like remotely sensed imagery where each EA will get its own unbiased indicator value.
-
-There are at least 3 ways to achieve this same complete areal coverage (i.e. spatialisation) that is relatively unbiased, from originally point-based indicator values:
-
-a.  Predict (project) values from a statistical model that accounts for the effect of environmental variation on the variables
+Similarly, a value of 0 means that the variable is in the worst possible state, for example that the species or ecosystem function in question is completely lost from the ecosystem.
+The reference condition needs to be defined for each ecosystem condition assessment separately, but SEEA EA gives some suggestions, such as an an ecosystem with no or minimal anthropogenic disturbance.
+
+The SEEA EA framework recommends that ecosystem condition indicators give an unbiased representation of the condition for a given area [table 1 in @czucz_selection_2021; see also @unitednations2021 §2.87], and that condition indicators are recorded uniquely for each ecosystem asset (EA) so that the condition in the ecosystem accounting area (EAA) can be inferred from an area weighted mean based on the relative sizes of the EAs [@unitednations2021 §5.54].
+Ecosystem assets are defined as "ecological entities" (meaning areas) about which information is sought and about which statistics are compiled [@unitednations2021].
+This recommendation for representativity in the indicators means that spatially biased data are ill suited, especially if sampling intensity varies along gradients of anthropogenic pressures and hence ecosystem condition.
+This recommendation for the spatial resolution also means that when a condition metric is sampled coarsely, for example in an opportunistic and spatially biased field campaign, then it is very likely that you will not have a enough data points to to reliably estimate an indicator value for each EA.
+This problem does not exist for complete wall-to-wall data, like remotely sensed imagery, where each EA will get its own unbiased indicator value.
+There are at least 3 ways to achieve this same complete areal coverage from what is originally point-based indicator values:
+
+a.  Predict (project) indicator values from a statistical model that accounts for the effect of environmental variation on the variables
 b.  Take a central tendency (e.g. the mean) from the sample population and project it to all EAs of a given ecosystem type.
-c.  Take a central tendency from stratified sections of the sample polulation and projection them to the EAs, adhering to the corresponding strata.
+c.  Take a central tendency from a stratified sections of the sample population and projection them to the EAs, adhering to the corresponding strata.
 
@@
 The need for an unbiased estimation of indicator values is unquestionable, but nonetheless, this requirement puts a large limitation on what types of data one can use.
-Ecosystem condition assessments are generally limited by data availability, and the choice of variables and indicators to include in assessments are more often than not a pragmatic and opportunistic one which is unlikely to reflect the full scope of the ecosystem condition characteristics.
+Ecosystem condition assessments are generally limited by data availability, and the choice of variables and indicators to include in assessments are more often than not a pragmatic and opportunistic one, which is unlikely to reflect the full scope of the ecosystem condition characteristics.
 Note that the same is true for thematic biases.
-For example reflected in the scarcity of data included on insects or soil biota, even though most will agree they represent key ecosystem characteristics.
+This is for example reflected in the scarcity of data commonly included on insects or soil biota, even though most will agree they represent key ecosystem characteristics.
 Also having data from only one or a subset of nature types inside what is defined as the ecosystem in the assessment, is a typical thematic bias in ecosystem accounting.
@@
 Being able to make use of spatially biased data would greatly alleviate data shortage problems in ecosystem condition accounts.
-One way to achieve this is modelling (option b in the list above).
+One way to achieve this is modelling (option *a* in the list above).
 Models can describe the general associations between the spatially sampled data and the context (e.g.
 the set of environmental variables) where it was sampled, and use these relationships to predict and project indicator values to areas that where not originally sampled.
-Depending on the data that goes into these models, they can be very reliable and make good indicators.
-This is especially true when the ecosystem assets are large (e.g. regions or nations).
-But when they are small, like the scale of a municipality, and when the indicator is more likely to be used as the evidence base in concrete physical land use planning, then the inherent level of uncertainty from such models becomes unacceptable.
+Depending on the data that goes into these models, they can make very precice and good indicators.
+This is especially true when the EAAs are large (e.g. regions or nations).
+But when they are small, like the scale of a municipality, and when the indicator is more likely to be used as the evidence base in concrete physical land use planning, then the inherent level of uncertainty from the spatial extrapolation of such models becomes unacceptable.
 
 In this study we explore the potential for using a stratified aggregation technique to make use of spatially biased field data in ecosystem condition accounting.
-We demonstrate this technique using a generic GIS-based workflow for compiling ecosystem condition accounts that can be applied at any spatial scale, and we highlight the opportunities for local use-cases of this workflow by contrasting our findings across three neighbouring municipalities in Norway.
-The main question is how much generalisation can we perform on the data we have before the resulting indicator looses its practical value in local governance processes.
-We end by interviewing end-users from the relevant municipalities about the perceived benefits and shortcomings of our condition indicators.
+We demonstrate this technique on three indicators from the same spatially biased nature type mapping data set from Norwegian mires.
+We highlight the opportunities for local use-cases of this GIS-based workflow by contrasting our findings across three neighboring municipalities in Norway.
 
@@
 | id  | Variable code        | Variable name                                    | Measurement unit                                                                         | Description                                                                                                                                                       | Reference            |
-|-----------|-----------|-----------|-----------|----------------|-----------|
+|----------|----------|----------|----------|------------------------|----------|
 | 1   | 7FA                  | Prevalence of alien species                      | Unit-less, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                                                                                | @runehalvorsen2019   |
@@
 In conclusion, we have demonstrated a new GIS based workflow for constructing ecosystem condition indicators from spatially biased variables using a method of stratified aggregation by an ordinal gradient of Homogeneous Impact Areas (HIA).
-In addition, we show how Bayesian updating can help produce reliable uncertainty estimates for condition indicators that suffer from low sampling sizes where traditional variances would be impossible.
+In addition, we show how Bayesian updating can help produce uncertainty estimates for condition indicators that suffer from low sampling sizes where traditional variances would be unreliable.
 Our workflow can also help guide local field efforts to in the future get better and less resource demanding local ecosystem condition accounts.
@@
 
-We wich to thank Kwaku Peprah Adjei for statistical help, and Hanno Sandvik for code review.
+We wish to thank Kwaku Peprah Adjei for statistical help, and Hanno Sandvik for code review.
 

### 2024-04-17 manuscript/Appendix A.qmd
edits, refs ect

@@
 date: last-modified
+bibliography: bibliography.bib
 format:
@@
 
+| HIA class | Lower limit | Upper limit |
+|-----------|-------------|-------------|
+| 0         | 0           | \<1         |
+| 1         | 1           | \<6         |
+| 2         | 6           | \<12        |
+| 3         | 12          | 15.44       |
+: Data ranges used to categorise the Norwegian Land Use Intensity Index (the Infrastructure Index) [@erikstad2023] in the new categorical variable Homogeneous Impact Areas (HIA). {#tbl-ranges}
+
 {{< pagebreak >}}
+
 ```{r tbl-EEA}
@@
 (tbl <- readRDS("../output/EEA-tbl.RDS"))
+
 ```
+
+# References {.unnumbered}
+
+::: {#refs}
+:::

### 2024-04-17 manuscript/Manuscript.qmd
edits, refs ect

@@
 abstract: |
-  Ecosystem Condition Accounts (ECA) rely on spatially representative indicators for condition.
+  Ecosystem Condition Accounts (ECA) should reflect the integrity or quality of all nature inside the scope of the account, and therefore rely on spatially representative indicators for condition.
 All ECAs are subject to data constraints in some way.
 Therefore, being able to make use of spatially biased data sets would be very valuable.
-For national ECAs, modelling aproaches can in some cases be used to control for sampling bias.
-For local ECAs however, this is often not an option as it involved spatial extrapolation using data from outside the ecosystem accounting area.
-In this study we develop three ecosystem condition indicators from the same spatially biased data set from nature type mapping of Norwegian mires.
-These indicators are Alien species, Trenching and Antropogenic Disturbance to Soil and Vegetation (ADSV).
+For national ECAs, modelling aproaches can in some cases be used to control for sampling biases.
+For local ECAs however, like at the scale of individual municipalities of developemetn projects, this is often not an option as it involved spatial extrapolation using data from outside the ecosystem accounting area.
+In this study we develop three ecosystem condition indicators from the same spatially biased data set based on nature type mapping of Norwegian mires.
+The indicators are Alien species, Trenching and Anthropogenic Disturbance to Soil and Vegetation.
 We test our approach in three municipalities in south Norway.
 A spatial variable was created with four steps of increasing of infrastructure prevalence that we refer to as Homogeneous Impact Areas (HIAs).
-The spatially aggregated indicator values were calculated explicity for each HIA (ie.
-a stratification).
 To facilitate religable estimation of indicator uncertainties, even in cases with very low sample sizes, we use a Bayesian updating method and produce probability distributions for the area-weighted mean indicator values in each HIA seperately.
-Then we use area-weighted resampling and produce similar propabiliy distributions for each municipality.
-With this Bayesian updating and resampling approach, small sample sizes can be compensated for by correspondingly large uncertainty ranges, as long as the full dataset is large enough to estimate the true population standard deviation.
-This paper demonstrates the use of a GIS-based workflow to control for some of the most problematic biases in an oppurtunistic field survey.
-The workflow can be used at any scale.
-By calculating indicator values for unique spatial strata, local governemnts or others can target their data aquisition towards strata with low sample sizes, and thus achieve higher cost effectiveness and ultuimately better spatial indicator coverage.
+Then we use area-weighted resampling and produce indicator propabiliy distributions for each municipality.
+With this Bayesian updating and resampling approach, small sample sizes can be compensated by correspondingly large uncertainty ranges, as long as the full dataset is large enough to estimate the true population standard deviation.
+This paper demonstrates the use of a GIS-based workflow to control for some of the most problematic biases in an oppurtunistic field survey so that the data can be used for indicator in ECAs.
+The workflow can be used at any scale, including national scale.
+Because indicator values are calculatind for unique spatial strata, local governments or others can target their data aquisition towards strata with low sample sizes, and thus achieve higher cost effectiveness and ultimately better spatial indicator coverage.
 keywords: 
@@
   - disturbance
+  - ecosystem accounting
+  - ecosystem condition
+  - horizontal aggregation
   - indicator
-  - ecosystem condition
-  - wetlands
   - mire
   - peatlands
-  - ecosystem accounting
   - SEEA EA
+  - wetlands
 date: last-modified
@@
 
-# infrastructure index
+# infrastructure index (i.e. land use intensity index)
 path_infrastructure <- ifelse(dir == "C:",
@@
 A statistical standard for ecosystem accounting, including ecosystem condition accounting, was developed by the UN and adopted by the UN Statistical Commission in 2021 and is called SEEA EA (@unitednations2021; System of Environmental-Economic Accounting - Ecosystem Accounting).
-The standard, or framework, is a set of rules, principles and best practices for compiling Ecosystem accounts, mainly aimed at national accounts.
+The standard, or framework, is a set of rules, principles and best practices for compiling ecosystem accounts, mainly aimed at national accounts.
 
@@
 This normalisation ensures that indicators are more comparable because an indicator value of 1 will mean the same for all indicators, i.e.
-that the variable equals the upper reference value which again reflect the value of the variables under the reference condition.
+that the variable equals the upper reference value which again reflect the value of the variables under the reference (pristine) condition.
 Similarly, a value of 0 means that the variable is in the worst possible state, for example that the species or ecosystem function in question is completely lost from the ecosystem.
@@
 
-The SEEA EA framework recommends that ecosystem condition indicators give an unbiased representation of the condition for a given area [table 1 in @czucz_selection_2021; see also @unitednations2021 §2.87], and that condition indicators are recorded uniquely for each ecosystem asset (EA) so that the condition in the ecosystem accounting area (EAA) can be inferred from an area weighted mean based on the relative sizes of the EAs [@unitednations2021 §5.54].
+The SEEA EA framework recommends that ecosystem condition indicators give an unbiased representation of the condition for a given area [table 1 in @czucz_selection_2021; see also @unitednations2021 §2.87], and that condition indicators are recorded uniquely for each ecosystem asset (EA) so that the condition in the ecosystem accounting area (EAA) can be inferred from an area weighted mean based on the relative extents of the EAs [@unitednations2021 §5.54].
 Ecosystem assets are defined as "ecological entities" (meaning areas) about which information is sought and about which statistics are compiled [@unitednations2021].
-This recommendation for representativity in the indicators means that spatially biased data are ill suited, especially if sampling intensity varies along gradients of anthropogenic pressures and hence ecosystem condition.
+This recommendation for representativity in the indicators means that spatially biased data are ill suited, especially if sampling intensity varies systematically along gradients of anthropogenic pressures and hence ecosystem condition.
 This recommendation for the spatial resolution also means that when a condition metric is sampled coarsely, for example in an opportunistic and spatially biased field campaign, then it is very likely that you will not have a enough data points to to reliably estimate an indicator value for each EA.
-This problem does not exist for complete wall-to-wall data, like remotely sensed imagery, where each EA will get its own unbiased indicator value.
-There are at least 3 ways to achieve this same complete areal coverage from what is originally point-based indicator values:
+This problem does not exist for complete wall-to-wall data, like remotely sensed imagery, where each EA will have ebough data to estaimte its unique indicator value.
+This challenge if spatial generalisation is not unique to ecosystem condition accounts, and it is much discussed in the ecosystem service accounting community of practice via the related term *value transfer*[@unitednations2022; @ncaves; @grammatikopoulou2023; @barton2023].
+There are at least three ways to achieve this same complete areal coverage from indicator values that originally are spatially incomplete:
 
@@
 b.  Take a central tendency (e.g. the mean) from the sample population and project it to all EAs of a given ecosystem type.
-c.  Take a central tendency from a stratified sections of the sample population and projection them to the EAs, adhering to the corresponding strata.
-
-This challenge if spatial generalisation is not unique to ecosystem condition accounts, and it is much discussed in the ecosystem service accounting community of practice via the related term *value transfer*[@unitednations2022; @ncaves; @grammatikopoulou2023; @barton2023].
+c.  Take a central tendency from stratified sections of the sample population and projection them to the EAs, adhering to the corresponding strata.
 
@@
 One way to achieve this is modelling (option *a* in the list above).
-Models can describe the general associations between the spatially sampled data and the context (e.g.
-the set of environmental variables) where it was sampled, and use these relationships to predict and project indicator values to areas that where not originally sampled.
+Statistical models can describe the general associations between the sampled data and the context where it was sampled (e.g.
+a set of environmental variables), and use these relationships to predict and project indicator values to areas that where not originally sampled.
 Depending on the data that goes into these models, they can make very precice and good indicators.
@@
 
--   Calcareous southern fens (minerotrophic mires) \>500 m^2^
-
--   Calcareous northern fens \>1000 m^2^
+-   Calcareous southern fens (minerotrophic mires) \> 500 m^2^
+
+-   Calcareous northern fens \> 1000 m^2^
 
@@
 We included data from 2018 to 2023.
-In this paper we assume the survey is representative for the entire mire ecosystem in Norway.
-Although it is possible that smaller or less calcareous mires will score systematically different than the ones that are surveyed we do not think this is so much the case for our variables (see below).
+In this paper we assume the survey is representative for the entire mire ecosystem in Norway (i.e. that there is no thematic bias).
+Although it is possible that smaller or less calcareous mires will score systematically different than the ones that are surveyed, we do not think this is so much the case for our variables (see below).
 However, we do assume that alien plants are slightly less common on bogs relative to fens, and therefore that this variable will be biased since that variable for only recorded in fens.
@@
 From the survey data set we identified six relevant variables (variable 1-6 in @tbl-variables) which we attribute to three different ecosystem condition characteristics that describe the typical behavior of open mire ecosystems in the reference condition (Figure S1).
-Variable 1, 7FA, represents the abundance of alien species, mainly plants.\
+Variable 1, 7FA, represents the abundance of alien species, mainly plants.
 Variables 2-5 (7SE, PRSL, 7TK and PRTK) describe very related aspects, and are attributed to the same ecosystem condition characteristic (vegetation intactness) and so they where combined into a single indicator called anthropogenic disturbance to soil and vegetation (ADSV; Figure S1).
@@
 
-The variable 7TK was rescaled into the indicator named *Alien species*, 7GR-GI was rescaled into the indicator *Trenching*, and ADSV was rescaled and kept the name *ADSV*.
-For Alien species and ADSV, the lower and upper reference values were defined as 100% and 0%, respectively.
+The variable 7TK was rescaled into the indicator named *Alien species*, 7GR-GI was rescaled into the indicator *Trenching*, and ADSV was rescaled and kept the name *ADSV* (Figure S1).
+For *Alien species* and *ADSV*, the lower and upper reference values were defined as 100% and 0%, respectively.
 The threshold for *good ecosystem condition* was defined as 10%, which was then mapped to the value 0.6 on the rescaled indicators, thus creating a non-linear rescaling of the variable (Figure S2).
@@
 A value of 2 indicates observable change within the range expected for the same mapping unit, and a value of 3 indicates a mire transitioning into a neighbouring (ecologically speaking) mapping unit.
-See Figure S1 and @fig-workflowExample for a schematic workflow for the indicator development.
-
-According to the SEEA ECT, *Alien species* can be attributed to ECT class B1 - Compositional state characteristics and *ADSV* and *Trenching* attributed to ECT class A1 - Physical state characteristics [@czucz_common_2021].
+See @fig-workflowExample and S1 for schematic workflows for the indicator development.
+*Alien species* was attributed to Ecosystem Condition Typology (ECT) class B1 - Compositional state characteristics and *ADSV,* and *Trenching* was attributed to ECT class A1 - Physical state characteristics [@czucz_common_2021].
 
@@
 The current study builds on the work in that report.
-
-We used the entire national data set for assessing the relationship between the HIA levels and the indicator values.
-We then sub-setted the national data set and chose three municipalities in south Norway to test out the indicators.
-The municipalities differ in several aspects, such as the amount of mire area, the total area surveyed, and the prevalence of infrastructure (@tbl-munis, @fig-studyLocation2, Figures S3, S4).
 
@@
 | id  | Variable code        | Variable name                                    | Measurement unit                                                                         | Description                                                                                                                                                       | Reference            |
-|----------|----------|----------|----------|------------------------|----------|
+|------------|------------|------------|------------|------------|------------|
 | 1   | 7FA                  | Prevalence of alien species                      | Unit-less, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                                                                                | @runehalvorsen2019   |
@@
 This index is a continuous variable that represents the frequency of different infrastructure types inside 500 m radius circles around each 100×100 m pixel [@erikstad2023].
-We then categorised this continuous variable into four classes (0–3) using the following value ranges: 0 = \<1; 1 ≤ 1 \> 6; 6 ≤ 2 \>12; 3 ≤ 12.
-These ranges were chosen as they produced a sensible and relatively linear area classification when visualised along an urban-wilderness gradient, whilst at the same time delineating sufficient areas for each range across Norway.
+We then categorised this continuous variable into four classes (0–3) using ranges chosen to produce a sensible and relatively linear area classification when visualised along an urban-wilderness gradient, whilst at the same time delineating sufficient areas for each strata across Norway (Table S1).
 We then aggregated the data to 1×1 km pixels to ease computations, and vectorized it.
-
 The name HIAs is not a perfect representation of the information found in the infrastructure index, but we introduce this name here as a general term which is aligned with the concept of Homogeneous Ecosystem Areas, *sensu* @vallecillo_eu-wide_2022.
 
+We used the entire national data set for assessing the relationship between the HIA levels and the indicator values.
+We then sub-setted the national data set and chose three municipalities in south Norway to test out the indicators.
+The municipalities differ in several aspects, such as the amount of mire area, the total area surveyed, and the prevalence of infrastructure (@tbl-munis, @fig-studyLocation2, S3, S4).
 We then took the nature type polygons with the indicator values and intersected with the HIA map and a map of municipal outlines in Norway.
 The relationship between the HIA classes and indicator values was examined visually.
-For each HIA and municipality combination we then created a probability distribution for the area weighted mean of the indicator values using Bayesian updating of a uniform prior between 0 and 1, informed by the standard deviation (SD) of the indicator values in the national data set [@fig-workflowExample C; Appendix C, line XXX].
-We refer to this step as the first of two *aggregation* steps (Fig. S1).
+For each HIA and municipality combination we then created a probability distribution for the area weighted mean of the indicator values using Bayesian updating of a uniform prior between 0 and 1, informed by the standard deviation (SD) of the indicator values in the national data set (@fig-workflowExample C; Appendix C, line 138 <!--# update line number manually when re-rendering appendix C -->).
+We refer to this step as the first of two *aggregation* steps (Figure S1).
 The resulting distributions are assumed to be normally distributed, and we therefore simply carried the mean and SD from the posterior distributions over to individual polygons in the ecosystem delineation map, for each HIA class separately.
-This step is referred to as *spreading* the data (Fig. S1).
+This step is referred to as *spreading* the data (Figure S1).
 For each polygon we then sampled random numbers from a normal distribution with this same mean and SD.
 The number of m^2^ dictated the number of samples for each polygon, thus ensuring that large polygons ended up counting more towards the indicator value in the given municipality.
-We then drew 1000 random values from this vector of possible indicator values and created a probability distribution for the indicators for each municipality (the second aggregation step; Fig. S1).
+We then drew 1000 random values from this vector of possible indicator values and created a probability distribution for the indicators for each municipality (the second aggregation step; Figure S1).
 When there were no indicator values for a given HIA class, we ignored that class also in the municipal estimate.
 
+All the data preparation and analyses were done in RStudio [@positteam2024] and R version 4.3.0 [@rcoreteam2023], relying heavily on the packages tidyverse [@wickham2019] , sf [@pebesma2018] and stars [@pebesma2023].
+
 # Results
 
-## Indicator validty and trends in Norway
+## Indicator validity and trends in Norway
 
@@
 The indicators mostly showed a worsening of condition with increasing presence of human infrastructure (@fig-validation).
-This relationship was much stronger for the indicator Trenching, and weakest for the indicator Alien species.
-Alien species also had the highest indicator values overall, with most mires having no alien species recorded.
-For Trenching, 57% of mires in HIA-3 had some trenches, whereas in HIA-0 this number was 10%.
+This relationship was much stronger for the indicator *Trenching*, and weakest for the indicator *Alien species*.
+*Alien species* also had the highest indicator values overall, with most mires having no alien species recorded.
+For *Trenching*, 57% of mires in HIA-3 had some trenches, whereas in HIA-0 this number was 10%.
 
@@
 
-Only in Gran municipality was there a significant difference in the Trenching indicator between HIAs, with HIA-2 having worse condition than HIA-1 (@fig-forest).
-Conversely, the ADSV indicator in Gran showed worse condition in HIA-1 compared to HIA-2.
+Only in Gran municipality was there a significant difference in the *Trenching* indicator between HIAs, with HIA-2 having worse condition than HIA-1 (@fig-forest).
+Conversely, the *ADSV* indicator in Gran showed worse condition in HIA-1 compared to HIA-2.
 Nordre-Follo had considerably fewer data points compared to Gran, and especially to Nord-Aurdal who had the most data points, and this paucity of data is reflected in the wide credible intervals for all three indicators in Nordre Follo.
-The credible intervals are widest for the Trenching indicator, and this is because the national data set showed more variation in the underlying variable 7GR-GI compared to the other variables 1-5 (@tbl-variables), and this information is informing the Bayesian updating process (see @sec-methods).
+The credible intervals are widest for the *Trenching* indicator, and this is because the national data set showed more variation in the underlying variable 7GR-GI compared to the other variables 1-5 (@tbl-variables), and this information is informing the Bayesian updating process (see @sec-methods).
 Besides Nordre Follo, which do not contain any HIA-0 areas (i.e. relatively wild areas), the three municipalities had areas in all HIA classes.
@@
 Therefore, when we transferred the mean and SD for each HIA (@fig-forest) over to the ecosystem delineation map (@fig-spread-example), some mire polygons in the ecosystem delineation map did not get assigned any data, and were therefore not included (i.e.
-did not have any influence on) the following aggregation to ecosystem accounting areas (EAA).
+did not have any influence on) the following aggregation to ecosystem accounting areas (EAA) (@fig-workflowExample, E) .
 At the municipality level (i.e. the EAA level) the three municipalities show some differentiation.
-Nordre Follo had the highest (the best) indicator values for ADSV, but the lowest for Trenching, which was also the only instance of an indicator crossing the threshold from good to deteriorated condition (@fig-EEA-ridgePlot; Table S1).
+Nordre Follo had the highest (the best) indicator values for ADSV, but the lowest for Trenching, which was also the only instance of an indicator crossing the threshold from good to deteriorated condition (@fig-EEA-ridgePlot; Table S2).
 
@@
 #| fig.cap:
-#| - "Simplified schematic showing the workflow (Figures A-E) for horizontal aggregation of a spatially biased ecosystem condition indicator.
-Figure A shows a spatially explicit indicator.
+#| - "Simplified schematic showing the workflow (Panes A-E) for spatial generalisation of a spatially biased ecosystem condition indicator.
+Pane A shows a spatially explicit, patchy and spatially biased indicator.
+The outline is the ecosystem accounting area (EEA).
+Pane B shows the location of four homogeneous impact areas (HIAs) inside the EEA.
+The indicator values in Pane A are used, in combination with the HIA map in Pane B, to update a uniform prior and produce a posterior probability distribution for the mean area woeghted indicator value for each HIA (Pane C; here simplified to only show the mean (circles) and 95% credible intervals (colored bands)).
+The bayesian updating is informed by the standard deviation for the indicator in a much bigger national data set.
+The width of the posterior distribution responds both to variation in the data and to the sample size, giving  a realistic measure for the uncertainty around the indicator even with a singe observation.
+The distributions in Pane C are assumed normally distributed.
+Note that because in this example there were no indicator data for HIA-3, indicator values are only aggregated for HIAs 0–2.
+In Pane D the mean and SD from the posterior distributions are transferred to individual polygons in an ecosystem delineation map, for each HIA seperately.
 The colors indicator different indicator values (highlighted in the inset).
-The outline is the ecosystem accounting area (EEA).
-Figure B shows the location of four homogeneous impact areas (HIAs) inside the EEA.
-The indicator values in Figure A are used, in combination with the HIA map in Figure B, to update a uniform prior and produce a posterior probability distribution for the mean indicator value for each HIA (Figure C; here simplified to only show the mean (circles) and 95% credible intervals (colored bands)).
-The bayesian updating is informed by the standard deviation for the indicator in a much bigger national data set.
-Unlike an arithmetic approach, the width of the posterior distribution responds both to variation in the data and to the sample size, giving  a religable measure for the uncertainty around the indicator even with a singe observation.
-The distributions in Figure C are assumed normally distributed.
-Note that because in this example there were no indicator data for HIA-3, indicator values are only aggregated for HIAs 0–2.
-In Figure D the mean and SD from the posterior distributions are transferred to individual polygons in an ecosystem delineation map, for each HIA seperately.
 For each ecosystem occurence (i.e.
-ecosystem assets) in Figure D, we draw one random value for each square meter of area from a normal distribution with the mean and SD that is associated with that polygon.
-In Figure E we randomly sample 1000 values from the entire vector of samples in the previous step, and we get an area weighted probability distribution for the mean indicator value for the EEA."
+ecosystem assets) in Pane D, we draw one random value for each square meter of area from a normal distribution with the mean and SD that is associated with that polygon.
+In Pane E we have randomly sampled 1000 values from the entire vector of samples in the previous step, and we get an area weighted probability distribution for the mean indicator value for the EEA."
 
@@
 #| - "Proportion of localities with different indicator values for three 
-#| ecosystem condition indicators along a gradient of infrastructure
-#| densities. The data is from a national vegetation type survey in Norway."
+#| ecosystem condition indicators along an ordinal gradient of increasing
+#| infrastructure densities (Homogeneous Impact Areas 0–3). 
+#| The data is from a national vegetation type survey in Norway."
 
@@
 #| - "Indicator values (circles = mean; bars = 95% credible intervals) for three
-#| mire ecosystem condition indicators in three Norwegian municipalities. The indicator values 
+#| mire ecosystem condition indicators (rows) 
+#| in three Norwegian municipalities (columns).
+#| The indicator values 
 #| are calculated uniquely for each Homogeneous Impact Area (HIA) in each 
@@
 
+# Histogram smoothing makes it look like there are more data points beyong the lower and upper percentiles than there really is. The 2.5% percentile sometimes looks like the 25% percentile.
+
 knitr::include_graphics("../images/ridgePlot.jpg")
@@
 In this paper we have demonstrated a generalisable GIS-based workflow for making use of spatially biased field data to produce representative ecosystem condition indicators at a local municipal scale (@fig-workflowExample).
-This approach is valuable especially for local assessments were we do not want to use data from outside the EAA inform or influence our estimates for condition indicators inside the EAA.
+This approach is useful especially for local assessments were we do not want to use data from outside the EAA inform or influence our estimates for condition indicators inside the EAA.
 The method relies on stratifying the field data by Homogeneous Impact Areas (HIAs).
-Our HIA was an ordinal gradient from low to high prevalence of infrastructure, and we found a relationship between increasing infrastructure and indicator values ranging from weak in the case of the Alien species indicator, to strong, for the indicator for Trenching in mires (the ADSV indicator falling somewhere in the middle).
-This implies that the stratification was warranted in the latter case (Trenching), i.e that field data from urban or sub-urban mires could not be said to represent the status of mires in more remote wilderness areas.
-In the case for Alien species, stratifying the variable by the HIAs can be predicted to have had a less directional effect on average, but we did find a difference between HIAs 1 and 2 in Nord-Aurdal municipality, indicating that the impact from alien species differs between these two strata of varying infrastructure (@fig-forest).
-The drawback that we see from using this stratification technique when the HIA variable is not a good predictor for variation in the indicator metric, is that we end up with less data points, and therefore higher uncertainties around our indicator estimates, than we otherwise would.
+Our HIA was an ordinal gradient from low to high prevalence of infrastructure, and we found a relationship of increasing infrastructure and decreasing indicator values ranging from weak, in the case of the *Alien species* indicator, to strong, for the indicator for *Trenching* in mires (the *ADSV* indicator falling somewhere in the middle) (@fig-validation) .
+This implies that the stratification was warranted in the latter case (*Trenching*), i.e that field data from urban or sub-urban mires could not be said to represent the status of mires in more remote wilderness areas.
+In the case for *Alien species*, stratifying the variable by the HIAs can be predicted to have had a less directional effect on average, but we did find a difference between HIA-1 and 2 in Nord-Aurdal municipality, indicating that the impact from alien species differs between these two strata of varying infrastructure (@fig-forest).
+The drawback that we see from using this stratification technique when the HIA variable is not a good predictor for variation in the indicator, is that we end up with less data points, and therefore higher uncertainties around our indicator estimates, than we otherwise would.
 However, we do not see this as a major problem.
-Also, since our HIA areas are somewhat cohesive, forming a clumped or aggregated pattern, stratifying by it will on average lead to more local data being used for each local indicator estimate, with less geographic extrapolation.
+Also, since our HIA areas are somewhat cohesive, forming a clumped or aggregated pattern, stratifying by it will on average lead to relatively more local data being used for each local indicator estimate, with less geographic extrapolation.
 Therefor, we see little risk in using the stratification process also on indicators where the relationship between the indicators and the HIA are weak, although we stress that not accounting for other, potentially more relevant predictors of indicator values, would leave the indicator as spatially biased as the original variable.
@@
 However, the variance is not reliable for low sample sizes.
-For example, even if a numerical population has a large variance, it is possible to sample, by chance alone, two or three numbers that are very close to each other, and this sample would then display a low variance which is not reflective of the true variance.
+For example, even if a numerical population has a large variance, it is still possible to sample, by chance alone, two or three numbers that are very close to each other, and this sample would then display a low variance which is not reflective of the true variance.
 Knowing this, an often used approach is to set a sample size limit, and refrain from estimating indicator values unless there are more than a given number of data points.
@@
 Assessments at national scales can make use of data sets, and hence indicators, that local assessments can not.
-For example national forest inventories or other national area representative field-based nature monitoring provide enough data points for a reliable estimates of indicator values only at large spatial scales.
+For example national forest inventories or other national area representative field-based nature monitoring provide enough data points for a reliable estimates of indicator values only at relatively large spatial scales.
 Local ecosystem accounts instead rely more on remote sensing or field measurements.
@@
 Fieldwork will therefore still be vital for collection information for ecosystem accounts also in the future.
+However, field work is labour intensive and expensive.
 To reduce costs, municipalities can stratify their field sampling according to the HIAs for efficient field efforts at local scales.
@@
 This data could be made very valuable for local ecosystem condition accounts, but currently there are several think stopping this.
-Firstly, the data collection is mainly on the occurrence of specific nature type or species of interest, and true condition parameters are not part of the information required by the authorities.
-In contrast, other nations have implemented EIA or EIA equivalent systems that have much greater synergistic properties towards ecosystem accounting, such as the Biodiversity Net Gain program in England\[\^1\].
-This program enforces developers to record and compile data on ecosystem extent and condition at very fine resolution in order to document they effect on nature and ensure a positive net gain from their activities.
-Relevant ecosystem condition variables are recorded for each delineated ecosytsen areas, providing an explicit link between ecosystem type and condition.
-Second, there is no standardised reporting for EIA in Norway, and no mandatory publication pathway, and therefore no way to obtain a unified data set of variables recorded across different EIAs.
+Firstly, there is no standardised reporting for EIA in Norway, and no mandatory publication pathway, and therefore no way to obtain a unified data set of variables recorded across different EIAs.
 If data is made publicly available in a national data base they could be used to design indicators that are informative at a multiple scales, also larger than the project level scale at which they were recorded.
-
-\[\^1\]: <https://www.gov.uk/guidance/understanding-biodiversity-net-gain> \[accessed 12.04.2024\]
+Second, the data collection is mainly on the occurrence of specific nature type or species of interest, and true condition parameters are not part of the information required by the authorities.
+In contrast, other nations have implemented EIA or EIA equivalent systems that have much greater synergistic properties towards ecosystem accounting, such as the Biodiversity Net Gain program in England [@departmentforenvironmentfoodruralaffairs].
+This program enforces developers to record and compile data on ecosystem extent and condition at very fine resolution in order to document their effect on nature and ensure a positive net gain from their activities.
+Relevant ecosystem condition variables are recorded for each delineated ecosystem areas, providing an explicit link between ecosystem type and condition.
 
@@
 The indicator values for our three indicators across the three municipalities showed in general a good condition in the mire ecosystems (@fig-EEA-ridgePlot).
-The exception was for variable Trenching in Nordre Follo, where the indicator value was 0.25 which is well below the threshold for good ecosystem condition (0.6).
+The exception was for variable *Trenching* in Nordre Follo, where the indicator value was 0.25; well below the threshold for good ecosystem condition (0.6).
 This estimate is based on just 4–5 data points (i.e. mires), and this is also reflected in the wide credible intervals \[0.04 – 0.50\].
-Nordre Follo has surveyed more of its total land area, but has less mire area to begin with (@tbl-munis) , and that is the reason for this low sample size.
-We believe this estimate is correct, in light of the wide credible intervals, and that it could reflect the true prevalence of trenching in Nordre Follo.
+Nordre Follo has surveyed more of its total land area compared to the other two municipalities, but has less mire area to begin with (@tbl-munis) , and that is the reason for this low sample size.
+We believe this indicator estimate is trustworthy, in light of the wide credible intervals, and that it could reflect the true prevalence of trenching in Nordre Follo.
 Future work could test this assumption by simulating the effect of varying the sample size for an EAA where all the mires are sampled.
@@
 Still, we know that they are subject to multiple threats, especially land conversion into agriculture, forestry or development, and we see this reflected in the Norwegian Red List for Nature types where many ombrotrophic bog types, and rich southern minerotrophic fens are listed [@artsdatabanken2018].
-Land conversion of mires is often initiated with trenching to lower the water table, but note that our Trenching indicator doen not document the effect of trenching on old mires that are today transitioned into another ecosystem.
+Land conversion of mires is often initiated with trenching to lower the water table, but note that our *Trenching* indicator does not document the effect of trenching on old mires that are today transitioned into another ecosystem.
 This is in line with the UN standard for ecosystem accounting, where these effects should be captured by the extent accounts [@unitednations2021].
@@
 
-Alien species risk is not considered a threat to any of the Norwegian mire types , and our results confirm this, at least for the present [@fig-EEA-ridgePlot].
-Motorised vehicle traffic and other form of human transportation are also threats listed for some mire types in the red list assessment, and we see some reduction in the ADSV indicator, most so in Gran and in areas closer to infrastructure [@fig-forest].
-Still, the indicator values are consistently above the threshold value [@fig-EEA-ridgePlot].
+Alien species risk is not considered a threat to any of the Norwegian mire types , and our results confirm this, at least for the present (@fig-EEA-ridgePlot).
+Motorised vehicle traffic and other forms of human transportation are also threats listed for some mire types in the red list assessment, and we see some reduction in the ADSV indicator, most so in Gran and in areas closer to infrastructure (@fig-forest).
+Still, the indicator values are consistently above the threshold value (@fig-EEA-ridgePlot).
 
@@
 This manuscript is written in Quarto, and the source file (Appendix B) also contains code for all the analyses underlying this study, including data exploration and cleaning.
-For a rendered version of the source file, with all code and calculation visible, see Appendix C. The source files and the data are also located on GitHub (https://github.com/anders-kolstad/HIAs).
+For a rendered version of the source file, with all code and calculation visible, see Appendix C.
+The source files and the data are also located on GitHub (https://github.com/anders-kolstad/HIAs) and will be archived on Zenodo with a stable doi at time of submission.
 The exception is the large nature type survey data which was downloaded locally, but which is freely available (see @norwegian_environmental_agency_naturtyper_2024).

### 2024-05-13 manuscript/Manuscript.qmd
Update Manuscript.qmd
Mariannes endringsforslag

@@
 For national ECAs, modelling aproaches can in some cases be used to control for sampling biases.
-For local ECAs however, like at the scale of individual municipalities of developemetn projects, this is often not an option as it involved spatial extrapolation using data from outside the ecosystem accounting area.
+For local ECAs however, like at the scale of individual municipalities of developement projects, this is often not an option as it involved spatial extrapolation using data from outside the ecosystem accounting area.
 In this study we develop three ecosystem condition indicators from the same spatially biased data set based on nature type mapping of Norwegian mires.
@@
 A spatial variable was created with four steps of increasing of infrastructure prevalence that we refer to as Homogeneous Impact Areas (HIAs).
-To facilitate religable estimation of indicator uncertainties, even in cases with very low sample sizes, we use a Bayesian updating method and produce probability distributions for the area-weighted mean indicator values in each HIA seperately.
+To facilitate reliable estimation of indicator uncertainties, even in cases with very low sample sizes, we use a Bayesian updating method and produce probability distributions for the area-weighted mean indicator values in each HIA seperately.
 Then we use area-weighted resampling and produce indicator propabiliy distributions for each municipality.
 With this Bayesian updating and resampling approach, small sample sizes can be compensated by correspondingly large uncertainty ranges, as long as the full dataset is large enough to estimate the true population standard deviation.
-This paper demonstrates the use of a GIS-based workflow to control for some of the most problematic biases in an oppurtunistic field survey so that the data can be used for indicator in ECAs.
+This paper demonstrates the use of a GIS-based workflow to control for some of the most problematic biases in an oppurtunistic field survey so that the data can be used for indicators in ECAs.
 The workflow can be used at any scale, including national scale.
@@
 
-The SEEA EA framework recommends that ecosystem condition indicators give an unbiased representation of the condition for a given area [table 1 in @czucz_selection_2021; see also @unitednations2021 §2.87], and that condition indicators are recorded uniquely for each ecosystem asset (EA) so that the condition in the ecosystem accounting area (EAA) can be inferred from an area weighted mean based on the relative extents of the EAs [@unitednations2021 §5.54].
-Ecosystem assets are defined as "ecological entities" (meaning areas) about which information is sought and about which statistics are compiled [@unitednations2021].
+The SEEA EA framework recommends that ecosystem condition indicators give an unbiased representation of the condition for a given area [table 1 in @czucz_selection_2021; see also @unitednations2021 §2.87], and that condition indicators are recorded uniquely for each ecosystem asset (EA) so that the condition in the ecosystem accounting area (EAA) can be inferred from an area weighted mean based on the relative extents of the ecosystem assets [@unitednations2021 §5.54].
+Ecosystem assets (EAs) are defined as "ecological entities" (meaning areas) about which information is sought and about which statistics are compiled [@unitednations2021].
 This recommendation for representativity in the indicators means that spatially biased data are ill suited, especially if sampling intensity varies systematically along gradients of anthropogenic pressures and hence ecosystem condition.
-This recommendation for the spatial resolution also means that when a condition metric is sampled coarsely, for example in an opportunistic and spatially biased field campaign, then it is very likely that you will not have a enough data points to to reliably estimate an indicator value for each EA.
+The recommendation for spatial resolution also means that when a condition metric is sampled coarsely, for example in an opportunistic and spatially biased field campaign, then it is very likely that you will not have a enough data points to reliably estimate an indicator value for each EA.
 This problem does not exist for complete wall-to-wall data, like remotely sensed imagery, where each EA will have ebough data to estaimte its unique indicator value.
-This challenge if spatial generalisation is not unique to ecosystem condition accounts, and it is much discussed in the ecosystem service accounting community of practice via the related term *value transfer*[@unitednations2022; @ncaves; @grammatikopoulou2023; @barton2023].
+This challenge if spatial generalisation is not unique to ecosystem condition accounts, and it is much discussed in the ecosystem service accounting community of practice via the related term *value transfer* [@unitednations2022; @ncaves; @grammatikopoulou2023; @barton2023].
 There are at least three ways to achieve this same complete areal coverage from indicator values that originally are spatially incomplete:
@@
 
-The need for an unbiased estimation of indicator values is unquestionable, but nonetheless, this requirement puts a large limitation on what types of data one can use.
-Ecosystem condition assessments are generally limited by data availability, and the choice of variables and indicators to include in assessments are more often than not a pragmatic and opportunistic one, which is unlikely to reflect the full scope of the ecosystem condition characteristics.
+The need for an unbiased estimation of indicator values is unquestionable, but nonetheless, this requirement puts a large limitation on what types of data one can use in ecosystem condition accounts.
+Ecosystem condition assessments are generally limited by data availability, and the choice of variables and indicators to include in assessments is more often than not a pragmatic and opportunistic one, which is unlikely to reflect the full scope of the ecosystem condition characteristics.
 Note that the same is true for thematic biases.
@@
 a set of environmental variables), and use these relationships to predict and project indicator values to areas that where not originally sampled.
-Depending on the data that goes into these models, they can make very precice and good indicators.
+Depending on the data that goes into these models, they can make very precice and good indicators (REF).
 This is especially true when the EAAs are large (e.g. regions or nations).
@@
 Although it is possible that smaller or less calcareous mires will score systematically different than the ones that are surveyed, we do not think this is so much the case for our variables (see below).
-However, we do assume that alien plants are slightly less common on bogs relative to fens, and therefore that this variable will be biased since that variable for only recorded in fens.
+However, we do assume that alien plants are slightly less common on bogs relative to fens, and therefore that this variable will be biased since that variable is only recorded in fens.
 The other variables are recorded in all delineated mires.
@@
 Variable 1, 7FA, represents the abundance of alien species, mainly plants.
-Variables 2-5 (7SE, PRSL, 7TK and PRTK) describe very related aspects, and are attributed to the same ecosystem condition characteristic (vegetation intactness) and so they where combined into a single indicator called anthropogenic disturbance to soil and vegetation (ADSV; Figure S1).
+Variables 2-5 (7SE, PRSL, 7TK and PRTK) describe very related aspects, and are attributed to the same ecosystem condition characteristic (vegetation intactness) and so they were combined into a single indicator called anthropogenic disturbance to soil and vegetation (ADSV; Figure S1).
 These variables were were originally recorded along binned frequency ranges @norwegian_environmental_agency_protocol .
@@
 To turn variables into indicators we scaled them using three normative reference values each: an upper (best possible condition), a lower (worst possible condition), and a threshold value that defines the breaking point between good and poor condition [@jakobsson2020].
-The reference values makes up a numerical representation of the reference condition.
+The reference values make up a numerical representation of the reference condition.
 We define the reference condition as one where ecosystems are subject to little or no human influence, with a climate as in the period 1961-1990 and a native species pool similar as today [@jakobsson2021].
 
-The variable 7TK was rescaled into the indicator named *Alien species*, 7GR-GI was rescaled into the indicator *Trenching*, and ADSV was rescaled and kept the name *ADSV* (Figure S1).
+The variable 7FA was rescaled into the indicator named *Alien species*, 7GR-GI was rescaled into the indicator *Trenching*, and ADSV was rescaled and kept the name *ADSV* (Figure S1).
 For *Alien species* and *ADSV*, the lower and upper reference values were defined as 100% and 0%, respectively.
@@
 See @fig-workflowExample and S1 for schematic workflows for the indicator development.
-*Alien species* was attributed to Ecosystem Condition Typology (ECT) class B1 - Compositional state characteristics and *ADSV,* and *Trenching* was attributed to ECT class A1 - Physical state characteristics [@czucz_common_2021].
+*Alien species* was attributed to Ecosystem Condition Typology (ECT) class B1 - Compositional state characteristics and *ADSV,* and *Trenching* were attributed to ECT class A1 - Physical state characteristics [@czucz_common_2021].
 
@@
 We then categorised this continuous variable into four classes (0–3) using ranges chosen to produce a sensible and relatively linear area classification when visualised along an urban-wilderness gradient, whilst at the same time delineating sufficient areas for each strata across Norway (Table S1).
-We then aggregated the data to 1×1 km pixels to ease computations, and vectorized it.
+We then aggregated the data to 1×1 km pixels to ease computations, and vectorised it.
 The name HIAs is not a perfect representation of the information found in the infrastructure index, but we introduce this name here as a general term which is aligned with the concept of Homogeneous Ecosystem Areas, *sensu* @vallecillo_eu-wide_2022.
@@
 The municipalities differ in several aspects, such as the amount of mire area, the total area surveyed, and the prevalence of infrastructure (@tbl-munis, @fig-studyLocation2, S3, S4).
-We then took the nature type polygons with the indicator values and intersected with the HIA map and a map of municipal outlines in Norway.
+The nature type polygons with the indicator values were intersected with the HIA map and a map of municipal outlines in Norway.
 The relationship between the HIA classes and indicator values was examined visually.
@@
 Pane B shows the location of four homogeneous impact areas (HIAs) inside the EEA.
-The indicator values in Pane A are used, in combination with the HIA map in Pane B, to update a uniform prior and produce a posterior probability distribution for the mean area woeghted indicator value for each HIA (Pane C; here simplified to only show the mean (circles) and 95% credible intervals (colored bands)).
-The bayesian updating is informed by the standard deviation for the indicator in a much bigger national data set.
-The width of the posterior distribution responds both to variation in the data and to the sample size, giving  a realistic measure for the uncertainty around the indicator even with a singe observation.
+The indicator values in Pane A are used, in combination with the HIA map in Pane B, to update a uniform prior and produce a posterior probability distribution for the mean area weighted indicator value for each HIA (Pane C; here simplified to only show the mean (circles) and 95% credible intervals (colored bands)).
+The Bayesian updating is informed by the standard deviation for the indicator in a much bigger national data set.
+The width of the posterior distribution responds both to variation in the data and to the sample size, giving  a realistic measure for the uncertainty around the indicator even with a single observation.
 The distributions in Pane C are assumed normally distributed.
@@
 In Pane D the mean and SD from the posterior distributions are transferred to individual polygons in an ecosystem delineation map, for each HIA seperately.
-The colors indicator different indicator values (highlighted in the inset).
+The colors indicate different indicator values (highlighted in the inset).
 For each ecosystem occurence (i.e.
@@
 #| infrastructure densities (Homogeneous Impact Areas 0–3). 
-#| The data is from a national vegetation type survey in Norway."
+#| The data is from a national nature type survey in Norway."
 
@@
 #| homogenous impact area where it lies. Colors are chosen to best reflect 
-#| categorical differences and exagerates the absolute difference between 
+#| categorical differences and exaggerates the absolute difference between 
 #| areas. The inset it just a visual aid. Coordinate reference system is 
@@
 #| - "Distributions for three ecosystem condition indicators in the Norwegian 
-#| municiaplities. The color gradient reflects the value of the x-axis. 
-#| The dotted vertical line represent the threshold for what is considered 
+#| municipalities. The color gradient reflects the value of the x-axis. 
+#| The dotted vertical line represents the threshold for what is considered 
 #| reduced ecosystem condtion (< 0.6). Vertical lines under the density curves
@@
 In this paper we have demonstrated a generalisable GIS-based workflow for making use of spatially biased field data to produce representative ecosystem condition indicators at a local municipal scale (@fig-workflowExample).
-This approach is useful especially for local assessments were we do not want to use data from outside the EAA inform or influence our estimates for condition indicators inside the EAA.
+This approach is useful especially for local assessments were we do not want to use data from outside the EAA to inform or influence our estimates for condition indicators inside the EAA.
 The method relies on stratifying the field data by Homogeneous Impact Areas (HIAs).
@@
 Also, since our HIA areas are somewhat cohesive, forming a clumped or aggregated pattern, stratifying by it will on average lead to relatively more local data being used for each local indicator estimate, with less geographic extrapolation.
-Therefor, we see little risk in using the stratification process also on indicators where the relationship between the indicators and the HIA are weak, although we stress that not accounting for other, potentially more relevant predictors of indicator values, would leave the indicator as spatially biased as the original variable.
+Therefore, we see little risk in using the stratification process also on indicators where the relationship between the indicators and the HIA are weak, although we stress that not accounting for other, potentially more relevant predictors of indicator values, would leave the indicator as spatially biased as the original variable.
 
@@
 We have also demonstrated how Bayesian updating and re-sampling can be used instead of traditional arithmetic aggregation of indicators, from EAs to EAAs, to produce probability density distributions for indicator values.
-A major benefit of this approach in our example was that it allowed us to present reliable uncertainty ranges for our indicator estimates even with very low sampling sizes, thus making it possible to estimate indicator values for a much greater areas than if we were forced to set a minimum sample size threshold.
+A major benefit of this approach in our example was that it allowed us to present reliable uncertainty ranges for our indicator estimates even with very low sampling sizes, thus making it possible to estimate indicator values for a much greater area than if we were forced to set a minimum sample size threshold.
 An indicator estimate is of little value unless it is accompanied by a measure of uncertainty.
@@
 Assessments at national scales can make use of data sets, and hence indicators, that local assessments can not.
-For example national forest inventories or other national area representative field-based nature monitoring provide enough data points for a reliable estimates of indicator values only at relatively large spatial scales.
+For example national forest inventories or other national area representative field-based nature monitoring provide enough data points for reliable estimates of indicator values only at relatively large spatial scales.
 Local ecosystem accounts instead rely more on remote sensing or field measurements.
-Remote sensing holds great promise for delivering valuable information to ecosystem accounts across all scales, yet there are some ecosystem characteristics that can not easily be sensed remotey.
-This include for example species identities, and hence indicators on topics such as alien species or occurrence of threatened or certain keystone species.
-Fieldwork will therefore still be vital for collection information for ecosystem accounts also in the future.
-However, field work is labour intensive and expensive.
-To reduce costs, municipalities can stratify their field sampling according to the HIAs for efficient field efforts at local scales.
+Remote sensing holds great promise for delivering valuable information to ecosystem accounts across all scales, yet there are some ecosystem characteristics that can not easily be sensed remotely.
+This includes for example species identities, and hence indicators on topics such as alien species or occurrence of threatened or certain keystone species.
+Fieldwork will therefore still be vital for collection of information for ecosystem accounts also in the future.
+
+To increase efficacy of fieldwork at local scales - and hence, reduce costs - our results demonstrate that municipalities can stratify their field sampling according to the HIAs.
 For example, Gran and Nord-Aurdal have not recorded any field data from their most urban areas (HIA 3), and could make these areas a priority in upcoming field campaigns in order to get more complete spatial indicator coverage.
 
-We used a national data set of nature type surveys that follow a shared protocol.
-There can be other data set, that may not be compiled nationally, but which could be relevant.
+We used a national data set of nature type surveys that follow a shared protocol (REF).
+There can be other data sets, that may not be compiled nationally, but which could be relevant to XXX.
 In some cases, a local ecosystem condition account is mostly interested in describing and monitoring the status within its own EAA, but great benefits may be had if accounts can also be compared between EAA, say between municipalities.
@@
 In Norway, ecosystem impact assessments (EIA) are frequently being produced in conjunction with planned development projects.
-This data could be made very valuable for local ecosystem condition accounts, but currently there are several think stopping this.
+This data could be made very valuable for local ecosystem condition accounts, but currently there are several things stopping this.
 Firstly, there is no standardised reporting for EIA in Norway, and no mandatory publication pathway, and therefore no way to obtain a unified data set of variables recorded across different EIAs.
-If data is made publicly available in a national data base they could be used to design indicators that are informative at a multiple scales, also larger than the project level scale at which they were recorded.
+If data is made publicly available in a national data base they could be used to design indicators that are informative at multiple scales, also larger than the project level scale at which they were recorded.
 Second, the data collection is mainly on the occurrence of specific nature type or species of interest, and true condition parameters are not part of the information required by the authorities.
@@
 The indicator values for our three indicators across the three municipalities showed in general a good condition in the mire ecosystems (@fig-EEA-ridgePlot).
-The exception was for variable *Trenching* in Nordre Follo, where the indicator value was 0.25; well below the threshold for good ecosystem condition (0.6).
+The exception was for variable *Trenching* in Nordre Follo, where the indicator value was 0.25; well below the threshold for good ecosystem condition (0.6) (REF Jakobsson 2020, 2021).
 This estimate is based on just 4–5 data points (i.e. mires), and this is also reflected in the wide credible intervals \[0.04 – 0.50\].
-Nordre Follo has surveyed more of its total land area compared to the other two municipalities, but has less mire area to begin with (@tbl-munis) , and that is the reason for this low sample size.
+Nordre Follo has surveyed more of its total land area compared to the other two municipalities, but has less mire area to begin with (@tbl-munis), which is the reason for this low sample size.
 We believe this indicator estimate is trustworthy, in light of the wide credible intervals, and that it could reflect the true prevalence of trenching in Nordre Follo.
@@
 
-Alien species risk is not considered a threat to any of the Norwegian mire types , and our results confirm this, at least for the present (@fig-EEA-ridgePlot).
+Alien species risk is not considered a threat to any of the Norwegian mire types, and our results confirm this, at least for the present (@fig-EEA-ridgePlot).
 Motorised vehicle traffic and other forms of human transportation are also threats listed for some mire types in the red list assessment, and we see some reduction in the ADSV indicator, most so in Gran and in areas closer to infrastructure (@fig-forest).
-Still, the indicator values are consistently above the threshold value (@fig-EEA-ridgePlot).
+Still, the indicator values are consistently above the threshold value for good ecological condition (@fig-EEA-ridgePlot).
 
@@
 **Matthew Grainger**: ...
-**Marianne Evju**: ...
+**Marianne Evju**: Conceptualization, Writing - Review & Editing
 

### 2024-06-05 manuscript/Manuscript.qmd
follow up

@@
     journal:
-      name: Ecological Indicators
+      name: EcoEvoRxiv
       formatting: preprint
@@
 
-The SEEA EA framework recommends that ecosystem condition indicators give an unbiased representation of the condition for a given area [table 1 in @czucz_selection_2021; see also @unitednations2021 §2.87], and that condition indicators are recorded uniquely for each ecosystem asset (EA) so that the condition in the ecosystem accounting area (EAA) can be inferred from an area weighted mean based on the relative extents of the ecosystem assets [@unitednations2021 §5.54].
-Ecosystem assets (EAs) are defined as "ecological entities" (meaning areas) about which information is sought and about which statistics are compiled [@unitednations2021].
+The SEEA EA framework recommends that ecosystem condition indicators give an unbiased representation of the condition for a given area [table 1 in @czucz_selection_2021; see also @unitednations2021 §2.87], and that condition indicators are recorded uniquely for each ecosystem asset so that the condition in the ecosystem accounting area (EAA) can be inferred from an area weighted mean based on the relative extents of the ecosystem assets [@unitednations2021 §5.54].
+Ecosystem assets are defined as "ecological entities" (meaning areas) about which information is sought and about which statistics are compiled [@unitednations2021].
 This recommendation for representativity in the indicators means that spatially biased data are ill suited, especially if sampling intensity varies systematically along gradients of anthropogenic pressures and hence ecosystem condition.
-The recommendation for spatial resolution also means that when a condition metric is sampled coarsely, for example in an opportunistic and spatially biased field campaign, then it is very likely that you will not have a enough data points to reliably estimate an indicator value for each EA.
-This problem does not exist for complete wall-to-wall data, like remotely sensed imagery, where each EA will have ebough data to estaimte its unique indicator value.
+The recommendation for spatial resolution also means that when a condition metric is sampled coarsely, for example in an opportunistic and spatially biased field campaign, then it is very likely that you will not have a enough data points to reliably estimate an indicator value for each ecosystem asset.
+This problem does not exist for complete wall-to-wall data, like remotely sensed imagery, where each ecosystem asset will have enough data to estimate its unique indicator value.
 This challenge if spatial generalisation is not unique to ecosystem condition accounts, and it is much discussed in the ecosystem service accounting community of practice via the related term *value transfer* [@unitednations2022; @ncaves; @grammatikopoulou2023; @barton2023].
@@
 a.  Predict (project) indicator values from a statistical model that accounts for the effect of environmental variation on the variables
-b.  Take a central tendency (e.g. the mean) from the sample population and project it to all EAs of a given ecosystem type.
-c.  Take a central tendency from stratified sections of the sample population and projection them to the EAs, adhering to the corresponding strata.
+b.  Take a central tendency (e.g. the mean) from the sample population and project it to all ecosystem assets of a given ecosystem type.
+c.  Take a central tendency from stratified sections of the sample population and projection them to the ecosystem assets, adhering to the corresponding strata.
 
@@
 a set of environmental variables), and use these relationships to predict and project indicator values to areas that where not originally sampled.
-Depending on the data that goes into these models, they can make very precice and good indicators (REF).
+Depending on the data that goes into these models, they can make very precice and good indicators.
 This is especially true when the EAAs are large (e.g. regions or nations).
@@
 
-We have also demonstrated how Bayesian updating and re-sampling can be used instead of traditional arithmetic aggregation of indicators, from EAs to EAAs, to produce probability density distributions for indicator values.
+We have also demonstrated how Bayesian updating and re-sampling can be used instead of traditional arithmetic aggregation of indicators, from ecosystem assets to EAAs, to produce probability density distributions for indicator values.
 A major benefit of this approach in our example was that it allowed us to present reliable uncertainty ranges for our indicator estimates even with very low sampling sizes, thus making it possible to estimate indicator values for a much greater area than if we were forced to set a minimum sample size threshold.
@@
 
-We used a national data set of nature type surveys that follow a shared protocol (REF).
-There can be other data sets, that may not be compiled nationally, but which could be relevant to XXX.
-In some cases, a local ecosystem condition account is mostly interested in describing and monitoring the status within its own EAA, but great benefits may be had if accounts can also be compared between EAA, say between municipalities.
+We used a national data set of nature type surveys that follow a shared protocol [@miljødirektoratet].
+There can be other data sets, that may not be compiled nationally, but which could be relevant to local ECAs.
+In some cases, a local ECA is mostly interested in describing and monitoring the status within its own EAA, but great benefits may be had if accounts can also be compared between EAA, say between municipalities.
 This kind of information can give grounds for a more regional assessment, for example looking at the effects of policies on ecosystem condition trajectories.
@@
 The indicator values for our three indicators across the three municipalities showed in general a good condition in the mire ecosystems (@fig-EEA-ridgePlot).
-The exception was for variable *Trenching* in Nordre Follo, where the indicator value was 0.25; well below the threshold for good ecosystem condition (0.6) (REF Jakobsson 2020, 2021).
+The exception was for variable *Trenching* in Nordre Follo, where the indicator value was 0.25; well below the threshold for good ecosystem condition (0.6).
 This estimate is based on just 4–5 data points (i.e. mires), and this is also reflected in the wide credible intervals \[0.04 – 0.50\].

### 2024-06-06 manuscript/Manuscript.qmd
mostly typos and formatting

@@
 Therefore, being able to make use of spatially biased data sets would be very valuable.
-For national ECAs, modelling aproaches can in some cases be used to control for sampling biases.
-For local ECAs however, like at the scale of individual municipalities of developement projects, this is often not an option as it involved spatial extrapolation using data from outside the ecosystem accounting area.
+For national ECAs, modelling approaches can in some cases be used to control for sampling biases.
+For local ECAs however, like at the scale of individual municipalities of development projects, this is often not an option as it involves spatial extrapolation using data from outside the ecosystem accounting area.
 In this study we develop three ecosystem condition indicators from the same spatially biased data set based on nature type mapping of Norwegian mires.
@@
 We test our approach in three municipalities in south Norway.
-A spatial variable was created with four steps of increasing of infrastructure prevalence that we refer to as Homogeneous Impact Areas (HIAs).
-To facilitate reliable estimation of indicator uncertainties, even in cases with very low sample sizes, we use a Bayesian updating method and produce probability distributions for the area-weighted mean indicator values in each HIA seperately.
-Then we use area-weighted resampling and produce indicator propabiliy distributions for each municipality.
+We discretised a spatial variable representing infrastructure prevalence, and we refer to this new map as Homogeneous Impact Areas (HIAs).
+To facilitate reliable estimation of indicator uncertainties, even in cases with very low sample sizes, we use a Bayesian updating method and produce probability distributions for the area-weighted mean indicator values in each HIA separately.
+Then we use area-weighted resampling and produce indicator probability distributions for each municipality.
 With this Bayesian updating and resampling approach, small sample sizes can be compensated by correspondingly large uncertainty ranges, as long as the full dataset is large enough to estimate the true population standard deviation.
-This paper demonstrates the use of a GIS-based workflow to control for some of the most problematic biases in an oppurtunistic field survey so that the data can be used for indicators in ECAs.
+This paper demonstrates the use of a GIS-based workflow to control for some of the most problematic biases in an opportunistic field survey so that the data can be used for indicators in ECAs.
 The workflow can be used at any scale, including national scale.
-Because indicator values are calculatind for unique spatial strata, local governments or others can target their data aquisition towards strata with low sample sizes, and thus achieve higher cost effectiveness and ultimately better spatial indicator coverage.
+Because indicator values are calculated for unique spatial strata, local governments or others can target their data acquisition towards strata with low sample sizes, and thus achieve higher cost effectiveness and ultimately better spatial indicator coverage.
 keywords: 
@@
   message: false
+header-includes:
+- |
+  \usepackage{pdflscape}
+  \newcommand{\blandscape}{\begin{landscape}}
+  \newcommand{\elandscape}{\end{landscape}}
 ---
@@
 
-a.  Predict (project) indicator values from a statistical model that accounts for the effect of environmental variation on the variables
+a.  Predict (project) indicator values from a statistical model that accounts for the effect of environmental variation on the variables.
 b.  Take a central tendency (e.g. the mean) from the sample population and project it to all ecosystem assets of a given ecosystem type.
@@
 Statistical models can describe the general associations between the sampled data and the context where it was sampled (e.g.
-a set of environmental variables), and use these relationships to predict and project indicator values to areas that where not originally sampled.
-Depending on the data that goes into these models, they can make very precice and good indicators.
+a set of environmental variables) and use these relationships to predict and project indicator values to areas that where not originally sampled.
+Depending on the data that goes into these models, they can make very precise and good indicators.
 This is especially true when the EAAs are large (e.g. regions or nations).
@@
 The surveys are commissioned with the goal of producing data relevant for immediate land-use decisions, and are therefore spatially biased, typically towards areas with high human impact or expected impact.
-In addition there is a thematic and size bias in the sampling protocol.
+In addition, there is a thematic and size bias in the sampling protocol.
 For example, for the forest ecosystem, rare, endangered or calcareous forest types are delineated, whereas more common or ordinary forest types are not.
@@
 Variables 2-5 (7SE, PRSL, 7TK and PRTK) describe very related aspects, and are attributed to the same ecosystem condition characteristic (vegetation intactness) and so they were combined into a single indicator called anthropogenic disturbance to soil and vegetation (ADSV; Figure S1).
-These variables were were originally recorded along binned frequency ranges @norwegian_environmental_agency_protocol .
-Because the data was strongly right skewed we used the lower limit for each frequency range to convert them into percentages.
+These variables were were originally recorded along binned frequency ranges [@norwegian_environmental_agency_protocol].
+Because the data was strongly right skewed, we used the lower limit for each frequency range to convert them into percentages.
 This was done by summing the four variables after they had been converted to percentages.
-This was not a perfect solution, especially since some localities only had one of the variables recorded, but we chose this, rather than for example using a *worst rule* principle, in order to better separate the localities in terms of their indicator values.
-Variable 6, 7GR-GI, is different from the preceding variables in @tbl-variables in that it includes an estimation of future effects that the observed trenches are projected to have on mire vegetation, function or structure over time.
+This was not a perfect solution, especially since some localities only had one of the variables recorded, but we chose this, rather than for example using a *worst rule* principle, to better separate the localities in terms of their indicator values.
+Variable 6, 7GR-GI, is different from the preceding variables in @tbl-variables in that it includes an estimation of future effects that the observed trenches are projected to have on mire vegetation, function, or structure over time.
 This is not a favourable trait in a metric used to evaluate the ecosystem condition as it is today, yet we include it here nonetheless because there is a general shortage of data on mire hydrology, which is a fundamental part of mire integrity.
@@
 
-We used an ecosystem delineation map for open mires in southern Norway, produced using remotely sensed data and a deep learning model [@bakkestuen_delineation_2023].
-This model estimates 12.7% of the area in southern Norway is mire [@bakkestuen_delineation_2023].
-Mires are ecologically and socially important in Norway simply due to their large extent, and due to their role in climate mitigation as mires store a large amount of carbon.
-There has not been a national assessment of the ecosystem condition of mires in Norway, but the authors recently contributed to a report which presented several new indicators that can be used in future national assessments for this ecosystem [@nybo_indikatorer_2023; see also @kolstad2023]).
-The current study builds on the work in that report.
-
 ```{r fig-studyLocation2}
@@
 #| The survey coverage maps are in grey, and the mapped mires are
-#| in red, with polygon borders made extra thich to make them visible,
+#| in red, with polygon borders made extra thick to make them visible,
 #| but then also exaggerating their size.
@@
 #| is an ordinal gradient from 0–3 with increasing presence of human 
-#| instrastructure."
+#| infrastructure."
 
@@
 
-{{< pagebreak >}}
-
-| id  | Variable code        | Variable name                                    | Measurement unit                                                                         | Description                                                                                                                                                       | Reference            |
-|------------|------------|------------|------------|------------|------------|
-| 1   | 7FA                  | Prevalence of alien species                      | Unit-less, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                                                                                | @runehalvorsen2019   |
-| 2   | 7SE                  | Human caused abration or abration-caused erosion | Unit-less, ordinal, non-linear scale from 1 to 4.
-| Measures the frequency of imagined 4 m^2^ quadrats laid over the area that has some sign of abration                                                              | @runehalvorsen2019   |
-| 3   | PRSL                 | *as above*                                       | Unit-less, ordinal, non-linear scale from 0 to 7.
-| Same as 7SE, but recorded at a higher resolution                                                                                                                  | @miljødirektoratet   |
-| 4   | 7TK                  | Tracks from large vehicles                       | Unit-less, ordinal, non-linear scale from 1 to 4.
-| Measures the frequency of imagined 100 m^2^ quadrats laid over the area that has some signs of vehicle tracks                                                     | @runehalvorsen2019   |
-| 5   | PRTK                 | *as above*                                       | Unit-less, ordinal, non-linear scale from 0 to 7.
-| Same as 7TK, but recorded at a higher resolution                                                                                                                  | @miljødirektoratet   |
-| 6   | 7GR-GI               | Trenching intensity                              | Unit-less, ordinal scale from 1 to 5                                                     | Describes the effect that drainage ditches is estimate to have on the species composition and environmental variables ones the system reached its new equilibrium | @miljødirektoratet   |
-| 7   | Infrastructure Index | Infrastructure Index                             | Unit-less linear scale from 0 to 13.2                                                    | Unit-less index ranging from from 0 to 13.2                                                                                                                       | @erikstad_index_2023 |
-| 8   | HIA                  | Homogeneous Impact Area                          | Ordinal, non-linear scale from 1 to 4                                                    | A categorical representation of the infrastructure index                                                                                                          | this paper           |
-
-: Variables used in this study {#tbl-variables}
-
-{{< pagebreak >}}
-
-| Municipality  | Total terrestrial area (km^2^)                                            | \% of terrestrial area surveyed                                            | \% open mires in relation to total terrestrial area                         | Total mire area (km^2^)                                                  | %of mire area inside survey area                                                      | Number of mire polygons in survey                                 | Mean Infrastructure Index value                                         |
-|---------|---------|---------|---------|---------|---------|---------|---------|
-| Nordre Follo  | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(t_area_km)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(dk_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mirePercent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mire_km2)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mireInSurvery_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(n)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(meanHIA)` |
-| `Gran`        | `r muni_tbl |> filter(Municipality == "Gran") |> pull(t_area_km)`         | `r muni_tbl |> filter(Municipality == "Gran") |> pull(dk_percent)`         | `r muni_tbl |> filter(Municipality == "Gran") |> pull(mirePercent)`         | `r muni_tbl |> filter(Municipality == "Gran") |> pull(mire_km2)`         | `r muni_tbl |> filter(Municipality == "Gran") |> pull(mireInSurvery_percent)`         | `r muni_tbl |> filter(Municipality == "Gran") |> pull(n)`         | `r muni_tbl |> filter(Municipality == "Gran") |> pull(meanHIA)`         |
-| `Nord-Aurdal` | `r muni_tbl |> filter(Municipality == "Nord-Aurdal") |> pull(t_area_km)`  | `r muni_tbl |> filter(Municipality == "Nord-Aurdal") |> pull(dk_percent)`  | `r muni_tbl |> filter(Municipality == "Nord-Aurdal") |> pull(mirePercent)`  | `r muni_tbl |> filter(Municipality == "Nord-Aurdal") |> pull(mire_km2)`  | `r muni_tbl |> filter(Municipality == "Nord-Aurdal") |> pull(mireInSurvery_percent)`  | `r muni_tbl |> filter(Municipality == "Nord-Aurdal") |> pull(n)`  | `r muni_tbl |> filter(Municipality == "Nord-Aurdal") |> pull(meanHIA)`  |
-
-: Information for the three target municipalities in Norway {#tbl-munis}
+\pagebreak
+
+We used an ecosystem delineation map for open mires in southern Norway, produced using remotely sensed data and a deep learning model [@bakkestuen_delineation_2023].
+This model estimates 12.7% of the area in southern Norway is mire [@bakkestuen_delineation_2023].
+Mires are ecologically and socially important in Norway simply due to their large extent, and due to their role in climate mitigation as mires store a large amount of carbon.
+There has not been a national assessment of the ecosystem condition of mires in Norway, but the authors recently contributed to a report which presented several new indicators that can be used in future national assessments for this ecosystem [@nybo_indikatorer_2023; see also @kolstad2023]).
+The current study builds on the work in that report.
 
 Because the nature type survey data are spatially biased, we cannot assume that they are area-representative.
-In an attempt to overcome this issue we divided the area of Norway into four non-overlapping Homogeneous Impact Areas (HIAs) based on an infrastructure index for the year 2022 (@tbl-variables).
+In an attempt to overcome this issue, we divided the area of Norway into four non-overlapping Homogeneous Impact Areas (HIAs) based on an infrastructure index for the year 2022 (@tbl-variables).
 This index is a continuous variable that represents the frequency of different infrastructure types inside 500 m radius circles around each 100×100 m pixel [@erikstad2023].
-We then categorised this continuous variable into four classes (0–3) using ranges chosen to produce a sensible and relatively linear area classification when visualised along an urban-wilderness gradient, whilst at the same time delineating sufficient areas for each strata across Norway (Table S1).
-We then aggregated the data to 1×1 km pixels to ease computations, and vectorised it.
+We then categorised this continuous variable into four classes (0–3) using ranges chosen to produce a sensible and relatively linear area classification when visualised along an urban-wilderness gradient, whilst at the same time delineating sufficient areas for each stratum across Norway (Table S1).
+We then aggregated the data to 1×1 km pixels to ease computations and vectorised it.
 The name HIAs is not a perfect representation of the information found in the infrastructure index, but we introduce this name here as a general term which is aligned with the concept of Homogeneous Ecosystem Areas, *sensu* @vallecillo_eu-wide_2022.
@@
 We used the entire national data set for assessing the relationship between the HIA levels and the indicator values.
-We then sub-setted the national data set and chose three municipalities in south Norway to test out the indicators.
+We then subsetted the national data set and chose three municipalities in south Norway to test out the indicators.
 The municipalities differ in several aspects, such as the amount of mire area, the total area surveyed, and the prevalence of infrastructure (@tbl-munis, @fig-studyLocation2, S3, S4).
@@
 
-All the data preparation and analyses were done in RStudio [@positteam2024] and R version 4.3.0 [@rcoreteam2023], relying heavily on the packages tidyverse [@wickham2019] , sf [@pebesma2018] and stars [@pebesma2023].
+All the data preparation and analyses were done in RStudio [@positteam2024] and R version 4.3.0 [@rcoreteam2023], relying heavily on the packages tidyverse [@wickham2019], sf [@pebesma2018], and stars [@pebesma2023].
+
+\pagebreak
+
+\blandscape
+
+| id  | Variable code         | Variable name                                    | Measurement unit                                                                         | Description                                                                                                                                                       | Reference                                |
+|------------|------------|------------|------------|--------------|------------|
+| 1   | 7FA                   | Prevalence of alien species                      | Unit-less, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                                                                                | @runehalvorsen2019                       |
+| 2   | 7SE                   | Human caused abrasion or abrasion-caused erosion | Unit-less, ordinal, non-linear scale from 1 to 4.
+| Measures the frequency of imagined 4 m^2^ quadrats laid over the area that has some sign of abrasion                                                              | @runehalvorsen2019                       |
+| 3   | PRSL                  | *as above*                                       | Unit-less, ordinal, non-linear scale from 0 to 7.
+| Same as 7SE, but recorded at a higher resolution                                                                                                                  | @norwegian_environmental_agency_protocol |
+| 4   | 7TK                   | Tracks from large vehicles                       | Unit-less, ordinal, non-linear scale from 1 to 4.
+| Measures the frequency of imagined 100 m^2^ quadrats laid over the area that has some signs of vehicle tracks                                                     | @runehalvorsen2019                       |
+| 5   | PRTK                  | *as above*                                       | Unit-less, ordinal, non-linear scale from 0 to 7.
+| Same as 7TK, but recorded at a higher resolution                                                                                                                  | @norwegian_environmental_agency_protocol |
+| 6   | 7GR-GI                | Trenching intensity                              | Unit-less, ordinal scale from 1 to 5                                                     | Describes the effect that drainage ditches is estimate to have on the species composition and environmental variables ones the system reached its new equilibrium | @norwegian_environmental_agency_protocol |
+| 7   | Infra-structure Index | Infrastructure Index                             | Unit-less linear scale from 0 to 13.2                                                    | Unit-less index ranging from from 0 to 13.2                                                                                                                       | @erikstad_index_2023                     |
+| 8   | HIA                   | Homogeneous Impact Area                          | Ordinal, non-linear scale from 1 to 4                                                    | A categorical representation of the infrastructure index                                                                                                          | this paper                               |
+
+: Variables used in this study {#tbl-variables} {tbl-colwidths="\[5,10,15,25,25,20\]"}
+
+{{< pagebreak >}}
+
+| Municipality  | Total terrestrial area (km^2^)                                            | \% of terrestrial area surveyed                                            | \% open mires in relation to total terrestrial area                         | Total mire area (km^2^)                                                  | %of mire area inside survey area                                                      | Number of mire polygons in survey                                 | Mean Infrastructure Index value                                         |
+|---------|---------|---------|---------|---------|---------|---------|---------|
+| Nordre Follo  | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(t_area_km)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(dk_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mirePercent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mire_km2)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mireInSurvery_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(n)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(meanHIA)` |
+| `Gran`        | `r muni_tbl |> filter(Municipality == "Gran") |> pull(t_area_km)`         | `r muni_tbl |> filter(Municipality == "Gran") |> pull(dk_percent)`         | `r muni_tbl |> filter(Municipality == "Gran") |> pull(mirePercent)`         | `r muni_tbl |> filter(Municipality == "Gran") |> pull(mire_km2)`         | `r muni_tbl |> filter(Municipality == "Gran") |> pull(mireInSurvery_percent)`         | `r muni_tbl |> filter(Municipality == "Gran") |> pull(n)`         | `r muni_tbl |> filter(Municipality == "Gran") |> pull(meanHIA)`         |
+| `Nord-Aurdal` | `r muni_tbl |> filter(Municipality == "Nord-Aurdal") |> pull(t_area_km)`  | `r muni_tbl |> filter(Municipality == "Nord-Aurdal") |> pull(dk_percent)`  | `r muni_tbl |> filter(Municipality == "Nord-Aurdal") |> pull(mirePercent)`  | `r muni_tbl |> filter(Municipality == "Nord-Aurdal") |> pull(mire_km2)`  | `r muni_tbl |> filter(Municipality == "Nord-Aurdal") |> pull(mireInSurvery_percent)`  | `r muni_tbl |> filter(Municipality == "Nord-Aurdal") |> pull(n)`  | `r muni_tbl |> filter(Municipality == "Nord-Aurdal") |> pull(meanHIA)`  |
+
+: Information for the three target municipalities in Norway {#tbl-munis}
+
+\elandscape
 
@@
 Pane B shows the location of four homogeneous impact areas (HIAs) inside the EEA.
-The indicator values in Pane A are used, in combination with the HIA map in Pane B, to update a uniform prior and produce a posterior probability distribution for the mean area weighted indicator value for each HIA (Pane C; here simplified to only show the mean (circles) and 95% credible intervals (colored bands)).
+The indicator values in Pane A are used, in combination with the HIA map in Pane B, to update a uniform prior and produce a posterior probability distribution for the mean area weighted indicator value for each HIA (Pane C; here simplified to only show the mean (circles) and 95% credible intervals (coloured bands)).
 The Bayesian updating is informed by the standard deviation for the indicator in a much bigger national data set.
@@
 Note that because in this example there were no indicator data for HIA-3, indicator values are only aggregated for HIAs 0–2.
-In Pane D the mean and SD from the posterior distributions are transferred to individual polygons in an ecosystem delineation map, for each HIA seperately.
-The colors indicate different indicator values (highlighted in the inset).
-For each ecosystem occurence (i.e.
+In Pane D the mean and SD from the posterior distributions are transferred to individual polygons in an ecosystem delineation map, for each HIA separately.
+The colours indicate different indicator values (highlighted in the inset).
+For each ecosystem occurrence (i.e.
 ecosystem assets) in Pane D, we draw one random value for each square meter of area from a normal distribution with the mean and SD that is associated with that polygon.
@@
 #| - "An indicator for mire trenching shown for Nord-Aurdal municipality. 
-#| Individual mire polygons are colored by the mean indicator values for the
-#| homogenous impact area where it lies. Colors are chosen to best reflect 
+#| Individual mire polygons are coloured by the mean indicator values for the
+#| homogenous impact area where it lies. Colours are chosen to best reflect 
 #| categorical differences and exaggerates the absolute difference between 
@@
 #| - "Distributions for three ecosystem condition indicators in the Norwegian 
-#| municipalities. The color gradient reflects the value of the x-axis. 
+#| municipalities. The colour gradient reflects the value of the x-axis. 
 #| The dotted vertical line represents the threshold for what is considered 
@@
 
-# Histogram smoothing makes it look like there are more data points beyong the lower and upper percentiles than there really is. The 2.5% percentile sometimes looks like the 25% percentile.
+# Histogram smoothing makes it look like there are more data points beyond the lower and upper percentiles than there really is. The 2.5% percentile sometimes looks like the 25% percentile.
 
@@
 In this paper we have demonstrated a generalisable GIS-based workflow for making use of spatially biased field data to produce representative ecosystem condition indicators at a local municipal scale (@fig-workflowExample).
-This approach is useful especially for local assessments were we do not want to use data from outside the EAA to inform or influence our estimates for condition indicators inside the EAA.
+This approach is useful especially for local assessments where we do not want to use data from outside the EAA to inform or influence our estimates for condition indicators inside the EAA.
 The method relies on stratifying the field data by Homogeneous Impact Areas (HIAs).
 Our HIA was an ordinal gradient from low to high prevalence of infrastructure, and we found a relationship of increasing infrastructure and decreasing indicator values ranging from weak, in the case of the *Alien species* indicator, to strong, for the indicator for *Trenching* in mires (the *ADSV* indicator falling somewhere in the middle) (@fig-validation) .
-This implies that the stratification was warranted in the latter case (*Trenching*), i.e that field data from urban or sub-urban mires could not be said to represent the status of mires in more remote wilderness areas.
+This implies that the stratification was warranted in the latter case (*Trenching*), i.e.
+that field data from urban or sub-urban mires could not be said to represent the status of mires in more remote wilderness areas.
 In the case for *Alien species*, stratifying the variable by the HIAs can be predicted to have had a less directional effect on average, but we did find a difference between HIA-1 and 2 in Nord-Aurdal municipality, indicating that the impact from alien species differs between these two strata of varying infrastructure (@fig-forest).
@@
 For example, even if a numerical population has a large variance, it is still possible to sample, by chance alone, two or three numbers that are very close to each other, and this sample would then display a low variance which is not reflective of the true variance.
-Knowing this, an often used approach is to set a sample size limit, and refrain from estimating indicator values unless there are more than a given number of data points.
+Knowing this, one approach is to set a sample size limit, and refrain from estimating indicator values unless there are more than a given number of data points.
 In our test municipality Nordre Follo, we had only 1-3 data points for each HIA strata, yet we were able to use these data regardless because the Bayesian updating produces a reliable, and relatively broad, uncertainty range for the estimate.
@@
 Similarly, reporting of Ecosystem Accounts to Eurostat, expected to become mandatory for all EU and EEA countries from 2026, is unlikely to require sub-national resolution.
-Assessments at national scales can make use of data sets, and hence indicators, that local assessments can not.
-For example national forest inventories or other national area representative field-based nature monitoring provide enough data points for reliable estimates of indicator values only at relatively large spatial scales.
+Assessments at national scales can make use of data sets, and hence indicators, that local assessments cannot.
+For example, national forest inventories or other national area representative field-based nature monitoring provide enough data points for reliable estimates of indicator values only at relatively large spatial scales.
 Local ecosystem accounts instead rely more on remote sensing or field measurements.
-Remote sensing holds great promise for delivering valuable information to ecosystem accounts across all scales, yet there are some ecosystem characteristics that can not easily be sensed remotely.
+Remote sensing holds great promise for delivering valuable information to ecosystem accounts across all scales, yet there are some ecosystem characteristics that cannot easily be sensed remotely.
 This includes for example species identities, and hence indicators on topics such as alien species or occurrence of threatened or certain keystone species.
@@
 To increase efficacy of fieldwork at local scales - and hence, reduce costs - our results demonstrate that municipalities can stratify their field sampling according to the HIAs.
-For example, Gran and Nord-Aurdal have not recorded any field data from their most urban areas (HIA 3), and could make these areas a priority in upcoming field campaigns in order to get more complete spatial indicator coverage.
-
-We used a national data set of nature type surveys that follow a shared protocol [@miljødirektoratet].
+For example, Gran and Nord-Aurdal have not recorded any field data from their most urban areas (HIA 3) and could make these areas a priority in upcoming field campaigns in order to get more complete spatial indicator coverage.
+
+We used a national data set of nature type surveys that follow a shared protocol [@norwegian_environmental_agency_protocol].
 There can be other data sets, that may not be compiled nationally, but which could be relevant to local ECAs.
@@
 Firstly, there is no standardised reporting for EIA in Norway, and no mandatory publication pathway, and therefore no way to obtain a unified data set of variables recorded across different EIAs.
-If data is made publicly available in a national data base they could be used to design indicators that are informative at multiple scales, also larger than the project level scale at which they were recorded.
+If data is made publicly available in a national data base, they could be used to design indicators that are informative at multiple scales, also larger than the project level scale at which they were recorded.
 Second, the data collection is mainly on the occurrence of specific nature type or species of interest, and true condition parameters are not part of the information required by the authorities.
@@
 Nordre Follo has surveyed more of its total land area compared to the other two municipalities, but has less mire area to begin with (@tbl-munis), which is the reason for this low sample size.
-We believe this indicator estimate is trustworthy, in light of the wide credible intervals, and that it could reflect the true prevalence of trenching in Nordre Follo.
+We believe this indicator estimate is trustworthy, considering the wide credible intervals, and that it could reflect the true prevalence of trenching in Nordre Follo.
 Future work could test this assumption by simulating the effect of varying the sample size for an EAA where all the mires are sampled.
@@
 Mire condition in Norway has not been assessed at a national scale.
-Still, we know that they are subject to multiple threats, especially land conversion into agriculture, forestry or development, and we see this reflected in the Norwegian Red List for Nature types where many ombrotrophic bog types, and rich southern minerotrophic fens are listed [@artsdatabanken2018].
-Land conversion of mires is often initiated with trenching to lower the water table, but note that our *Trenching* indicator does not document the effect of trenching on old mires that are today transitioned into another ecosystem.
+Still, we know that they are subject to multiple threats, especially land conversion into agriculture, forestry, or development, and we see this reflected in the Norwegian Red List for Nature types where many ombrotrophic bog types, and rich southern minerotrophic fens are listed [@artsdatabanken2018].
+Land conversion of mires is often initiated with trenching to lower the water table but note that our *Trenching* indicator does not document the effect of trenching on old mires that are today transitioned into another ecosystem.
 This is in line with the UN standard for ecosystem accounting, where these effects should be captured by the extent accounts [@unitednations2021].
@@
 **Anders L. Kolstad**: Conceptualization, Methodology, Software, Validation, Formal analysis, Data Curation, Writing - Original Draft, Visualization, Project administration, Funding acquisition.
-**Matthew Grainger**: ...
-**Marianne Evju**: Conceptualization, Writing - Review & Editing
+**Matthew Grainger**: Validation, Writing - Review & Editing.
+**Marianne Evju**: Conceptualization, Writing - Review & Editing.
 
@@
 The source files and the data are also located on GitHub (https://github.com/anders-kolstad/HIAs) and will be archived on Zenodo with a stable doi at time of submission.
-The exception is the large nature type survey data which was downloaded locally, but which is freely available (see @norwegian_environmental_agency_naturtyper_2024).
+The exception is the large nature type survey data, which was downloaded locally, but which is freely available (see @norwegian_environmental_agency_naturtyper_2024).
 

### 2024-06-06 manuscript/Manuscript.qmd
fix Moen ref

@@
 The relationship between the HIA classes and indicator values was examined visually.
-For each HIA and municipality combination we then created a probability distribution for the area weighted mean of the indicator values using Bayesian updating of a uniform prior between 0 and 1, informed by the standard deviation (SD) of the indicator values in the national data set (@fig-workflowExample C; Appendix C, line 138 <!--# update line number manually when re-rendering appendix C -->).
+For each HIA and municipality combination we then created a probability distribution for the area weighted mean of the indicator values using Bayesian updating of a uniform prior between 0 and 1, informed by the standard deviation (SD) of the indicator values in the national data set (@fig-workflowExample C; Appendix C, line 134 <!--# update line number manually when re-rendering appendix C -->).
 We refer to this step as the first of two *aggregation* steps (Figure S1).

### 2024-06-06 manuscript/Appendix B - source code.qmd
rename appendices

--- 2024-06-06 manuscript/Manuscript.qmd
@@
 | id  | Variable code         | Variable name                                    | Measurement unit                                                                         | Description                                                                                                                                                       | Reference                                |
-|------------|------------|------------|------------|--------------|------------|
+|------------|------------|------------|------------|------------|------------|
 | 1   | 7FA                   | Prevalence of alien species                      | Unit-less, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                                                                                | @runehalvorsen2019                       |
@@
 For a rendered version of the source file, with all code and calculation visible, see Appendix C.
-The source files and the data are also located on GitHub (https://github.com/anders-kolstad/HIAs) and will be archived on Zenodo with a stable doi at time of submission.
+The source files and the data are also located on GitHub (https://github.com/anders-kolstad/HIAs) with an archived stable version on Zenodo (doi:10.5281/zenodo.11504703).
 The exception is the large nature type survey data, which was downloaded locally, but which is freely available (see @norwegian_environmental_agency_naturtyper_2024).

### 2025-02-26 manuscript/Manuscript.qmd
text

@@
 These metrics are quantified and ideally monitored over time to reflect the status and trends in condition.
-Indicators are variables that are normalised (rescaled) against upper and lower reference values to become bound between the values 0 and 1.
+Indicators are variables that are normalised (referred to as rescaling in SEEA EA) against upper (X~100~) and lower (X~0~)reference levels to become bound between the values 0 and 1 [@czúcz].
 This normalisation ensures that indicators are more comparable because an indicator value of 1 will mean the same for all indicators, i.e.
-that the variable equals the upper reference value which again reflect the value of the variables under the reference (pristine) condition.
+that the variable equals the upper reference level, which again reflect the value of the variables under the reference condition.
 Similarly, a value of 0 means that the variable is in the worst possible state, for example that the species or ecosystem function in question is completely lost from the ecosystem.
-The reference condition needs to be defined for each ecosystem condition assessment separately, but SEEA EA gives some suggestions, such as an an ecosystem with no or minimal anthropogenic disturbance.
-
-The SEEA EA framework recommends that ecosystem condition indicators give an unbiased representation of the condition for a given area [table 1 in @czucz_selection_2021; see also @unitednations2021 §2.87], and that condition indicators are recorded uniquely for each ecosystem asset so that the condition in the ecosystem accounting area (EAA) can be inferred from an area weighted mean based on the relative extents of the ecosystem assets [@unitednations2021 §5.54].
+The reference condition needs to be defined for each ecosystem type and each ecosystem condition assessment separately, but SEEA EA gives some suggestions, such as an an ecosystem with no or minimal anthropogenic disturbance.
+
+The SEEA EA recommends that ecosystem condition indicators give an unbiased representation of the condition for a given area [table 1 in @czucz_selection_2021; see also @unitednations2021 §2.87], and that condition indicators are recorded, or estimated, uniquely for each ecosystem asset so that the condition in the ecosystem accounting area (EAA) can be inferred from an area weighted mean based on the relative extents of the ecosystem assets [@unitednations2021 §5.54].
 Ecosystem assets are defined as "ecological entities" (meaning areas) about which information is sought and about which statistics are compiled [@unitednations2021].
 This recommendation for representativity in the indicators means that spatially biased data are ill suited, especially if sampling intensity varies systematically along gradients of anthropogenic pressures and hence ecosystem condition.
-The recommendation for spatial resolution also means that when a condition metric is sampled coarsely, for example in an opportunistic and spatially biased field campaign, then it is very likely that you will not have a enough data points to reliably estimate an indicator value for each ecosystem asset.
-This problem does not exist for complete wall-to-wall data, like remotely sensed imagery, where each ecosystem asset will have enough data to estimate its unique indicator value.
-This challenge if spatial generalisation is not unique to ecosystem condition accounts, and it is much discussed in the ecosystem service accounting community of practice via the related term *value transfer* [@unitednations2022; @ncaves; @grammatikopoulou2023; @barton2023].
-There are at least three ways to achieve this same complete areal coverage from indicator values that originally are spatially incomplete:
+The recommendation for the spatial support, i.e.
+the spatial grain, also means that when a condition metric is sampled coarsely, for example in an opportunistic and spatially biased field campaign, then it is very likely that you will not have a enough data points to reliably estimate an indicator value for each ecosystem asset.
+This problem does not exist for exhaustive, or complete, wall-to-wall data, like remotely sensed imagery, where each ecosystem asset will have enough data to estimate its unique indicator value.
+This challenge if spatial generalisation <!--# use term upscaling and spatial aggregation -->is not unique to ecosystem condition accounts, and it is much discussed in the ecosystem service accounting community of practice via the related term *value transfer* [@unitednations2022; @ncaves; @grammatikopoulou2023; @barton2023].
+There are at least three ways <!--# see UN guidelines annex D and Bierkens et al 2002 for many other ways --> to achieve this same complete areal coverage <!--# but we can get EAA estimated without inter-extrapolation.
+Need to frame this more using Bierkens et al --> from indicator values that originally are spatially incomplete:
 
@@
 | id | Variable code | Variable name | Measurement unit | Description | Reference |
-|------------|------------|------------|------------|--------------|------------|
+|------------|------------|------------|------------|------------|------------|
 | 1 | 7FA | Prevalence of alien species | Unit-less, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species | @runehalvorsen2019 |
@@
 | Measures the frequency of imagined 4 m^2^ quadrats laid over the area that has some sign of abrasion | @runehalvorsen2019 |
-| 3   | PRSL                  | *as above*                                       | Unit-less, ordinal, non-linear scale from 0 to 7.
-| Same as 7SE, but recorded at a higher resolution                                                                                                                  | @norwegian_environmental_agency_protocol |
+| 3 | PRSL | *as above* | Unit-less, ordinal, non-linear scale from 0 to 7. | Same as 7SE, but recorded at a higher resolution | @norwegian_environmental_agency_protocol |
 | 4 | 7TK | Tracks from large vehicles | Unit-less, ordinal, non-linear scale from 1 to 4.
 | Measures the frequency of imagined 100 m^2^ quadrats laid over the area that has some signs of vehicle tracks | @runehalvorsen2019 |
-| 5   | PRTK                  | *as above*                                       | Unit-less, ordinal, non-linear scale from 0 to 7.
-| Same as 7TK, but recorded at a higher resolution                                                                                                                  | @norwegian_environmental_agency_protocol |
+| 5 | PRTK | *as above* | Unit-less, ordinal, non-linear scale from 0 to 7. | Same as 7TK, but recorded at a higher resolution | @norwegian_environmental_agency_protocol |
 | 6 | 7GR-GI | Trenching intensity | Unit-less, ordinal scale from 1 to 5 | Describes the effect that drainage ditches is estimate to have on the species composition and environmental variables ones the system reached its new equilibrium | @norwegian_environmental_agency_protocol |
@@
 
+We acknowledge funding from the Research Council of Norway through the project EcoGaps (contract #320042).
+
 # Data availability {.unnumbered}

### 2025-04-22 manuscript/Manuscript.qmd

--- 2024-06-06 manuscript/Manuscript.qmd
@@
 library(kableExtra)
+library(here)
 
@@
 # # downloaded from https://kartkatalog.geonorge.no/metadata/naturtyper-miljoedirektoratets-instruks/eb48dd19-03da-41e1-afd9-7ebc3079265c
-path_naturetypes <- "../data/survey.gdb"
+path_naturetypes <- "P:/41001581_egenutvikling_anders_kolstad/data/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb"
 
 # municipality outline
-path_muni <- "../data/Basisdata_0000_Norge_25833_Kommuner_FGDB.gdb"
+path_muni <- here("data/Basisdata_0000_Norge_25833_Kommuner_FGDB.gdb")
 
@@
 # summary file (https://github.com/NINAnor/ecosystemCondition/blob/main/data/naturetypes/natureType_summary.rds)
-naturetypes_summary <- readRDS("../data/natureType_summary.rds")
+naturetypes_summary <- readRDS(here("data/natureType_summary.rds"))
 
@@
 st_layers(path_naturetypes)
+Sys.setlocale("LC_ALL","Norwegian")
 naturetypes <- sf::st_read(dsn = path_naturetypes, layer = "naturtyper_nin_omr")
@@
 # Outline of norway (coastline)
-outline <- sf::read_sf("../data/outlineOfNorway_EPSG25833.shp") |>
+outline <- sf::read_sf(here("data/outlineOfNorway_EPSG25833.shp")) |>
   st_transform(myCRS)
@@
   filter(kommunenummer %in% c(
-    "3020", # Nordre Follo
+    "3207", # Nordre Follo
     "3451", # Nord-Aurdal
@@
   mutate(Municipality = case_when(
-    kommunenummer == "3020" ~ "Nordre Follo",
+    kommunenummer == "3207" ~ "Nordre Follo",
     kommunenummer == "3451" ~ "Nord-Aurdal",
@@
 nf <- muni3 |>
-  filter(kommunenummer == "3020")
+  filter(kommunenummer == "3207")
 na <- muni3 |>
@@
 We wish to thank Kwaku Peprah Adjei for statistical help, and Hanno Sandvik for code review.
+We acknowledge partial funding from the Research Council of Norway project EcoGaps (contract #320042).
 
